### 2022-09-10 paper/sections/1_introduction.tex
Initial work on paper

\section{Introduction}
Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.

Dependently typed languages are different from other languages because they allow types to be parameterized by values. This allows more rigorous reasoning over types and the values that are inhabited by a type. This expressiveness also makes dependent type systems more complicated to implement. Especially, deciding equality of types requires evaluation of the terms they are parameterized by.

This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language. We want to investigate whether typical features of dependently typed language can be encoded concisely in Statix.

### 2022-09-14 paper/sections/1_introduction.tex
Work on paper

@@
 We want to investigate whether typical features of dependently typed language can be encoded concisely in Statix.
+The goal is not to show that Statix can implement it, but that implementing it is easier in Statix than in a general-purpose programming language.
+
+We will first show the base language and explain the way that Statix was used to implement this language. Next, we will explore several features and see how well they can be expressed in Statix.
+
+Features
+- Base language (sec 2)
+- Name collision avoidance (sec 3)
+- Editor services (sec 4)
+- Inference (sec 5)
+- Data types (sec 6)

### 2022-09-16 paper_new/200-base-language.tex
More work on paper

% !TEX root = document.tex

\chapter{\label{chap:baselang}Base Language}

The base language that was implemented is the Calculus of Constructions \autocite{Coquand_Huet_1988}, with a syntax somewhat similar to that of Haskell. One extra feature was added that is not present in the Calculus of Constructions, that is, let bindings.

\section{Syntax}

The syntax of the base language is defined in SDF3, the syntax definition language of Spoofax. The definition is very similar to that of a simply typed lambda calculus, except that types and expressions are a single sort.

\begin{lstlisting}
context-free sorts
	Expr

context-free syntax
	Expr.Let = [let [ID] = [Expr]; [Expr]]
	Expr.Type = "Type"
	Expr.Var = ID
	Expr.FnType = ID ":" Expr "->" Expr {right}
	Expr.FnConstruct = "\\" ID ":" Expr "." Expr
	Expr.FnDestruct = Expr Expr {left}
	Expr = "(" Expr ")" {bracket}

context-free priorities
	Expr.Type > Expr.Var > Expr.FnType > Expr.FnDestruct 
		> Expr.FnConstruct > Expr.Let
\end{lstlisting}

\section{Static Analysis}

\section{How scope graphs are used}

To type-check the base language, we need to use Statix, and scope graphs. This section describes how scope graphs are used.

The scope graph only has a single type of edge, we call them \verb|P| edges (parent edges). It also only has a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NameType|, which stores the type of a name, or a \verb|NameSubst|, which stores a substitution corresponding to a name.

\begin{lstlisting}
signature sorts
	NameEntry
constructors
	NameType : Expr -> NameEntry
	NameSubst : scope * Expr -> NameEntry
relations
	name : ID -> NameEntry
name-resolution	labels P
\end{lstlisting}

These are all the definitions we will need to type-check programs. We will create a name with \verb|NameType| as entry for function parameters, and we will use \verb|NameSubst| for let bindings and function application in beta reduction.

\section{Garbage}

- Base language is calculus of constructions with lets + type assertions (move type assertions to sec 5?)
- Describe syntax of the base language?
- Describe rules of base language (statix syntax or mathy?)
- Scope graphs for substitutions + scopes

### 2022-09-25 paper/100-introduction.tex
Rename paper new -> paper

--- 2022-09-14 paper/sections/1_introduction.tex
@@
-\section{Introduction}
+% !TEX root = document.tex
+
+\chapter{\label{chap:introduction}Introduction}
+
 Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages.
@@
 - Name collision avoidance (sec 3)
-- Editor services (sec 4)
+- Language parametric services (sec 4)
 - Inference (sec 5)

### 2022-09-25 paper/200-base-language.tex
Rename paper new -> paper

--- 2022-09-16 paper_new/200-base-language.tex
@@
 
-To type-check the base language, we need to use Statix, and scope graphs. This section describes how scope graphs are used.
+To type-check the base language, we need to scope graphs. This section describes how scope graphs are used.
 
-The scope graph only has a single type of edge, we call them \verb|P| edges (parent edges).
+The scope graph only has a single type of edge, called \verb|P| (parent) edges.
 It also only has a single relation, called \verb|name|.
@@
 These are all the definitions we will need to type-check programs.
-We will create a name with \verb|NameType| as entry for function parameters, and we will use \verb|NameSubst| for let bindings and function application in beta reduction.
+
+\section{Typechecking programs}
+
+We will define a statix relation \verb|typeOfExpr| that takes a scope and an expression and typechecks the scope in the expression.
+It returns two expressions, the first being the expression that was input (this will be changed when we add inference), and the second being the type of the input expression.
+
+\begin{lstlisting}
+typeOfExpr : scope * Expr -> (Expr * Expr)
+\end{lstlisting}
+
+\subsection{Let bindings and variables}
+
+Let bindings are typechecked by first typechecking the value that was bound, and then putting this as a substitution in the scope graph. This new scope is then used to typecheck the body.
+
+\begin{lstlisting}
+typeOfExpr_(s, l@Let(n, v, b)) = (Let(n, v', b'), b_type) :- {s'}
+	typeOfExpr(s, v) == (v', _),
+	scopePutSubst(s, n, (s, v')) == s',
+	typeOfExpr(s', b) == (b', b_type).
+\end{lstlisting}
+
+We can then
 

### 2022-09-26 paper/200-base-language.tex
A little writing

@@
 
-To type-check the base language, we need to scope graphs. This section describes how scope graphs are used.
+To type-check the base language, we need to scope graph, this section describes how scope graphs are used.
 
@@
 
-These are all the definitions we will need to type-check programs.
+These are all the definitions we will need to type-check programs. Next, we will introduce some Statix relations that can be used to interact with these scope graphs:
+
+\begin{lstlisting}
+scopePutType : scope * ID * Expr -> scope
+scopePutSubst : scope * ID * (scope * Expr) -> scope
+scopeGetName : scope * ID -> (NameEntry * scope)
+scopeGetNames : scope * ID -> list((path * (ID * NameEntry)))
+empty_scope : -> scopes
+\end{lstlisting}
 

### 2022-09-28 paper/200-base-language.tex
Write some part of base language

@@
 
-\section{Static Analysis}
-
 \section{How scope graphs are used}
@@
 scopePutSubst : scope * ID * (scope * Expr) -> scope
-scopeGetName : scope * ID -> (NameEntry * scope)
+scopeGetName : scope * ID -> NameEntry
 scopeGetNames : scope * ID -> list((path * (ID * NameEntry)))
-empty_scope : -> scopes
+empty_scope : -> scope
 \end{lstlisting}
+
+The \verb|scopePutType| and \verb|scopePutSubsts| relations generate a new scope given a parent scope and a type or substitution respectively.
+To query the scope graph, use \verb|scopeGetName| or \verb|scopeGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found.
+Finally, \verb|empty_scope| returns a fresh empty scope.
 
@@
 
-We will define a statix relation \verb|typeOfExpr| that takes a scope and an expression and typechecks the scope in the expression.
-It returns two expressions, the first being the expression that was input (this will be changed when we add inference), and the second being the type of the input expression.
+We will define a statix relation \verb|typeOfExpr| that takes a scope and an expression and typechecks the scope in the expression. It returns the type of the expression.
 
 \begin{lstlisting}
-typeOfExpr : scope * Expr -> (Expr * Expr)
+typeOfExpr : scope * Expr -> Expr
 \end{lstlisting}
 
-\subsection{Let bindings and variables}
+\subsection{Typechecking Type, Let bindings and Variables}
+
+The \verb|Type| expression, being the Type of Types, has itself as its Type.
+
+\begin{lstlisting}
+typeOfExpr(_, Type()) = Type().
+\end{lstlisting}
 
@@
 \begin{lstlisting}
-typeOfExpr_(s, l@Let(n, v, b)) = (Let(n, v', b'), b_type) :- {s'}
-	typeOfExpr(s, v) == (v', _),
-	scopePutSubst(s, n, (s, v')) == s',
-	typeOfExpr(s', b) == (b', b_type).
+typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
+	typeOfExpr(s, v) == _,
+	scopePutSubst(s, n, (s, v)) == s'.
 \end{lstlisting}
 
-We can then
+We can then typecheck variables by querying the scope graph using \verb|scopeGetName|.
+Then, we need to define a relation \verb|typeOfNameEntry| that takes the \verb|NameEntry|, if it is a \verb|NameType| returns the type, and if it is a \verb|NameSubst| computes the type of the substitution \footnote{Note that as an optimization, to avoid recomputing the type each time the variable is used, we could compute the type of the NameSubst when we create the NameEntry.
+To keep the implementation as simple as possible, we didn't do this for now.}.
 
-\section{Garbage}
+\begin{lstlisting}
+typeOfExpr(s, Var(id)) = typeOfNameEntry(scopeGetName(s, id)).
+typeOfNameEntry : NameEntry -> Expr
+typeOfNameEntry(NameType(T)) = T.
+typeOfNameEntry(NameSubst(se, e)) = typeOfExpr(se, e).
+\end{lstlisting}
 
-- Base language is calculus of constructions with lets + type assertions (move type assertions to sec 5?)
-- Describe syntax of the base language?
-- Describe rules of base language (statix syntax or mathy?)
-- Scope graphs for substitutions + scopes
+\subsection{Typechecking Functions}
+
+Functions consist of three different expressions, \verb|FnType|, the type of functions, \verb|FnConstruct|, the constructor for functions (a lambda function) and \verb|FnDestruct|, the destructor for functions (function application).
+
+ The type of a \verb|FnType| expression is \verb|Type|, however we do need to typecheck the subexpressions.
+The argument type can be typechecked using the same scope, but because this is a dependently typed language, the return type may contain references to the name of the argument.
+Thus, we need to add the type of the variable to the scope before typechecking the return type.
+We then check that both the argument type and the return type are types, by checking that their type is beta equal to \verb|Type|.
+
+\begin{lstlisting}
+typeOfExpr(s, FnType(arg_name, arg_type, rtrn_type)) = Type() :- {s'}
+    expectBetaEq((empty_scope(), typeOfExpr(s, arg_type)), (empty_scope(), Type())),
+    scopePutType(s, arg_name, betaReduce((s, arg_type))) == s',
+	expectBetaEq((empty_scope(), typeOfExpr(s', rtrn_type)), (empty_scope(), Type())).
+\end{lstlisting}
+
+Next, ...
+
+\subsection{Beta Reductions}
+//todo

### 2022-09-29 paper/000-abstract.tex
Work on writing

% !TEX root = document.tex

When working with the Spoofax workbench, the Statix meta-language can be used for the specification of Static Semantics. To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems. However, no attempts have been made to express dependently typed languages in Statix. Dependently typed languages are unique in that types may depend on values. Type-checking them comes with unique challenges, such as the need to evaluate terms during type-checking. We present how to make a dependent language by implementing the calculus of constructions in Statix. Next, we discuss the difficulties that come from using names, rather than De Bruijn indices. Finally, we explore features such as type Inference and inductive data types.

### 2022-09-29 paper/200-base-language.tex
Work on writing

@@
 
+\subsection{Beta Reductions}
+A unique requirement for dependently typed languages is beta reduction during typechecking, since types may require evaluation to compare.
+In this section, we will define three rules, \verb|expectBetaEq|, which compares two expressions and checks if they're equal, \verb|betaReduce|, which fully evaluates an expression, and \verb|betaReduceHead|, which evaluates the head of an expression.
+
+Starting with the latter, the signature of \verb|betaReduceHead| is the following, taking in an expression with a scope, and also returning and expression with a possibly different scope.
+
+\begin{lstlisting}
+betaReduceHead : (scope * Expr) -> (scope * Expr)
+\end{lstlisting}
+
 \subsection{Typechecking Functions}
@@
 Next, ...
-
-\subsection{Beta Reductions}
-//todo

### 2022-10-04 paper/200-base-language.tex
Work on paper

@@
 
-\subsection{Typechecking Type, Let bindings and Variables}
+\subsection{Beta Reductions}
+A unique requirement for dependently typed languages is beta reduction during typechecking, since types may require evaluation to compare.
+In this section, we will define three rules, \verb|expectBetaEq|, which compares two expressions and checks if they're equal, \verb|betaReduce|, which fully evaluates an expression, and \verb|betaReduceHead|, which evaluates the head of an expression.
+The signatures of the relations are the following:
 
-The \verb|Type| expression, being the Type of Types, has itself as its Type.
+\begin{lstlisting}
+betaReduceHead : (scope * Expr) -> (scope * Expr)
+
+betaReduce : (scope * Expr) -> Expr
+betaReduce(e) = betaReduce_(betaReduceHead(e)).
+betaReduce_ : (scope * Expr) -> Expr
+
+expectBetaEq : (scope * Expr) * (scope * Expr)
+expectBetaEq(e1, e2) :- expectBetaEq_(betaReduceHead(e1), betaReduceHead(e2)).
+expectBetaEq_ : (scope * Expr) * (scope * Expr)
+\end{lstlisting}
+
+The definition of \verb|betaReduce| is to first beta reduce the head, and then to match on the resulting head.
+It could be defined separately, but defining it like this prevents a lot of code duplication.
+Similarly, the definition of \verb|expectBetaEq| is to first beta reduce the head of both expressions, and then compare their heads using the \verb|expectBetaEq_| relation.
+The relations \verb|betaReduceHead|,  \verb|betaReduce_|, and \verb|expectBetaEq_|  will be defined incrementally as we build up the language over the next few sections.
+
+\subsection{Typechecking Type}
+
+The \verb|Type| expression, being the type of types, has itself as its type.
 
@@
 
-Let bindings are typechecked by first typechecking the value that was bound, and then putting this as a substitution in the scope graph. This new scope is then used to typecheck the body.
+Furthermore, \verb|Type| cannot be beta-reduced, and comparing two types is trivial.
+Because \verb|expectBetaEq| already executed \verb|betaReduceHead| on both sides, the only way for them to be equal is for both to be exactly \verb|Type()|.
+
+\begin{lstlisting}
+betaReduceHead((s, Type())) = (s, Type()).
+betaReduce_((_, Type())) = Type().
+expectBetaEq_((_, Type()), (_, Type())).
+\end{lstlisting}
+
+\subsection{Let bindings and Variables}
+
+Let bindings are typechecked by first typechecking the value that was bound, and then putting the value as a substitution in the scope graph. This new scope is then used to typecheck the body.
 
@@
 Then, we need to define a relation \verb|typeOfNameEntry| that takes the \verb|NameEntry|, if it is a \verb|NameType| returns the type, and if it is a \verb|NameSubst| computes the type of the substitution \footnote{Note that as an optimization, to avoid recomputing the type each time the variable is used, we could compute the type of the NameSubst when we create the NameEntry.
-To keep the implementation as simple as possible, we didn't do this for now.}.
+To keep the implementation as simple as possible, we didn't do this.}.
 
@@
 
-\subsection{Beta Reductions}
-A unique requirement for dependently typed languages is beta reduction during typechecking, since types may require evaluation to compare.
-In this section, we will define three rules, \verb|expectBetaEq|, which compares two expressions and checks if they're equal, \verb|betaReduce|, which fully evaluates an expression, and \verb|betaReduceHead|, which evaluates the head of an expression.
-
-Starting with the latter, the signature of \verb|betaReduceHead| is the following, taking in an expression with a scope, and also returning and expression with a possibly different scope.
+Next, we need to define beta reduction for let bindings.
+This is simple, we only need to define the \verb|betaReduceHead| relation on them.
+The other relations don't need definitions of let bindings, since \verb|betaReduceHead| can never return a let binding.
+Note that the definition recursively calls \verb|betaReduceHead|, since the body needs to be put into head-normal form as well.
 
 \begin{lstlisting}
-betaReduceHead : (scope * Expr) -> (scope * Expr)
+betaReduceHead((s, Let(n, v, b))) = betaReduceHead((scopePutSubst(s, n, (s, v)), b)).
 \end{lstlisting}
 
-\subsection{Typechecking Functions}
+Finally, we need to define beta reduction for variables.
+This is a bit more complex.
+We call \verb|scopeGetNames|, which returns a list of scope graph results.
+If we can find a substitution for the variable, we keep beta reducing the head of this substitution.
+Otherwise, we return the variable.
+Since \verb|betaReduceHead| may return a variable, we also need to define the other relations.
+
+\begin{lstlisting}
+betaReduceHead((s, e@Var(id))) = betaReduceHeadVar(id, scopeGetNames(s, id)).
+betaReduceHeadVar : ID * list((path * (ID * NameEntry))) -> (scope * Expr)
+betaReduceHeadVar(_, [(_, (_, NameSubst(sw, w))) | _]) = betaReduceHead((sw, w)).
+betaReduceHeadVar(id, _) = (empty_scope(), Var(id)).
+
+betaReduce_((_, Var(n))) = Var(n).
+expectBetaEq_((_, Var(n)), (_, Var(n))).
+\end{lstlisting}
+
+\subsection{Typechecking Function Types}
 
@@
 Thus, we need to add the type of the variable to the scope before typechecking the return type.
-We then check that both the argument type and the return type are types, by checking that their type is beta equal to \verb|Type|.
+The argument type needs to be fully beta reduced, since we don't give types a scope.\footnote{Whether giving types a scope has benefits is an opportunity for future research.} We then check that both the argument type and the return type are types, by checking that their type is beta equal to \verb|Type|.
 
@@
 
-Next, ...
+Beta reduction for function types is defined relatively trivially, as follows
+
+\begin{lstlisting}
+betaReduceHead((s, e@FnType(_, _, _))) = (s, e).
+betaReduce_((s, FnType(arg_name, arg_type, rtrn))) = 
+	FnType(arg_name, betaReduce((s, arg_type)), betaReduce((s, rtrn))).
+\end{lstlisting}
+
+Next, we need to define \verb|expectBetaEq_| for function types.
+This is not trivial, since we need to keep alpha equality into account.
+For example, the following two types are considered equal (they are both functions that take in a Type, and return a value of that Type\footnote{It is not possible to implement these functions, but expressing their type is still possible}):
+
+\begin{lstlisting}
+x: Type -> x
+y: Type -> y
+\end{lstlisting}
+
+The solution is to introduce a new constructor for expressions, \verb|AlphaEqVars : ID * ID -> Expr|.
+In the first function type body, we will substitute \verb|x| with \verb|AlphaEqVars(x, y)|, and in the second function type body we do the same for \verb|y|.
+We then check if the bodies are equal.
+
+\begin{lstlisting}
+expectBetaEq_((s1, FnType(arg_name1, arg_type1, body1)), 
+              (s2, FnType(arg_name2, arg_type2, body2))) :- {q}
+	expectBetaEq((s1, arg_type1), (s2, arg_type2)),
+	q == AlphaEqVars(arg_name1, arg_name2),
+	expectBetaEq(
+		(scopePutSubst(s1, arg_name1, (empty_scope(), q)), body1), 
+		(scopePutSubst(s2, arg_name2, (empty_scope(), q)), body2)).
+\end{lstlisting}
+
+\subsection{Typechecking Functions}

### 2022-10-04 paper/300-name-collisions.tex
Work on paper

% !TEX root = document.tex

\chapter{\label{chap:namecolls}Solving Name Collisions}

The implementation of the base language shown in section \ref{chap:baselang} has one big problem, that is name collisions. This section will explore several ways of solving these collisions. An example of such a collision is the following: What is the type of this expression (a polymorphic identity function)?
\begin{lstlisting}
\T : Type. \T : T. T
\end{lstlisting}

The algorithm so far would tell you it is \verb|T : Type -> T : T -> T|. Given the scoping rules of the language, that is equivalent to \verb|T : Type -> x : T -> x|. However, the correct answer would be \verb|T : Type -> x : T -> T|. There is no way of expressing this type without renaming a variable.

\section{In depth: Why does this happen?}

In this section, we will step through the steps that happen during the type checking of the term above, to explain why the incorrect type signature is returned. To find the type, the following is evaluated:

\begin{lstlisting}
typeOfExpr(_, FnConstruct("T", Type(), FnConstruct("T", Var("T"), Var("T"))))
\end{lstlisting}

\noindent
This creates a new node in the scope graph, and then type checks the body with this scope.

\begin{lstlisting}
typeOfExpr(s1, FnConstruct("T", Var("T"), Var("T")))

(T : Type)
   [s1]
\end{lstlisting}

\noindent
The same thing happens, the body of the \verb|FnConstruct| is typechecked with a new scope.

\begin{lstlisting}
typeOfExpr(s2, Var("T"))

(T : Type)      (T : Var("T"))
	[s1]<-----------[s2]
\end{lstlisting}

Finally, we need to find the type of \verb|T|. This finds the lexically closest definition of \verb|T| (the one in s2), which is correct. But the type of \verb|T| is \verb|T|, which does NOT refer to the lexically closest \verb|T|, but instead to the \verb|T| in s1. This situation, in which a type can contain a reference to a variable that is shadowed, is the problem. We need to find a way to make sure that shadowing like this can never happen.

\section{Solutions}

\subsection{De Bruijn Indices}

Almost all compilers that typecheck dependently typed languages use de Bruijn representation for variables \cite{TODO agda lean etc}. Using de Bruijn indices in statix is possible, but sacrifices a lot. It would require a transformation on the AST before typechecking. Modifying the AST like this causes problems, because it changes AST nodes. All editor services that rely on \verb|.ref| annotations, such as renaming, can no longer be used. It also loses a lot of the benefits of using Spoofax, since using scope graphs relies on using names.

\subsection{Uniquifying names}

The first solution that was attempted was having a pre-analysis transformation that gives each variable a unique name. This doesn't work for a variety of reasons. The simplest being, it doesn't actually solve the problem. Names can be duplicated during beta reduction of terms, so we still don't have the guarantee that each variable has a unique name. Furthermore, this is a pre-analysis transformation, so similarly to using de Bruijn indices, it breaks editor services such as renaming.

\subsection{Renaming terms dynamically}

\subsection{Using scopes to distinguish names}

### 2022-10-05 paper/300-name-collisions.tex
Work on paper

@@
 \noindent
-The same thing happens, the body of the \verb|FnConstruct| is typechecked with a new scope.
+The same thing happens, the body of the \verb|FnConstruct| is typechecked with a new scope. Note that the T in the type of the second T is ambiguous.
 
@@
 
-\section{Solutions}
+\section{Alternative Solutions}
 
@@
 
-\subsection{Using scopes to distinguish names}
+Anytime that we introduce a new name in a type, we could check if the name already exists in the environment, and if it does, choose a different unique name.
+This approach is possible but tedious to implement in Statix.
+It requires a new relation to traverse through the type and rename.
+It also increases the time complexity, as constant traversals of the type are needed.
+
+\section{Using scopes to distinguish names}
+
+The solution we found to work best in the end is to change the definition of \verb|ID|.
+To be precise, at the grammar level we have two sorts, \verb|RID| is a "Raw ID", being just a string.
+\verb|ID| will have two constructors, one being \verb|Syn|, a syntactical \verb|ID|.
+The second one is \verb|ScopedName|, it is defined in statix, so there is no syntax for it.
+It will be generated by \verb|typeOfExpr|.
+
+\begin{lstlisting}
+context-free sorts ID
+lexical sorts RID
+context-free syntax
+   ID.Syn = RID
+signature constructors
+  ScopedName : scope * RID -> ID
+\end{lstlisting}
+
+The \verb|ScopedName| constructor has a scope and a raw ID.
+The scope is used to uniquely identify the name.
+The main idea is that whenever we encounter a syntactical name, we replace it with a scoped name, so it unambiguous.
+The scope graph will never have a syntactical name in it.
+However, when querying the scope graph for a syntactical name, we return the lexically closest name.
+
+\subsection{The example revisited}
+
+In this section, we will step through the steps that happen during the type checking of the term above, with name collisions solved.
+To find the type, the following is evaluated, note that the names are now wrapped in a \verb|Syn| constructor:
+
+\begin{lstlisting}
+typeOfExpr(_, FnConstruct(Syn("T"), Type(), 
+	FnConstruct(Syn("T"), Var(Syn("T")), Var(Syn("T")))))
+\end{lstlisting}
+
+\noindent
+The name in the \verb|FnConstruct| is replaced with a scoped name. The scope of the name is the scope that the name is first defined in. We then type check the body with this scope.
+
+\begin{lstlisting}
+typeOfExpr(s1, FnConstruct(Syn("T"), Var(Syn("T")), Var(Syn("T"))))
+
+(ScopedName(s1, T) : Type)
+	       [s1]
+\end{lstlisting}
+
+\noindent
+The same thing happens, the body of the \verb|FnConstruct| is typechecked with a new scope. Note that the type of the new \verb|T| now specifies which \verb|T| it means, so it is no longer ambiguous.
+
+\begin{lstlisting}
+typeOfExpr(s2, Var(Syn("T")))
+
+(ScopedName(s1, T) : Type)     (ScopedName(s2, T) : Var(ScopedName(s1, T)))
+	       [s1]<----------------------------------[s2]
+\end{lstlisting}
+
+Finally, we need to find the type of \verb|T|.
+This finds the lexically closest definition of \verb|T| (the one in s2), as defined earlier.
+The type of this \verb|T| is \verb|ScopedName(s1, T)|, which explicitly defined which \verb|T| it is.
+A name can now never shadow another name, since each scope uniquely identifies a name.
+The final type of the expression is now:
+
+\begin{lstlisting}
+FnType(ScopedName(s1, T), Type(), FnType(ScopedName(s2, T), Var(ScopedName(s1, T)), Var(ScopedName(s1, T))))
+\end{lstlisting}
+
+\subsection{Improving the readability of types}

### 2022-10-12 paper-evcs/oasics-v2021-sample-article.tex
Start work on EVCS paper

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{oasics-v2021}
%This is a template for producing OASIcs articles. 
%See oasics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideOASIcs %uncomment to remove references to OASIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Dependently Typed Languages in Statix} %TODO Please add

\author{Jonathan Brouwer}{Delft University of Technology, The Netherlands \and \url{http://jonathanb.nl}}{j.t.brouwer@student.tudelft.nl}{}{}

\author{Jesper Cockx}{Delft University of Technology, The Netherlands \and \url{http://jesper.sikanda.be}}{j.g.h.cockx@tudelft.nl}{}{}

\author{Aron Zwaan}{Delft University of Technology, The Netherlands \and \url{http://aronzwaan.github.io}}{a.s.zwaan@tudelft.nl}{}{}

\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}}

\keywords{Spoofax, Statix, Dependent Types} %TODO mandatory; please add comma-separated list of keywords

\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti. 
\end{abstract}

\section{Introduction}

Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of Static Semantics. To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems. However, no attempts have been made to express dependently typed languages in Statix.

Dependently typed languages are different from other languages because they allow types to be parameterized by values. This allows more rigorous reasoning over types and the values that are inhabited by a type. This expressiveness also makes dependent type systems more complicated to implement. Especially, deciding equality of types requires evaluation of the terms they are parameterized by.

This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language. We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix. The goal is not to show that Statix can implement it, but that implementing it is easier in Statix than in a general-purpose programming language.

We will first show the base language and explain the way that Statix was used to implement this language. Next, we will explore several features and see how well they can be expressed in Statix.

\section{Background}

\bibliography{oasics-v2021-sample-article}

\end{document}

### 2022-10-12 paper-evcs/oasics-v2021-sample-article.tex
Add gitignore
Work on paper

@@
 
-\section{Background}
+\section{Calculus of Constructions}
+
+The base language that was implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube.
+One extra feature was added that is not present in the Calculus of Constructions, let bindings.
+
+\subsection{How scope graphs are used}
+
+To type-check the base language, we need to store information about the names we have encountered.
+There are two different situations, names that we encounter that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.
+
+   we need scope graphs. This section describes how scope graphs are used.
+
+The scope graph only has a single type of edge, called \verb|P| (parent) edges.
+It also only has a single relation, called \verb|name|.
+This name stores a \verb|NameEntry|, which can be either a \verb|NameType|, which stores the type of a name, or a \verb|NameSubst|, which stores a substitution corresponding to a name.
+
+\begin{lstlisting}
+	signature sorts
+	NameEntry
+	constructors
+	NameType : Expr -> NameEntry
+	NameSubst : scope * Expr -> NameEntry
+	relations
+	name : ID -> NameEntry
+	name-resolution	labels P
+\end{lstlisting}
+
+These are all the definitions we will need to type-check programs. Next, we will introduce some Statix relations that can be used to interact with these scope graphs:
+
+\begin{lstlisting}
+	scopePutType : scope * ID * Expr -> scope
+	scopePutSubst : scope * ID * (scope * Expr) -> scope
+	scopeGetName : scope * ID -> NameEntry
+	scopeGetNames : scope * ID -> list((path * (ID * NameEntry)))
+	empty_scope : -> scope
+\end{lstlisting}
+
+The \verb|scopePutType| and \verb|scopePutSubsts| relations generate a new scope given a parent scope and a type or substitution respectively.
+To query the scope graph, use \verb|scopeGetName| or \verb|scopeGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found.
+Finally, \verb|empty_scope| returns a fresh empty scope.
+
+System F, which allows a term to depend on a type (also known as Type Polymorphism or Type-Generic Functions)  has been implemented before
+
+The Calculus of Constructions is the top of the lambda cube. System F, which
 

### 2022-10-12 paper-evcs/oasics-v2021-sample-article.tex
Meeting result

@@
 \Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}}
+\begin{CCSXML}
+    <ccs2012>
+    <concept>
+    <concept_id>10011007.10011006.10011039.10011311</concept_id>
+    <concept_desc>Software and its engineering~Semantics</concept_desc>
+    <concept_significance>500</concept_significance>
+    </concept>
+    <concept>
+    <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+    <concept_desc>Software and its engineering~Functional languages</concept_desc>
+    <concept_significance>300</concept_significance>
+    </concept>
+    <concept>
+    <concept_id>10011007.10011006.10011041</concept_id>
+    <concept_desc>Software and its engineering~Compilers</concept_desc>
+    <concept_significance>300</concept_significance>
+    </concept>
+    </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Semantics}
+\ccsdesc[300]{Software and its engineering~Functional languages}
+\ccsdesc[300]{Software and its engineering~Compilers}
 
@@
 To type-check the base language, we need to store information about the names we have encountered.
-There are two different situations, names that we encounter that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.
+There are two different situations, names that we encounter that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}
 
-   we need scope graphs. This section describes how scope graphs are used.
-
+Both of these need to be stored in the scope graph.
 The scope graph only has a single type of edge, called \verb|P| (parent) edges.
@@
 \begin{lstlisting}
-	signature sorts
-	NameEntry
-	constructors
-	NameType : Expr -> NameEntry
+signature sorts	NameEntry
+constructors    NameType : Expr -> NameEntry
                 NameSubst : scope * Expr -> NameEntry
-	relations
-	name : ID -> NameEntry
+relations       name : ID -> NameEntry
 name-resolution	labels P
@@
 
-System F, which allows a term to depend on a type (also known as Type Polymorphism or Type-Generic Functions)  has been implemented before
-
-The Calculus of Constructions is the top of the lambda cube. System F, which
+\subsection{}
 

### 2022-10-12 paper-evcs/oasics-v2021-sample-article.tex
More work on aosics papers

@@
 
+% Added because the included oasics-logo-bw.pdf is pdf 1.6 and this was giving a warning
+% Feel free to remove this if it breaks anything
+\pdfminorversion=6
+
 \bibliographystyle{plainurl}% the mandatory bibstyle
@@
 
-\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\authorrunning{J. Brouwer, J. Cockx and A. Zwaan}
+
+\Copyright{Jonathan Brouwer, Jesper Cockx and Aron Zwaan} 
 \begin{CCSXML}
@@
 
-Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages.
+Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages\cite{spoofax}.
 When working with the Spoofax workbench, the Statix meta-language can be used for the specification of Static Semantics.
@@
 
-Dependently typed languages are different from other languages because they allow types to be parameterized by values.
+Dependently typed languages are different from other languages, because they allow types to be parameterized by values.
 This allows more rigorous reasoning over types and the values that are inhabited by a type.
@@
 
-The \verb|scopePutType| and \verb|scopePutSubsts| relations generate a new scope given a parent scope and a type or substitution respectively.
+The \verb|scopePutType| and \verb|scopePutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively.
 To query the scope graph, use \verb|scopeGetName| or \verb|scopeGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found.
@@
 
-\subsection{}
+\subsection{Beta Reductions}
+
+A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare.
+Beta reduction is done using Krivine abstract machines\cite{krivine}.
+We define a rule \verb|betaReduceHead|, that takes a scoped expression and a stack of applications.
+The scope acts as the environment from \cite{krivine} paper, using \verb|NameSubst| to store substitutions.
+Furthermore, we can define \verb|expectBetaEq|, which asserts that two terms are equal.
+This calls \verb|expectBetaEq_| with \verb|betaReduceHead| applied.
+
+\begin{lstlisting}
+betaReduceHead : (scope * Expr) * list((scope * Expr)) -> (scope * Expr)
+expectBetaEq   : (scope * Expr) * (scope * Expr)
+expectBetaEq(e1, e2) :- 
+    expectBetaEq_(betaReduceHead(e1, []), betaReduceHead(e2, [])).
+expectBetaEq_  : (scope * Expr) * (scope * Expr)
+\end{lstlisting}
+
+\subsection{Type-checking programs}
+
+We will define a Statix relation \verb|typeOfExpr| that takes a scope and an expression and type-checks the scope in the expression. It returns the type of the expression.
+
+\begin{lstlisting}
+typeOfExpr : scope * Expr -> Expr
+\end{lstlisting}
 

### 2022-10-13 paper-evcs/oasics-v2021-sample-article.tex
Rename scope relations + work on paper

@@
 \pdfminorversion=6
+
+\usepackage{mathpartir}
 
@@
 
-\subsection{How scope graphs are used}
+\subsection{Scope Graphs}
 
@@
 It also only has a single relation, called \verb|name|.
-This name stores a \verb|NameEntry|, which can be either a \verb|NameType|, which stores the type of a name, or a \verb|NameSubst|, which stores a substitution corresponding to a name.
+This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a substitution corresponding to a name.
 
@@
 signature sorts	NameEntry
-constructors    NameType : Expr -> NameEntry
-                NameSubst : scope * Expr -> NameEntry
+constructors    NType : Expr -> NameEntry
+                NSubst : scope * Expr -> NameEntry
 relations       name : ID -> NameEntry
@@
 \begin{lstlisting}
-scopePutType  : scope * ID * Expr -> scope
-scopePutSubst : scope * ID * (scope * Expr) -> scope
-scopeGetName  : scope * ID -> NameEntry
-scopeGetNames : scope * ID -> list((path * (ID * NameEntry)))
-empty_scope   : -> scope
+sPutType  : scope * ID * Expr -> scope
+sPutSubst : scope * ID * (scope * Expr) -> scope
+sGetName  : scope * ID -> NameEntry
+sGetNames : scope * ID -> list((path * (ID * NameEntry)))
+sEmpty    : -> scope
 \end{lstlisting}
 
-The \verb|scopePutType| and \verb|scopePutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively.
-To query the scope graph, use \verb|scopeGetName| or \verb|scopeGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found.
-Finally, \verb|empty_scope| returns a fresh empty scope.
+The \verb|sPutType| and \verb|sPutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively.
+To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found.
+Finally, \verb|sEmpty| returns a fresh empty scope.
 
@@
 We define a rule \verb|betaReduceHead|, that takes a scoped expression and a stack of applications.
-The scope acts as the environment from \cite{krivine} paper, using \verb|NameSubst| to store substitutions.
+The scope acts as the environment from \cite{krivine} paper, using \verb|NSubst| to store substitutions.
 Furthermore, we can define \verb|expectBetaEq|, which asserts that two terms are equal.
@@
 \begin{lstlisting}
-betaReduceHead : (scope * Expr) * list((scope * Expr)) -> (scope * Expr)
+betaReduceHead : (scope * Expr) * list((scope * Expr))
+                                     -> (scope * Expr)
 expectBetaEq   : (scope * Expr) * (scope * Expr)
@@
 
+We can then start defining concepts in this language. First, we present the inference rules of the language:
+
+\begin{figure}
+
+\begin{mathpar}
+\inferrule{ }{s \vdash Type() : Type()}
+
+\inferrule{ sPutSubst(s, n, (s, v)) \vdash b : t }{s \vdash Let(n, v, b) : t }
+
+\inferrule{ sGetName(s, n) = NType(t) }{ s \vdash Var(n) : t }
+
+\inferrule{ sGetName(s, n) = NSubst(se, e) \and se \vdash e : t }{ s \vdash Var(n) : t }
+
+\inferrule{ s \vdash a : at \and at \underset{\beta}{=} Type() \and a \underset{\beta}{\Rightarrow} a' \and sPutType(s, n, a') \vdash b : bt \and bt \underset{\beta}{=} Type() } { s \vdash FnType(n, a, b) : Type() }
+
+\inferrule{ s \vdash a : at \and at \underset{\beta}{=} Type() \and a \underset{\beta}{\Rightarrow} a' \and sPutType(s, n, a') \vdash b : bt } { s \vdash FnConstruct(n, a, b) : FnType(n, a', bt) }
+
+\end{mathpar}
+\caption{Rules for type-checking the Calculus of Constructions}
+\label{fig:type-check-rules}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+\inferrule{ }{ s, Type(), [] \Rightarrow s, Type() }
+
+\inferrule{ sPutSubst(s, n, (s, v)) \vdash b : t }{s \vdash Let(n, v, b) : t }
+\end{mathpar}
+\caption{Rules for beta reducing the Calculus of Constructions}
+\label{fig:beta-head-rules}
+\end{figure}
+
 \bibliography{oasics-v2021-sample-article}

### 2022-10-13 paper-evcs/oasics-v2021-sample-article.tex
More work on paper

@@
 
+% Package is used to create inference rules
 \usepackage{mathpartir}
 
-\bibliographystyle{plainurl}% the mandatory bibstyle
+\bibliographystyle{plainurl}
 
-\title{Dependently Typed Languages in Statix} %TODO Please add
+\title{Dependently Typed Languages in Statix}
 
@@
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Praesent convallis orci arcu, eu mollis dolor.
-Aliquam eleifend suscipit lacinia.
-Maecenas quam mi, porta ut lacinia sed, convallis ac dui.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse potenti.
+When working with the Spoofax workbench, the Statix meta-language can be used for the specification of Static Semantics.
+Statix aims to cover a broad range of languages and type-systems.
+However, no attempts have been made to express dependently typed languages in Statix.
+Type-checking dependent languages is challenging since there is a need to evaluate terms during type-checking.
+This paper presents how to make a dependent language by implementing the calculus of constructions in Statix.
 \end{abstract}
@@
 
+The implementation of the language is available at \url{https://github.com/JonathanBrouwer/master-thesis/}. The \verb|final-simple| branch contains the implementation that is relevant for this paper.
+
+\section{Background}
+
+There is no reason to assume that implementing the Calculus of Constructions in Statix is easy.
+For example, implementing Hindley Milner type inference is Statix is not possible, since generalization requires knowing which meta-variables are free, which is a language feature that Statix does not support.
+
 \section{Calculus of Constructions}
 
-The base language that was implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube.
+\subsection{The language}
+
+The base language that was implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}.
 One extra feature was added that is not present in the Calculus of Constructions, let bindings.
+Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
+
+\begin{lstlisting}
+Type        : Expr
+Let         : ID * Expr * Expr -> Expr
+Var         : ID -> Expr
+FnType      : ID * Expr * Expr -> Expr
+FnConstruct : ID * Expr * Expr -> Expr
+FnDestruct  : Expr * Expr -> Expr
+\end{lstlisting}
 
@@
 Beta reduction is done using Krivine abstract machines\cite{krivine}.
-We define a rule \verb|betaReduceHead|, that takes a scoped expression and a stack of applications.
+We define a rule \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
 The scope acts as the environment from \cite{krivine} paper, using \verb|NSubst| to store substitutions.
@@
                                      -> (scope * Expr)
+betaReduce : (scope * Expr) -> Expr
 expectBetaEq   : (scope * Expr) * (scope * Expr)
@@
 
-\subsection{Type-checking programs}
+\subsection{Type-checking the Calculus of Constructions}
+
+\begin{figure}
+
+	\begin{mathpar}	
+		\inferrule{ 
+		} {s \vdash Type() : Type()}
+
+		\inferrule{ 
+			s \vdash v : vt
+			\\sPutSubst(s, n, (s, v)) \vdash b : t 
+		}{s \vdash Let(n, v, b) : t }
+
+		\\
+
+		\inferrule{ 
+			sGetName(s, n) = NType(t) 
+		}{ s \vdash Var(n) : t }
+
+		\inferrule{ 
+			sGetName(s, n) = NSubst(se, e) 
+			\\ se \vdash e : t 
+		}{ s \vdash Var(n) : t }
+
+		\inferrule{ 
+			s \vdash a : at 
+			\\ at \underset{\beta}{=} Type() 
+			\\ s \vdash a \underset{\beta}{\Rightarrow} a' 
+			\\\\ sPutType(s, n, a') \vdash b : bt 
+			\\ bt \underset{\beta}{=} Type() 
+		} { s \vdash FnType(n, a, b) : Type() }
+
+		\inferrule{ 
+			s \vdash a : at 
+			\\ at \underset{\beta}{=} Type() 
+			\\ s \vdash a \underset{\beta}{\Rightarrow} a' 
+			\\\\ sPutType(s, n, a') \vdash b : bt
+		} { s \vdash FnConstruct(n, a, b) : FnType(n, a', bt) }
+
+		\inferrule{ 
+			s \vdash f : ft 
+			\\ s \vdash ft \underset{\beta h}{\Rightarrow} s' \vdash FnType(da, dt, db) 
+			\\\\ s \vdash a : at 
+			\\ at \underset{\beta}{=} sf \vdash dt
+			\\ s' \vdash db \underset{\beta}{\Rightarrow} db'
+		} { s \vdash FnDestruct(f, a) : db' }
+
+	\end{mathpar}
+	\caption{Rules for type-checking the Calculus of Constructions}
+	\label{fig:type-check-rules}
+\end{figure}
 
@@
 \end{lstlisting}
+We can then start defining type-checking rules for the language. We use some syntax in order to make the rules a bit more compact. 
+\begin{enumerate}
+	\item $s \vdash e : t$ is the same as \verb|typeOfExpr(s, e) == t|
+	\item $s1 \vdash e1 \underset{\beta}{=} s2 \vdash e2$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
+	\item $s1 \vdash e1 \underset{\beta h}{\Rightarrow} s2 \vdash e2$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)|
+	\item $s1 \vdash e1 \underset{\beta h}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
+	\item $sEmpty \vdash e$ is the same as $e$ (empty scopes can be left out)
+\end{enumerate}
 
-We can then start defining concepts in this language. First, we present the inference rules of the language:
+The inference rules above can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
+\begin{lstlisting}
+typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
+    typeOfExpr(s, v) == vt, sPutSubst(s, n, (s, v)) == s'.
+\end{lstlisting}
 
-\begin{figure}
+%\begin{figure}
+%\begin{mathpar}
+%\inferrule{ }{ s, Type(), [] \Rightarrow s, Type() }
+%
+%\inferrule{ sPutSubst(s, n, (s, v)) \vdash b : t }{s \vdash Let(n, v, b) : t }
+%\end{mathpar}
+%\caption{Rules for beta reducing the Calculus of Constructions}
+%\label{fig:beta-head-rules}
+%\end{figure}
 
-\begin{mathpar}
-\inferrule{ }{s \vdash Type() : Type()}
+\section{Extensions}
 
-\inferrule{ sPutSubst(s, n, (s, v)) \vdash b : t }{s \vdash Let(n, v, b) : t }
+\subsection{Avoiding variable capture}
+One problem with the current implementation is that it does not avoid variable capture.
+Variable capture happens when a term becomes bound because of a wrong substitution.
+For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type.
+\T : T.
+T| is \verb|T : Type -> T : T -> T|.
+This is not the correct type, and even worse, the correct type is not possible to express without renaming!
 
-\inferrule{ sGetName(s, n) = NType(t) }{ s \vdash Var(n) : t }
+The problem is that there are multiple names, and there is no way to distinguish between them. This can be solved by using scopes to distinguish names.
 
-\inferrule{ sGetName(s, n) = NSubst(se, e) \and se \vdash e : t }{ s \vdash Var(n) : t }
+\subsection{Inference}
 
-\inferrule{ s \vdash a : at \and at \underset{\beta}{=} Type() \and a \underset{\beta}{\Rightarrow} a' \and sPutType(s, n, a') \vdash b : bt \and bt \underset{\beta}{=} Type() } { s \vdash FnType(n, a, b) : Type() }
+Ideally,
 
-\inferrule{ s \vdash a : at \and at \underset{\beta}{=} Type() \and a \underset{\beta}{\Rightarrow} a' \and sPutType(s, n, a') \vdash b : bt } { s \vdash FnConstruct(n, a, b) : FnType(n, a', bt) }
+\subsection{Data Types}
 
-\end{mathpar}
-\caption{Rules for type-checking the Calculus of Constructions}
-\label{fig:type-check-rules}
-\end{figure}
+\section{Related Work}
+The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
+In section 2.5 of Scopes as Types \cite{scopes_as_types}, an implementation of System F is shown that does lazy substitutions, by using scopes as types.
+It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.
 
-\begin{figure}
-\begin{mathpar}
-\inferrule{ }{ s, Type(), [] \Rightarrow s, Type() }
+Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language.
+The pi-forall language\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper.
+In principle, the implementations are very similar.
+For example, figure 3 of pi-forall is similar to figure \ref{fig:type-check-rules} from this paper.
+The primary difference is that they use a bidirectional type system, whereas this paper does not.
 
-\inferrule{ sPutSubst(s, n, (s, v)) \vdash b : t }{s \vdash Let(n, v, b) : t }
-\end{mathpar}
-\caption{Rules for beta reducing the Calculus of Constructions}
-\label{fig:beta-head-rules}
-\end{figure}
+\section{Conclusion}
+
+We have demonstrated that the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph.
+We have also presented a few extensions to the Calculus of Constructions and discussed how they could be implemented.
 

### 2022-10-14 paper-evcs/oasics-v2021-sample-article.tex
Work on paper

@@
 
-The implementation of the language is available at \url{https://github.com/JonathanBrouwer/master-thesis/}. The \verb|final-simple| branch contains the implementation that is relevant for this paper.
+The implementation of the language is available on github. The \verb|final-simple| branch contains the implementation that is relevant for this paper. \\
+\url{https://github.com/JonathanBrouwer/master-thesis/}
 
@@
 	\item $s1 \vdash e1 \underset{\beta h}{\Rightarrow} s2 \vdash e2$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)|
-	\item $s1 \vdash e1 \underset{\beta h}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
+	\item $s1 \vdash e1 \underset{\beta}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
 	\item $sEmpty \vdash e$ is the same as $e$ (empty scopes can be left out)
@@
 
-%\begin{figure}
-%\begin{mathpar}
-%\inferrule{ }{ s, Type(), [] \Rightarrow s, Type() }
-%
-%\inferrule{ sPutSubst(s, n, (s, v)) \vdash b : t }{s \vdash Let(n, v, b) : t }
-%\end{mathpar}
-%\caption{Rules for beta reducing the Calculus of Constructions}
-%\label{fig:beta-head-rules}
-%\end{figure}
+\section{Extensions}
+This section will discuss some further ideas that can be explored to build upon what has already been shown.
+These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.
 
-\section{Extensions}
-
-\subsection{Avoiding variable capture}
+\paragraph*{Avoiding variable capture}
 One problem with the current implementation is that it does not avoid variable capture.
@@
 This is not the correct type, and even worse, the correct type is not possible to express without renaming!
+The problem is that there are multiple names, and there is no way to distinguish between them.
+This can be solved by using scopes to distinguish names.
 
-The problem is that there are multiple names, and there is no way to distinguish between them. This can be solved by using scopes to distinguish names.
+\paragraph*{Term Inference}
+In dependently typed language, we can infer the value of types.
+Ideally, we would like to infer the most general type possible.
+However, this kind of analysis is not possible in Statix because you cannot reason over whether meta-variables are instantiated or not.
+We implemented an approximation to inference that can infer most types.
 
-\subsection{Inference}
+\paragraph*{Inductive Data Types}
+Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types.
+Both of these are also fully possible to express in Statix.
 
-Ideally,
-
-\subsection{Data Types}
+\paragraph*{Semantic Code Completion}
+Finally, we explored how semantic code completion presented in \cite{codecompletion} applies to dependently typed languages.
+It functions works well, only showing completions that are semantically possible.
 

### 2022-10-14 paper-evcs/oasics-v2021-sample-article.tex
More work on evcs paper

@@
 \usepackage{mathpartir}
+
+% Commands used
+\newcommand{\bhr}{\underset{\beta h}{\Rightarrow}}
 
@@
 
-\section{Background}
-
-There is no reason to assume that implementing the Calculus of Constructions in Statix is easy.
-For example, implementing Hindley Milner type inference is Statix is not possible, since generalization requires knowing which meta-variables are free, which is a language feature that Statix does not support.
-
 \section{Calculus of Constructions}
@@
 Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
+The abstract syntax of the language is available in figure \ref{fig:syntax}.
 
+\begin{figure}[h]
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{The syntax for the base language. FnConstruct is a lambda function, FnDestruct is application of a lambda function.}
+\label{fig:syntax}
+\end{figure}
 
@@
 A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare.
-Beta reduction is done using Krivine abstract machines\cite{krivine}.
-We define a rule \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
+Beta reduction is done using a Krivine abstract machine\cite{krivine}.
+
+We need to define multiple relations that will be used later for type-checking.
+First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
 The scope acts as the environment from \cite{krivine} paper, using \verb|NSubst| to store substitutions.
-Furthermore, we can define \verb|expectBetaEq|, which asserts that two terms are equal.
-This calls \verb|expectBetaEq_| with \verb|betaReduceHead| applied.
+All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
+We use the syntax $s1 \vdash e1 \bhr s2 \vdash e2$ to express \verb|betaReduceHead((s1, e1)) == (s2, e2)|.
+Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
+One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
+
+Additionally, we define \verb|betaReduce| fully beta reduces a term.
+It works by first calling \verb|betaReduceHead| and then matching on the head, calling \verb|betaReduce| on the sub-expressions of the head recursively.
+
+Finally, we define \verb|expectBetaEq|.
+This rule first beta reduces the heads of both sides, and then compares them.
+If the head is not the same, the rule fails.
+Otherwise, it recurses on the sub-expressions.
+One special case is when comparing two \verb|FnConstruct|s.
+Here we need to take into account alpha equality, that it, two expressions which only differ in the names that they use should be considered equal.
+We implement this by substituting in the body of the functions, replacing their argument names with placeholders.
+
+\begin{figure}[h]
+	\begin{mathpar}	
+		\inferrule{ 
+		} {s \vdash Type(), [] \bhr s \vdash Type()}
+
+		\inferrule{ 
+		}{s \vdash Let(n, v, b), [] \bhr sPutSubst(s, n, (s, v)) \vdash b }
+
+		\inferrule{ 
+			sGetName(s, n) = NSubst(se, e) 
+			\\ se \vdash e, t \bhr se', e' 
+		}{ s \vdash Var(n), t \bhr se', e' }
+
+		\inferrule{ 
+			sGetName(s, n) = NType(t) 
+		}{ s \vdash Var(n), t \bhr rebuild(s, Var(n), t)  }
+
+		\inferrule{ 
+		} { s \vdash FnType(n, a, b), [] \bhr s \vdash FnType(n, a, b) }
+
+		\inferrule{ 
+		} { s \vdash FnConstruct(n, a, b), [] \bhr s \vdash FnConstruct(n, a, b) }
+
+		\inferrule{ 
+			sPutSubst(s, n, a) \vdash b, ts \bhr s' \vdash e'
+		} { s \vdash FnConstruct(n, a, b), t :: ts \bhr s' \vdash e' }
+
+		\inferrule{ 
+			s \vdash f, a :: ts \bhr s' \vdash e'
+		} { s \vdash FnDestruct(f, a), ts \bhr s' \vdash e' }
+
+	\end{mathpar}
+	\caption{Rules for beta head reducing the Calculus of Constructions}
+	\label{fig:beta-head-reduce-rules}
+\end{figure}
+
+\newpage
+\subsection{Type-checking the Calculus of Constructions}
+
+We will define a Statix relation \verb|typeOfExpr| that takes a scope and an expression and type-checks the scope in the expression. It returns the type of the expression.
 
 \begin{lstlisting}
-betaReduceHead : (scope * Expr) * list((scope * Expr))
-                                     -> (scope * Expr)
-betaReduce : (scope * Expr) -> Expr
-expectBetaEq   : (scope * Expr) * (scope * Expr)
-expectBetaEq(e1, e2) :- 
-    expectBetaEq_(betaReduceHead(e1, []), betaReduceHead(e2, [])).
-expectBetaEq_  : (scope * Expr) * (scope * Expr)
+typeOfExpr : scope * Expr -> Expr
+\end{lstlisting}
+We can then start defining type-checking rules for the language. We use some syntax in order to make the rules a bit more compact. 
+\begin{enumerate}
+	\item $s \vdash e : t$ is the same as \verb|typeOfExpr(s, e) == t|
+	\item $s1 \vdash e1 \underset{\beta}{=} s2 \vdash e2$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
+	\item $s1 \vdash e1 \bhr s2 \vdash e2$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)|
+	\item $s1 \vdash e1 \underset{\beta}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
+	\item $sEmpty \vdash e$ is the same as $e$ (empty scopes can be left out)
+\end{enumerate}
+
+The inference rules above can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
+\begin{lstlisting}
+typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
+    typeOfExpr(s, v) == vt, sPutSubst(s, n, (s, v)) == s'.
 \end{lstlisting}
 
-\subsection{Type-checking the Calculus of Constructions}
-
-\begin{figure}
-
+\begin{figure}[h]
 	\begin{mathpar}	
@@
 			s \vdash f : ft 
-			\\ s \vdash ft \underset{\beta h}{\Rightarrow} s' \vdash FnType(da, dt, db) 
+			\\ s \vdash ft \bhr s' \vdash FnType(da, dt, db) 
 			\\\\ s \vdash a : at 
@@
 
-We will define a Statix relation \verb|typeOfExpr| that takes a scope and an expression and type-checks the scope in the expression. It returns the type of the expression.
-
-\begin{lstlisting}
-typeOfExpr : scope * Expr -> Expr
-\end{lstlisting}
-We can then start defining type-checking rules for the language. We use some syntax in order to make the rules a bit more compact. 
-\begin{enumerate}
-	\item $s \vdash e : t$ is the same as \verb|typeOfExpr(s, e) == t|
-	\item $s1 \vdash e1 \underset{\beta}{=} s2 \vdash e2$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $s1 \vdash e1 \underset{\beta h}{\Rightarrow} s2 \vdash e2$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)|
-	\item $s1 \vdash e1 \underset{\beta}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
-	\item $sEmpty \vdash e$ is the same as $e$ (empty scopes can be left out)
-\end{enumerate}
-
-The inference rules above can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
-\begin{lstlisting}
-typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
-    typeOfExpr(s, v) == vt, sPutSubst(s, n, (s, v)) == s'.
-\end{lstlisting}
-
+\newpage
 \section{Extensions}

### 2022-10-14 paper-evcs/oasics-v2021-sample-article.tex
Minor changes

@@
 \begin{abstract}
-When working with the Spoofax workbench, the Statix meta-language can be used for the specification of Static Semantics.
+When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
 Statix aims to cover a broad range of languages and type-systems.
@@
 Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages\cite{spoofax}.
-When working with the Spoofax workbench, the Statix meta-language can be used for the specification of Static Semantics.
+When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
 To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems.
@@
 
-The implementation of the language is available on github. The \verb|final-simple| branch contains the implementation that is relevant for this paper. \\
+The implementation of the language is available on GitHub. The \verb|final-simple| branch contains the implementation that is relevant for this paper. \\
 \url{https://github.com/JonathanBrouwer/master-thesis/}
@@
 
-The base language that was implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}.
+The base language that has been implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}.
 One extra feature was added that is not present in the Calculus of Constructions, let bindings.
@@
 It also only has a single relation, called \verb|name|.
-This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a substitution corresponding to a name.
+This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.
 
-\begin{lstlisting}
-signature sorts	NameEntry
-constructors    NType : Expr -> NameEntry
-                NSubst : scope * Expr -> NameEntry
-relations       name : ID -> NameEntry
-name-resolution	labels P
-\end{lstlisting}
-
-These are all the definitions we will need to type-check programs. Next, we will introduce some Statix relations that can be used to interact with these scope graphs:
+Next, we will introduce some Statix relations that can be used to interact with these scope graphs:
 
@@
 \end{lstlisting}
-
 The \verb|sPutType| and \verb|sPutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively.
@@
 
+\newpage
 \subsection{Beta Reductions}
@@
 All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
-We use the syntax $s1 \vdash e1 \bhr s2 \vdash e2$ to express \verb|betaReduceHead((s1, e1)) == (s2, e2)|.
+We use the syntax $s1 \vdash e1, t \bhr s2 \vdash e2$ to express \verb|betaReduceHead((s1, e1)) == (s2, e2)|.
 Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
@@
 		\inferrule{ 
-		}{s \vdash Let(n, v, b), [] \bhr sPutSubst(s, n, (s, v)) \vdash b }
+			sPutSubst(s, n, (s, v)) \vdash b, t \bhr s' \vdash e'
+		}{s \vdash Let(n, v, b), t \bhr s' \vdash e' }
 

### 2022-10-25 paper-evcs/paper.tex
Fix checking multiple parameters

--- 2022-10-14 paper-evcs/oasics-v2021-sample-article.tex
@@
 \begin{abstract}
-When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
-Statix aims to cover a broad range of languages and type-systems.
-However, no attempts have been made to express dependently typed languages in Statix.
-Type-checking dependent languages is challenging since there is a need to evaluate terms during type-checking.
-This paper presents how to make a dependent language by implementing the calculus of constructions in Statix.
+Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone.
+The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system.
+However, so far Statix has not yet been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking.
+In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
+While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.
 \end{abstract}

### 2022-10-26 paper-evcs/paper.tex
Process feedback on paper

@@
 Dependently typed languages are different from other languages, because they allow types to be parameterized by values.
-This allows more rigorous reasoning over types and the values that are inhabited by a type.
+This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree.
 This expressiveness also makes dependent type systems more complicated to implement.
@@
 We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix.
-The goal is not to show that Statix can implement it, but that implementing it is easier in Statix than in a general-purpose programming language.
+The goal is not to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.
 
-We will first show the base language and explain the way that Statix was used to implement this language. Next, we will explore several features and see how well they can be expressed in Statix.
+In this paper we show that Statix is already usable as a tool for implementing dependent types.
+There were some challenges in implementing the dependent type system, which this paper also discusses.
+Finally, we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
 
-The implementation of the language is available on GitHub. The \verb|final-simple| branch contains the implementation that is relevant for this paper. \\
+The implementation of the language is available on GitHub.\\
 \url{https://github.com/JonathanBrouwer/master-thesis/}
@@
 
+In this section, we will describe how to implement a dependently typed language in Statix.
+In section \ref{sec:coc-syntax} we will describe the syntax of the language, then in section \ref{sec:coc-scopes} we will describe how scope graphs are used to type check the language.
+Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type-check the language.
+
 \subsection{The language}
+\label{sec:coc-syntax}
 
@@
 
+An example program is the following, which defines a polymorphic identity function and applies it to a function:
+
+\begin{lstlisting}
+let f = \T: Type. \x: T. x;
+f (_: Type -> Type) (\x: Type. x)
+\end{lstlisting}
+The AST of this program would be:
+\begin{lstlisting}
+Let(
+  "f", 
+  FnConstruct("T", Type(), FnConstruct("x", Var("T"), Var("x"))),
+  FnDestruct(
+	FnDestruct(Var("f"), FnConstruct("_", Type(), Type())),
+	FnConstruct("x", Type(), Var("x"))
+  )
+)
+\end{lstlisting}
+
 \subsection{Scope Graphs}
+\label{sec:coc-scopes}
 
-To type-check the base language, we need to store information about the names we have encountered.
-There are two different situations, names that we encounter that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}
+To type-check the base language, we need to store information about the names that are in scope at each point in the program.
+There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}
 
-Both of these need to be stored in the scope graph.
-The scope graph only has a single type of edge, called \verb|P| (parent) edges.
+In Statix, all this information can be stored in a \verb|scope graph|\cite{scope_graphs}, which is a feature of Statix.
+We only use a single type of edge, called \verb|P| (parent) edges.
 It also only has a single relation, called \verb|name|.
@@
 
-\newpage
 \subsection{Beta Reductions}
+\label{sec:coc-dynsyms}
 
 A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare.
-Beta reduction is done using a Krivine abstract machine\cite{krivine}.
+
+We implement beta reduction using a Krivine abstract machine\cite{krivine}.
+This turned out to be the more natural way of expressing this over substitution-based evaluation relation.
+We originally tried to implement a substitution-based evaluation relation, which works fine for the base language.
+However, it runs into trouble when implementing inductive data types, more information about this will be in the full master thesis.
+Furthermore, abstract machines are usually more efficient than substitution-based approaches.
+
+Note that the evaluator uses names rather than De Bruijn indices.
+In conventional dependently typed languages, evaluation is often done using De Bruijn indices.
+However, since the aim of this paper is to use the features of Statix, and scope graphs work based on names, we chose to use names.
 
@@
 			sGetName(s, n) = NSubst(se, e) 
-			\\ se \vdash e, t \bhr se', e' 
-		}{ s \vdash Var(n), t \bhr se', e' }
+			\\ se \vdash e, t \bhr se' \vdash  e' 
+		}{ s \vdash Var(n), t \bhr se' \vdash e' }
 
@@
 
-\newpage
 \subsection{Type-checking the Calculus of Constructions}
+\label{sec:coc-typecheck}
 
@@
 \end{lstlisting}
-We can then start defining type-checking rules for the language. We use some syntax in order to make the rules a bit more compact. 
+We can then start defining type-checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
 \begin{enumerate}
@@
 
-\newpage
-\section{Extensions}
-This section will discuss some further ideas that can be explored to build upon what has already been shown.
-These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.
-
-\paragraph*{Avoiding variable capture}
+\subsection{Avoiding variable capture}
 One problem with the current implementation is that it does not avoid variable capture.
@@
 
+\section{Extensions}
+This section will discuss some further ideas that can be explored to build upon what has already been shown.
+These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.
+
 \paragraph*{Term Inference}
@@
 We implemented an approximation to inference that can infer most types.
+\begin{lstlisting}
+(\x: _. x) true
+\end{lstlisting}
 
@@
 Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types.
-Both of these are also fully possible to express in Statix.
+We can also generate eliminators for these data types to use their values.
+All of this has been implemented in Statix.
+\begin{lstlisting}
+data Maybe (T : Type) = 
+    None : Maybe T,
+    Some : _: T -> Maybe T;
+\end{lstlisting}
 
@@
 
-\bibliography{oasics-v2021-sample-article}
+\bibliography{paper}
 

### 2022-10-26 paper-evcs/paper.tex
Add a few more sentences
Improve syntax of inference rules

@@
 % Commands used
-\newcommand{\bhr}{\underset{\beta h}{\Rightarrow}}
+\newcommand{\scope}[2]{\langle#1 \: | \: #2\rangle}
+\newcommand{\bhr}[3]{ #1 \: #2 \underset{\beta h}{\Rightarrow} #3 }
+\newcommand{\toe}[3]{ \scope{#1}{#2} : #3 }
+\newcommand{\bred}[3]{ \scope{#1}{#2} \underset{\beta}{\Rightarrow} #3 }
+\newcommand{\beq}[2]{ #1 \underset{\beta}{=} #2 }
 
@@
 However, since the aim of this paper is to use the features of Statix, and scope graphs work based on names, we chose to use names.
+It would also stop us from using editor services that rely on \verb|.ref| annotations, such as renaming.
 
@@
 All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
-We use the syntax $s1 \vdash e1, t \bhr s2 \vdash e2$ to express \verb|betaReduceHead((s1, e1)) == (s2, e2)|.
+We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|.
 Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
@@
 	\begin{mathpar}	
+		% Type rule
 		\inferrule{ 
-		} {s \vdash Type(), [] \bhr s \vdash Type()}
+		} { 
+			\bhr
+			{ \scope{s}{Type} }
+			{ [] }
+			{ \scope{s}{Type}} 
+		}
 
+		% Let rule
 		\inferrule{ 
-			sPutSubst(s, n, (s, v)) \vdash b, t \bhr s' \vdash e'
-		}{s \vdash Let(n, v, b), t \bhr s' \vdash e' }
+			\bhr 
+			{ \scope{ sPutSubst(s, n, (s, v))}{b} } 
+			{ t }
+			{ \scope{ s' } { e' } }
+		}{
+			\bhr
+			{ \scope{ s }{ Let(n, v, b) } }
+			{ t }
+			{ \scope { s' } { e' } }
+		}
 
+		% Var rule - NSubst
 		\inferrule{ 
 			sGetName(s, n) = NSubst(se, e) 
-			\\ se \vdash e, t \bhr se' \vdash  e' 
-		}{ s \vdash Var(n), t \bhr se' \vdash e' }
+			\\ \bhr
+			{\scope{se}{e}}
+			{t}
+			{\scope{se'}{e'}} 
+		}{
+			\bhr
+			{\scope{s}{Var(n)}}
+			{t}
+			{\scope{se'}{e'}} 
+		}
 
+		% Var rule - NType
 		\inferrule{ 
 			sGetName(s, n) = NType(t) 
-		}{ s \vdash Var(n), t \bhr rebuild(s, Var(n), t)  }
+		}{ 
+			\bhr
+			{\scope{s}{Var(n)}}
+			{t}
+			{rebuild(s, Var(n), t)} 
+		}
 
+		% FnType rule
 		\inferrule{ 
-		} { s \vdash FnType(n, a, b), [] \bhr s \vdash FnType(n, a, b) }
+		} { 
+			\bhr
+			{\scope{s}{FnType(n, a, b)}}
+			{[]}
+			{\scope{s}{FnType(n, a, b)}} 
+		}
 
+		% FnConstruct rule - No args
 		\inferrule{ 
-		} { s \vdash FnConstruct(n, a, b), [] \bhr s \vdash FnConstruct(n, a, b) }
+		} { 
+			\bhr
+			{\scope{s}{FnConstruct(n, a, b)}}
+			{[]}
+			{\scope{s}{FnConstruct(n, a, b)}} 
+		}
 
+		% FnConstruct rule - Args
 		\inferrule{ 
-			sPutSubst(s, n, a) \vdash b, ts \bhr s' \vdash e'
-		} { s \vdash FnConstruct(n, a, b), t :: ts \bhr s' \vdash e' }
+			\bhr
+			{\scope{sPutSubst(s, n, a)}{b}}
+			{ts}
+			{\scope{s'}{e'}} 
+		} { 
+			\bhr
+			{\scope{s}{FnConstruct(n, a, b)}}
+			{(t::ts)}
+			{\scope{s'}{e'}} 
+		}
 
+		% FnDestruct rule
 		\inferrule{ 
-			s \vdash f, a :: ts \bhr s' \vdash e'
-		} { s \vdash FnDestruct(f, a), ts \bhr s' \vdash e' }
-
+			\bhr
+			{\scope{s}{f}}
+			{(a::ts)}
+			{\scope{s'}{e'}} 
+		} { 
+			\bhr
+			{\scope{s}{FnDestruct(f, a)}}
+			{ts}
+			{\scope{s'}{e'}} 
+		}
 	\end{mathpar}
@@
 \begin{enumerate}
-	\item $s \vdash e : t$ is the same as \verb|typeOfExpr(s, e) == t|
-	\item $s1 \vdash e1 \underset{\beta}{=} s2 \vdash e2$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $s1 \vdash e1 \bhr s2 \vdash e2$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)|
-	\item $s1 \vdash e1 \underset{\beta}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
-	\item $sEmpty \vdash e$ is the same as $e$ (empty scopes can be left out)
+	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
+	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
+	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
+	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
+	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
 \end{enumerate}
@@
 	\begin{mathpar}	
+		% Type rule
 		\inferrule{ 
-		} {s \vdash Type() : Type()}
+		} {
+			\toe{s}{Type()}{Type()}
+		}
 
+		% Let rule
 		\inferrule{ 
-			s \vdash v : vt
-			\\sPutSubst(s, n, (s, v)) \vdash b : t 
-		}{s \vdash Let(n, v, b) : t }
+			\toe{s}{v}{vt}
+			\\ \toe{sPutSubst(s, n, (s, v))}{b}{t}
+		}{
+			\toe{s}{Let(n, v, b)}{t}
+		}
 
@@
 
+		% Var rule - NType
 		\inferrule{ 
 			sGetName(s, n) = NType(t) 
-		}{ s \vdash Var(n) : t }
+		}{ 
+			\toe{s}{Var(n)}{t}
+		}
 
+		% Var rule - NSubst
 		\inferrule{ 
 			sGetName(s, n) = NSubst(se, e) 
-			\\ se \vdash e : t 
-		}{ s \vdash Var(n) : t }
+			\\ \toe{se}{e}{t}
+		}{
+		 	\toe{s}{Var(n)}{t}
+		}
 
+		% FnType rule
 		\inferrule{ 
-			s \vdash a : at 
-			\\ at \underset{\beta}{=} Type() 
-			\\ s \vdash a \underset{\beta}{\Rightarrow} a' 
-			\\\\ sPutType(s, n, a') \vdash b : bt 
-			\\ bt \underset{\beta}{=} Type() 
-		} { s \vdash FnType(n, a, b) : Type() }
+			\toe{s}{a}{at}
+			\\ \beq{at}{Type()} 
+			\\ \bred{s}{a}{a'}
+			\\\\  \toe{sPutType(s, n, a')}{b}{bt}
+			\\ \beq{bt}{Type()}
+		} { 
+			\toe{s}{FnType(n, a, b)}{Type()}
+		}
 
+		% Force these to be on the same line, using negative spaces
+		\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!
+
+		% FnConstruct rule
 		\inferrule{ 
-			s \vdash a : at 
-			\\ at \underset{\beta}{=} Type() 
-			\\ s \vdash a \underset{\beta}{\Rightarrow} a' 
-			\\\\ sPutType(s, n, a') \vdash b : bt
-		} { s \vdash FnConstruct(n, a, b) : FnType(n, a', bt) }
+			\toe{s}{a}{at}
+			\\ \beq{at}{Type()}
+			\\ \bred{s}{a}{a'} 
+			\\\\ \toe{sPutType(s, n, a')}{b}{bt}
+		} { 
+			\toe{s}{FnConstruct(n, a, b)}{FnType(n, a', bt)}
+		}
 
+		% FnDestruct rule
 		\inferrule{ 
-			s \vdash f : ft 
-			\\ s \vdash ft \bhr s' \vdash FnType(da, dt, db) 
-			\\\\ s \vdash a : at 
-			\\ at \underset{\beta}{=} sf \vdash dt
-			\\ s' \vdash db \underset{\beta}{\Rightarrow} db'
-		} { s \vdash FnDestruct(f, a) : db' }
+			\toe{s}{f}{ft}
+			\\ \bhr
+				{\scope{s}{ft}}
+				{[]}
+				{\scope{s'}{FnType(da, dt, db)}} 
+			\\\\ \toe{s}{a}{at} 
+			\\ \beq{at}{\scope{sf}{dt}}
+			\\ \bred{s'}{db}{db'}
+		} { 
+			\toe{s}{FnDestruct(f, a)}{db'}
+	 	}
 

### 2022-10-26 paper-evcs/paper.tex
Minor changes
Add \co formatting
Process Ola's feedback

@@
 \newcommand{\beq}[2]{ #1 \underset{\beta}{=} #2 }
+\newcommand{\co}[1]{ \mathsf{#1} }
 
@@
 
-In this paper we show that Statix is already usable as a tool for implementing dependent types.
+In section \ref{sec:coc} of this paper we show that Statix is already usable as a tool for implementing dependent types.
 There were some challenges in implementing the dependent type system, which this paper also discusses.
-Finally, we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
+Finally, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
 
@@
 \section{Calculus of Constructions}
+\label{sec:coc}
 
@@
 
-We implement beta reduction using a Krivine abstract machine\cite{krivine}.
+We implemented beta reduction using a Krivine abstract machine\cite{krivine}.
 This turned out to be the more natural way of expressing this over substitution-based evaluation relation.
-We originally tried to implement a substitution-based evaluation relation, which works fine for the base language.
-However, it runs into trouble when implementing inductive data types, more information about this will be in the full master thesis.
-Furthermore, abstract machines are usually more efficient than substitution-based approaches.
+We originally tried to implement the latter, which works fine for the base language.
+However, it runs into trouble when implementing inductive data types; more information about this will be in the full master thesis.
+An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.
 
@@
 First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
-The scope acts as the environment from \cite{krivine} paper, using \verb|NSubst| to store substitutions.
+The scope acts as the environment from \cite{krivine}, using \verb|NSubst| to store substitutions.
 All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
@@
 
-Additionally, we define \verb|betaReduce| fully beta reduces a term.
+Additionally, we define \verb|betaReduce| which fully beta reduces a term.
 It works by first calling \verb|betaReduceHead| and then matching on the head, calling \verb|betaReduce| on the sub-expressions of the head recursively.
@@
 One special case is when comparing two \verb|FnConstruct|s.
-Here we need to take into account alpha equality, that it, two expressions which only differ in the names that they use should be considered equal.
+Here we need to take into account alpha equality: two expressions which only differ in the names that they use should be considered equal.
 We implement this by substituting in the body of the functions, replacing their argument names with placeholders.
@@
 			\bhr
-			{ \scope{s}{Type} }
+			{ \scope{s}{\co{Type}()} }
 			{ [] }
-			{ \scope{s}{Type}} 
+			{ \scope{s}{\co{Type}()}} 
 		}
@@
 			\bhr 
-			{ \scope{ sPutSubst(s, n, (s, v))}{b} } 
+			{ \scope{ \co{sPutSubst}(s, n, (s, v))}{b} } 
 			{ t }
@@
 			\bhr
-			{ \scope{ s }{ Let(n, v, b) } }
+			{ \scope{ s }{ \co{Let}(n, v, b) } }
 			{ t }
@@
 		\inferrule{ 
-			sGetName(s, n) = NSubst(se, e) 
+			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
 			\\ \bhr
@@
 			\bhr
-			{\scope{s}{Var(n)}}
+			{\scope{s}{\co{Var}(n)}}
 			{t}
@@
 		\inferrule{ 
-			sGetName(s, n) = NType(t) 
+			\co{sGetName}(s, n) = \co{NType}(t) 
 		}{ 
 			\bhr
-			{\scope{s}{Var(n)}}
+			{\scope{s}{\co{Var}(n)}}
 			{t}
-			{rebuild(s, Var(n), t)} 
+			{\co{rebuild}(s, \co{Var}(n), t)} 
 		}
@@
 			\bhr
-			{\scope{s}{FnType(n, a, b)}}
+			{\scope{s}{\co{FnType}(n, a, b)}}
 			{[]}
-			{\scope{s}{FnType(n, a, b)}} 
+			{\scope{s}{\co{FnType}(n, a, b)}} 
 		}
@@
 			\bhr
-			{\scope{s}{FnConstruct(n, a, b)}}
+			{\scope{s}{\co{FnConstruct}(n, a, b)}}
 			{[]}
-			{\scope{s}{FnConstruct(n, a, b)}} 
+			{\scope{s}{\co{FnConstruct}(n, a, b)}} 
 		}
@@
 			\bhr
-			{\scope{sPutSubst(s, n, a)}{b}}
+			{\scope{\co{sPutSubst}(s, n, a)}{b}}
 			{ts}
@@
 			\bhr
-			{\scope{s}{FnConstruct(n, a, b)}}
+			{\scope{s}{\co{FnConstruct}(n, a, b)}}
 			{(t::ts)}
@@
 			\bhr
-			{\scope{s}{FnDestruct(f, a)}}
+			{\scope{s}{\co{FnDestruct}(f, a)}}
 			{ts}
@@
 		} {
-			\toe{s}{Type()}{Type()}
+			\toe{s}{\co{Type}()}{\co{Type}()}
 		}
@@
 			\toe{s}{v}{vt}
-			\\ \toe{sPutSubst(s, n, (s, v))}{b}{t}
+			\\ \toe{\co{sPutSubst}(s, n, (s, v))}{b}{t}
 		}{
-			\toe{s}{Let(n, v, b)}{t}
+			\toe{s}{\co{Let}(n, v, b)}{t}
 		}
@@
 		\inferrule{ 
-			sGetName(s, n) = NType(t) 
+			\co{sGetName}(s, n) = \co{NType}(t) 
 		}{ 
-			\toe{s}{Var(n)}{t}
+			\toe{s}{\co{Var}(n)}{t}
 		}
@@
 		\inferrule{ 
-			sGetName(s, n) = NSubst(se, e) 
+			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
 			\\ \toe{se}{e}{t}
 		}{
-		 	\toe{s}{Var(n)}{t}
+		 	\toe{s}{\co{Var}(n)}{t}
 		}
@@
 			\toe{s}{a}{at}
-			\\ \beq{at}{Type()} 
+			\\ \beq{at}{\co{Type}()} 
 			\\ \bred{s}{a}{a'}
-			\\\\  \toe{sPutType(s, n, a')}{b}{bt}
-			\\ \beq{bt}{Type()}
+			\\\\  \toe{\co{sPutType}(s, n, a')}{b}{bt}
+			\\ \beq{bt}{\co{Type}()}
 		} { 
-			\toe{s}{FnType(n, a, b)}{Type()}
+			\toe{s}{\co{FnType}(n, a, b)}{\co{Type}()}
 		}
@@
 			\toe{s}{a}{at}
-			\\ \beq{at}{Type()}
+			\\ \beq{at}{\co{Type}()}
 			\\ \bred{s}{a}{a'} 
-			\\\\ \toe{sPutType(s, n, a')}{b}{bt}
+			\\\\ \toe{\co{sPutType}(s, n, a')}{b}{bt}
 		} { 
-			\toe{s}{FnConstruct(n, a, b)}{FnType(n, a', bt)}
+			\toe{s}{\co{FnConstruct}(n, a, b)}{\co{FnType}(n, a', bt)}
 		}
@@
 				{[]}
-				{\scope{s'}{FnType(da, dt, db)}} 
+				{\scope{s'}{\co{FnType}(da, dt, db)}} 
 			\\\\ \toe{s}{a}{at} 
@@
 		} { 
-			\toe{s}{FnDestruct(f, a)}{db'}
+			\toe{s}{\co{FnDestruct}(f, a)}{db'}
 	 	}
@@
 This can be solved by using scopes to distinguish names.
+A full explanation of this will be in the master thesis.
 
 \section{Extensions}
+\label{sec:extensions}
 This section will discuss some further ideas that can be explored to build upon what has already been shown.
@@
 \paragraph*{Semantic Code Completion}
-Finally, we explored how semantic code completion presented in \cite{codecompletion} applies to dependently typed languages.
-It functions works well, only showing completions that are semantically possible.
+Finally, we explored how semantic code completion presented in \cite{codecompletion} applies to dependently typed languages. It works well, only showing completions that are semantically possible.
+
+For example, in the following code it would only suggest expressions that can be booleans:
+\begin{lstlisting}
+let f = \b: Bool. b;
+f {{Expr}}
+\end{lstlisting}
 
@@
 The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
-In section 2.5 of Scopes as Types \cite{scopes_as_types}, an implementation of System F is shown that does lazy substitutions, by using scopes as types.
+In section 2.5 of Scopes as Types \cite{scopes_as_types}, an implementation of System F is presented; it does lazy substitutions, by using scopes as types.
 It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.
@@
 In principle, the implementations are very similar.
-For example, figure 3 of pi-forall is similar to figure \ref{fig:type-check-rules} from this paper.
+For example, the inference rules presented in \cite{pi_forall} are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper.
 The primary difference is that they use a bidirectional type system, whereas this paper does not.

### 2022-10-27 paper-evcs/paper.tex
Add Jespers text

@@
 \newcommand{\co}[1]{ \mathsf{#1} }
+
+\DeclareUnicodeCharacter{03BB}{L}
+\DeclareUnicodeCharacter{03A0}{M}
 
@@
 
+There exist several so-called \emph{logical frameworks}, tools designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}.
+Since these tools are designed specifically for the task, implementing the type system takes less effort in them compared to Spoofax, but for other tasks such as defining a parser or editor services they are not as well equipped.
+Some logical frameworks such as Twelf and Dedukti support Miller's \emph{higher-order pattern unification}~\cite{Miller89}, which can be used as a more powerful way of inferring implicit arguments than the first-order unification built into Statix.
+Andromeda 2 also supports \emph{extensionality rules} that can match on the type of an equality.
+We expect that adding extensionality rules to our implementation would be possible to do in Statix, but we leave an actual implementation to future work.
+
 \section{Conclusion}

### 2022-10-28 paper-evcs/paper.tex
Fix eliminators of recursive datatypes
Aron's feedback

@@
 \newcommand{\co}[1]{ \mathsf{#1} }
-
-\DeclareUnicodeCharacter{03BB}{L}
-\DeclareUnicodeCharacter{03A0}{M}
 
@@
 The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system.
-However, so far Statix has not yet been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking.
+However, so far Statix has not been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking.
 In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
@@
 
-Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages\cite{spoofax}.
+Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages \cite{spoofax}.
 When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
-To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems.
+To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type systems.
 However, no attempts have been made to express dependently typed languages in Statix.
@@
 We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix.
-The goal is not to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.
+The goal is not only to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.
 
@@
 Finally, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
-
-The implementation of the language is available on GitHub.\\
-\url{https://github.com/JonathanBrouwer/master-thesis/}
 
@@
 In section \ref{sec:coc-syntax} we will describe the syntax of the language, then in section \ref{sec:coc-scopes} we will describe how scope graphs are used to type check the language.
-Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type-check the language.
+Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type check the language.
+The implementation of the language is available on GitHub.
+\url{https://github.com/JonathanBrouwer/master-thesis/}
 
@@
 The base language that has been implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}.
-One extra feature was added that is not present in the Calculus of Constructions, let bindings.
+One extra feature was added that is not present in the Calculus of Constructions: let bindings.
 Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
@@
 \end{lstlisting}
-The AST of this program would be:
+The AST of this program in Aterm format would be:
 \begin{lstlisting}
@@
 
-To type-check the base language, we need to store information about the names that are in scope at each point in the program.
+To type check the base language, we need to store information about the names that are in scope at each point in the program.
 There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}
 
-In Statix, all this information can be stored in a \verb|scope graph|\cite{scope_graphs}, which is a feature of Statix.
+In Statix, all this information can be stored in a \emph{scope graph} \cite{scope_graphs}, which is a feature of Statix.
+It is a graph consisting of nodes for scopes, labeled edges for visibility relations, scoped declarations for a relation, and queries for references.
 We only use a single type of edge, called \verb|P| (parent) edges.
@@
 
-Next, we will introduce some Statix relations that can be used to interact with these scope graphs:
+Next, we will introduce some Statix predicates that can be used to interact with these scope graphs:
 
@@
 \end{lstlisting}
-The \verb|sPutType| and \verb|sPutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively.
-To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found.
+The \verb|sPutType| and \verb|sPutSubsts| predicates generate a new scope given a parent scope and a type or a substitution respectively.
+To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of NameEntries respectively that the query found.
 Finally, \verb|sEmpty| returns a fresh empty scope.
+
+We will define a \emph{scoped expression}, as a pair of a scope and an expression. The scope acts as the environment of the expression, containing all of the context needed to evaluate the expression.
 
@@
 
-A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare.
+A unique requirement for dependently typed languages is beta reduction during type checking, since types may require evaluation to compare.
 
 We implemented beta reduction using a Krivine abstract machine\cite{krivine}.
+The machine can head evaluate lambda expressions with a call-by-name semantics.
+It works by keeping a stack of all arguments that have not been applied yet.
 This turned out to be the more natural way of expressing this over substitution-based evaluation relation.
@@
 
-Note that the evaluator uses names rather than De Bruijn indices.
 In conventional dependently typed languages, evaluation is often done using De Bruijn indices.
-However, since the aim of this paper is to use the features of Statix, and scope graphs work based on names, we chose to use names.
-It would also stop us from using editor services that rely on \verb|.ref| annotations, such as renaming.
+However, we chose to use names rather than De Bruijn indices, because scope graphs work based on names rather.
+Using De Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations, such as renaming.
 
-We need to define multiple relations that will be used later for type-checking.
-First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
+We need to define multiple predicates that will be used later for type checking.
+First, the primary predicate is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
 The scope acts as the environment from \cite{krivine}, using \verb|NSubst| to store substitutions.
@@
 Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
-One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
+One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
 
@@
 		\inferrule{ 
-		} { 
+		} { relation
 			\bhr
@@
 
-\subsection{Type-checking the Calculus of Constructions}
+\subsection{Type checking the Calculus of Constructions}
 \label{sec:coc-typecheck}
 
-We will define a Statix relation \verb|typeOfExpr| that takes a scope and an expression and type-checks the scope in the expression. It returns the type of the expression.
+We will define a Statix predicate \verb|typeOfExpr| that takes a scope and an expression and type checks the scope in the expression. It returns the type of the expression.
 
@@
 \end{lstlisting}
-We can then start defining type-checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
+We can then start defining type checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
 \begin{enumerate}
@@
 	\end{mathpar}
-	\caption{Rules for type-checking the Calculus of Constructions}
+	\caption{Rules for type checking the Calculus of Constructions}
 	\label{fig:type-check-rules}
@@
 \subsection{Avoiding variable capture}
-One problem with the current implementation is that it does not avoid variable capture.
+One problem with the implementation presented in the previous sections is that it does not avoid variable capture.
 Variable capture happens when a term becomes bound because of a wrong substitution.
@@
 T| is \verb|T : Type -> T : T -> T|.
-This is not the correct type, and even worse, the correct type is not possible to express without renaming!
+This is not the correct type, and even worse, it is not possible to express the correct type without renaming!
 The problem is that there are multiple names, and there is no way to distinguish between them.
@@
 \paragraph*{Term Inference}
-In dependently typed language, we can infer the value of types.
+In dependently typed languages, the value of types can often be inferred.
 Ideally, we would like to infer the most general type possible.
@@
 We implemented an approximation to inference that can infer most types.
+For example, the type of \verb|x| in the function below can be inferred, because it is applied to \verb|true|, which is a boolean.
 \begin{lstlisting}
@@
 \paragraph*{Semantic Code Completion}
-Finally, we explored how semantic code completion presented in \cite{codecompletion} applies to dependently typed languages. It works well, only showing completions that are semantically possible.
+Finally, we explored how semantic code completion presented by Pelsmaeker et al.
+\cite{codecompletion} applies to dependently typed languages.
+It works well, only showing completions that are semantically possible.
 
-For example, in the following code it would only suggest expressions that can be booleans:
+For example, in the following code it would only suggest expressions that can be booleans (including variables that are booleans):
 \begin{lstlisting}
@@
 The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
-In section 2.5 of Scopes as Types \cite{scopes_as_types}, an implementation of System F is presented; it does lazy substitutions, by using scopes as types.
+Van Antwerpen et al.
+\cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types.
 It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.

### 2022-11-02 paper-evcs/paper.tex
Work
Add orcids

@@
 
-\author{Jonathan Brouwer}{Delft University of Technology, The Netherlands \and \url{http://jonathanb.nl}}{j.t.brouwer@student.tudelft.nl}{}{}
+\author{Jonathan Brouwer}{Delft University of Technology, The Netherlands \and \url{http://jonathanb.nl}}{j.t.brouwer@student.tudelft.nl}{https://orcid.org/0000-0002-2469-548X}{}
 
-\author{Jesper Cockx}{Delft University of Technology, The Netherlands \and \url{http://jesper.sikanda.be}}{j.g.h.cockx@tudelft.nl}{}{}
+\author{Jesper Cockx}{Delft University of Technology, The Netherlands \and \url{http://jesper.sikanda.be}}{j.g.h.cockx@tudelft.nl}{https://orcid.org/0000-0003-3862-4073}{}
 
-\author{Aron Zwaan}{Delft University of Technology, The Netherlands \and \url{http://aronzwaan.github.io}}{a.s.zwaan@tudelft.nl}{}{}
+\author{Aron Zwaan}{Delft University of Technology, The Netherlands \and \url{http://aronzwaan.github.io}}{a.s.zwaan@tudelft.nl}{https://orcid.org/0000-0002-1818-4245}{}
 
@@
 \end{lstlisting}
-The AST of this program in Aterm format would be:
+The AST of this program in Aterm format\cite{aterm} would be:
 \begin{lstlisting}

### 2022-11-07 paper-thesis/chapters/00-abstract.tex
Create paper-thesis

--- 2022-10-27 paper-evcs/paper.tex
@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{oasics-v2021}
-%This is a template for producing OASIcs articles. 
-%See oasics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
-
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideOASIcs %uncomment to remove references to OASIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
-% Added because the included oasics-logo-bw.pdf is pdf 1.6 and this was giving a warning
-% Feel free to remove this if it breaks anything
-\pdfminorversion=6
-
-% Package is used to create inference rules
-\usepackage{mathpartir}
-
-% Commands used
-\newcommand{\scope}[2]{\langle#1 \: | \: #2\rangle}
-\newcommand{\bhr}[3]{ #1 \: #2 \underset{\beta h}{\Rightarrow} #3 }
-\newcommand{\toe}[3]{ \scope{#1}{#2} : #3 }
-\newcommand{\bred}[3]{ \scope{#1}{#2} \underset{\beta}{\Rightarrow} #3 }
-\newcommand{\beq}[2]{ #1 \underset{\beta}{=} #2 }
-\newcommand{\co}[1]{ \mathsf{#1} }
-
-\DeclareUnicodeCharacter{03BB}{L}
-\DeclareUnicodeCharacter{03A0}{M}
-
-\bibliographystyle{plainurl}
-
-\title{Dependently Typed Languages in Statix}
-
-\author{Jonathan Brouwer}{Delft University of Technology, The Netherlands \and \url{http://jonathanb.nl}}{j.t.brouwer@student.tudelft.nl}{}{}
-
-\author{Jesper Cockx}{Delft University of Technology, The Netherlands \and \url{http://jesper.sikanda.be}}{j.g.h.cockx@tudelft.nl}{}{}
-
-\author{Aron Zwaan}{Delft University of Technology, The Netherlands \and \url{http://aronzwaan.github.io}}{a.s.zwaan@tudelft.nl}{}{}
-
-\authorrunning{J. Brouwer, J. Cockx and A. Zwaan}
-
-\Copyright{Jonathan Brouwer, Jesper Cockx and Aron Zwaan} 
-\begin{CCSXML}
-    <ccs2012>
-    <concept>
-    <concept_id>10011007.10011006.10011039.10011311</concept_id>
-    <concept_desc>Software and its engineering~Semantics</concept_desc>
-    <concept_significance>500</concept_significance>
-    </concept>
-    <concept>
-    <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
-    <concept_desc>Software and its engineering~Functional languages</concept_desc>
-    <concept_significance>300</concept_significance>
-    </concept>
-    <concept>
-    <concept_id>10011007.10011006.10011041</concept_id>
-    <concept_desc>Software and its engineering~Compilers</concept_desc>
-    <concept_significance>300</concept_significance>
-    </concept>
-    </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Semantics}
-\ccsdesc[300]{Software and its engineering~Functional languages}
-\ccsdesc[300]{Software and its engineering~Compilers}
-
-\keywords{Spoofax, Statix, Dependent Types} %TODO mandatory; please add comma-separated list of keywords
-
-\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-\maketitle
-
 \begin{abstract}
@@
 \end{abstract}
-
-\section{Introduction}
-
-Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages\cite{spoofax}.
-When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
-To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems.
-However, no attempts have been made to express dependently typed languages in Statix.
-
-Dependently typed languages are different from other languages, because they allow types to be parameterized by values.
-This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree.
-This expressiveness also makes dependent type systems more complicated to implement.
-Especially, deciding equality of types requires evaluation of the terms they are parameterized by.
-
-This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language.
-We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix.
-The goal is not to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.
-
-In section \ref{sec:coc} of this paper we show that Statix is already usable as a tool for implementing dependent types.
-There were some challenges in implementing the dependent type system, which this paper also discusses.
-Finally, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
-
-The implementation of the language is available on GitHub.\\
-\url{https://github.com/JonathanBrouwer/master-thesis/}
-
-\section{Calculus of Constructions}
-\label{sec:coc}
-
-In this section, we will describe how to implement a dependently typed language in Statix.
-In section \ref{sec:coc-syntax} we will describe the syntax of the language, then in section \ref{sec:coc-scopes} we will describe how scope graphs are used to type check the language.
-Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type-check the language.
-
-\subsection{The language}
-\label{sec:coc-syntax}
-
-The base language that has been implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}.
-One extra feature was added that is not present in the Calculus of Constructions, let bindings.
-Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
-The abstract syntax of the language is available in figure \ref{fig:syntax}.
-
-\begin{figure}[h]
-\begin{lstlisting}
-Type        : Expr
-Let         : ID * Expr * Expr -> Expr
-Var         : ID -> Expr
-FnType      : ID * Expr * Expr -> Expr
-FnConstruct : ID * Expr * Expr -> Expr
-FnDestruct  : Expr * Expr -> Expr
-\end{lstlisting}
-\caption{The syntax for the base language. FnConstruct is a lambda function, FnDestruct is application of a lambda function.}
-\label{fig:syntax}
-\end{figure}
-
-An example program is the following, which defines a polymorphic identity function and applies it to a function:
-
-\begin{lstlisting}
-let f = \T: Type. \x: T. x;
-f (_: Type -> Type) (\x: Type. x)
-\end{lstlisting}
-The AST of this program would be:
-\begin{lstlisting}
-Let(
-  "f", 
-  FnConstruct("T", Type(), FnConstruct("x", Var("T"), Var("x"))),
-  FnDestruct(
-	FnDestruct(Var("f"), FnConstruct("_", Type(), Type())),
-	FnConstruct("x", Type(), Var("x"))
-  )
-)
-\end{lstlisting}
-
-\subsection{Scope Graphs}
-\label{sec:coc-scopes}
-
-To type-check the base language, we need to store information about the names that are in scope at each point in the program.
-There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}
-
-In Statix, all this information can be stored in a \verb|scope graph|\cite{scope_graphs}, which is a feature of Statix.
-We only use a single type of edge, called \verb|P| (parent) edges.
-It also only has a single relation, called \verb|name|.
-This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.
-
-Next, we will introduce some Statix relations that can be used to interact with these scope graphs:
-
-\begin{lstlisting}
-sPutType  : scope * ID * Expr -> scope
-sPutSubst : scope * ID * (scope * Expr) -> scope
-sGetName  : scope * ID -> NameEntry
-sGetNames : scope * ID -> list((path * (ID * NameEntry)))
-sEmpty    : -> scope
-\end{lstlisting}
-The \verb|sPutType| and \verb|sPutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively.
-To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found.
-Finally, \verb|sEmpty| returns a fresh empty scope.
-
-\subsection{Beta Reductions}
-\label{sec:coc-dynsyms}
-
-A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare.
-
-We implemented beta reduction using a Krivine abstract machine\cite{krivine}.
-This turned out to be the more natural way of expressing this over substitution-based evaluation relation.
-We originally tried to implement the latter, which works fine for the base language.
-However, it runs into trouble when implementing inductive data types; more information about this will be in the full master thesis.
-An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.
-
-Note that the evaluator uses names rather than De Bruijn indices.
-In conventional dependently typed languages, evaluation is often done using De Bruijn indices.
-However, since the aim of this paper is to use the features of Statix, and scope graphs work based on names, we chose to use names.
-It would also stop us from using editor services that rely on \verb|.ref| annotations, such as renaming.
-
-We need to define multiple relations that will be used later for type-checking.
-First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
-The scope acts as the environment from \cite{krivine}, using \verb|NSubst| to store substitutions.
-All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
-We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|.
-Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
-One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
-
-Additionally, we define \verb|betaReduce| which fully beta reduces a term.
-It works by first calling \verb|betaReduceHead| and then matching on the head, calling \verb|betaReduce| on the sub-expressions of the head recursively.
-
-Finally, we define \verb|expectBetaEq|.
-This rule first beta reduces the heads of both sides, and then compares them.
-If the head is not the same, the rule fails.
-Otherwise, it recurses on the sub-expressions.
-One special case is when comparing two \verb|FnConstruct|s.
-Here we need to take into account alpha equality: two expressions which only differ in the names that they use should be considered equal.
-We implement this by substituting in the body of the functions, replacing their argument names with placeholders.
-
-\begin{figure}[h]
-	\begin{mathpar}	
-		% Type rule
-		\inferrule{ 
-		} { 
-			\bhr
-			{ \scope{s}{\co{Type}()} }
-			{ [] }
-			{ \scope{s}{\co{Type}()}} 
-		}
-
-		% Let rule
-		\inferrule{ 
-			\bhr 
-			{ \scope{ \co{sPutSubst}(s, n, (s, v))}{b} } 
-			{ t }
-			{ \scope{ s' } { e' } }
-		}{
-			\bhr
-			{ \scope{ s }{ \co{Let}(n, v, b) } }
-			{ t }
-			{ \scope { s' } { e' } }
-		}
-
-		% Var rule - NSubst
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
-			\\ \bhr
-			{\scope{se}{e}}
-			{t}
-			{\scope{se'}{e'}} 
-		}{
-			\bhr
-			{\scope{s}{\co{Var}(n)}}
-			{t}
-			{\scope{se'}{e'}} 
-		}
-
-		% Var rule - NType
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NType}(t) 
-		}{ 
-			\bhr
-			{\scope{s}{\co{Var}(n)}}
-			{t}
-			{\co{rebuild}(s, \co{Var}(n), t)} 
-		}
-
-		% FnType rule
-		\inferrule{ 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnType}(n, a, b)}}
-			{[]}
-			{\scope{s}{\co{FnType}(n, a, b)}} 
-		}
-
-		% FnConstruct rule - No args
-		\inferrule{ 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
-			{[]}
-			{\scope{s}{\co{FnConstruct}(n, a, b)}} 
-		}
-
-		% FnConstruct rule - Args
-		\inferrule{ 
-			\bhr
-			{\scope{\co{sPutSubst}(s, n, a)}{b}}
-			{ts}
-			{\scope{s'}{e'}} 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
-			{(t::ts)}
-			{\scope{s'}{e'}} 
-		}
-
-		% FnDestruct rule
-		\inferrule{ 
-			\bhr
-			{\scope{s}{f}}
-			{(a::ts)}
-			{\scope{s'}{e'}} 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnDestruct}(f, a)}}
-			{ts}
-			{\scope{s'}{e'}} 
-		}
-	\end{mathpar}
-	\caption{Rules for beta head reducing the Calculus of Constructions}
-	\label{fig:beta-head-reduce-rules}
-\end{figure}
-
-\subsection{Type-checking the Calculus of Constructions}
-\label{sec:coc-typecheck}
-
-We will define a Statix relation \verb|typeOfExpr| that takes a scope and an expression and type-checks the scope in the expression. It returns the type of the expression.
-
-\begin{lstlisting}
-typeOfExpr : scope * Expr -> Expr
-\end{lstlisting}
-We can then start defining type-checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
-\begin{enumerate}
-	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
-	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
-	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
-	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
-\end{enumerate}
-
-The inference rules above can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
-\begin{lstlisting}
-typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
-    typeOfExpr(s, v) == vt, sPutSubst(s, n, (s, v)) == s'.
-\end{lstlisting}
-
-\begin{figure}[h]
-	\begin{mathpar}	
-		% Type rule
-		\inferrule{ 
-		} {
-			\toe{s}{\co{Type}()}{\co{Type}()}
-		}
-
-		% Let rule
-		\inferrule{ 
-			\toe{s}{v}{vt}
-			\\ \toe{\co{sPutSubst}(s, n, (s, v))}{b}{t}
-		}{
-			\toe{s}{\co{Let}(n, v, b)}{t}
-		}
-
-		\\
-
-		% Var rule - NType
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NType}(t) 
-		}{ 
-			\toe{s}{\co{Var}(n)}{t}
-		}
-
-		% Var rule - NSubst
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
-			\\ \toe{se}{e}{t}
-		}{
-		 	\toe{s}{\co{Var}(n)}{t}
-		}
-
-		% FnType rule
-		\inferrule{ 
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()} 
-			\\ \bred{s}{a}{a'}
-			\\\\  \toe{\co{sPutType}(s, n, a')}{b}{bt}
-			\\ \beq{bt}{\co{Type}()}
-		} { 
-			\toe{s}{\co{FnType}(n, a, b)}{\co{Type}()}
-		}
-
-		% Force these to be on the same line, using negative spaces
-		\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!
-
-		% FnConstruct rule
-		\inferrule{ 
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()}
-			\\ \bred{s}{a}{a'} 
-			\\\\ \toe{\co{sPutType}(s, n, a')}{b}{bt}
-		} { 
-			\toe{s}{\co{FnConstruct}(n, a, b)}{\co{FnType}(n, a', bt)}
-		}
-
-		% FnDestruct rule
-		\inferrule{ 
-			\toe{s}{f}{ft}
-			\\ \bhr
-				{\scope{s}{ft}}
-				{[]}
-				{\scope{s'}{\co{FnType}(da, dt, db)}} 
-			\\\\ \toe{s}{a}{at} 
-			\\ \beq{at}{\scope{sf}{dt}}
-			\\ \bred{s'}{db}{db'}
-		} { 
-			\toe{s}{\co{FnDestruct}(f, a)}{db'}
-	 	}
-
-	\end{mathpar}
-	\caption{Rules for type-checking the Calculus of Constructions}
-	\label{fig:type-check-rules}
-\end{figure}
-
-\subsection{Avoiding variable capture}
-One problem with the current implementation is that it does not avoid variable capture.
-Variable capture happens when a term becomes bound because of a wrong substitution.
-For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type.
-\T : T.
-T| is \verb|T : Type -> T : T -> T|.
-This is not the correct type, and even worse, the correct type is not possible to express without renaming!
-The problem is that there are multiple names, and there is no way to distinguish between them.
-This can be solved by using scopes to distinguish names.
-A full explanation of this will be in the master thesis.
-
-\section{Extensions}
-\label{sec:extensions}
-This section will discuss some further ideas that can be explored to build upon what has already been shown.
-These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.
-
-\paragraph*{Term Inference}
-In dependently typed language, we can infer the value of types.
-Ideally, we would like to infer the most general type possible.
-However, this kind of analysis is not possible in Statix because you cannot reason over whether meta-variables are instantiated or not.
-We implemented an approximation to inference that can infer most types.
-\begin{lstlisting}
-(\x: _. x) true
-\end{lstlisting}
-
-\paragraph*{Inductive Data Types}
-Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types.
-We can also generate eliminators for these data types to use their values.
-All of this has been implemented in Statix.
-\begin{lstlisting}
-data Maybe (T : Type) = 
-    None : Maybe T,
-    Some : _: T -> Maybe T;
-\end{lstlisting}
-
-\paragraph*{Semantic Code Completion}
-Finally, we explored how semantic code completion presented in \cite{codecompletion} applies to dependently typed languages. It works well, only showing completions that are semantically possible.
-
-For example, in the following code it would only suggest expressions that can be booleans:
-\begin{lstlisting}
-let f = \b: Bool. b;
-f {{Expr}}
-\end{lstlisting}
-
-\section{Related Work}
-The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
-In section 2.5 of Scopes as Types \cite{scopes_as_types}, an implementation of System F is presented; it does lazy substitutions, by using scopes as types.
-It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.
-
-Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language.
-The pi-forall language\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper.
-In principle, the implementations are very similar.
-For example, the inference rules presented in \cite{pi_forall} are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper.
-The primary difference is that they use a bidirectional type system, whereas this paper does not.
-
-There exist several so-called \emph{logical frameworks}, tools designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}.
-Since these tools are designed specifically for the task, implementing the type system takes less effort in them compared to Spoofax, but for other tasks such as defining a parser or editor services they are not as well equipped.
-Some logical frameworks such as Twelf and Dedukti support Miller's \emph{higher-order pattern unification}~\cite{Miller89}, which can be used as a more powerful way of inferring implicit arguments than the first-order unification built into Statix.
-Andromeda 2 also supports \emph{extensionality rules} that can match on the type of an equality.
-We expect that adding extensionality rules to our implementation would be possible to do in Statix, but we leave an actual implementation to future work.
-
-\section{Conclusion}
-
-We have demonstrated that the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph.
-We have also presented a few extensions to the Calculus of Constructions and discussed how they could be implemented.
-
-\bibliography{paper}
-
-\end{document}

### 2022-11-07 paper-thesis/chapters/01-introduction.tex
Create paper-thesis

--- 2022-10-28 paper-evcs/paper.tex
@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{oasics-v2021}
-%This is a template for producing OASIcs articles. 
-%See oasics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
-
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideOASIcs %uncomment to remove references to OASIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
-% Added because the included oasics-logo-bw.pdf is pdf 1.6 and this was giving a warning
-% Feel free to remove this if it breaks anything
-\pdfminorversion=6
-
-% Package is used to create inference rules
-\usepackage{mathpartir}
-
-% Commands used
-\newcommand{\scope}[2]{\langle#1 \: | \: #2\rangle}
-\newcommand{\bhr}[3]{ #1 \: #2 \underset{\beta h}{\Rightarrow} #3 }
-\newcommand{\toe}[3]{ \scope{#1}{#2} : #3 }
-\newcommand{\bred}[3]{ \scope{#1}{#2} \underset{\beta}{\Rightarrow} #3 }
-\newcommand{\beq}[2]{ #1 \underset{\beta}{=} #2 }
-\newcommand{\co}[1]{ \mathsf{#1} }
-
-\bibliographystyle{plainurl}
-
-\title{Dependently Typed Languages in Statix}
-
-\author{Jonathan Brouwer}{Delft University of Technology, The Netherlands \and \url{http://jonathanb.nl}}{j.t.brouwer@student.tudelft.nl}{}{}
-
-\author{Jesper Cockx}{Delft University of Technology, The Netherlands \and \url{http://jesper.sikanda.be}}{j.g.h.cockx@tudelft.nl}{}{}
-
-\author{Aron Zwaan}{Delft University of Technology, The Netherlands \and \url{http://aronzwaan.github.io}}{a.s.zwaan@tudelft.nl}{}{}
-
-\authorrunning{J. Brouwer, J. Cockx and A. Zwaan}
-
-\Copyright{Jonathan Brouwer, Jesper Cockx and Aron Zwaan} 
-\begin{CCSXML}
-    <ccs2012>
-    <concept>
-    <concept_id>10011007.10011006.10011039.10011311</concept_id>
-    <concept_desc>Software and its engineering~Semantics</concept_desc>
-    <concept_significance>500</concept_significance>
-    </concept>
-    <concept>
-    <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
-    <concept_desc>Software and its engineering~Functional languages</concept_desc>
-    <concept_significance>300</concept_significance>
-    </concept>
-    <concept>
-    <concept_id>10011007.10011006.10011041</concept_id>
-    <concept_desc>Software and its engineering~Compilers</concept_desc>
-    <concept_significance>300</concept_significance>
-    </concept>
-    </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Semantics}
-\ccsdesc[300]{Software and its engineering~Functional languages}
-\ccsdesc[300]{Software and its engineering~Compilers}
-
-\keywords{Spoofax, Statix, Dependent Types} %TODO mandatory; please add comma-separated list of keywords
-
-\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone.
-The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system.
-However, so far Statix has not been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking.
-In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
-While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.
-\end{abstract}
-
-\section{Introduction}
+\chapter{Introduction}
 
@@
 
-In section \ref{sec:coc} of this paper we show that Statix is already usable as a tool for implementing dependent types.
-There were some challenges in implementing the dependent type system, which this paper also discusses.
-Finally, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
-
-\section{Calculus of Constructions}
-\label{sec:coc}
-
-In this section, we will describe how to implement a dependently typed language in Statix.
-In section \ref{sec:coc-syntax} we will describe the syntax of the language, then in section \ref{sec:coc-scopes} we will describe how scope graphs are used to type check the language.
-Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type check the language.
-The implementation of the language is available on GitHub.
-\url{https://github.com/JonathanBrouwer/master-thesis/}
-
-\subsection{The language}
-\label{sec:coc-syntax}
-
-The base language that has been implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}.
-One extra feature was added that is not present in the Calculus of Constructions: let bindings.
-Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
-The abstract syntax of the language is available in figure \ref{fig:syntax}.
-
-\begin{figure}[h]
-\begin{lstlisting}
-Type        : Expr
-Let         : ID * Expr * Expr -> Expr
-Var         : ID -> Expr
-FnType      : ID * Expr * Expr -> Expr
-FnConstruct : ID * Expr * Expr -> Expr
-FnDestruct  : Expr * Expr -> Expr
-\end{lstlisting}
-\caption{The syntax for the base language. FnConstruct is a lambda function, FnDestruct is application of a lambda function.}
-\label{fig:syntax}
-\end{figure}
-
-An example program is the following, which defines a polymorphic identity function and applies it to a function:
-
-\begin{lstlisting}
-let f = \T: Type. \x: T. x;
-f (_: Type -> Type) (\x: Type. x)
-\end{lstlisting}
-The AST of this program in Aterm format would be:
-\begin{lstlisting}
-Let(
-  "f", 
-  FnConstruct("T", Type(), FnConstruct("x", Var("T"), Var("x"))),
-  FnDestruct(
-	FnDestruct(Var("f"), FnConstruct("_", Type(), Type())),
-	FnConstruct("x", Type(), Var("x"))
-  )
-)
-\end{lstlisting}
-
-\subsection{Scope Graphs}
-\label{sec:coc-scopes}
-
-To type check the base language, we need to store information about the names that are in scope at each point in the program.
-There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}
-
-In Statix, all this information can be stored in a \emph{scope graph} \cite{scope_graphs}, which is a feature of Statix.
-It is a graph consisting of nodes for scopes, labeled edges for visibility relations, scoped declarations for a relation, and queries for references.
-We only use a single type of edge, called \verb|P| (parent) edges.
-It also only has a single relation, called \verb|name|.
-This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.
-
-Next, we will introduce some Statix predicates that can be used to interact with these scope graphs:
-
-\begin{lstlisting}
-sPutType  : scope * ID * Expr -> scope
-sPutSubst : scope * ID * (scope * Expr) -> scope
-sGetName  : scope * ID -> NameEntry
-sGetNames : scope * ID -> list((path * (ID * NameEntry)))
-sEmpty    : -> scope
-\end{lstlisting}
-The \verb|sPutType| and \verb|sPutSubsts| predicates generate a new scope given a parent scope and a type or a substitution respectively.
-To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of NameEntries respectively that the query found.
-Finally, \verb|sEmpty| returns a fresh empty scope.
-
-We will define a \emph{scoped expression}, as a pair of a scope and an expression. The scope acts as the environment of the expression, containing all of the context needed to evaluate the expression.
-
-\subsection{Beta Reductions}
-\label{sec:coc-dynsyms}
-
-A unique requirement for dependently typed languages is beta reduction during type checking, since types may require evaluation to compare.
-
-We implemented beta reduction using a Krivine abstract machine\cite{krivine}.
-The machine can head evaluate lambda expressions with a call-by-name semantics.
-It works by keeping a stack of all arguments that have not been applied yet.
-This turned out to be the more natural way of expressing this over substitution-based evaluation relation.
-We originally tried to implement the latter, which works fine for the base language.
-However, it runs into trouble when implementing inductive data types; more information about this will be in the full master thesis.
-An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.
-
-In conventional dependently typed languages, evaluation is often done using De Bruijn indices.
-However, we chose to use names rather than De Bruijn indices, because scope graphs work based on names rather.
-Using De Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations, such as renaming.
-
-We need to define multiple predicates that will be used later for type checking.
-First, the primary predicate is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
-The scope acts as the environment from \cite{krivine}, using \verb|NSubst| to store substitutions.
-All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
-We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|.
-Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
-One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
-
-Additionally, we define \verb|betaReduce| which fully beta reduces a term.
-It works by first calling \verb|betaReduceHead| and then matching on the head, calling \verb|betaReduce| on the sub-expressions of the head recursively.
-
-Finally, we define \verb|expectBetaEq|.
-This rule first beta reduces the heads of both sides, and then compares them.
-If the head is not the same, the rule fails.
-Otherwise, it recurses on the sub-expressions.
-One special case is when comparing two \verb|FnConstruct|s.
-Here we need to take into account alpha equality: two expressions which only differ in the names that they use should be considered equal.
-We implement this by substituting in the body of the functions, replacing their argument names with placeholders.
-
-\begin{figure}[h]
-	\begin{mathpar}	
-		% Type rule
-		\inferrule{ 
-		} { 
-			\bhr
-			{ \scope{s}{\co{Type}()} }
-			{ [] }
-			{ \scope{s}{\co{Type}()}} 
-		}
-
-		% Let rule
-		\inferrule{ 
-			\bhr 
-			{ \scope{ \co{sPutSubst}(s, n, (s, v))}{b} } 
-			{ t }
-			{ \scope{ s' } { e' } }
-		}{
-			\bhr
-			{ \scope{ s }{ \co{Let}(n, v, b) } }
-			{ t }
-			{ \scope { s' } { e' } }
-		}
-
-		% Var rule - NSubst
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
-			\\ \bhr
-			{\scope{se}{e}}
-			{t}
-			{\scope{se'}{e'}} 
-		}{
-			\bhr
-			{\scope{s}{\co{Var}(n)}}
-			{t}
-			{\scope{se'}{e'}} 
-		}
-
-		% Var rule - NType
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NType}(t) 
-		}{ 
-			\bhr
-			{\scope{s}{\co{Var}(n)}}
-			{t}
-			{\co{rebuild}(s, \co{Var}(n), t)} 
-		}
-
-		% FnType rule
-		\inferrule{ 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnType}(n, a, b)}}
-			{[]}
-			{\scope{s}{\co{FnType}(n, a, b)}} 
-		}
-
-		% FnConstruct rule - No args
-		\inferrule{ 
-		} { relation
-			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
-			{[]}
-			{\scope{s}{\co{FnConstruct}(n, a, b)}} 
-		}
-
-		% FnConstruct rule - Args
-		\inferrule{ 
-			\bhr
-			{\scope{\co{sPutSubst}(s, n, a)}{b}}
-			{ts}
-			{\scope{s'}{e'}} 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
-			{(t::ts)}
-			{\scope{s'}{e'}} 
-		}
-
-		% FnDestruct rule
-		\inferrule{ 
-			\bhr
-			{\scope{s}{f}}
-			{(a::ts)}
-			{\scope{s'}{e'}} 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnDestruct}(f, a)}}
-			{ts}
-			{\scope{s'}{e'}} 
-		}
-	\end{mathpar}
-	\caption{Rules for beta head reducing the Calculus of Constructions}
-	\label{fig:beta-head-reduce-rules}
-\end{figure}
-
-\subsection{Type checking the Calculus of Constructions}
-\label{sec:coc-typecheck}
-
-We will define a Statix predicate \verb|typeOfExpr| that takes a scope and an expression and type checks the scope in the expression. It returns the type of the expression.
-
-\begin{lstlisting}
-typeOfExpr : scope * Expr -> Expr
-\end{lstlisting}
-We can then start defining type checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
-\begin{enumerate}
-	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
-	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
-	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
-	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
-\end{enumerate}
-
-The inference rules above can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
-\begin{lstlisting}
-typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
-    typeOfExpr(s, v) == vt, sPutSubst(s, n, (s, v)) == s'.
-\end{lstlisting}
-
-\begin{figure}[h]
-	\begin{mathpar}	
-		% Type rule
-		\inferrule{ 
-		} {
-			\toe{s}{\co{Type}()}{\co{Type}()}
-		}
-
-		% Let rule
-		\inferrule{ 
-			\toe{s}{v}{vt}
-			\\ \toe{\co{sPutSubst}(s, n, (s, v))}{b}{t}
-		}{
-			\toe{s}{\co{Let}(n, v, b)}{t}
-		}
-
-		\\
-
-		% Var rule - NType
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NType}(t) 
-		}{ 
-			\toe{s}{\co{Var}(n)}{t}
-		}
-
-		% Var rule - NSubst
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
-			\\ \toe{se}{e}{t}
-		}{
-		 	\toe{s}{\co{Var}(n)}{t}
-		}
-
-		% FnType rule
-		\inferrule{ 
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()} 
-			\\ \bred{s}{a}{a'}
-			\\\\  \toe{\co{sPutType}(s, n, a')}{b}{bt}
-			\\ \beq{bt}{\co{Type}()}
-		} { 
-			\toe{s}{\co{FnType}(n, a, b)}{\co{Type}()}
-		}
-
-		% Force these to be on the same line, using negative spaces
-		\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!
-
-		% FnConstruct rule
-		\inferrule{ 
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()}
-			\\ \bred{s}{a}{a'} 
-			\\\\ \toe{\co{sPutType}(s, n, a')}{b}{bt}
-		} { 
-			\toe{s}{\co{FnConstruct}(n, a, b)}{\co{FnType}(n, a', bt)}
-		}
-
-		% FnDestruct rule
-		\inferrule{ 
-			\toe{s}{f}{ft}
-			\\ \bhr
-				{\scope{s}{ft}}
-				{[]}
-				{\scope{s'}{\co{FnType}(da, dt, db)}} 
-			\\\\ \toe{s}{a}{at} 
-			\\ \beq{at}{\scope{sf}{dt}}
-			\\ \bred{s'}{db}{db'}
-		} { 
-			\toe{s}{\co{FnDestruct}(f, a)}{db'}
-	 	}
-
-	\end{mathpar}
-	\caption{Rules for type checking the Calculus of Constructions}
-	\label{fig:type-check-rules}
-\end{figure}
-
-\subsection{Avoiding variable capture}
-One problem with the implementation presented in the previous sections is that it does not avoid variable capture.
-Variable capture happens when a term becomes bound because of a wrong substitution.
-For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type.
-\T : T.
-T| is \verb|T : Type -> T : T -> T|.
-This is not the correct type, and even worse, it is not possible to express the correct type without renaming!
-The problem is that there are multiple names, and there is no way to distinguish between them.
-This can be solved by using scopes to distinguish names.
-A full explanation of this will be in the master thesis.
-
-\section{Extensions}
-\label{sec:extensions}
-This section will discuss some further ideas that can be explored to build upon what has already been shown.
-These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.
-
-\paragraph*{Term Inference}
-In dependently typed languages, the value of types can often be inferred.
-Ideally, we would like to infer the most general type possible.
-However, this kind of analysis is not possible in Statix because you cannot reason over whether meta-variables are instantiated or not.
-We implemented an approximation to inference that can infer most types.
-For example, the type of \verb|x| in the function below can be inferred, because it is applied to \verb|true|, which is a boolean.
-\begin{lstlisting}
-(\x: _. x) true
-\end{lstlisting}
-
-\paragraph*{Inductive Data Types}
-Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types.
-We can also generate eliminators for these data types to use their values.
-All of this has been implemented in Statix.
-\begin{lstlisting}
-data Maybe (T : Type) = 
-    None : Maybe T,
-    Some : _: T -> Maybe T;
-\end{lstlisting}
-
-\paragraph*{Semantic Code Completion}
-Finally, we explored how semantic code completion presented by Pelsmaeker et al.
-\cite{codecompletion} applies to dependently typed languages.
-It works well, only showing completions that are semantically possible.
-
-For example, in the following code it would only suggest expressions that can be booleans (including variables that are booleans):
-\begin{lstlisting}
-let f = \b: Bool. b;
-f {{Expr}}
-\end{lstlisting}
-
-\section{Related Work}
-The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
-Van Antwerpen et al.
-\cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types.
-It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.
-
-Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language.
-The pi-forall language\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper.
-In principle, the implementations are very similar.
-For example, the inference rules presented in \cite{pi_forall} are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper.
-The primary difference is that they use a bidirectional type system, whereas this paper does not.
-
-There exist several so-called \emph{logical frameworks}, tools designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}.
-Since these tools are designed specifically for the task, implementing the type system takes less effort in them compared to Spoofax, but for other tasks such as defining a parser or editor services they are not as well equipped.
-Some logical frameworks such as Twelf and Dedukti support Miller's \emph{higher-order pattern unification}~\cite{Miller89}, which can be used as a more powerful way of inferring implicit arguments than the first-order unification built into Statix.
-Andromeda 2 also supports \emph{extensionality rules} that can match on the type of an equality.
-We expect that adding extensionality rules to our implementation would be possible to do in Statix, but we leave an actual implementation to future work.
-
-\section{Conclusion}
-
-We have demonstrated that the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph.
-We have also presented a few extensions to the Calculus of Constructions and discussed how they could be implemented.
-
-\bibliography{paper}
-
-\end{document}
+TODO navigation

### 2022-11-07 paper-thesis/chapters/02-coc.tex
Create paper-thesis

--- 2022-11-02 paper-evcs/paper.tex
@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{oasics-v2021}
-%This is a template for producing OASIcs articles. 
-%See oasics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
-
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideOASIcs %uncomment to remove references to OASIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
-% Added because the included oasics-logo-bw.pdf is pdf 1.6 and this was giving a warning
-% Feel free to remove this if it breaks anything
-\pdfminorversion=6
-
-% Package is used to create inference rules
-\usepackage{mathpartir}
-
-% Commands used
-\newcommand{\scope}[2]{\langle#1 \: | \: #2\rangle}
-\newcommand{\bhr}[3]{ #1 \: #2 \underset{\beta h}{\Rightarrow} #3 }
-\newcommand{\toe}[3]{ \scope{#1}{#2} : #3 }
-\newcommand{\bred}[3]{ \scope{#1}{#2} \underset{\beta}{\Rightarrow} #3 }
-\newcommand{\beq}[2]{ #1 \underset{\beta}{=} #2 }
-\newcommand{\co}[1]{ \mathsf{#1} }
-
-\bibliographystyle{plainurl}
-
-\title{Dependently Typed Languages in Statix}
-
-\author{Jonathan Brouwer}{Delft University of Technology, The Netherlands \and \url{http://jonathanb.nl}}{j.t.brouwer@student.tudelft.nl}{https://orcid.org/0000-0002-2469-548X}{}
-
-\author{Jesper Cockx}{Delft University of Technology, The Netherlands \and \url{http://jesper.sikanda.be}}{j.g.h.cockx@tudelft.nl}{https://orcid.org/0000-0003-3862-4073}{}
-
-\author{Aron Zwaan}{Delft University of Technology, The Netherlands \and \url{http://aronzwaan.github.io}}{a.s.zwaan@tudelft.nl}{https://orcid.org/0000-0002-1818-4245}{}
-
-\authorrunning{J. Brouwer, J. Cockx and A. Zwaan}
-
-\Copyright{Jonathan Brouwer, Jesper Cockx and Aron Zwaan} 
-\begin{CCSXML}
-    <ccs2012>
-    <concept>
-    <concept_id>10011007.10011006.10011039.10011311</concept_id>
-    <concept_desc>Software and its engineering~Semantics</concept_desc>
-    <concept_significance>500</concept_significance>
-    </concept>
-    <concept>
-    <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
-    <concept_desc>Software and its engineering~Functional languages</concept_desc>
-    <concept_significance>300</concept_significance>
-    </concept>
-    <concept>
-    <concept_id>10011007.10011006.10011041</concept_id>
-    <concept_desc>Software and its engineering~Compilers</concept_desc>
-    <concept_significance>300</concept_significance>
-    </concept>
-    </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Semantics}
-\ccsdesc[300]{Software and its engineering~Functional languages}
-\ccsdesc[300]{Software and its engineering~Compilers}
-
-\keywords{Spoofax, Statix, Dependent Types} %TODO mandatory; please add comma-separated list of keywords
-
-\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone.
-The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system.
-However, so far Statix has not been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking.
-In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
-While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.
-\end{abstract}
-
-\section{Introduction}
-
-Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages \cite{spoofax}.
-When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
-To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type systems.
-However, no attempts have been made to express dependently typed languages in Statix.
-
-Dependently typed languages are different from other languages, because they allow types to be parameterized by values.
-This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree.
-This expressiveness also makes dependent type systems more complicated to implement.
-Especially, deciding equality of types requires evaluation of the terms they are parameterized by.
-
-This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language.
-We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix.
-The goal is not only to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.
-
-In section \ref{sec:coc} of this paper we show that Statix is already usable as a tool for implementing dependent types.
-There were some challenges in implementing the dependent type system, which this paper also discusses.
-Finally, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
-
-\section{Calculus of Constructions}
-\label{sec:coc}
+\chapter{Calculus of Constructions}
+\label{chap:baselang}
 
@@
 
-\subsection{The language}
+\section{The language}
 \label{sec:coc-syntax}
@@
 
-\subsection{Scope Graphs}
+\section{Scope Graphs}
 \label{sec:coc-scopes}
@@
 
-\subsection{Beta Reductions}
+\section{Beta Reductions}
 \label{sec:coc-dynsyms}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 	\begin{mathpar}
@@
 
-\subsection{Type checking the Calculus of Constructions}
+\section{Type checking the Calculus of Constructions}
 \label{sec:coc-typecheck}
@@
 \end{figure}
-
-\subsection{Avoiding variable capture}
-One problem with the implementation presented in the previous sections is that it does not avoid variable capture.
-Variable capture happens when a term becomes bound because of a wrong substitution.
-For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type.
-\T : T.
-T| is \verb|T : Type -> T : T -> T|.
-This is not the correct type, and even worse, it is not possible to express the correct type without renaming!
-The problem is that there are multiple names, and there is no way to distinguish between them.
-This can be solved by using scopes to distinguish names.
-A full explanation of this will be in the master thesis.
-
-\section{Extensions}
-\label{sec:extensions}
-This section will discuss some further ideas that can be explored to build upon what has already been shown.
-These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.
-
-\paragraph*{Term Inference}
-In dependently typed languages, the value of types can often be inferred.
-Ideally, we would like to infer the most general type possible.
-However, this kind of analysis is not possible in Statix because you cannot reason over whether meta-variables are instantiated or not.
-We implemented an approximation to inference that can infer most types.
-For example, the type of \verb|x| in the function below can be inferred, because it is applied to \verb|true|, which is a boolean.
-\begin{lstlisting}
-(\x: _. x) true
-\end{lstlisting}
-
-\paragraph*{Inductive Data Types}
-Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types.
-We can also generate eliminators for these data types to use their values.
-All of this has been implemented in Statix.
-\begin{lstlisting}
-data Maybe (T : Type) = 
-    None : Maybe T,
-    Some : _: T -> Maybe T;
-\end{lstlisting}
-
-\paragraph*{Semantic Code Completion}
-Finally, we explored how semantic code completion presented by Pelsmaeker et al.
-\cite{codecompletion} applies to dependently typed languages.
-It works well, only showing completions that are semantically possible.
-
-For example, in the following code it would only suggest expressions that can be booleans (including variables that are booleans):
-\begin{lstlisting}
-let f = \b: Bool. b;
-f {{Expr}}
-\end{lstlisting}
-
-\section{Related Work}
-The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
-Van Antwerpen et al.
-\cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types.
-It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.
-
-Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language.
-The pi-forall language\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper.
-In principle, the implementations are very similar.
-For example, the inference rules presented in \cite{pi_forall} are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper.
-The primary difference is that they use a bidirectional type system, whereas this paper does not.
-
-There exist several so-called \emph{logical frameworks}, tools designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}.
-Since these tools are designed specifically for the task, implementing the type system takes less effort in them compared to Spoofax, but for other tasks such as defining a parser or editor services they are not as well equipped.
-Some logical frameworks such as Twelf and Dedukti support Miller's \emph{higher-order pattern unification}~\cite{Miller89}, which can be used as a more powerful way of inferring implicit arguments than the first-order unification built into Statix.
-Andromeda 2 also supports \emph{extensionality rules} that can match on the type of an equality.
-We expect that adding extensionality rules to our implementation would be possible to do in Statix, but we leave an actual implementation to future work.
-
-\section{Conclusion}
-
-We have demonstrated that the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph.
-We have also presented a few extensions to the Calculus of Constructions and discussed how they could be implemented.
-
-\bibliography{paper}
-
-\end{document}

### 2022-11-07 paper-thesis/chapters/03-variable-capture.tex
Create paper-thesis

--- 2022-10-05 paper/300-name-collisions.tex
@@
 
-\chapter{\label{chap:namecolls}Solving Name Collisions}
+\chapter{\label{chap:namecolls}Avoiding Variable Capturing}
+
+TODO Add reference to types and programming languages
 
@@
 \end{lstlisting}
+TODO Latex library for scope graphs
 
@@
 
-Almost all compilers that typecheck dependently typed languages use de Bruijn representation for variables \cite{TODO agda lean etc}.
+Almost all compilers that typecheck dependently typed languages use de Bruijn representation for variables.
 Using de Bruijn indices in statix is possible, but sacrifices a lot.
@@
 \begin{lstlisting}
-FnType(ScopedName(s1, T), Type(), FnType(ScopedName(s2, T), Var(ScopedName(s1, T)), Var(ScopedName(s1, T))))
+FnType(ScopedName(s1, T), Type(),
+	FnType(ScopedName(s2, T), Var(ScopedName(s1, T)), Var(ScopedName(s1, T))))
 \end{lstlisting}
 
-\subsection{Improving the readability of types}
+\section{Improving the readability of types}

### 2022-11-07 paper-thesis/chapters/98-relatedwork.tex
Create paper-thesis

--- 2022-10-27 paper-evcs/paper.tex
@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{oasics-v2021}
-%This is a template for producing OASIcs articles. 
-%See oasics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
-
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideOASIcs %uncomment to remove references to OASIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
-% Added because the included oasics-logo-bw.pdf is pdf 1.6 and this was giving a warning
-% Feel free to remove this if it breaks anything
-\pdfminorversion=6
-
-% Package is used to create inference rules
-\usepackage{mathpartir}
-
-% Commands used
-\newcommand{\scope}[2]{\langle#1 \: | \: #2\rangle}
-\newcommand{\bhr}[3]{ #1 \: #2 \underset{\beta h}{\Rightarrow} #3 }
-\newcommand{\toe}[3]{ \scope{#1}{#2} : #3 }
-\newcommand{\bred}[3]{ \scope{#1}{#2} \underset{\beta}{\Rightarrow} #3 }
-\newcommand{\beq}[2]{ #1 \underset{\beta}{=} #2 }
-\newcommand{\co}[1]{ \mathsf{#1} }
-
-\DeclareUnicodeCharacter{03BB}{L}
-\DeclareUnicodeCharacter{03A0}{M}
-
-\bibliographystyle{plainurl}
-
-\title{Dependently Typed Languages in Statix}
-
-\author{Jonathan Brouwer}{Delft University of Technology, The Netherlands \and \url{http://jonathanb.nl}}{j.t.brouwer@student.tudelft.nl}{}{}
-
-\author{Jesper Cockx}{Delft University of Technology, The Netherlands \and \url{http://jesper.sikanda.be}}{j.g.h.cockx@tudelft.nl}{}{}
-
-\author{Aron Zwaan}{Delft University of Technology, The Netherlands \and \url{http://aronzwaan.github.io}}{a.s.zwaan@tudelft.nl}{}{}
-
-\authorrunning{J. Brouwer, J. Cockx and A. Zwaan}
-
-\Copyright{Jonathan Brouwer, Jesper Cockx and Aron Zwaan} 
-\begin{CCSXML}
-    <ccs2012>
-    <concept>
-    <concept_id>10011007.10011006.10011039.10011311</concept_id>
-    <concept_desc>Software and its engineering~Semantics</concept_desc>
-    <concept_significance>500</concept_significance>
-    </concept>
-    <concept>
-    <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
-    <concept_desc>Software and its engineering~Functional languages</concept_desc>
-    <concept_significance>300</concept_significance>
-    </concept>
-    <concept>
-    <concept_id>10011007.10011006.10011041</concept_id>
-    <concept_desc>Software and its engineering~Compilers</concept_desc>
-    <concept_significance>300</concept_significance>
-    </concept>
-    </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Semantics}
-\ccsdesc[300]{Software and its engineering~Functional languages}
-\ccsdesc[300]{Software and its engineering~Compilers}
-
-\keywords{Spoofax, Statix, Dependent Types} %TODO mandatory; please add comma-separated list of keywords
-
-\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone.
-The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system.
-However, so far Statix has not yet been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking.
-In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
-While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.
-\end{abstract}
-
-\section{Introduction}
-
-Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages\cite{spoofax}.
-When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
-To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems.
-However, no attempts have been made to express dependently typed languages in Statix.
-
-Dependently typed languages are different from other languages, because they allow types to be parameterized by values.
-This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree.
-This expressiveness also makes dependent type systems more complicated to implement.
-Especially, deciding equality of types requires evaluation of the terms they are parameterized by.
-
-This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language.
-We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix.
-The goal is not to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.
-
-In section \ref{sec:coc} of this paper we show that Statix is already usable as a tool for implementing dependent types.
-There were some challenges in implementing the dependent type system, which this paper also discusses.
-Finally, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
-
-The implementation of the language is available on GitHub.\\
-\url{https://github.com/JonathanBrouwer/master-thesis/}
-
-\section{Calculus of Constructions}
-\label{sec:coc}
-
-In this section, we will describe how to implement a dependently typed language in Statix.
-In section \ref{sec:coc-syntax} we will describe the syntax of the language, then in section \ref{sec:coc-scopes} we will describe how scope graphs are used to type check the language.
-Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type-check the language.
-
-\subsection{The language}
-\label{sec:coc-syntax}
-
-The base language that has been implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}.
-One extra feature was added that is not present in the Calculus of Constructions, let bindings.
-Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
-The abstract syntax of the language is available in figure \ref{fig:syntax}.
-
-\begin{figure}[h]
-\begin{lstlisting}
-Type        : Expr
-Let         : ID * Expr * Expr -> Expr
-Var         : ID -> Expr
-FnType      : ID * Expr * Expr -> Expr
-FnConstruct : ID * Expr * Expr -> Expr
-FnDestruct  : Expr * Expr -> Expr
-\end{lstlisting}
-\caption{The syntax for the base language. FnConstruct is a lambda function, FnDestruct is application of a lambda function.}
-\label{fig:syntax}
-\end{figure}
-
-An example program is the following, which defines a polymorphic identity function and applies it to a function:
-
-\begin{lstlisting}
-let f = \T: Type. \x: T. x;
-f (_: Type -> Type) (\x: Type. x)
-\end{lstlisting}
-The AST of this program would be:
-\begin{lstlisting}
-Let(
-  "f", 
-  FnConstruct("T", Type(), FnConstruct("x", Var("T"), Var("x"))),
-  FnDestruct(
-	FnDestruct(Var("f"), FnConstruct("_", Type(), Type())),
-	FnConstruct("x", Type(), Var("x"))
-  )
-)
-\end{lstlisting}
-
-\subsection{Scope Graphs}
-\label{sec:coc-scopes}
-
-To type-check the base language, we need to store information about the names that are in scope at each point in the program.
-There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}
-
-In Statix, all this information can be stored in a \verb|scope graph|\cite{scope_graphs}, which is a feature of Statix.
-We only use a single type of edge, called \verb|P| (parent) edges.
-It also only has a single relation, called \verb|name|.
-This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.
-
-Next, we will introduce some Statix relations that can be used to interact with these scope graphs:
-
-\begin{lstlisting}
-sPutType  : scope * ID * Expr -> scope
-sPutSubst : scope * ID * (scope * Expr) -> scope
-sGetName  : scope * ID -> NameEntry
-sGetNames : scope * ID -> list((path * (ID * NameEntry)))
-sEmpty    : -> scope
-\end{lstlisting}
-The \verb|sPutType| and \verb|sPutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively.
-To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found.
-Finally, \verb|sEmpty| returns a fresh empty scope.
-
-\subsection{Beta Reductions}
-\label{sec:coc-dynsyms}
-
-A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare.
-
-We implemented beta reduction using a Krivine abstract machine\cite{krivine}.
-This turned out to be the more natural way of expressing this over substitution-based evaluation relation.
-We originally tried to implement the latter, which works fine for the base language.
-However, it runs into trouble when implementing inductive data types; more information about this will be in the full master thesis.
-An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.
-
-Note that the evaluator uses names rather than De Bruijn indices.
-In conventional dependently typed languages, evaluation is often done using De Bruijn indices.
-However, since the aim of this paper is to use the features of Statix, and scope graphs work based on names, we chose to use names.
-It would also stop us from using editor services that rely on \verb|.ref| annotations, such as renaming.
-
-We need to define multiple relations that will be used later for type-checking.
-First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
-The scope acts as the environment from \cite{krivine}, using \verb|NSubst| to store substitutions.
-All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
-We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|.
-Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
-One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
-
-Additionally, we define \verb|betaReduce| which fully beta reduces a term.
-It works by first calling \verb|betaReduceHead| and then matching on the head, calling \verb|betaReduce| on the sub-expressions of the head recursively.
-
-Finally, we define \verb|expectBetaEq|.
-This rule first beta reduces the heads of both sides, and then compares them.
-If the head is not the same, the rule fails.
-Otherwise, it recurses on the sub-expressions.
-One special case is when comparing two \verb|FnConstruct|s.
-Here we need to take into account alpha equality: two expressions which only differ in the names that they use should be considered equal.
-We implement this by substituting in the body of the functions, replacing their argument names with placeholders.
-
-\begin{figure}[h]
-	\begin{mathpar}	
-		% Type rule
-		\inferrule{ 
-		} { 
-			\bhr
-			{ \scope{s}{\co{Type}()} }
-			{ [] }
-			{ \scope{s}{\co{Type}()}} 
-		}
-
-		% Let rule
-		\inferrule{ 
-			\bhr 
-			{ \scope{ \co{sPutSubst}(s, n, (s, v))}{b} } 
-			{ t }
-			{ \scope{ s' } { e' } }
-		}{
-			\bhr
-			{ \scope{ s }{ \co{Let}(n, v, b) } }
-			{ t }
-			{ \scope { s' } { e' } }
-		}
-
-		% Var rule - NSubst
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
-			\\ \bhr
-			{\scope{se}{e}}
-			{t}
-			{\scope{se'}{e'}} 
-		}{
-			\bhr
-			{\scope{s}{\co{Var}(n)}}
-			{t}
-			{\scope{se'}{e'}} 
-		}
-
-		% Var rule - NType
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NType}(t) 
-		}{ 
-			\bhr
-			{\scope{s}{\co{Var}(n)}}
-			{t}
-			{\co{rebuild}(s, \co{Var}(n), t)} 
-		}
-
-		% FnType rule
-		\inferrule{ 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnType}(n, a, b)}}
-			{[]}
-			{\scope{s}{\co{FnType}(n, a, b)}} 
-		}
-
-		% FnConstruct rule - No args
-		\inferrule{ 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
-			{[]}
-			{\scope{s}{\co{FnConstruct}(n, a, b)}} 
-		}
-
-		% FnConstruct rule - Args
-		\inferrule{ 
-			\bhr
-			{\scope{\co{sPutSubst}(s, n, a)}{b}}
-			{ts}
-			{\scope{s'}{e'}} 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
-			{(t::ts)}
-			{\scope{s'}{e'}} 
-		}
-
-		% FnDestruct rule
-		\inferrule{ 
-			\bhr
-			{\scope{s}{f}}
-			{(a::ts)}
-			{\scope{s'}{e'}} 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnDestruct}(f, a)}}
-			{ts}
-			{\scope{s'}{e'}} 
-		}
-	\end{mathpar}
-	\caption{Rules for beta head reducing the Calculus of Constructions}
-	\label{fig:beta-head-reduce-rules}
-\end{figure}
-
-\subsection{Type-checking the Calculus of Constructions}
-\label{sec:coc-typecheck}
-
-We will define a Statix relation \verb|typeOfExpr| that takes a scope and an expression and type-checks the scope in the expression. It returns the type of the expression.
-
-\begin{lstlisting}
-typeOfExpr : scope * Expr -> Expr
-\end{lstlisting}
-We can then start defining type-checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
-\begin{enumerate}
-	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
-	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
-	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
-	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
-\end{enumerate}
-
-The inference rules above can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
-\begin{lstlisting}
-typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
-    typeOfExpr(s, v) == vt, sPutSubst(s, n, (s, v)) == s'.
-\end{lstlisting}
-
-\begin{figure}[h]
-	\begin{mathpar}	
-		% Type rule
-		\inferrule{ 
-		} {
-			\toe{s}{\co{Type}()}{\co{Type}()}
-		}
-
-		% Let rule
-		\inferrule{ 
-			\toe{s}{v}{vt}
-			\\ \toe{\co{sPutSubst}(s, n, (s, v))}{b}{t}
-		}{
-			\toe{s}{\co{Let}(n, v, b)}{t}
-		}
-
-		\\
-
-		% Var rule - NType
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NType}(t) 
-		}{ 
-			\toe{s}{\co{Var}(n)}{t}
-		}
-
-		% Var rule - NSubst
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
-			\\ \toe{se}{e}{t}
-		}{
-		 	\toe{s}{\co{Var}(n)}{t}
-		}
-
-		% FnType rule
-		\inferrule{ 
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()} 
-			\\ \bred{s}{a}{a'}
-			\\\\  \toe{\co{sPutType}(s, n, a')}{b}{bt}
-			\\ \beq{bt}{\co{Type}()}
-		} { 
-			\toe{s}{\co{FnType}(n, a, b)}{\co{Type}()}
-		}
-
-		% Force these to be on the same line, using negative spaces
-		\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!
-
-		% FnConstruct rule
-		\inferrule{ 
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()}
-			\\ \bred{s}{a}{a'} 
-			\\\\ \toe{\co{sPutType}(s, n, a')}{b}{bt}
-		} { 
-			\toe{s}{\co{FnConstruct}(n, a, b)}{\co{FnType}(n, a', bt)}
-		}
-
-		% FnDestruct rule
-		\inferrule{ 
-			\toe{s}{f}{ft}
-			\\ \bhr
-				{\scope{s}{ft}}
-				{[]}
-				{\scope{s'}{\co{FnType}(da, dt, db)}} 
-			\\\\ \toe{s}{a}{at} 
-			\\ \beq{at}{\scope{sf}{dt}}
-			\\ \bred{s'}{db}{db'}
-		} { 
-			\toe{s}{\co{FnDestruct}(f, a)}{db'}
-	 	}
-
-	\end{mathpar}
-	\caption{Rules for type-checking the Calculus of Constructions}
-	\label{fig:type-check-rules}
-\end{figure}
-
-\subsection{Avoiding variable capture}
-One problem with the current implementation is that it does not avoid variable capture.
-Variable capture happens when a term becomes bound because of a wrong substitution.
-For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type.
-\T : T.
-T| is \verb|T : Type -> T : T -> T|.
-This is not the correct type, and even worse, the correct type is not possible to express without renaming!
-The problem is that there are multiple names, and there is no way to distinguish between them.
-This can be solved by using scopes to distinguish names.
-A full explanation of this will be in the master thesis.
-
-\section{Extensions}
-\label{sec:extensions}
-This section will discuss some further ideas that can be explored to build upon what has already been shown.
-These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.
-
-\paragraph*{Term Inference}
-In dependently typed language, we can infer the value of types.
-Ideally, we would like to infer the most general type possible.
-However, this kind of analysis is not possible in Statix because you cannot reason over whether meta-variables are instantiated or not.
-We implemented an approximation to inference that can infer most types.
-\begin{lstlisting}
-(\x: _. x) true
-\end{lstlisting}
-
-\paragraph*{Inductive Data Types}
-Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types.
-We can also generate eliminators for these data types to use their values.
-All of this has been implemented in Statix.
-\begin{lstlisting}
-data Maybe (T : Type) = 
-    None : Maybe T,
-    Some : _: T -> Maybe T;
-\end{lstlisting}
-
-\paragraph*{Semantic Code Completion}
-Finally, we explored how semantic code completion presented in \cite{codecompletion} applies to dependently typed languages. It works well, only showing completions that are semantically possible.
-
-For example, in the following code it would only suggest expressions that can be booleans:
-\begin{lstlisting}
-let f = \b: Bool. b;
-f {{Expr}}
-\end{lstlisting}
-
-\section{Related Work}
+\chapter{Related Work}
 The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
-In section 2.5 of Scopes as Types \cite{scopes_as_types}, an implementation of System F is presented; it does lazy substitutions, by using scopes as types.
+Van Antwerpen et al.
+\cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types.
 It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.
@@
 We expect that adding extensionality rules to our implementation would be possible to do in Statix, but we leave an actual implementation to future work.
-
-\section{Conclusion}
-
-We have demonstrated that the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph.
-We have also presented a few extensions to the Calculus of Constructions and discussed how they could be implemented.
-
-\bibliography{paper}
-
-\end{document}

### 2022-11-07 paper-thesis/chapters/99-conclusion.tex
Create paper-thesis

--- 2022-10-27 paper-evcs/paper.tex
@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{oasics-v2021}
-%This is a template for producing OASIcs articles. 
-%See oasics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
-
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideOASIcs %uncomment to remove references to OASIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
-% Added because the included oasics-logo-bw.pdf is pdf 1.6 and this was giving a warning
-% Feel free to remove this if it breaks anything
-\pdfminorversion=6
-
-% Package is used to create inference rules
-\usepackage{mathpartir}
-
-% Commands used
-\newcommand{\scope}[2]{\langle#1 \: | \: #2\rangle}
-\newcommand{\bhr}[3]{ #1 \: #2 \underset{\beta h}{\Rightarrow} #3 }
-\newcommand{\toe}[3]{ \scope{#1}{#2} : #3 }
-\newcommand{\bred}[3]{ \scope{#1}{#2} \underset{\beta}{\Rightarrow} #3 }
-\newcommand{\beq}[2]{ #1 \underset{\beta}{=} #2 }
-\newcommand{\co}[1]{ \mathsf{#1} }
-
-\DeclareUnicodeCharacter{03BB}{L}
-\DeclareUnicodeCharacter{03A0}{M}
-
-\bibliographystyle{plainurl}
-
-\title{Dependently Typed Languages in Statix}
-
-\author{Jonathan Brouwer}{Delft University of Technology, The Netherlands \and \url{http://jonathanb.nl}}{j.t.brouwer@student.tudelft.nl}{}{}
-
-\author{Jesper Cockx}{Delft University of Technology, The Netherlands \and \url{http://jesper.sikanda.be}}{j.g.h.cockx@tudelft.nl}{}{}
-
-\author{Aron Zwaan}{Delft University of Technology, The Netherlands \and \url{http://aronzwaan.github.io}}{a.s.zwaan@tudelft.nl}{}{}
-
-\authorrunning{J. Brouwer, J. Cockx and A. Zwaan}
-
-\Copyright{Jonathan Brouwer, Jesper Cockx and Aron Zwaan} 
-\begin{CCSXML}
-    <ccs2012>
-    <concept>
-    <concept_id>10011007.10011006.10011039.10011311</concept_id>
-    <concept_desc>Software and its engineering~Semantics</concept_desc>
-    <concept_significance>500</concept_significance>
-    </concept>
-    <concept>
-    <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
-    <concept_desc>Software and its engineering~Functional languages</concept_desc>
-    <concept_significance>300</concept_significance>
-    </concept>
-    <concept>
-    <concept_id>10011007.10011006.10011041</concept_id>
-    <concept_desc>Software and its engineering~Compilers</concept_desc>
-    <concept_significance>300</concept_significance>
-    </concept>
-    </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Semantics}
-\ccsdesc[300]{Software and its engineering~Functional languages}
-\ccsdesc[300]{Software and its engineering~Compilers}
-
-\keywords{Spoofax, Statix, Dependent Types} %TODO mandatory; please add comma-separated list of keywords
-
-\nolinenumbers %uncomment to disable line numbering
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone.
-The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system.
-However, so far Statix has not yet been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking.
-In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
-While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.
-\end{abstract}
-
-\section{Introduction}
-
-Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages\cite{spoofax}.
-When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
-To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems.
-However, no attempts have been made to express dependently typed languages in Statix.
-
-Dependently typed languages are different from other languages, because they allow types to be parameterized by values.
-This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree.
-This expressiveness also makes dependent type systems more complicated to implement.
-Especially, deciding equality of types requires evaluation of the terms they are parameterized by.
-
-This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language.
-We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix.
-The goal is not to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.
-
-In section \ref{sec:coc} of this paper we show that Statix is already usable as a tool for implementing dependent types.
-There were some challenges in implementing the dependent type system, which this paper also discusses.
-Finally, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
-
-The implementation of the language is available on GitHub.\\
-\url{https://github.com/JonathanBrouwer/master-thesis/}
-
-\section{Calculus of Constructions}
-\label{sec:coc}
-
-In this section, we will describe how to implement a dependently typed language in Statix.
-In section \ref{sec:coc-syntax} we will describe the syntax of the language, then in section \ref{sec:coc-scopes} we will describe how scope graphs are used to type check the language.
-Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type-check the language.
-
-\subsection{The language}
-\label{sec:coc-syntax}
-
-The base language that has been implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}.
-One extra feature was added that is not present in the Calculus of Constructions, let bindings.
-Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
-The abstract syntax of the language is available in figure \ref{fig:syntax}.
-
-\begin{figure}[h]
-\begin{lstlisting}
-Type        : Expr
-Let         : ID * Expr * Expr -> Expr
-Var         : ID -> Expr
-FnType      : ID * Expr * Expr -> Expr
-FnConstruct : ID * Expr * Expr -> Expr
-FnDestruct  : Expr * Expr -> Expr
-\end{lstlisting}
-\caption{The syntax for the base language. FnConstruct is a lambda function, FnDestruct is application of a lambda function.}
-\label{fig:syntax}
-\end{figure}
-
-An example program is the following, which defines a polymorphic identity function and applies it to a function:
-
-\begin{lstlisting}
-let f = \T: Type. \x: T. x;
-f (_: Type -> Type) (\x: Type. x)
-\end{lstlisting}
-The AST of this program would be:
-\begin{lstlisting}
-Let(
-  "f", 
-  FnConstruct("T", Type(), FnConstruct("x", Var("T"), Var("x"))),
-  FnDestruct(
-	FnDestruct(Var("f"), FnConstruct("_", Type(), Type())),
-	FnConstruct("x", Type(), Var("x"))
-  )
-)
-\end{lstlisting}
-
-\subsection{Scope Graphs}
-\label{sec:coc-scopes}
-
-To type-check the base language, we need to store information about the names that are in scope at each point in the program.
-There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}
-
-In Statix, all this information can be stored in a \verb|scope graph|\cite{scope_graphs}, which is a feature of Statix.
-We only use a single type of edge, called \verb|P| (parent) edges.
-It also only has a single relation, called \verb|name|.
-This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.
-
-Next, we will introduce some Statix relations that can be used to interact with these scope graphs:
-
-\begin{lstlisting}
-sPutType  : scope * ID * Expr -> scope
-sPutSubst : scope * ID * (scope * Expr) -> scope
-sGetName  : scope * ID -> NameEntry
-sGetNames : scope * ID -> list((path * (ID * NameEntry)))
-sEmpty    : -> scope
-\end{lstlisting}
-The \verb|sPutType| and \verb|sPutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively.
-To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found.
-Finally, \verb|sEmpty| returns a fresh empty scope.
-
-\subsection{Beta Reductions}
-\label{sec:coc-dynsyms}
-
-A unique requirement for dependently typed languages is beta reduction during type-checking, since types may require evaluation to compare.
-
-We implemented beta reduction using a Krivine abstract machine\cite{krivine}.
-This turned out to be the more natural way of expressing this over substitution-based evaluation relation.
-We originally tried to implement the latter, which works fine for the base language.
-However, it runs into trouble when implementing inductive data types; more information about this will be in the full master thesis.
-An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.
-
-Note that the evaluator uses names rather than De Bruijn indices.
-In conventional dependently typed languages, evaluation is often done using De Bruijn indices.
-However, since the aim of this paper is to use the features of Statix, and scope graphs work based on names, we chose to use names.
-It would also stop us from using editor services that rely on \verb|.ref| annotations, such as renaming.
-
-We need to define multiple relations that will be used later for type-checking.
-First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
-The scope acts as the environment from \cite{krivine}, using \verb|NSubst| to store substitutions.
-All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
-We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|.
-Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
-One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
-
-Additionally, we define \verb|betaReduce| which fully beta reduces a term.
-It works by first calling \verb|betaReduceHead| and then matching on the head, calling \verb|betaReduce| on the sub-expressions of the head recursively.
-
-Finally, we define \verb|expectBetaEq|.
-This rule first beta reduces the heads of both sides, and then compares them.
-If the head is not the same, the rule fails.
-Otherwise, it recurses on the sub-expressions.
-One special case is when comparing two \verb|FnConstruct|s.
-Here we need to take into account alpha equality: two expressions which only differ in the names that they use should be considered equal.
-We implement this by substituting in the body of the functions, replacing their argument names with placeholders.
-
-\begin{figure}[h]
-	\begin{mathpar}	
-		% Type rule
-		\inferrule{ 
-		} { 
-			\bhr
-			{ \scope{s}{\co{Type}()} }
-			{ [] }
-			{ \scope{s}{\co{Type}()}} 
-		}
-
-		% Let rule
-		\inferrule{ 
-			\bhr 
-			{ \scope{ \co{sPutSubst}(s, n, (s, v))}{b} } 
-			{ t }
-			{ \scope{ s' } { e' } }
-		}{
-			\bhr
-			{ \scope{ s }{ \co{Let}(n, v, b) } }
-			{ t }
-			{ \scope { s' } { e' } }
-		}
-
-		% Var rule - NSubst
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
-			\\ \bhr
-			{\scope{se}{e}}
-			{t}
-			{\scope{se'}{e'}} 
-		}{
-			\bhr
-			{\scope{s}{\co{Var}(n)}}
-			{t}
-			{\scope{se'}{e'}} 
-		}
-
-		% Var rule - NType
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NType}(t) 
-		}{ 
-			\bhr
-			{\scope{s}{\co{Var}(n)}}
-			{t}
-			{\co{rebuild}(s, \co{Var}(n), t)} 
-		}
-
-		% FnType rule
-		\inferrule{ 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnType}(n, a, b)}}
-			{[]}
-			{\scope{s}{\co{FnType}(n, a, b)}} 
-		}
-
-		% FnConstruct rule - No args
-		\inferrule{ 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
-			{[]}
-			{\scope{s}{\co{FnConstruct}(n, a, b)}} 
-		}
-
-		% FnConstruct rule - Args
-		\inferrule{ 
-			\bhr
-			{\scope{\co{sPutSubst}(s, n, a)}{b}}
-			{ts}
-			{\scope{s'}{e'}} 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
-			{(t::ts)}
-			{\scope{s'}{e'}} 
-		}
-
-		% FnDestruct rule
-		\inferrule{ 
-			\bhr
-			{\scope{s}{f}}
-			{(a::ts)}
-			{\scope{s'}{e'}} 
-		} { 
-			\bhr
-			{\scope{s}{\co{FnDestruct}(f, a)}}
-			{ts}
-			{\scope{s'}{e'}} 
-		}
-	\end{mathpar}
-	\caption{Rules for beta head reducing the Calculus of Constructions}
-	\label{fig:beta-head-reduce-rules}
-\end{figure}
-
-\subsection{Type-checking the Calculus of Constructions}
-\label{sec:coc-typecheck}
-
-We will define a Statix relation \verb|typeOfExpr| that takes a scope and an expression and type-checks the scope in the expression. It returns the type of the expression.
-
-\begin{lstlisting}
-typeOfExpr : scope * Expr -> Expr
-\end{lstlisting}
-We can then start defining type-checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
-\begin{enumerate}
-	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
-	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
-	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
-	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
-\end{enumerate}
-
-The inference rules above can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
-\begin{lstlisting}
-typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
-    typeOfExpr(s, v) == vt, sPutSubst(s, n, (s, v)) == s'.
-\end{lstlisting}
-
-\begin{figure}[h]
-	\begin{mathpar}	
-		% Type rule
-		\inferrule{ 
-		} {
-			\toe{s}{\co{Type}()}{\co{Type}()}
-		}
-
-		% Let rule
-		\inferrule{ 
-			\toe{s}{v}{vt}
-			\\ \toe{\co{sPutSubst}(s, n, (s, v))}{b}{t}
-		}{
-			\toe{s}{\co{Let}(n, v, b)}{t}
-		}
-
-		\\
-
-		% Var rule - NType
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NType}(t) 
-		}{ 
-			\toe{s}{\co{Var}(n)}{t}
-		}
-
-		% Var rule - NSubst
-		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
-			\\ \toe{se}{e}{t}
-		}{
-		 	\toe{s}{\co{Var}(n)}{t}
-		}
-
-		% FnType rule
-		\inferrule{ 
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()} 
-			\\ \bred{s}{a}{a'}
-			\\\\  \toe{\co{sPutType}(s, n, a')}{b}{bt}
-			\\ \beq{bt}{\co{Type}()}
-		} { 
-			\toe{s}{\co{FnType}(n, a, b)}{\co{Type}()}
-		}
-
-		% Force these to be on the same line, using negative spaces
-		\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!
-
-		% FnConstruct rule
-		\inferrule{ 
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()}
-			\\ \bred{s}{a}{a'} 
-			\\\\ \toe{\co{sPutType}(s, n, a')}{b}{bt}
-		} { 
-			\toe{s}{\co{FnConstruct}(n, a, b)}{\co{FnType}(n, a', bt)}
-		}
-
-		% FnDestruct rule
-		\inferrule{ 
-			\toe{s}{f}{ft}
-			\\ \bhr
-				{\scope{s}{ft}}
-				{[]}
-				{\scope{s'}{\co{FnType}(da, dt, db)}} 
-			\\\\ \toe{s}{a}{at} 
-			\\ \beq{at}{\scope{sf}{dt}}
-			\\ \bred{s'}{db}{db'}
-		} { 
-			\toe{s}{\co{FnDestruct}(f, a)}{db'}
-	 	}
-
-	\end{mathpar}
-	\caption{Rules for type-checking the Calculus of Constructions}
-	\label{fig:type-check-rules}
-\end{figure}
-
-\subsection{Avoiding variable capture}
-One problem with the current implementation is that it does not avoid variable capture.
-Variable capture happens when a term becomes bound because of a wrong substitution.
-For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type.
-\T : T.
-T| is \verb|T : Type -> T : T -> T|.
-This is not the correct type, and even worse, the correct type is not possible to express without renaming!
-The problem is that there are multiple names, and there is no way to distinguish between them.
-This can be solved by using scopes to distinguish names.
-A full explanation of this will be in the master thesis.
-
-\section{Extensions}
-\label{sec:extensions}
-This section will discuss some further ideas that can be explored to build upon what has already been shown.
-These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.
-
-\paragraph*{Term Inference}
-In dependently typed language, we can infer the value of types.
-Ideally, we would like to infer the most general type possible.
-However, this kind of analysis is not possible in Statix because you cannot reason over whether meta-variables are instantiated or not.
-We implemented an approximation to inference that can infer most types.
-\begin{lstlisting}
-(\x: _. x) true
-\end{lstlisting}
-
-\paragraph*{Inductive Data Types}
-Another common feature in dependently typed language is support for inductive data types, including support for parameterized and indexed data types.
-We can also generate eliminators for these data types to use their values.
-All of this has been implemented in Statix.
-\begin{lstlisting}
-data Maybe (T : Type) = 
-    None : Maybe T,
-    Some : _: T -> Maybe T;
-\end{lstlisting}
-
-\paragraph*{Semantic Code Completion}
-Finally, we explored how semantic code completion presented in \cite{codecompletion} applies to dependently typed languages. It works well, only showing completions that are semantically possible.
-
-For example, in the following code it would only suggest expressions that can be booleans:
-\begin{lstlisting}
-let f = \b: Bool. b;
-f {{Expr}}
-\end{lstlisting}
-
-\section{Related Work}
-The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
-In section 2.5 of Scopes as Types \cite{scopes_as_types}, an implementation of System F is presented; it does lazy substitutions, by using scopes as types.
-It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.
-
-Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language.
-The pi-forall language\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper.
-In principle, the implementations are very similar.
-For example, the inference rules presented in \cite{pi_forall} are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper.
-The primary difference is that they use a bidirectional type system, whereas this paper does not.
-
-There exist several so-called \emph{logical frameworks}, tools designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}.
-Since these tools are designed specifically for the task, implementing the type system takes less effort in them compared to Spoofax, but for other tasks such as defining a parser or editor services they are not as well equipped.
-Some logical frameworks such as Twelf and Dedukti support Miller's \emph{higher-order pattern unification}~\cite{Miller89}, which can be used as a more powerful way of inferring implicit arguments than the first-order unification built into Statix.
-Andromeda 2 also supports \emph{extensionality rules} that can match on the type of an equality.
-We expect that adding extensionality rules to our implementation would be possible to do in Statix, but we leave an actual implementation to future work.
-
-\section{Conclusion}
+\chapter{Conclusion}
 
@@
 We have also presented a few extensions to the Calculus of Constructions and discussed how they could be implemented.
-
-\bibliography{paper}
-
-\end{document}

### 2022-11-07 paper-thesis/chapters/03-variable-capture.tex
Create more structure

@@
 
-The implementation of the base language shown in section \ref{chap:baselang} has one big problem, that is name collisions.
+The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is name collisions.
 This section will explore several ways of solving these collisions.
@@
 \section{Improving the readability of types}
+
+Because the expression above, with \verb|ScopedName|s, is not particularly readable, we add a new post-analysis Stratego pass that converts the \verb|ScopedName|s to ticked named.
+For example, the above would be transformed to:
+
+\begin{lstlisting}
+FnType(T, Type(), FnType(T', Var(T)), Var(T)))
+\end{lstlisting}
+Ticks are added to names where necessary. We do this by following these rules:
+
+\begin{enumerate}
+	\item When we encounter a \verb|ScopedName| for the first time, we keep adding ticks to the name until we find a name that has not been used before. 
+	\item We define a dynamic rule \verb|Rename :: string -> string| and we store the new name we generated using this rule.
+	\item When we encounter a \verb|ScopedName|, use the \verb|Rename| rule to find what the name was transformed to.
+\end{enumerate}

### 2022-11-07 paper-thesis/chapters/04-adding-booleans.tex
Create more structure

\chapter{\label{chap:bools}Booleans and other "simple" constructs that really need a better title}

In this chapter, we will add booleans, postulates and type asserts to the language. The goal of this is to show that this language is easy to extend, and to add some features that make testing the language easier.

\section{Booleans}

This section describes how to add Booleans to the language. We will add the type of booleans `Bool`, `true`, `false` and `if`, allowing for programs such as the following:

\begin{lstlisting}
let and = \x: Bool. \y: Bool. if x then y else false end
\end{lstlisting}

\section{Postulate and Type Assert}

### 2022-11-07 paper-thesis/chapters/04-adding-booleans.tex
More writing

@@
 
-This section describes how to add Booleans to the language. We will add the type of booleans `Bool`, `true`, `false` and `if`, allowing for programs such as the following:
+This section describes how to add Booleans to the language.
+We will add the type of booleans \verb|Bool|, \verb|true|, \verb|false| and \verb|if|.
+The \verb|if| expression is not dependent, it expects both branches to have the same type.
+An example of a program with booleans:
 
@@
 \end{lstlisting}
+The constructors for the language are:
 
-\section{Postulate and Type Assert}
+\begin{lstlisting}
+BoolType : Expr
+BoolFalse : Expr
+BoolTrue : Expr
+BoolIf : Expr * Expr * Expr -> Expr
+\end{lstlisting}
+Then, the rules for beta reducing the language are in figure \ref{fig:bool-rules-beta}.
+There is one particularly interesting case, that is how to beta-reduce if statements.
+Converting this rule to Statix is not entirely trivial, since you need to choose which rule to apply based on what \verb|c| evaluates to.
+A new rule is needed, which has 3 cases.
+One for if \verb|c| evaluates to true, one for if \verb|c| evaluates to false, and finally a third case for other cases (such as when \verb|c| is a variable that does not have a substitution).
+Remember that \verb|rebuild| is the rule introduced in chapter \ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
+These rules are stated below:
+
+\begin{lstlisting}
+betaReduceHead((s, BoolIf(c, b1, b2)), t) = 
+  betaReduceHeadIf(s, betaReduceHead((s, c), []), c, b1, b2, t).
+betaReduceHeadIf(s, (_, BoolTrue()), _, b1, _, t) = 
+  betaReduceHead((s, b1), t).
+betaReduceHeadIf(s, (_, BoolFalse()), _, _, b2, t) = 
+  betaReduceHead((s, b2), t).
+betaReduceHeadIf(s, _, c, b1, b2, t) = 
+  rebuild((s, BoolIf(c, b1, b2)), t).
+\end{lstlisting}
+
+\begin{figure}[ht]
+	\begin{mathpar}
+		\inferrule{
+		} {
+			\bhr
+			{ \scope{s}{\co{BoolTrue}()} }
+			{ [] }
+			{ \scope{s}{\co{BoolTrue}()}}
+		}
+
+		\inferrule{
+		} {
+			\bhr
+			{ \scope{s}{\co{BoolFalse}()} }
+			{ [] }
+			{ \scope{s}{\co{BoolFalse}()}}
+		}
+
+		\inferrule{
+		} {
+			\bhr
+			{ \scope{s}{\co{BoolType}()} }
+			{ [] }
+			{ \scope{s}{\co{BoolType}()}}
+		}
+
+		\inferrule{
+			\bhr
+			{ \scope{s}{c} }
+			{ [] }
+			{ \scope{s'}{BoolTrue()} }
+			\\\bhr
+			{ \scope{s}{b1} }
+			{ t }
+			{ \scope{s''}{b1'} }
+		} {
+			\bhr
+			{ \scope{s}{\co{BoolIf}(c, b1, b2)} }
+			{ t }
+			{ \scope{s}{b1'}}
+		}
+
+		\inferrule{
+			\bhr
+			{ \scope{s}{c} }
+			{ [] }
+			{ \scope{s'}{BoolFalse()} }
+			\\\bhr
+			{ \scope{s}{b2} }
+			{ t }
+			{ \scope{s''}{b2'} }
+		} {
+			\bhr
+			{ \scope{s}{\co{BoolIf}(c, b1, b2)} }
+			{ t }
+			{ \scope{s}{b2'}}
+		}
+	\end{mathpar}
+	\caption{Rules for beta head reducing booleans}
+	\label{fig:bool-rules-beta}
+\end{figure}
+
+\section{Postulate}
+
+Next we will add \verb|Postulate| to the language.
+
+\section{Type Assert}
+
+Finally, we will add \verb|TypeAssert| to the language.

### 2022-11-07 paper-thesis/chapters/05-adding-inference.tex
More writing

\chapter{Inference}

\section{Background: What is inference?}

Inference

\section{General inference in Statix}
\label{general-inference}

\section{Greedy inference}

Instead, we implement a strategy that I would like to call greedy inference (TODO: does this have a name?).

\section{In statix}

First, we introduce a new constructor, \verb|Infer : Expr -> Expr|. This will be used to store metavariables which we want to infer. This is a hack we use, ideally we would like to query statix to see if a metavariable is initialized, but as discussed in section \ref{general-inference} this is not possible.

The constructor is introduced when we encounter a \verb|_| variable, a sign that something needs to be inferred. When type-checking this, we create an infer constructor. The constraint at the bottom cannot be solved immediately, since the \verb|q| variable is unbound, but if \verb|q| is bound later the constraint will be evaluated.

\begin{lstlisting}
typeOfExpr_(s, Var(Syn("_"))) = (Infer(q), qt) :-
	(_, qt) == typeOfExpr(sEmpty(), q).
\end{lstlisting}
When an \verb|Infer| needs to be type-checked, the logic is very similar, except that we don't need to generate a new metavariable:

\begin{lstlisting}
typeOfExpr_(s, Infer(q)) = (Infer(q), t) :-
	typeOfExpr_(s, q) == (_, t).
\end{lstlisting}
When we encounter a \verb|Infer| in \verb|betaReduceHead|, we keep it intact, because we still want to know that it is a \verb|Infer| in the \verb|expectBetaEq| rule.

The \verb|expectBetaEq| is where it gets interesting. We have a rule for each constructor in the language. For simple cases like \verb|BoolTrue|, we assert that \verb|e2| must be equal.
\begin{lstlisting}
expectBetaEq_((s1, e1@BoolTrue()), (_, Infer(e2))) :- e1 == e2.
\end{lstlisting}
For more complicated constructors such as \verb|FnType|, we generate new metavariables for the subexpressions, and assert they it must be beta-equal to the provided ones:
\begin{lstlisting}
expectBetaEq_(
	(s1, e1@FnType(arg_name1, arg_type1, body1)), 
	(_, Infer(e2))) :- {arg_type2 body2}
  e2 == FnType(arg_name1, Infer(arg_type2), Infer(body2)),
  expectBetaEq_((s1, e1), (sEmpty(), e2)).
\end{lstlisting}

Finally, there is a case for if we encounter two \verb|Infer|s. Ideally, we would look at whether one infer is instantiated, and apply the rules above. However, this is not possible since we cannot do such queries in Statix. Instead, we say that they must be exactly equal and we hope for the best. There are situations where this rule fails (where e1 and e2 are beta-equal but not identical), but this works well enough for practical use.
\begin{lstlisting}
expectBetaEq_((_, Infer(e1)), (_, Infer(e2))) :-
    e1 == e2.
\end{lstlisting}

\section{Analysis: How well does it work?}

### 2022-11-08 rust-coc/src/main.rs
Play around with dedukti and rust

@@ -1,0 +1,105 @@
+use std::rc::Rc;
+
+type RExpr = Rc<Box<Expr>>;
+type REnv = Rc<Box<Env>>;
+
+enum Expr {
+    Type,
+    Var(usize),
+    Let(RExpr, RExpr),
+    FnType(RExpr, RExpr),
+    FnConstruct(RExpr, RExpr),
+    FnDestruct(RExpr, RExpr),
+}
+
+enum Env {
+    NType(RExpr, Option<REnv>),
+    NSubst(SExpr, Option<REnv>),
+}
+
+impl Env {
+    fn depth(&self) -> usize {
+        match self {
+            Env::NType(_, t) => 1 + t.as_ref().map(|t| t.depth()).unwrap_or(0),
+            Env::NSubst(_, t) => 1 + t.as_ref().map(|t| t.depth()).unwrap_or(0),
+        }
+    }
+
+    fn parent(&self) -> &Option<REnv> {
+        match self {
+            Env::NType(_, t) => t,
+            Env::NSubst(_, t) => t,
+        }
+    }
+}
+
+fn s_put_type(s: REnv, e: RExpr) -> REnv {
+    Rc::new(Box::new(Env::NType(e, Some(s))))
+}
+
+fn s_put_subst(s: REnv, e: SExpr) -> REnv {
+    Rc::new(Box::new(Env::NSubst(e, Some(s))))
+}
+
+#[derive(Clone)]
+struct SExpr(REnv, RExpr);
+
+fn bhr(e_orig: SExpr) -> SExpr {
+    let mut e = e_orig.clone();
+    let mut stack: Vec<SExpr> = Vec::new();
+
+    loop {
+        match e.1.as_ref().as_ref() {
+            Expr::Type => break,
+            Expr::Var(mut i) => {
+                let mut env = &e.0;
+                while i > 0 {
+                    env = env.parent().as_ref().unwrap();
+                    i -= 1;
+                }
+                match &***env {
+                    Env::NType(_, _) => break,
+                    Env::NSubst(se, _) => e = se.clone(),
+                }
+            }
+            Expr::Let(v, b) => {
+                e = SExpr(
+                    s_put_subst(e.0.clone(), SExpr(e.0, v.clone())),
+                    b.clone(),
+                )
+            }
+            Expr::FnType(_, _) => break,
+            Expr::FnConstruct(_, _) if stack.len() == 0 => break,
+            Expr::FnConstruct(_, b) => {
[…]

### 2022-11-08 haskell-coc/app/Main.hs
Haskell-coc work

@@ -1,0 +1,44 @@
+module Main (main) where
+
+import Data.Maybe (fromMaybe)
+
+
+data Expr =
+  Type
+  | Let Expr Expr
+  | Var Int
+  | FnType Expr Expr
+  | FnConstruct Expr Expr
+  | FnDestruct Expr Expr
+  deriving Show
+
+type SExpr = (Env, Expr)
+
+type Env = [EnvEntry]
+
+data EnvEntry = NType Expr | NSubst SExpr
+  deriving Show
+
+sPutSubst :: Env -> SExpr -> Env
+sPutSubst env v = NSubst v : env
+
+sPutType :: Env -> Expr -> Env
+sPutType env v = NType v : env
+
+brh :: SExpr -> SExpr
+brh e = fromMaybe e (brh_ e [])
+
+brh_ :: SExpr -> [SExpr] -> Maybe SExpr
+brh_ (s, Type) [] = Just (s, Type)
+brh_ (s, Let v b) tt = brh_ (sPutSubst s (s, v), b) tt
+brh_ (s, Var i) tt = case s !! i of
+  NType _ -> if null tt then Just (s, Var i) else Nothing
+  NSubst se -> brh_ se tt
+brh_ (s, FnType a b) [] = Just (s, FnType a b)
+brh_ (s, FnConstruct a b) [] = Just (s, FnConstruct a b)
+brh_ (s, FnConstruct _ b) (t : tt) = brh_ (sPutSubst s t, b) tt
+brh_ (s, FnDestruct f a) tt = brh_ (s, f) ((s, a) : tt)
+
+
+main :: IO ()
+main = putStr ""

### 2022-11-09 haskell-coc/out/production/haskell-coc/Main.hs
Work on haskell coc

@@ -39,6 +39,58 @@ brh_ (s, FnConstruct a b) [] = Just (s, FnConstruct a b)
 brh_ (s, FnConstruct _ b) (t : tt) = brh_ (sPutSubst s t, b) tt
 brh_ (s, FnDestruct f a) tt = brh_ (s, f) ((s, a) : tt)
 
+br :: SExpr -> Expr
+br se = case brh se of
+  (_, Type) -> Type
+  (_, Var i) -> Var i
+  (s, FnType a b) -> FnType (br (s, a)) (br (s, b))
+  (s, FnConstruct a b) -> FnConstruct (br (s, a)) (br (s, b))
+  (s, FnDestruct a b) -> FnDestruct (br (s, a)) (br (s, b))
+
+beq :: SExpr -> SExpr -> Either String ()
+beq e1 e2 = case (brh e1, brh e2) of
+  ((_, Type), (_, Type)) -> pure ()
+  ((_, Var i), (_, Var j)) -> if i == j then pure () else Left $ "Expected variable " ++ show j ++ ", but got variable " ++ show i
+  ((s1, FnType a1 b1), (s2, FnType a2 b2)) -> do
+     beq (s1, a1) (s2, a2)
+     beq (sPutType s1 a1, b1) (sPutType s2 a2, b2)
+  ((s1, FnConstruct a1 b1), (s2, FnConstruct a2 b2)) -> do
+     beq (s1, a1) (s2, a2)
+     beq (s1, b1) (s2, b2)
+  ((s1, FnDestruct a1 b1), (s2, FnDestruct a2 b2)) -> do
+    beq (s1, a1) (s2, a2)
+    beq (s1, b1) (s2, b2)
+  _ -> Left $ "Expected " ++ show e2 ++ ", but got " ++ show e1
+
+tc :: SExpr -> Either String Expr
+tc (_, Type) = pure Type
+tc (s, Var i) = case s !! i of
+  NType t -> pure t
+  NSubst es -> tc es
+tc (s, Let v b) = do
+  tc (s, v)
+  tc (sPutSubst s (s, v), b)
+tc (s, FnType a b) = do
+  at <- tc (s, a)
+  beq ([], at) ([], Type)
+  let at' = br (s, a)
+  bt <- tc (sPutType s at', b)
+  beq ([], bt) ([], Type)
+  pure Type
+tc (s, FnConstruct a b) = do
+  at <- tc (s, a)
+  beq ([], at) ([], Type)
[…]

### 2022-11-09 paper-thesis/chapters/03-variable-capture.tex
Work on the haskell-coc implementation

@@
 
-The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is name collisions.
-This section will explore several ways of solving these collisions.
-An example of such a collision is the following: What is the type of this expression (a polymorphic identity function)?
+The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is variable capture.
+This section will explore several ways of solving this.
+An example is the following: What is the type of this expression (a polymorphic identity function)?
 \begin{lstlisting}
@@
 \begin{lstlisting}
-typeOfExpr(_, FnConstruct("T", Type(), FnConstruct("T", Var("T"), Var("T"))))
+typeOfExpr(_, FnConstruct("T", Type(), 
+    FnConstruct("T", Var("T"), Var("T")))
+)
 \end{lstlisting}
@@
 \noindent
-This creates a new node in the scope graph, and then type checks the body with this scope.
+This first handles the outer \verb|FnConstruct|, it creates a new node in the scope graph, and then type checks the body with this scope.
 
@@
 typeOfExpr(s1, FnConstruct("T", Var("T"), Var("T")))
-
-(T : Type)
-   [s1]
 \end{lstlisting}
-TODO Latex library for scope graphs
+\begin{tikzpicture}[scopegraph, node distance = 2.5em and 3.0em]
+\node[scope] (s1) {$s1$};
+\node[type, above of = s1] (s1l) {$\texttt{T : Type()}$};
+\draw (s1) edge[type] (s1l);
+\end{tikzpicture}
 
 \noindent
-The same thing happens, the body of the \verb|FnConstruct| is typechecked with a new scope. Note that the T in the type of the second T is ambiguous.
+The same thing happens, the body of the \verb|FnConstruct| is typechecked with a new scope.
+Note that \verb|Var("T")| in the type of the second \verb|T| is ambiguous, does it refer to the first or second node?
 
@@
 typeOfExpr(s2, Var("T"))
+\end{lstlisting}
+\begin{tikzpicture}[scopegraph, node distance = 2.5em and 2.5em]
+	\node[scope] (s1) {$s1$};
+	\node[type, above of = s1] (s1l) {$\texttt{T : Type()}$};
+	\draw (s1) edge[type] (s1l);
 
-(T : Type)      (T : Var("T"))
-	[s1]<-----------[s2]
-\end{lstlisting}
+	\node[scope, right of = s1, xshift=5em] (s2) {$s2$};
+	\node[type, above of = s2] (s2l) {$\texttt{T : Var("T")}$};
+	\draw (s2) edge[type] (s2l);
 
-Finally, we need to find the type of \verb|T|.
+	\draw (s2) edge[] (s1);
+\end{tikzpicture}
+
+Finally, we need to find the type of \verb|Var("T")|.
 This finds the lexically closest definition of \verb|T| (the one in s2), which is correct.

### 2022-11-09 paper-thesis/chapters/20-comparing-with-haskell.tex
Work on the haskell-coc implementation

\chapter{A comparison with conventional implementations}
\label{ch:comp-haskell}

In this chapter, we implement the language defined in chapter \ref{chap:baselang} in Haskell, and then compare the implementation with the implementation in Statix. We want the design of the implementation in Haskell to be similar to conventional implementations of dependently typed languages, so we can compare the Statix implementation with them.

\section{Defining the AST}
To implement the calculus of constructions in Haskell, we first define the \verb|Expr| datatype. We chose to define the language using De Bruijn indices, since this is the convention when implementing dependently typed languages, and the goal of this implementation in Haskell is to compare the Statix implementation with conventional ones. 
\begin{lstlisting}
data Expr =
	Type
	| Let Expr Expr
	| Var Int
	| FnType Expr Expr
	| FnConstruct Expr Expr
	| FnDestruct Expr Expr
\end{lstlisting}

\section{Defining environments}

Next, we need to decide how we want to store the environment. We still need to store both function arguments and substitutions, which we called \verb|NType| and \verb|NSubst| in chapter \ref{chap:baselang}. We will use the same names, and store the environment as a list, with the head of the list representing De Bruijn index 0. Finally, we define the type of a scoped expression \verb|SExpr|, as a tuple of an environment and an expression.
\begin{lstlisting}
data EnvEntry = NType Expr | NSubst SExpr
type Env = [EnvEntry]
type SExpr = (Env, Expr)
\end{lstlisting}

The way these environments are defined is isomorphic to the way we defined the way we use scope graphs in section \ref{sec:coc-scopes}. Nodes in the scope graph have at most one parent, and each node stores one entry, which is exactly the structure of a list. For example, the following scope graph and list have the same meaning:

\begin{lstlisting}
[NType(Var(0)), NType(Type())]
\end{lstlisting}
\begin{tikzpicture}[scopegraph, node distance = 2.5em and 2.5em]
	\node[scope] (s1) {$s1$};
	\node[type, above of = s1] (s1l) {$\texttt{T : Type()}$};
	\draw (s1) edge[type] (s1l);

	\node[scope, right of = s1, xshift=5em] (s2) {$s2$};
	\node[type, above of = s2] (s2l) {$\texttt{x : Var("T")}$};
	\draw (s2) edge[type] (s2l);

	\draw (s2) edge[] (s1);
\end{tikzpicture}

\section{Defining beta reduction}

### 2022-11-09 haskell-coc/app/Main.hs

@@ -1,96 +1,5 @@
 module Main (main) where
 
-import Data.Maybe (fromMaybe)
-
-
-data Expr =
-  Type
-  | Let Expr Expr
-  | Var Int
-  | FnType Expr Expr
-  | FnConstruct Expr Expr
-  | FnDestruct Expr Expr
-  deriving Show
-
-type SExpr = (Env, Expr)
-
-type Env = [EnvEntry]
-
-data EnvEntry = NType Expr | NSubst SExpr
-  deriving Show
-
-sPutSubst :: Env -> SExpr -> Env
-sPutSubst env v = NSubst v : env
-
-sPutType :: Env -> Expr -> Env
-sPutType env v = NType v : env
-
-brh :: SExpr -> SExpr
-brh e = fromMaybe e (brh_ e [])
-
-brh_ :: SExpr -> [SExpr] -> Maybe SExpr
-brh_ (s, Type) [] = Just (s, Type)
-brh_ (s, Let v b) tt = brh_ (sPutSubst s (s, v), b) tt
-brh_ (s, Var i) tt = case s !! i of
-  NType _ -> if null tt then Just (s, Var i) else Nothing
-  NSubst se -> brh_ se tt
-brh_ (s, FnType a b) [] = Just (s, FnType a b)
-brh_ (s, FnConstruct a b) [] = Just (s, FnConstruct a b)
-brh_ (s, FnConstruct _ b) (t : tt) = brh_ (sPutSubst s t, b) tt
-brh_ (s, FnDestruct f a) tt = brh_ (s, f) ((s, a) : tt)
-
-br :: SExpr -> Expr
-br se = case brh se of
-  (_, Type) -> Type
-  (_, Var i) -> Var i
-  (s, FnType a b) -> FnType (br (s, a)) (br (s, b))
-  (s, FnConstruct a b) -> FnConstruct (br (s, a)) (br (s, b))
-  (s, FnDestruct a b) -> FnDestruct (br (s, a)) (br (s, b))
-
-beq :: SExpr -> SExpr -> Either String ()
-beq e1 e2 = case (brh e1, brh e2) of
-  ((_, Type), (_, Type)) -> pure ()
-  ((_, Var i), (_, Var j)) -> if i == j then pure () else Left $ "Expected variable " ++ show j ++ ", but got variable " ++ show i
-  ((s1, FnType a1 b1), (s2, FnType a2 b2)) -> do
[…]

### 2022-11-09 haskell-coc/out/production/haskell-coc/Lib.hs

@@ -1,0 +1,104 @@
+module Lib (Expr (..), Env, SExpr, tc) where
+
+import Data.Maybe (fromMaybe)
+
+
+data Expr =
+  Type
+  | Let Expr Expr
+  | Var Int
+  | FnType Expr Expr
+  | FnConstruct Expr Expr
+  | FnDestruct Expr Expr
+  deriving (Show, Eq)
+
+type SExpr = (Env, Expr)
+
+type Env = [EnvEntry]
+
+data EnvEntry = NType Expr | NSubst SExpr
+  deriving Show
+
+sPutSubst :: Env -> SExpr -> Env
+sPutSubst env v = NSubst v : env
+
+sPutType :: Env -> Expr -> Env
+sPutType env v = NType v : env
+
+brh :: SExpr -> SExpr
+brh e = fromMaybe e (brh_ e [])
+
+brh_ :: SExpr -> [SExpr] -> Maybe SExpr
+brh_ (s, Type) [] = Just (s, Type)
+brh_ (s, Let v b) tt = brh_ (sPutSubst s (s, v), b) tt
+brh_ (s, Var i) tt = case s !! i of
+  NType _ -> if null tt then Just (s, Var i) else Nothing
+  NSubst se -> brh_ se tt
+brh_ (s, FnType a b) [] = Just (s, FnType a b)
+brh_ (s, FnConstruct a b) [] = Just (s, FnConstruct a b)
+brh_ (s, FnConstruct _ b) (t : tt) = brh_ (sPutSubst s t, b) tt
+brh_ (s, FnDestruct f a) tt = brh_ (s, f) ((s, a) : tt)
+
+br :: SExpr -> Expr
+br se = case brh se of
+  (_, Type) -> Type
+  (_, Var i) -> Var i
+  (s, FnType a b) -> FnType (br (s, a)) (br (sPutType s a, b))
+  (s, FnConstruct a b) -> FnConstruct (br (s, a)) (br (sPutType s a, b))
+  (s, FnDestruct a b) -> FnDestruct (br (s, a)) (br (s, b))
+
+beq :: SExpr -> SExpr -> Either String ()
+beq e1 e2 = case (brh e1, brh e2) of
+  ((_, Type), (_, Type)) -> pure ()
+  ((_, Var i), (_, Var j)) -> if i == j then pure () else Left $ "Expected variable " ++ show j ++ ", but got variable " ++ show i
[…]

### 2022-11-09 haskell-coc/out/test/haskell-coc/Spec.hs

@@ -1,2 +1,22 @@
+-- file Spec.hs
+import Test.Hspec
+import Lib
+
 main :: IO ()
-main = putStrLn "Test suite not yet implemented"
+main = hspec $ do
+  describe "Type checking" $ do
+    it "1" $ Type `hasType` Type
+    it "2" $ FnConstruct Type Type `hasType` FnType Type Type
+    it "3" $ FnType Type Type `hasType` Type
+    it "4" $ FnConstruct Type (FnConstruct (Var 0) (Var 1)) `hasType` FnType Type (FnType (Var 0) Type)
+    it "5" $ Let (FnConstruct Type Type) (Var 0) `hasType` FnType Type Type
+    it "6" $ Let Type (Let (FnConstruct (Var 0) (Var 1)) (Var 0)) `hasType` FnType Type Type
+    it "7" $ FnConstruct Type (Let (FnConstruct (Var 0) (Var 0)) (Var 0)) `hasType` FnType Type (FnType (Var 0) (Var 1))
+    it "8" $ FnConstruct Type (FnConstruct (Var 0) (Var 0)) `hasType` FnType Type (FnType (Var 0) (Var 1))
+    it "9" $ FnConstruct Type (FnDestruct (FnConstruct Type (Var 0)) (Var 0)) `hasType` FnType Type Type
+    it "10" $ FnConstruct Type (FnDestruct (FnConstruct Type (Var 1)) (Var 0)) `hasType` FnType Type Type
+    it "11" $ FnConstruct Type (FnConstruct (Var 0) (Var 0)) `hasType` FnType Type (FnType (Var 0) (Var 1))
+    it "12" $ FnDestruct (FnConstruct Type (FnConstruct (Var 0) (Var 0))) Type `hasType` FnType Type Type
+
+hasType :: Expr -> Expr -> Expectation
+hasType e t = tc ([], e) `shouldBe` Right t

### 2022-11-10 paper-thesis/chapters/20-adding-booleans.tex
Reindex chapters + writing

--- 2022-11-07 paper-thesis/chapters/04-adding-booleans.tex
@@
 One for if \verb|c| evaluates to true, one for if \verb|c| evaluates to false, and finally a third case for other cases (such as when \verb|c| is a variable that does not have a substitution).
-Remember that \verb|rebuild| is the rule introduced in chapter \ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
-These rules are stated below:
+These rules are stated below: (Remember that \verb|rebuild| is the rule introduced in chapter \ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s)
 
@@
 
+Next, the rules for type-checking booleans are in figure \ref{fig:bool-rules-typecheck}, which are relatively simple.
+The if expression checks that both branches have the same type, as it is a non-dependent if statement.
+
+\begin{figure}[ht]
+\begin{mathpar}
+	\inferrule{
+	} {
+		\toe{s}{BoolType()}{Type()}
+	}
+
+	\inferrule{
+	} {
+		\toe{s}{BoolTrue()}{BoolType()}
+	}
+
+	\inferrule{
+	} {
+		\toe{s}{BoolFalse()}{BoolType()}
+	}
+
+	\inferrule{
+		\toe{s}{c}{ct}
+		\\\beq{ct}{BoolType()}
+		\\\\
+		\\\toe{s}{b1}{tb1}
+		\\\toe{s}{b2}{tb2}
+		\\\beq{tb1}{tb2}
+	} {
+		\toe{s}{BoolIf(c, b1, b2)}{tb1}
+	}
+\end{mathpar}
+\caption{Rules for type checking booleans}
+\label{fig:bool-rules-typecheck}
+\end{figure}
+
 \section{Postulate}

### 2022-11-10 paper-thesis/chapters/21-adding-inference.tex
Reindex chapters + writing

--- 2022-11-07 paper-thesis/chapters/05-adding-inference.tex
@@
-\chapter{Inference}
+\chapter{Term Inference}
 
-\section{Background: What is inference?}
+Inference is an important feature of dependent programming languages, that allows redundant parts of programs to be left out.
+For example, it allows you to infer the arguments of a function, if they can be inferred by the arguments that follow, like here where the type of the argument can be inferred, since we pass it \verb|true| which is a boolean:
+\begin{lstlisting}
+let id = (\T : Type. \x: T. x);
+id _ true
+\end{lstlisting}
 
-Inference
+We call it \emph{term inference} rather than \emph{type inference}, because we can infer values other than types. For example, it can infer that the \verb|_| in this example must be \verb|true|:
+\begin{lstlisting}
+postulate f: Bool -> Type;
+postulate g: f true -> Type;
+\x: f _. g x
+\end{lstlisting}
 
-\section{General inference in Statix}
-\label{general-inference}
+\section{Different sorts of inference}
+\label{strength-inference}
 
-\section{Greedy inference}
+The strength of an inference algorithm is how good it is at inferring. Some algorithms are better than others. \cite{typeinference}
 
-Instead, we implement a strategy that I would like to call greedy inference (TODO: does this have a name?).
-
-\section{In statix}
+\section{Using Statix's first-order unification for inference}
 
@@
 This will be used to store metavariables which we want to infer.
-This is a hack we use, ideally we would like to query statix to see if a metavariable is initialized, but as discussed in section \ref{general-inference} this is not possible.
+This is a hack we use, ideally we would like to query statix to see if a metavariable is initialized, but as discussed in section \ref{strength-inference} this is not possible.
 
@@
 \end{lstlisting}
-
-\section{Analysis: How well does it work?}

### 2022-11-10 paper-thesis/chapters/22-adding-datatypes.tex
Reindex chapters + writing

\chapter{Inductive Data Types}

Another useful feature is support for inductive data types. An example of a simple inductive datatype is the \verb|Nat| type:
\begin{lstlisting}
data Nat : -> Type where
    zero : Nat,
    suc : Nat -> Nat;
\end{lstlisting}
We will design data types to have the same features as in Agda:
\begin{enumerate}
\item Parameters, which are datatypes that are polymorphic, and are required to be the same for all constructors, such as
\begin{lstlisting}
data Maybe (T : Type) : -> Type where
    None : Maybe T,
    Some : T -> Maybe T;
\end{lstlisting}
\item Indices, which are datatypes that are polymorphic, that may vary from constructor to constructor
\begin{lstlisting}
data Eq : (e1: Bool) (e2 : Bool) -> Type where
    refl : e : Bool -> Eq e e;
\end{lstlisting}
\item We can also combine parameters and indices, for example to create the generic \verb|Eq| type. Note that parameters and indices may also depend on earlier parameters and indices.
\begin{lstlisting}
data Eq (T : Type) : (e1 : T) (e2 : T) -> Type where
    refl : e : T -> Eq e e;
\end{lstlisting}
\item Data types must be strictly positive. This ensures that we cannot make non-terminating programs\footnote{After we add support for universes in chapter \ref{ch:universes}}. For example, the following data type is forbidden, because it refers to itself in a negative position. We will explain exactly how this check works in section \ref{ch:datatypes:positivity}.
\begin{lstlisting}
data Bad : -> Type where
	bad : (Bad -> Bool) -> Bad;
\end{lstlisting}

\section{In statix}

\section{Positivity Checking}
\label{ch:datatypes:positivity}

\end{enumerate}

### 2022-11-10 paper-thesis/chapters/21-adding-inference.tex
A section about inference

@@
 
-\section{Different sorts of inference}
+\section{Different algorithms for inference}
 \label{strength-inference}
 
-The strength of an inference algorithm is how good it is at inferring. Some algorithms are better than others. \cite{typeinference}
+There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inferences than others.
+However, we would like to avoid implementing an algorithm at all, instead using Statix's built-in first-order unification to do the type inference for us.
+Implementing an inference algorithm in Statix is theoretically possible but this would be a lot of ugly code, and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.
+
+The inference that we will build is a very simple unification system, where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the free variable to be equal to the other variable.
+This process is called \emph{first-order unification}.
+
+There are some situations in which this approach fails, but in most real-world scenarios it works perfectly.
+For example, it can infer both programs in the introduction of this chapter, but it fails to infer the following program:
+\begin{lstlisting}
+let f = _;
+\x : Type.
+\g: (_: (f x) -> Bool).
+g true
+\end{lstlisting}
+We know that \verb|f| is a function from \verb|Type -> Type|, but it fails to infer the value of \verb|f|.
+Because of the way that \verb|g| is used, the type checker asserts that $\beq{f x}{Bool}$.
+Since x is declared as a function argument and it is completely free, this means that for any \verb|x|, \verb|f x = Bool|.
+But the rule above is not powerful enough to derive all of this, so it fails.
 
 \section{Using Statix's first-order unification for inference}
+\label{statix-inference}
 

### 2022-11-10 paper-thesis/chapters/01-introduction.tex
A bit of more writing

@@
 
+\section{Contributions}
 This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language.
@@
 
+The technical contributions of this thesis are the following:
+\begin{itemize}
+	\item 
+\end{itemize}
+
+\section{Outline}
+
 TODO navigation

### 2022-11-10 paper-thesis/chapters/30-comparing-with-haskell.tex
A bit of more writing

--- 2022-11-09 paper-thesis/chapters/20-comparing-with-haskell.tex
@@
 \begin{lstlisting}
-[NType(Var(0)), NType(Type())]
+NType(Var(0)) :: NType(Type()) :: Nil
 \end{lstlisting}
@@
 
+The nodes in scope graphs may have multiple children, but we never query the children of a node.
+We only follow the edges, we don't go in the opposite direction.
+Similarly, part of a list may be shared, but this is fine in Haskell, since values are immutable.
+
 \section{Defining beta reduction}
+
+Now we want to define beta head reduction. Remember that in Statix, beta head reduction is a relation with the signature
+\begin{lstlisting}
+betaReduceHead : 
+	(scope * Expr) * list((scope * Expr)) -> (scope * Expr)
+\end{lstlisting}

### 2022-11-14 paper-thesis/chapters/000-abstract.tex
Work a little bit on thesis + fse

--- 2022-11-07 paper-thesis/chapters/00-abstract.tex
@@
-\begin{abstract}
 Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone.
@@
 While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.
-\end{abstract}

### 2022-11-14 paper-thesis/chapters/31-comparing-with-lambdapi.tex
Work a little bit on thesis + fse

\chapter{A comparison with implementations in logical frameworks}
\label{ch:comp-lambdapi}

### 2022-11-14 haskell-coc/out/production/haskell-coc/Lib.hs

@@ -1,4 +1,4 @@
-module Lib (Expr (..), Env, SExpr, tc) where
+module Lib (Expr (..), Env, EnvEntry (..), SExpr, tc, brh) where
 
 import Data.Maybe (fromMaybe)
 
@@ -17,7 +17,7 @@ type SExpr = (Env, Expr)
 type Env = [EnvEntry]
 
 data EnvEntry = NType Expr | NSubst SExpr
-  deriving Show
+  deriving (Show, Eq)
 
 sPutSubst :: Env -> SExpr -> Env
 sPutSubst env v = NSubst v : env

### 2022-11-14 haskell-coc/out/test/haskell-coc/Spec.hs

@@ -17,6 +17,7 @@ main = hspec $ do
     it "10" $ FnConstruct Type (FnDestruct (FnConstruct Type (Var 1)) (Var 0)) `hasType` FnType Type Type
     it "11" $ FnConstruct Type (FnConstruct (Var 0) (Var 0)) `hasType` FnType Type (FnType (Var 0) (Var 1))
     it "12" $ FnDestruct (FnConstruct Type (FnConstruct (Var 0) (Var 0))) Type `hasType` FnType Type Type
+    it "13" $ brh ([NType Type], Let (Var 0) (Var 0)) `shouldBe` ([NType Type], Var 0)
 
 hasType :: Expr -> Expr -> Expectation
 hasType e t = tc ([], e) `shouldBe` Right t

### 2022-11-15 paper-thesis/chapters/30-comparing-with-haskell.tex
Some writing

@@
 
+Finally, we define the two functions \verb|sPutSubst| and \verb|sPutType| to mimic the Statix relations with the same name.
+\begin{lstlisting}
+sPutSubst :: Env -> SExpr -> Env
+sPutSubst env v = NSubst v : env
+sPutType :: Env -> Expr -> Env
+sPutType env v = NType v : env
+\end{lstlisting}
+
 \section{Defining beta reduction}
@@
 \end{lstlisting}
+
+In Haskell, a slightly different structure was used.
+We defined two functions, one returning a \verb|Maybe SExpr| and the other checking if the result \verb|Nothing|, in which case it returns the original expression (unreduced).
+This structure could also be implemented in Statix, but Haskell makes it a bit easier for us by providing the \verb|Maybe| type and functions on it.
+\begin{lstlisting}
+brh :: SExpr -> SExpr
+brh e = fromMaybe e (brh_ e [])
+brh_ :: SExpr -> [SExpr] -> Maybe SExpr
+\end{lstlisting}

### 2022-11-21 haskell-coc/src/Lib.hs
Fix haskell-coc shift function

@@ -68,11 +68,11 @@ tc (s, Var i) = do
   vt <- case s !! i of
     NType t -> pure t
     NSubst es -> tc es
-  pure (shift (i + 1) vt)
+  pure (shift (i + 1) 0 vt)
 tc (s, Let v b) = do
   _ <- tc (s, v)
   bt <- tc (sPutSubst s (s, v), b)
-  pure (shift (-1) bt)
+  pure (shift (-1) 0 bt)
 tc (s, FnType a b) = do
   at <- tc (s, a)
   beq ([], at) ([], Type)
@@ -95,10 +95,12 @@ tc (s, FnDestruct f a) = do
   beq ([], at) (sf, dt)
   pure $ br (sPutSubst sf (s, a), db)
 
-shift :: Int -> Expr -> Expr
-shift _ Type = Type
-shift k (Var i) = Var (i + k)
-shift k (Let v b) = Let (shift k v) (shift k b)
-shift k (FnType a b) = FnType (shift k a) (shift k b)
-shift k (FnConstruct a b) = FnConstruct (shift k a) (shift k b)
-shift k (FnDestruct a b) = FnDestruct (shift k a) (shift k b)
+-- shift indices with `m` if >= than cutoff `c` in the term
+shift :: Int -> Int -> Expr -> Expr
+shift _ _ Type = Type
+shift k c (Var i) | i >= c = Var (i + k)
+shift _ _ (Var i) = Var i
+shift k c (Let v b) = Let (shift k c v) (shift k (c + 1) b)
+shift k c (FnType a b) = FnType (shift k c a) (shift k (c + 1) b)
+shift k c (FnConstruct a b) = FnConstruct (shift k c a) (shift k (c + 1) b)
+shift k c (FnDestruct a b) = FnDestruct (shift k c a) (shift k c b)

### 2022-11-22 paper-thesis/chapters/20-adding-booleans.tex
A section about inference

@@
-\chapter{\label{chap:bools}Booleans and other "simple" constructs that really need a better title}
+\chapter{\label{chap:bools}Extending the language}
 
@@
 		{ [] }
-		{ \scope{s'}{BoolTrue()} }
+		{ \scope{s'}{\co{BoolTrue}()} }
 		\\\bhr
@@
 		{ [] }
-		{ \scope{s'}{BoolFalse()} }
+		{ \scope{s'}{\co{BoolFalse}()} }
 		\\\bhr
@@
 	} {
-		\toe{s}{BoolType()}{Type()}
+		\toe{s}{\co{BoolType}()}{\co{Type}()}
 	}
@@
 	} {
-		\toe{s}{BoolTrue()}{BoolType()}
+		\toe{s}{\co{BoolTrue}()}{\co{BoolType}()}
 	}
@@
 	} {
-		\toe{s}{BoolFalse()}{BoolType()}
+		\toe{s}{\co{BoolFalse}()}{\co{BoolType}()}
 	}
@@
 		\toe{s}{c}{ct}
-		\\\beq{ct}{BoolType()}
+		\\\beq{ct}{\co{BoolType}()}
 		\\\\
@@
 	} {
-		\toe{s}{BoolIf(c, b1, b2)}{tb1}
+		\toe{s}{\co{BoolIf}(c, b1, b2)}{tb1}
 	}
@@
 Next we will add \verb|Postulate| to the language.
+A postulate declares that there is a variable with a certain type, without specifying a value.
+Through the view of the Curry-Howard correspondence, this is equivalent to an axiom.
+At the current stage, this is useful for testing the language.
+For example, this is a test with a postulate:
+
+\begin{lstlisting}
+postulate T : Type;
+if true then Bool else T end
+\end{lstlisting}
+The following rules are used to implement the feature, which translate cleanly to Statix:
+
+\begin{figure}[ht]
+	\begin{mathpar}
+		\inferrule{
+			\bhr
+			{\scope{s}{b}}
+			{a}
+			{\scope{s'}{b'}}
+		}{
+			\bhr
+			{\scope{s}{\co{Postulate}(n, t, b)}}
+			{a}
+			{\scope{s'}{b'}}
+		}
+
+		\inferrule{
+			\toe{s}{t}{tt}
+			\\\beq{tt}{\co{Type}()}
+			\\\\\bred{s}{t}{t'}
+			\\\toe{\co{sPutType}(s, n, t')}{b}{bt}
+		} {
+			\toe{s}{\co{Postulate}(n, t, b)}{bt}
+		}
+	\end{mathpar}
+	\caption{Rules for postulates}
+	\label{fig:postulate-rules}
+\end{figure}
 
@@
 Finally, we will add \verb|TypeAssert| to the language.
+
+\section{Extensibility of the approach}

### 2022-11-22 paper-thesis/chapters/20-adding-booleans.tex
Finish one chapter

@@
 For example, this is a test with a postulate:
-
 \begin{lstlisting}
@@
 Finally, we will add \verb|TypeAssert| to the language.
+This is another feature that is useful for testing.
+It takes an expression, and it asserts that the expression has a certain type.
+For example, we can do the following:
+\begin{lstlisting}
+postulate T : Type;
+true : if true then Bool else T end
+\end{lstlisting}
+
+Implementing this feature is also straight-forward, we add the following two rules, which translate cleanly to Statix:
+
+\begin{figure}[ht]
+	\begin{mathpar}
+		\inferrule{
+			\bhr
+			{\scope{s}{b}}
+			{a}
+			{\scope{s'}{b'}}
+		}{
+			\bhr
+			{\scope{s}{\co{TypeAssert}(b, t)}}
+			{a}
+			{\scope{s'}{b'}}
+		}
+
+		\inferrule{
+			\toe{s}{t}{tt}
+			\\\beq{tt}{\co{Type}()}
+			\\\\\toe{s}{b}{bt}
+			\\\beq
+			{bt}
+			{\scope{s}{t}}
+		} {
+			\toe{s}{\co{TypeAssert}(b, t)}{bt}
+		}
+	\end{mathpar}
+	\caption{Rules for type assertions}
+	\label{fig:typeassert-rules}
+\end{figure}
 
 \section{Extensibility of the approach}
+
+Now that a few features have been implemented, we can discuss how easy the language is to extend.
+The conclusion so far is that it is doable in a clean way.
+Specifically, the approach for each of the features above was:
+
+\begin{enumerate}
+	\item Define the parsing rules for the new feature
+	\item Create a new file, \verb|tp_bools.stx| and import this file in the main \verb|type_check.stx| file.
+	\item In this file, define a case for the \verb|betaReduceHead| rule for the constructors that were added. Also define cases for \verb|expectBetaEq| and \verb|betaReduce| if necessary.
+	\item In this file, define a case for the \verb|typeOfExpr| rule for the constructors that were added.
+\end{enumerate}
+
+This allows each feature to be isolated to its own file.
+If we decide that we don't like a feature after all, we can remove it simply by unimporting the file.
+This problem is similar to the expression problem\cite{expression_problem}, which Statix solves very well.
+
+In the following chapters, we're going to be extending the language further with these more interesting features using this approach:
+\begin{itemize}
+	\item Inference (chapter \ref{ch:inference})
+	\item Inductive Datatypes (chapter \ref{ch:datatypes})
+	\item Universes (chapter \ref{ch:universes})
+\end{itemize}

### 2022-11-22 paper-thesis/chapters/21-adding-inference.tex
Finish one chapter
Corrections

@@
 \chapter{Term Inference}
+\label{ch:inference}
 
@@
 There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inferences than others.
-However, we would like to avoid implementing an algorithm at all, instead using Statix's built-in first-order unification to do the type inference for us.
-Implementing an inference algorithm in Statix is theoretically possible but this would be a lot of ugly code, and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.
+However, we would like to avoid implementing an algorithm at all, instead using Statix' built-in first-order unification to do the type inference for us.
+Implementing an inference algorithm in Statix is theoretically possible but this would be a lot of ugly code (since Statix is not a general-purpose programming language), and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.
 
@@
 For example, it can infer both programs in the introduction of this chapter, but it fails to infer the following program:
+
 \begin{lstlisting}
@@
 
-\section{Using Statix's first-order unification for inference}
+\section{Using Statix' first-order unification for inference}
 \label{statix-inference}
@@
 This will be used to store metavariables which we want to infer.
-This is a hack we use, ideally we would like to query statix to see if a metavariable is initialized, but as discussed in section \ref{strength-inference} this is not possible.
+This is a hack we use, ideally we would like to query Statix to see if a metavariable is initialized, but as discussed in section \ref{strength-inference} this is not possible.
 

### 2022-11-29 paper-thesis/chapters/02-background-spoofax.tex
Work on agda.lp

% !TEX root = document.tex
\chapter{\label{chap:bg-spoofax}Background: Spoofax}

### 2022-12-07 paper-thesis/chapters/02-background-spoofax.tex
Background section for spoofax

@@
 \chapter{\label{chap:bg-spoofax}Background: Spoofax}
+
+The Spoofax Language Workbench \cite{spoofax} is a platform to develop domain-specific and
+general-purpose programming languages.
+Each aspect of the language is defined in one of three metalanguages, each with their own purpose.
+From these languages Spoofax generates a parser, type checker and other useful tools.
+
+The metalanguage SDF3 is used to define the syntax of the language \cite{sdf3}, which is then parsed using Spoofax's SGLR parser \cite{sdf3_parser}.
+The parser produces a abstract syntax tree in the ATerm format \cite{aterm}.
+A pretty printer is also generated based on this syntax definition.
+
+Next, the metalanguage Statix \cite{scopes_as_types} is used to define the static semantics of the language.
+It uses scope graphs to represent binding and typing information of the language.
+How this works is discussed in section \ref{sec:statix}.
+
+Finally, Stratego \cite{stratego} is used to define program transformations, using rewrite rules.
+This can be used for modifications to the AST, but also to build transformations to a different language (such as compiling to assembly).
+
+\section{\label{sec:statix} Statix and Scope graphs}
+
+In Statix, scope graphs are used to represent a programs name and type information \cite{scopes_as_types}. A scope graph is a graph that consists of the following components:
+
+\begin{itemize}
+	\item \textit{Scopes} represent "a region in a program that behaves uniformly with respect to name resolution".
+These scopes are modeled as nodes in the graph.
+Its graph representation is shown in \autoref{fig:graph-notation-scope}.
+	\item \textit{Labeled, directed edges} model visibility relations between scopes.
+For example, $\scopet{1} \scopeedget[\lblt{P}] \scopet{2}$ indicates that the graph contains an edge from \scopet{1} to \scopet{2} with label \lblt{P}.
+Its graph representation can be seen in \autoref{fig:graph-notation-edge}.
+	\item \textit{Declarations} model a datum (a piece of information, like the type of a variable) under a relation symbol in a scope.
+Its textual notation is $\scopet{1} \typeedget[rel] d$, meaning datum $d$ is declared in scope $\scopet{1}$ under relation \textit{rel}.
+Its pictorial equivalent is shown in \autoref{fig:graph-notation-data}.
+\end{itemize}
+
+\begin{figure}[!h]
+	\begin{framed}
+		\begin{subfigure}{0.2\linewidth}
+			\centering
+			\begin{tikzpicture}[scopegraph, node distance = 5em and 3.0em]
+				\node[scope] (s1) {$1$};
+			\end{tikzpicture}
+			\caption{Scope}
+			\label{fig:graph-notation-scope}
+		\end{subfigure}
+		\begin{subfigure}{0.4\linewidth}
+			\centering
+			\begin{tikzpicture}[scopegraph, node distance = 5em and 3.0em]
+				\node[scope] (s1) {$1$};
+				\node[scope, right of = s1] (s2) {$2$};
+				\draw (s1) edge[lbl={P}] (s2);
+			\end{tikzpicture}
+			\caption{Edge}
+			\label{fig:graph-notation-edge}
+		\end{subfigure}
+		\begin{subfigure}{0.4\linewidth}
+			\centering
+			\begin{tikzpicture}[scopegraph, node distance = 5em and 3.0em]
+				\node[scope] (s1) {$1$};
+				\node[type, right of = s1] (s2) {$\texttt{x : INT()}$};
+				\draw (s1) edge[type, lbl={rel}] (s2);
+			\end{tikzpicture}
+			\caption{Data}
+			\label{fig:graph-notation-data}
+		\end{subfigure}
+	\end{framed}
+	\caption{Scope Graph notation}
+\end{figure}
+
+Information can be retrieved from scope graphs using \textit{queries}.
+A query from a scope traverses the scope graph to find datums that match certain conditions.
+The result of a query is a list of (path, datum) tuples.
+A query has several parameters:
+\begin{itemize}
+	\item The relation to query. Only datums declared under this relation will be returned by the query.
+	\item Path well-formedness condition: a regular expression of labels that specifies which paths are well-formed.
+Only datums that are reached through a path whose edge labels are in the language described by the regular expression are included in the query result.
+	\item A match predicate, taking a single datum as input.
+Only datums that satisfy this predicate will be included in the query result.
+Its default value is \verb|true|, meaning that all datums under the specified relation that satisfy the path well-formedness condition are returned.
+	\item Result comparison parameters:
+	\begin{itemize}
+		\item A strict partial order on paths, described by less-than relations on labels. This relation defines a prefix-order on paths.
+		\item A `shadow' predicate, which takes two datums as input. Its default value is \verb|false|, meaning that no shadowing is performed.
+	\end{itemize}
+	When the result set contains multiple datums, and it holds for a datum $d$ that there exists another datum $d'$ which path is strictly smaller that $d$, and the shadow predicate holds, then $d$ is removed from the result set.
+\end{itemize}
+
+% CREDIT FOR THIS EXPLANATION TO ARON ZWAAN
+% HOW TO CREDIT?

### 2022-12-07 paper-thesis/chapters/03-background-dependent-types.tex
Work on section 3

% !TEX root = document.tex
\chapter{\label{chap:bg-dp}Background: Dependent Types}

This chapter will explain the concepts behind dependent types. If you are already familiar with Dependent Types, you can skip this chapter.

\section{What are dependent types?}

A dependent type system is a type system in which types may depend on values. This means it is possible to have a type such as \verb|Vec n|, which is a vector of exactly length \verb|n|. Note that the value of \verb|n| does not have to be known at compile time, it is more powerful than that.

The biggest advantage of dependent types is that they increase the expressiveness of a type system a lot, for example it is now possible to express a function that adds two vectors of the same length and returns a new vector of that length. 
\begin{lstlisting}
add_vec : (n : Nat) -> Vec n -> Vec n -> Vec n
\end{lstlisting}

\section{The Curry-Howard Isomorphism}

The Curry-Howard correspondence is a relationship that can be used to interpret typed programs as mathematical proofs \cite{chc}. This is done by representing false statements as empty types, and true statements as non-empty (inhabited) types.

For example, it is possible to define a type \verb|IsEven n|, that represent the statement "\verb|n| is even", and declare it in such a way that it is only inhabited if \verb|n| is even.

In order to use a type system to prove things it is important that the type system is sound, that is, it is impossible to prove a false statement. Using the Curry-Howard correspondence, that means it should be impossible to obtain an element of the empty type. We will explore this further in chapter \ref{ch:universes}.

\section{The Lambda Cube}

The lambda cube is a framework to categorize programming languages based on whether terms and types can depend on each other \cite{lambda_cube}. Specifically, it categorizes languages on three axis: \footnote{Note that the "Terms can depend on terms" axis is missing, this is because a language where terms cannot depend on terms cannot compute anything and is thus pretty useless.}

\begin{enumerate}
	\item Terms can depend on types. This corresponds to type-polymorphic (generic) functions. For example, it allows us to define the polymorphic identity function:
	\begin{lstlisting}
id : (T : Type) -> T -> T
	\end{lstlisting}
	\item Types can depend on types. This corresponds to type-polymorphic (generic) datatypes. For example, it allows to the define the \verb|List T| type, a list with items of type T.

	\item Types can depend on terms. This corresponds to dependent types.
\end{enumerate}

The \textit{Calculus of Constructions} is the language that has all three of these features, and it is what we will implement in this paper. If you want to learn more, I recommend playing around with the \textit{\href{https://agda.readthedocs.io/en/latest/}{Agda}} language, which is a real-world language that satisfies all three axis.

### 2022-12-23 paper-evcs/paper.tex
Play around with dedukti and rust
Work on the haskell-coc implementation
Process feedback

@@
 To type check the base language, we need to store information about the names that are in scope at each point in the program.
-There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}
+There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need the value of a binding to compare types, this is needed in dependently typed languages.}
 
@@
 sGetName  : scope * ID -> NameEntry
-sGetNames : scope * ID -> list((path * (ID * NameEntry)))
 sEmpty    : -> scope
 \end{lstlisting}
-The \verb|sPutType| and \verb|sPutSubsts| predicates generate a new scope given a parent scope and a type or a substitution respectively.
-To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of NameEntries respectively that the query found.
+The \verb|sPutType| and \verb|sPutSubst| predicates generate a new scope given a parent scope and a type or a substitution respectively.
+To query the scope graph use \verb|sGetName|, which will return a \verb|NameEntry| that the query found.
 Finally, \verb|sEmpty| returns a fresh empty scope.
@@
 A unique requirement for dependently typed languages is beta reduction during type checking, since types may require evaluation to compare.
+Beta reduction is the process of reduction a term to beta normal form, which is the state where no further beta reductions are possible \cite{tapl}.
+It works by matching a term of the form \verb|(\x.
+b) e|.
+and substituting \verb|x| in \verb|b| with \verb|e|.
+Beta reduction applies this rule anywhere in the term, whereas beta head-reduction only applies this rule at the head (outermost expression) of the term, and produces a term in beta-head normal form.
 
@@
 It works by keeping a stack of all arguments that have not been applied yet.
-This turned out to be the more natural way of expressing this over substitution-based evaluation relation.
+This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation.
 We originally tried to implement the latter, which works fine for the base language.
-However, it runs into trouble when implementing inductive data types; more information about this will be in the full master thesis.
+However, it runs into trouble when implementing inductive data types; more information about this will be in the full master's thesis\cite{thesis}.
 An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.
@@
 In conventional dependently typed languages, evaluation is often done using De Bruijn indices.
-However, we chose to use names rather than De Bruijn indices, because scope graphs work based on names rather.
-Using De Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations, such as renaming.
+However, we chose to use names rather than De Bruijn indices, because scope graphs work based on names.
+Using De Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations (which are Spoofax annotations that declare one name as being a use of another name that is a definition), such as renaming.
 
@@
 		\inferrule{ 
-		} { relation
+		} { 
 			\bhr
@@
 			\bhr
-			{\scope{\co{sPutSubst}(s, n, a)}{b}}
+			{\scope{\co{sPutSubst}(s, n, t)}{b}}
 			{ts}
@@
 
-We will define a Statix predicate \verb|typeOfExpr| that takes a scope and an expression and type checks the scope in the expression. It returns the type of the expression.
+We will define a Statix predicate \verb|typeOfExpr| that takes a scope and an expression and type checks the expression in the scope. It returns the type of the expression.
 
@@
 
-The inference rules above can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
+The inference rules in figure \ref{fig:type-check-rules} can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
 \begin{lstlisting}
@@
 			\\ \beq{at}{\scope{sf}{dt}}
-			\\ \bred{s'}{db}{db'}
+			\\ \bred{\co{sPutSubst}(s', n, (s, arg))}{db}{db'}
 		} { 
@@
 The problem is that there are multiple names, and there is no way to distinguish between them.
-This can be solved by using scopes to distinguish names.
-A full explanation of this will be in the master thesis.
+This problem needs to be addressed in any name-based approach.
+It can be solved by using scopes to distinguish names.
+A full explanation of this will be in the master's thesis\cite{thesis}.
 
@@
 This section will discuss some further ideas that can be explored to build upon what has already been shown.
-These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.
+These will be fully described in the master's thesis\cite{thesis} that this paper is based on, this is just a look into what is possible.
 

### 2023-01-09 paper-thesis/chapters/11-variable-capture.tex
Work on writing

--- 2022-11-09 paper-thesis/chapters/03-variable-capture.tex
@@
 
-TODO Add reference to types and programming languages
+The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is variable capture.
+Variable capture is the phenomenon of free variables in a term becoming bound when a naive substitution happens \cite{tapl}.
+This section will explore several ways of solving this.
 
-The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is variable capture.
-This section will explore several ways of solving this.
-An example is the following: What is the type of this expression (a polymorphic identity function)?
+An example term where this problem occurs is the following: What is the type of this expression (a polymorphic identity function)?
 \begin{lstlisting}
@@
 \end{lstlisting}
-
 The algorithm so far would tell you it is \verb|T : Type -> T : T -> T|.

### 2023-01-09 paper-thesis/chapters/20-adding-booleans.tex
Work on writing

@@
 	\item Create a new file, \verb|tp_bools.stx| and import this file in the main \verb|type_check.stx| file.
-	\item In this file, define a case for the \verb|betaReduceHead| rule for the constructors that were added. Also define cases for \verb|expectBetaEq| and \verb|betaReduce| if necessary.
-	\item In this file, define a case for the \verb|typeOfExpr| rule for the constructors that were added.
+	\item In the new file, define a case for the \verb|betaReduceHead| rule for the constructors that were added.
+Also define cases for \verb|expectBetaEq| and \verb|betaReduce| if necessary.
+It is necessary iff the new constructors are not always eliminated by \verb|betaReduceHead|.
+	\item In the new file, define a case for the \verb|typeOfExpr| rule for the constructors that were added.
 \end{enumerate}

### 2023-01-09 paper-thesis/chapters/21-adding-inference.tex
Work on writing

@@
 There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inferences than others.
-However, we would like to avoid implementing an algorithm at all, instead using Statix' built-in first-order unification to do the type inference for us.
-Implementing an inference algorithm in Statix is theoretically possible but this would be a lot of ugly code (since Statix is not a general-purpose programming language), and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.
-
-The inference that we will build is a very simple unification system, where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the free variable to be equal to the other variable.
-This process is called \emph{first-order unification}.
-
+One algorithm for unification is \emph{first-order unification}, where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the the value free variable to be equal to the value of the non-free variable.
 There are some situations in which this approach fails, but in most real-world scenarios it works perfectly.
@@
 \end{lstlisting}
+
 We know that \verb|f| is a function from \verb|Type -> Type|, but it fails to infer the value of \verb|f|.
@@
 Since x is declared as a function argument and it is completely free, this means that for any \verb|x|, \verb|f x = Bool|.
-But the rule above is not powerful enough to derive all of this, so it fails.
+But the rule above is not powerful enough to derive this, so it fails.
+
+\section{Inference in Statix}
+\label{statix-inference}
+
+We would like to avoid implementing an algorithm at all, instead using Statix' built-in first-order unification to do the type inference for us.
+Implementing an inference algorithm in Statix is theoretically possible but this would be a lot of ugly code (since Statix is not a general-purpose programming language), and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.
+
+However, we cannot immediately use Statix' built-in first-order unification (which acts in the meta language, Statix) to implement first-order unification in the object language.
+Ideally when implementing beta equality we would match on $\beq{e_1}{e_2}$ where $e_1$ is a free variable, but Statix does not allow for querying whether variables are free.
+
+Instead, we will be implementing a novel, less powerful form of first-order unification.
+This will work by explicitly denoting which variables \emph{could be} free, and explicitly handling these cases in a way that approximates first-order unification.
+We will denote this algorithm as \emph{approximated first-order unification (AFOU)}.
+
+\section{Implementing AFOU}
+\label{implementing-inference}
+
+First, we introduce a new constructor \verb|Infer : Expr -> Expr|, which denotes the variables which could be free.
+The constructor is introduced when we encounter a \verb|_| variable, a marker that something needs to be inferred.
+
+\begin{lstlisting}
+typeOfExpr(s, Var(Syn("_"))) = (Infer(q), qt) :-
+	(_, qt) == typeOfExpr(sEmpty(), q).
+\end{lstlisting}
+
+Note that the type of \verb|typeOfExpr| has changed, it is now
+
+\begin{lstlisting}
+typeOfExpr : scope * Expr -> Expr * Expr
+\end{lstlisting}
+
+\section{Analysis of the power of AFOU}
+\label{analysing-inference}
+
+\begin{comment}
 
 \section{Using Statix' first-order unification for inference}
-\label{statix-inference}
 
-First, we introduce a new constructor, \verb|Infer : Expr -> Expr|.
-This will be used to store metavariables which we want to infer.
-This is a hack we use, ideally we would like to query Statix to see if a metavariable is initialized, but as discussed in section \ref{strength-inference} this is not possible.
-
-The constructor is introduced when we encounter a \verb|_| variable, a sign that something needs to be inferred.
-When type-checking this, we create an infer constructor.
-The constraint at the bottom cannot be solved immediately, since the \verb|q| variable is unbound, but if \verb|q| is bound later the constraint will be evaluated.
-
-\begin{lstlisting}
-typeOfExpr_(s, Var(Syn("_"))) = (Infer(q), qt) :-
-	(_, qt) == typeOfExpr(sEmpty(), q).
-\end{lstlisting}
 When an \verb|Infer| needs to be type-checked, the logic is very similar, except that we don't need to generate a new metavariable:
@@
 \end{lstlisting}
+\end{comment}

### 2023-01-10 paper-thesis/chapters/21-adding-inference.tex
Work on thesis + evcs paper

@@
 The constructor is introduced when we encounter a \verb|_| variable, a marker that something needs to be inferred.
-
 \begin{lstlisting}
@@
 
-Note that the type of \verb|typeOfExpr| has changed, it is now
-
+Note that the type of \verb|typeOfExpr| has changed, it now returns two expressions, the first being the same expression that was passed in except with \verb|_| variables replaced with \verb|Infer| constructors, and the second being the type.
 \begin{lstlisting}

### 2023-01-11 paper-thesis/chapters/21-adding-inference.tex
More writing on inference

@@
 
+Next when we type-check an \verb|Infer| expression we just type-check the meta-variable inside.
+This will wait until the value of the meta-variable is known before type-checking, which is the behavior we want.
+\begin{lstlisting}
+typeOfExpr_(s, Infer(q)) = (Infer(q), t) :-
+	typeOfExpr_(s, q) == (_, t).
+\end{lstlisting}
+
+When beta reducing we deliberately keep the \verb|Infer| expression intact, since we want to keep the information that it is an expression that might have to be inferred during beta equality checks.
+\begin{lstlisting}
+betaReduce_((_, Infer(e))) = Infer(e).
+\end{lstlisting}
+
+Finally, the difficult part of handling inference in beta equality. There are four different cases that involve \verb|Infer| expressions in beta equality, these are:
+
+\begin{enumerate}
+	\item A value \verb|e1| on the left, an infer expression on the right. In this case we simply want to set the metavariable equal to \verb|e1|. For example:
+	\begin{lstlisting}
+expectBetaEq((s1, e1@Type(_)), (_, Infer(e2))) :- e1 == e2.
+expectBetaEq((s1, e1@BoolTrue()), (_, Infer(e2))) :- e1 == e2.
+expectBetaEq((s1, e1@BoolFalse()), (_, Infer(e2))) :- e1 == e2.
+expectBetaEq((s1, e1@BoolType()), (_, Infer(e2))) :- e1 == e2.
+	\end{lstlisting}
+	\item A complex expression \verb|e1| on the left, an infer expression on the right. 
+	\item An infer expression on the left, any expression on the right.
+	\item An infer expression on the left and on the right.
+\end{enumerate}
+
 \section{Analysis of the power of AFOU}

### 2023-01-11 paper-thesis/chapters/02-background-spoofax.tex
Work on section 3
x

@@
 \chapter{\label{chap:bg-spoofax}Background: Spoofax}
+
+This chapter will explain the concepts behind the Spoofax Language Workbench. If you are already familiar with Spoofax, you can skip this chapter.
+
+\section{Spoofax}
 
@@
 
-\section{\label{sec:statix} Statix and Scope graphs}
+\section{\label{sec:statix}Statix and Scope graphs}
+
+\textit{Note: This section is adapted from the Explanation in "Composable Type System Specification
+	using Heterogeneous Scope Graphs". \cite[sect. 4.1.2]{Zwaan21}}
 
@@
 \end{itemize}
-
-% CREDIT FOR THIS EXPLANATION TO ARON ZWAAN
-% HOW TO CREDIT?

### 2023-01-13 paper-thesis/chapters/21-adding-inference.tex
x
Finish inference chapter

@@
 There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inferences than others.
-One algorithm for unification is \emph{first-order unification}, where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the the value free variable to be equal to the value of the non-free variable.
+One algorithm for unification is \emph{first-order unification} (FOU), where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the the value free variable to be equal to the value of the non-free variable.
 There are some situations in which this approach fails, but in most real-world scenarios it works perfectly.
@@
 	\end{lstlisting}
+
 	\item A complex expression \verb|e1| on the left, an infer expression on the right.
-	\item An infer expression on the left, any expression on the right.
-	\item An infer expression on the left and on the right.
+In this case, we know what top-level constructor of the metavariable should be, but not necessarily the entire constructor (there might be infers in \verb|e1|).
+We introduce new \verb|Infer| expressions on the left, and call \verb|expectBetaEq| recursively.
+A simple example for \verb|BoolIf| is below:
+	\begin{lstlisting}
+expectBetaEq((s1, e1@BoolIf(c1, t1, b1)), (_, Infer(e2))) :-
+	{c2 t2 b2}
+	e2 == BoolIf(Infer(c2), Infer(t2), Infer(b2)),
+	expectBetaEq((s1, e1), (sEmpty(), e2)).
+	\end{lstlisting}
+	\verb|FnType| and \verb|FnConstruct| work similarly,
+
+	\item An infer expression on the left, any expression on the right. We should not duplicate the previous two rules, instead, we can just swap the two expressions and re-use the rules above.
+\begin{lstlisting}
+expectBetaEq((_, Infer(e1)), (s2, e2)) :-
+	expectBetaEq((s2, e2), (_, Infer(e1))).
+\end{lstlisting}
+
+	\item \label{inf-case4} An infer expression on both sides.
+This is where the approximation comes in.
+In normal first-order unification, we would see if either side is known, and possible apply one of the rules above depending on the result.
+This is not possible, so we're going to do something that approximates first-order unification: just set both sides to be equal.
+This is an approximation because this might fail if both sides are equal under beta equality but not identical.
+This approximation is analyzed in section \ref{analysing-inference}.
+\begin{lstlisting}
+expectBetaEq((_, Infer(e1)), (_, Infer(e2))) :-
+	e1 == e2.
+\end{lstlisting}
+
 \end{enumerate}
 
-\section{Analysis of the power of AFOU}
+\section{Analysis of the approximation}
 \label{analysing-inference}
 
-\begin{comment}
+The only difference between AFOU and FOU is case \ref{inf-case4}.
+Instead of asserting that both sides are equal under beta equality, we assert that both sides are identical.
+Both sides being identical implies that they are beta-equal, so this approximation is sound, meaning there is no program that AFOU can infer but FOU cannot.
 
-\section{Using Statix' first-order unification for inference}
+The approximation is not complete: There are situations where AFOU fails to infer a program that FOU can infer.
+However, these programs are rare.
+This happens when a program contains multiple infer points, two of these are asserted to be beta-equal, and they are not identical but they are beta-equal.
+\todo{Find program where this fails.}
 
-When an \verb|Infer| needs to be type-checked, the logic is very similar, except that we don't need to generate a new metavariable:
+\section{Analysis of the implementation}
+\label{analysing-inference-impl}
+
+This process of defining a rule for each constructor in the language is a bit tedious, and it easy to accidentally get wrong.
+Stratego (the transformation language of Spoofax) allows for \emph{generic term deconstruction} \cite{generictermdestr}, if Statix would also have this feature this would make writing the rules above in a generic way possible.
+Borrowing some of the syntax from Stratego, this would look like this:
 
 \begin{lstlisting}
-typeOfExpr_(s, Infer(q)) = (Infer(q), t) :-
-	typeOfExpr_(s, q) == (_, t).
+expectBetaEq((s1, e1@c#(xs)), (_, Infer(e2))) :-
+	e2 == c#(<all(Infer)>(xs)),
+	expectBetaEq((s1, e1), (sEmpty(), e2)).
 \end{lstlisting}
-When we encounter a \verb|Infer| in \verb|betaReduceHead|, we keep it intact, because we still want to know that it is a \verb|Infer| in the \verb|expectBetaEq| rule.
-
-The \verb|expectBetaEq| is where it gets interesting. We have a rule for each constructor in the language. For simple cases like \verb|BoolTrue|, we assert that \verb|e2| must be equal.
-\begin{lstlisting}
-expectBetaEq_((s1, e1@BoolTrue()), (_, Infer(e2))) :- e1 == e2.
-\end{lstlisting}
-For more complicated constructors such as \verb|FnType|, we generate new metavariables for the subexpressions, and assert they it must be beta-equal to the provided ones:
-\begin{lstlisting}
-expectBetaEq_(
-	(s1, e1@FnType(arg_name1, arg_type1, body1)), 
-	(_, Infer(e2))) :- {arg_type2 body2}
-  e2 == FnType(arg_name1, Infer(arg_type2), Infer(body2)),
-  expectBetaEq_((s1, e1), (sEmpty(), e2)).
-\end{lstlisting}
-
-Finally, there is a case for if we encounter two \verb|Infer|s.
-Ideally, we would look at whether one infer is instantiated, and apply the rules above.
-However, this is not possible since we cannot do such queries in Statix.
-Instead, we say that they must be exactly equal and we hope for the best.
-There are situations where this rule fails (where e1 and e2 are beta-equal but not identical), but this works well enough for practical use.
-\begin{lstlisting}
-expectBetaEq_((_, Infer(e1)), (_, Infer(e2))) :-
-    e1 == e2.
-\end{lstlisting}
-\end{comment}

### 2023-01-13 paper-thesis/chapters/22-adding-datatypes.tex
Finish one chapter
Write part of the datatypes chapter

@@
 \chapter{Inductive Data Types}
+\label{ch:datatypes}
 
@@
 \begin{enumerate}
+\item Data types may be recursive, that is the data type may take itself as one of the constructor arguments. This can be seen in the definition of \verb|Nat| above.
 \item Parameters, which are datatypes that are polymorphic, and are required to be the same for all constructors, such as
@@
 \end{lstlisting}
+\end{enumerate}
 
-\section{In statix}
+A difference between Agda and our implementation is that Agda has case matching as a native construct, whereas we chose to use eliminators \todo{Citation}.
+An eliminator is a function that can be used to case match on a data type.
+For example, the eliminator of the \verb|Nat| type above is:
+\begin{lstlisting}
+elim Nat : P: (v: Nat -> Type) -> P Z 
+	-> (x: Nat -> P x -> P (S x)) -> n: Nat -> P n
+\end{lstlisting}
+
+The general type of an eliminator for a data type \verb|N| is:
+\begin{lstlisting}
+parameters 
+-> P: (indices -> N params indices -> Type) 
+-> constructors
+-> indices
+-> v: N params indices
+-> P indices v
+\end{lstlisting}
+The meaning of all the arguments is:
+\begin{itemize}
+	\item \verb|parameters| are the parameters of the data type we want to eliminate, since these are constant among all constructors those can be at the start.
+	\item \verb|P| is the type that this eliminator will return. It may depend on the value \verb|v| that is eliminated.
+	\item \verb|constructors| is a function that eliminates each of the constructors of the datatype.
+To eliminate a constructor of the form \verb|C : args -> N params indices| it generates a function \verb|args -> P indices (C pars args)|.
+	\item \verb|indices| are the indices of the value that is to be eliminated.
+	\item \verb|v| is the value that is to be eliminated.
+	\item Finally, \verb|P indices v| is the result of the elimination.
+\end{itemize}
+
+\section{Defining inductive data types}
 
@@
 \label{ch:datatypes:positivity}
-
-\end{enumerate}

### 2023-01-14 paper-thesis/chapters/22-adding-datatypes.tex
Writing on the datatypes chapter

@@
 \label{ch:datatypes}
+
+\section{Introduction to Inductive Data Types}
+\label{ch:datatypes:intro}
 
@@
 
-\section{Defining inductive data types}
+\section{Type-checking data type declarations}
+\label{ch:datatypes:typechecking}
+
+The definition of \verb|type_check| for an inductive data type \verb|N| is, conceptually:
+\begin{enumerate}
+	\item Create a new scope \verb|s1| whose parent is the scope the declaration was in \verb|s0|.
+	\item Declare \verb|N : Params -> Indices -> Type| in \verb|s1|.
+	\item \label{datatype-decl}We use a new scope-graph relation \verb|datatype : ID -> Expr| and declare \\ \verb|N : DataType(name, params, indices, constructors)| in \verb|s1| so that we can access information about the data type later.
+
+	\item Create a chain of scopes starting in \verb|s1| that will contain a scope for each parameter.
+We will type-check each next parameter using the previous parameters scope, so that parameters can depend on previous parameters.
+Call the end of this chain \verb|s2|.
+
+	\item Create a chain of scopes starting in \verb|s2| that will contain a scope for each index.
+We will type-check each next index using the previous index' scope, so that one index can depend on parameters and previous indices.
+Call the end of this chain \verb|s3|.
+
+	\item Create a new scope \verb|s4| with \verb|s1| as the parent which will be the scope that the rest of the program will be type-checked with.
+Then type-check each constructor with \verb|s2| as scope (so that the constructors can depend on parameters, but cannot depend on indices), and declare the type of that constructor in \verb|s4|.
+The parameters are automatically added to the type of the constructor.
+\end{enumerate}
+For example, for the following inductive datatype:
+\begin{lstlisting}
+data Eq (T : Type) : (e1 : T) (e2 : T) -> Type where
+	refl : e : T -> Eq e e;
+...
+\end{lstlisting}
+Figure \ref{scope-graph-eq} shows the scope graph generated by this data type.
+\begin{figure}[!h]
+	\begin{framed}
+		\centering
+		\begin{tikzpicture}[scopegraph, node distance = 6.5em and 5em]
+			% S0 and S1 node
+			\node[scope] (s0) {$s0$};
+			\node[scope, below = 2em of s0] (s1) {$s1$};
+			\draw (s1) edge[lbl={P}] (s0);
+
+			% S1 declarations
+			\node[type, above right=0.5em and 3em of s1, text width=11cm] (ntype) {\texttt{Eq : (T : Type) -> (e1 : T) -> (e2 : T) -> Type)}};
+			\draw (ntype) edge[type, lbl={name}] (s1);
+			\node[type, below right=0em and 3em of s1, text width=11cm] (ndatatype) {\texttt{Eq : DataType("Eq", [("T", Type())], [("e1", Var("T")), ("e2", Var("T"))], [("refl", ...)])}};
+			\draw (ndatatype) edge[type, lbl={datatype}] (s1);
+
+			% Path to S3
+			\node[scope, below right=4em and 1em of s1] (s2a) {s2};
+			\draw (s2a) edge[lbl={P}] (s1);
+			\node[type, right=3em of s2a] (ntype) {\texttt{T : Type}};
+			\draw (ntype) edge[type, lbl={name}] (s2a);
+
+			\node[scope, below=1em of s2a] (s2b) {};
+			\draw (s2b) edge[lbl={P}] (s2a);
+			\node[type, right=3em of s2b] (ntype) {\texttt{e1 : T}};
+			\draw (ntype) edge[type, lbl={name}] (s2b);
+
+			\node[scope, below=1em of s2b] (s2c) {s3};
+			\draw (s2c) edge[lbl={P}] (s2b);
+			\node[type, right=3em of s2c] (ntype) {\texttt{e2 : T}};
+			\draw (ntype) edge[type, lbl={name}] (s2c);
+
+			% S4
+			\node[scope, below = 12em of s1] (s4) {$s4$};
+			\draw (s4) edge[lbl={P}] (s1);
+			\node[type, right=3em of s4] (ntype) {\texttt{refl : (T : Type) -> (e : T) -> Eq e e}};
+			\draw (ntype) edge[type, lbl={name}] (s4);
+		\end{tikzpicture}
+	\end{framed}
+	\caption{The scope graph generated by the Eq data type.}
+	\label{scope-graph-eq}
+\end{figure}
+
+\section{Type-checking eliminators}
+
+The type of an eliminator was discussed in section \ref{ch:datatypes:intro}. When we type check an \verb|elim N| expression, the following happens:
+\begin{enumerate}
+	\item Query the scope to find the data type declaration that we created in point \ref{datatype-decl} of section \ref{ch:datatypes:typechecking}.
+This gives us access to the parameters, indices and constructors of the data type that is being eliminated.
+	\item Using some tedious logic, create the type discussed in section \ref{ch:datatypes:intro}. This ends up being around 100 lines of Statix code.
+\end{enumerate}
+
+The second part that needs to be defined is beta reducing an eliminator. During this process, the following happens:
+\begin{enumerate}
+	\item Query the scope to find the data type declaration that we created in point \ref{datatype-decl} of section \ref{ch:datatypes:typechecking}.
+This gives us access to the parameters, indices and constructors of the data type that is being eliminated.
+(Same as during type checking)
+	\item Try to split the arguments applied to the eliminator into the groups defined in section \ref{ch:datatypes:intro}.
+If this fails (because there are not enough arguments), the eliminator cannot be reduced.
+	\item Beta head reduce \verb|v|, if this does not result in one of the constructors at the head then the eliminator cannot be reduced. 
+	\item 
+\end{enumerate}
 

### 2023-01-14 paper-thesis/chapters/40-discussing-statix.tex
Writing on the datatypes chapter

\chapter{The usability of Statix for defining dependently typed languages}

### 2023-01-16 paper-evcs/paper.tex
Changes based on Aron's feedback

@@
 
-\keywords{Spoofax, Statix, Dependent Types} %TODO mandatory; please add comma-separated list of keywords
+\keywords{Spoofax, Statix, Dependent Types, Scope Graphs, Calculus of Constructions}
+
+\supplementdetails[subcategory ={ Source Code }]{The implementation of the language}{https://doi.org/10.5281/zenodo.7541014}
 
@@
 %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventEditors{Ralf L\"{a}mmel, Peter D. Mosses, and Friedrich Steimann}
+\EventNoEds{3}
+\EventLongTitle{Eelco Visser Commemorative Symposium (EVCS 2023)}
+\EventShortTitle{EVCS 2023}
+\EventAcronym{EVCS}
+\EventYear{2023}
+\EventDate{April 5, 2023}
+\EventLocation{Delft, The Netherlands}
 \EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
+\SeriesVolume{109}
+\ArticleNo{30}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone.
-The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system.
-However, so far Statix has not been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking.
+The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of a type system.
+However, so far Statix has not been used to implement dependent types, which is a class of type systems which require evaluation of terms during type checking.
 In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
@@
 
-Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages \cite{spoofax}.
+Spoofax is a language workbench: \label{key}a collection of tools that enable the development of textual languages~\cite{spoofax}.
 When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
-To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type systems.
-However, no attempts have been made to express dependently typed languages in Statix.
+It is a declarative language that uses inference rules to define the static semantics.
+From these rules a type checker and other editor tools can be derived.
+Statix aims to cover a broad range of languages and type systems.
+However, no attempts have been made to express dependently typed languages in Statix until now.
 
@@
 This expressiveness also makes dependent type systems more complicated to implement.
-Especially, deciding equality of types requires evaluation of the terms they are parameterized by.
+Especially, deciding equality of types requires normalization of the terms they are parameterized by.
 
@@
 We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix.
-The goal is not only to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.
+The goal is not only to show that Statix can implement it, but also that the implementation is concise.
 
@@
 There were some challenges in implementing the dependent type system, which this paper also discusses.
-Finally, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
+Then, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.
 
@@
 Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type check the language.
-The implementation of the language is available on GitHub.
-\url{https://github.com/JonathanBrouwer/master-thesis/}
 
@@
 
-The base language that has been implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}.
+The base language that has been implemented is the Calculus of Constructions~\cite{Coquand_Huet_1988}.
 One extra feature was added that is not present in the Calculus of Constructions: let bindings.
 Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
-The abstract syntax of the language is available in figure \ref{fig:syntax}.
+The concrete syntax of the language is available in figure \ref{fig:syntax}.
 
@@
 \begin{lstlisting}
-Type        : Expr
-Let         : ID * Expr * Expr -> Expr
-Var         : ID -> Expr
-FnType      : ID * Expr * Expr -> Expr
-FnConstruct : ID * Expr * Expr -> Expr
-FnDestruct  : Expr * Expr -> Expr
+Expr.Let = "let" ID "=" Expr ";" Expr
+Expr.Type = "Type"
+Expr.Var = ID
+Expr.FnType = ID ":" Expr "->" Expr {right}
+Expr.FnConstruct = "\\" ID ":" Expr "." Expr
+Expr.FnDestruct = Expr Expr {left}
 \end{lstlisting}
-\caption{The syntax for the base language. FnConstruct is a lambda function, FnDestruct is application of a lambda function.}
+\caption{The concrete syntax for the base language. $\co{FnConstruct}$ is a lambda function, $\co{FnDestruct}$ is application of a lambda function.}
 \label{fig:syntax}
@@
 \end{lstlisting}
-The AST of this program in Aterm format\cite{aterm} would be:
-\begin{lstlisting}
-Let(
-  "f", 
-  FnConstruct("T", Type(), FnConstruct("x", Var("T"), Var("x"))),
-  FnDestruct(
-	FnDestruct(Var("f"), FnConstruct("_", Type(), Type())),
-	FnConstruct("x", Type(), Var("x"))
-  )
-)
-\end{lstlisting}
 
@@
 To type check the base language, we need to store information about the names that are in scope at each point in the program.
-There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need the value of a binding to compare types, this is needed in dependently typed languages.}
+There are two different cases, names that do not have a known value (only a type), which are names created by function arguments, and names that do have a known value, which are names created by let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need \emph{the value} of a binding to compare types, this is needed in dependently typed languages.}
 
-In Statix, all this information can be stored in a \emph{scope graph} \cite{scope_graphs}, which is a feature of Statix.
+In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}.
 It is a graph consisting of nodes for scopes, labeled edges for visibility relations, scoped declarations for a relation, and queries for references.
@@
 This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.
+The Statix definition of these concepts is given below:
+\begin{lstlisting}
+constructors
+    NameType : Expr -> NameEntry
+    NameSubst : scope * Expr -> NameEntry
+relations
+    name : ID -> NameEntry
+\end{lstlisting}
 
@@
 A unique requirement for dependently typed languages is beta reduction during type checking, since types may require evaluation to compare.
-Beta reduction is the process of reduction a term to beta normal form, which is the state where no further beta reductions are possible \cite{tapl}.
+Beta reduction is the process of reduction a term to its beta normal form, which is the state where no further beta reductions are possible~\cite{tapl}.
 It works by matching a term of the form \verb|(\x.
@@
 
-We implemented beta reduction using a Krivine abstract machine\cite{krivine}.
+We implemented beta reduction using a Krivine abstract machine~\cite{krivine}.
 The machine can head evaluate lambda expressions with a call-by-name semantics.
@@
 We originally tried to implement the latter, which works fine for the base language.
-However, it runs into trouble when implementing inductive data types; more information about this will be in the full master's thesis\cite{thesis}.
+However, it runs into trouble when implementing inductive data types; more information about this will be in the full master's thesis~\cite{thesis}.
 An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.
@@
 First, the primary predicate is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
-The scope acts as the environment from \cite{krivine}, using \verb|NSubst| to store substitutions.
+The scope acts as the environment from~\cite{krivine}, using \verb|NSubst| to store substitutions.
 All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
-We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|.
+We use the syntax $\bhr{\scope{s_1}{e_1}}{t}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|.
+The \verb|t| in this definition is the argument stack of the Krivine machine.
 Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
-One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
+One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and the stack of arguments (of the Krivine machine state) and converts it to an expression by adding \verb|FnDestruct|s.
+Its signature is:
+\begin{lstlisting}
+rebuild : (scope * Expr) * list((scope * Expr)) -> (scope * Expr)
+\end{lstlisting}
 
@@
 			\bhr 
-			{ \scope{ \co{sPutSubst}(s, n, (s, v))}{b} } 
+			{ \scope{ \co{sPutSubst}(s, x, (s, e))}{b} } 
 			{ t }
-			{ \scope{ s' } { e' } }
+			{ \scope{ s' } { b' } }
 		}{
 			\bhr
-			{ \scope{ s }{ \co{Let}(n, v, b) } }
+			{ \scope{ s }{ \co{Let}(x, e, b) } }
 			{ t }
-			{ \scope { s' } { e' } }
+			{ \scope { s' } { b' } }
 		}
@@
 		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
+			\co{sGetName}(s, x) = \co{NSubst}(s_e, e) 
 			\\ \bhr
-			{\scope{se}{e}}
+			{\scope{s_e}{e}}
 			{t}
-			{\scope{se'}{e'}} 
+			{\scope{s_{e'}}{e'}} 
 		}{
 			\bhr
-			{\scope{s}{\co{Var}(n)}}
+			{\scope{s}{\co{Var}(x)}}
 			{t}
-			{\scope{se'}{e'}} 
+			{\scope{s_{e'}}{e'}} 
 		}
@@
 		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NType}(t) 
+			\co{sGetName}(s, x) = \co{NType}(t) 
 		}{ 
 			\bhr
-			{\scope{s}{\co{Var}(n)}}
+			{\scope{s}{\co{Var}(x)}}
 			{t}
-			{\co{rebuild}(s, \co{Var}(n), t)} 
+			{\co{rebuild}(s, \co{Var}(x), t)} 
 		}
@@
 			\bhr
-			{\scope{s}{\co{FnType}(n, a, b)}}
+			{\scope{s}{\co{FnType}(x, a, b)}}
 			{[]}
-			{\scope{s}{\co{FnType}(n, a, b)}} 
+			{\scope{s}{\co{FnType}(x, a, b)}} 
 		}
@@
 			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
+			{\scope{s}{\co{FnConstruct}(x, a, b)}}
 			{[]}
-			{\scope{s}{\co{FnConstruct}(n, a, b)}} 
+			{\scope{s}{\co{FnConstruct}(x, a, b)}} 
 		}
@@
 			\bhr
-			{\scope{\co{sPutSubst}(s, n, t)}{b}}
+			{\scope{\co{sPutSubst}(s, x, t)}{b}}
 			{ts}
@@
 			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
+			{\scope{s}{\co{FnConstruct}(x, \_, b)}}
 			{(t::ts)}
@@
 	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
+	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1), t) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
 	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
@@
 		\inferrule{ 
-			\toe{s}{v}{vt}
-			\\ \toe{\co{sPutSubst}(s, n, (s, v))}{b}{t}
+			\toe{s}{v}{t_e}
+			\\ \toe{\co{sPutSubst}(s, x, (s, e))}{b}{t_b}
 		}{
-			\toe{s}{\co{Let}(n, v, b)}{t}
+			\toe{s}{\co{Let}(x, e, b)}{t_b}
 		}
@@
 		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NType}(t) 
+			\co{sGetName}(s, x) = \co{NType}(t) 
 		}{ 
-			\toe{s}{\co{Var}(n)}{t}
+			\toe{s}{\co{Var}(x)}{t}
 		}
@@
 		\inferrule{ 
-			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
-			\\ \toe{se}{e}{t}
+			\co{sGetName}(s, x) = \co{NSubst}(s_e, e) 
+			\\ \toe{s_e}{e}{t}
 		}{
-		 	\toe{s}{\co{Var}(n)}{t}
+		 	\toe{s}{\co{Var}(x)}{t}
 		}
@@
 		\inferrule{ 
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()} 
+			\toe{s}{a}{t_a}
+			\\ \beq{t_a}{\co{Type}()} 
 			\\ \bred{s}{a}{a'}
-			\\\\  \toe{\co{sPutType}(s, n, a')}{b}{bt}
-			\\ \beq{bt}{\co{Type}()}
+			\\\\  \toe{\co{sPutType}(s, x, a')}{b}{t_b}
+			\\ \beq{t_b}{\co{Type}()}
 		} { 
-			\toe{s}{\co{FnType}(n, a, b)}{\co{Type}()}
+			\toe{s}{\co{FnType}(x, a, b)}{\co{Type}()}
 		}
@@
 		\inferrule{ 
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()}
+			\toe{s}{a}{t_a}
+			\\ \beq{t_a}{\co{Type}()}
 			\\ \bred{s}{a}{a'} 
-			\\\\ \toe{\co{sPutType}(s, n, a')}{b}{bt}
+			\\\\ \toe{\co{sPutType}(s, x, a')}{b}{t_b}
 		} { 
-			\toe{s}{\co{FnConstruct}(n, a, b)}{\co{FnType}(n, a', bt)}
+			\toe{s}{\co{FnConstruct}(x, a, b)}{\co{FnType}(x, a', t_b)}
 		}
@@
 		\inferrule{ 
-			\toe{s}{f}{ft}
+			\toe{s}{f}{t_f}
 			\\ \bhr
-				{\scope{s}{ft}}
+				{\scope{s}{t_f}}
 				{[]}
-				{\scope{s'}{\co{FnType}(da, dt, db)}} 
-			\\\\ \toe{s}{a}{at} 
-			\\ \beq{at}{\scope{sf}{dt}}
-			\\ \bred{\co{sPutSubst}(s', n, (s, arg))}{db}{db'}
+				{\scope{s_f}{\co{FnType}(x, t_{da}, t_b)}} 
+			\\\\ \toe{s}{a}{t_a} 
+			\\ \beq{t_a}{\scope{s_f}{t_{da}}}
+			\\ \bred{\co{sPutSubst}(s_f, x, (s, a))}{t_b}{t_b'}
 		} { 
-			\toe{s}{\co{FnDestruct}(f, a)}{db'}
+			\toe{s}{\co{FnDestruct}(f, a)}{t_b'}
 	 	}
@@
 It can be solved by using scopes to distinguish names.
-A full explanation of this will be in the master's thesis\cite{thesis}.
+A full explanation of this will be in the master's thesis~\cite{thesis}.
 
@@
 This section will discuss some further ideas that can be explored to build upon what has already been shown.
-These will be fully described in the master's thesis\cite{thesis} that this paper is based on, this is just a look into what is possible.
+These will be fully described in the master's thesis~\cite{thesis} that this paper is based on, this is just an exploration of what is possible.
 
@@
 \paragraph*{Semantic Code Completion}
-Finally, we explored how semantic code completion presented by Pelsmaeker et al.
-\cite{codecompletion} applies to dependently typed languages.
+Finally, we explored how semantic code completion presented by Pelsmaeker et al.~\cite{codecompletion} applies to dependently typed languages.
 It works well, only showing completions that are semantically possible.
@@
 The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
-Van Antwerpen et al.
-\cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types.
+Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types.
 It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.
@@
 Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language.
-The pi-forall language\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper.
+The pi-forall language~\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper.
 In principle, the implementations are very similar.
-For example, the inference rules presented in \cite{pi_forall} are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper.
-The primary difference is that they use a bidirectional type system, whereas this paper does not.
+For example, the inference rules of pi-forall are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper.
+The primary difference is that they use a bidirectional type system~\cite{bidirectional}, whereas this paper uses Statix' unification.
 

### 2023-01-16 paper-evcs/paper.tex
Changes based on Aron's feedback v2

@@
 \newcommand{\co}[1]{ \mathsf{#1} }
+
+\makeatletter
+\newcommand\@eatpar{\@ifnextchar\par{\expandafter\@eatpar\@gobble}\relax}
+\newcommand{\figuresection}[2][]{%
+	\par%
+	{\sffamily\bfseries #2}\hfill{#1}%
+	\smallskip%
+	\@eatpar}
+\makeatother
 
@@
 When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
-It is a declarative language that uses inference rules to define the static semantics.
+It is a declarative language that uses inference rules to define static semantics.
 From these rules a type checker and other editor tools can be derived.
@@
 Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
-The concrete syntax of the language is available in figure \ref{fig:syntax}.
+The concrete syntax~\cite{sdf3} of the language is available in figure \ref{fig:syntax}.
+
+Please observe that the syntax definition does not have a separate sort for types, as types may be arbitrary expressions in a dependently typed language.
+Furthermore, \verb|FnType| assigns a name to its first argument to allow the return type of the function to depend on the argument type.
 
@@
 let f = \T: Type. \x: T. x;
-f (_: Type -> Type) (\x: Type. x)
+f (T: Type -> Type) (\x: Type. x)
 \end{lstlisting}
@@
 To type check the base language, we need to store information about the names that are in scope at each point in the program.
-There are two different cases, names that do not have a known value (only a type), which are names created by function arguments, and names that do have a known value, which are names created by let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need \emph{the value} of a binding to compare types, this is needed in dependently typed languages.}
+There are two different cases, names that do not have a known value (only a type), which are names created by function arguments and dependent function types, and names that do have a known value, which are names created by let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need \emph{the value} of a binding to compare types, this is needed in dependently typed languages.}
 
@@
 \begin{figure}[h]
+	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{t}{\scope{s_2}{e_2}}$}]{Beta head-reduction rules}
+
 	\begin{mathpar}	
@@
 \begin{figure}[h]
+	\figuresection[\fbox{$\toe{s}{e}{t}$}]{Type checking rules}
 	\begin{mathpar}	
+		\mprset{vskip=0.7ex}
 		% Type rule
@@
 		\inferrule{ 
-			\toe{s}{v}{t_e}
+			\toe{s}{e}{t_e}
 			\\ \toe{\co{sPutSubst}(s, x, (s, e))}{b}{t_b}
@@
 		% FnType rule
+
 		\inferrule{ 
@@
 
+		\mprset{vskip=0ex}
 	\end{mathpar}
@@
     None : Maybe T,
-    Some : _: T -> Maybe T;
+    Some : x: T -> Maybe T;
 \end{lstlisting}

### 2023-01-16 paper-evcs/paper.tex
Swap paragraph and figure
Changes based on Aron's feedback v3

@@
 Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
-The concrete syntax~\cite{sdf3} of the language is available in figure \ref{fig:syntax}.
-
-Please observe that the syntax definition does not have a separate sort for types, as types may be arbitrary expressions in a dependently typed language.
-Furthermore, \verb|FnType| assigns a name to its first argument to allow the return type of the function to depend on the argument type.
+The concrete syntax (written in SDF3~\cite{sdf3}) of the language is available in figure \ref{fig:syntax}.
 
@@
 
+Please observe that the syntax definition does not have a separate sort for types, as types may be arbitrary expressions in a dependently typed language.
+Furthermore, \verb|FnType| assigns a name to its first argument to allow the return type of the function to depend on the argument type.
+
 An example program is the following, which defines a polymorphic identity function and applies it to a function:
@@
 let f = \T: Type. \x: T. x;
-f (T: Type -> Type) (\x: Type. x)
+f (T: Type -> Type) (\y: Type. y)
 \end{lstlisting}
@@
 All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
-We use the syntax $\bhr{\scope{s_1}{e_1}}{t}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|.
-The \verb|t| in this definition is the argument stack of the Krivine machine.
+We use the syntax $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), ps) == (s2, e2)|.
+The \verb|p| in this definition is the argument stack of the Krivine machine.
 Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
@@
 \begin{figure}[h]
-	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{t}{\scope{s_2}{e_2}}$}]{Beta head-reduction rules}
+	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$}]{Beta head-reduction rules}
 
@@
 			{ \scope{ \co{sPutSubst}(s, x, (s, e))}{b} } 
-			{ t }
+			{ \overline{p} }
 			{ \scope{ s' } { b' } }
@@
 			{ \scope{ s }{ \co{Let}(x, e, b) } }
-			{ t }
+			{ \overline{p} }
 			{ \scope { s' } { b' } }
@@
 			{\scope{s_e}{e}}
-			{t}
+			{\overline{p}}
 			{\scope{s_{e'}}{e'}} 
@@
 			{\scope{s}{\co{Var}(x)}}
-			{t}
+			{\overline{p}}
 			{\scope{s_{e'}}{e'}} 
@@
 			{\scope{s}{\co{Var}(x)}}
-			{t}
-			{\co{rebuild}(s, \co{Var}(x), t)} 
+			{\overline{p}}
+			{\co{rebuild}(s, \co{Var}(x), \overline{p})} 
 		}
@@
 			\bhr
-			{\scope{\co{sPutSubst}(s, x, t)}{b}}
-			{ts}
+			{\scope{\co{sPutSubst}(s, x, p)}{b}}
+			{\overline{p}}
 			{\scope{s'}{e'}} 
@@
 			{\scope{s}{\co{FnConstruct}(x, \_, b)}}
-			{(t::ts)}
+			{(p::\overline{p})}
 			{\scope{s'}{e'}} 
@@
 			{\scope{s}{f}}
-			{(a::ts)}
+			{(a::\overline{p})}
 			{\scope{s'}{e'}} 
@@
 			{\scope{s}{\co{FnDestruct}(f, a)}}
-			{ts}
+			{\overline{p}}
 			{\scope{s'}{e'}} 
@@
 	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1), t) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
+	\item $\bhr{\scope{s1}{e1}}{\overline{p}}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1), ps) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
 	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
@@
 
-For example, in the following code it would only suggest expressions that can be booleans (including variables that are booleans):
+For example, in the following code it would only suggest expressions that can be booleans:
 \begin{lstlisting}
 let f = \b: Bool. b;
-f {{Expr}}
+f [[Expr]]
 \end{lstlisting}
+
+This would return the following suggestions: (Note that \verb|f| and \verb|Type| are not suggested, since they cannot have type boolean)
+\begin{itemize}
+	\item \verb|[[Expr]] [[Expr]]|
+	\item \verb|let [[ID]] = [[Expr]]; [[Expr]]|
+	\item \verb|true|
+	\item \verb|false|
+	\item \verb|if [[Expr]] then [[Expr]] else [[Expr]] end|
+\end{itemize}
 

### 2023-01-18 paper-thesis/chapters/12-editor-services.tex
Add the semantic code completion chapter

% !TEX root = document.tex
\chapter{\label{chap:editor-services}Semantic Code Completion}

We explored how semantic code completion presented by Pelsmaeker et al.~\cite{codecompletion} applies to dependently typed languages. Code completion is an editor service in IDEs that proposes code fragments for the user to insert at the caret position in their code.

\section{Setup required}
To set up editor services, we followed the steps at \todo{cite} \url{https://spoofax.dev/spoofax-pie/develop/guide/static-semantics/code-completion/}. To be precise, we followed the following steps:
\begin{enumerate}
	\item Add \verb|tego-runtime {}| and \verb|code-completion {}| to the \verb|spoofax.cfg| file, to enable code completion.
	\item Add the following rules to the \verb|main.str2| file, to pre-process and post-process the AST for code completion.
	\begin{lstlisting}
rules
	downgrade-placeholders = downgrade-placeholders-MyLang
	upgrade-placeholders   = upgrade-placeholders-MyLang
	is-inj                 = is-MyLang-inj-cons
	pp-partial             = pp-partial-MyLang-string
	pre-analyze            = explicate-injections-MyLang
	post-analyze           = implicate-injections-MyLang
	\end{lstlisting}
	\item For each rule define a predicate that accepts a placeholder where a syntactic sort is permitted. For our language, those are the following:
	\begin{lstlisting}
expectBetaEq((_, Expr-Plhdr()), _).
expectBetaEq(_, (_, Expr-Plhdr())).
betaReduceHead((_, Expr-Plhdr())) = _.
betaReduce((_, Expr-Plhdr())) = _.
typeOfExpr(_, Expr-Plhdr()) = _.
programOk(Start-Plhdr()).
	\end{lstlisting}

\end{enumerate}

\section{How well it works}
In order to get suggestions, we can now insert a placeholder \verb|[[Expr]]| and press \verb|ctrl + space| in the editor to get semantic suggestions.

It works well, only showing completions that are semantically relevant. For example, given the following code it only suggests expressions that can be booleans:
\begin{lstlisting}
let f = \b: Bool. b;
f [[Expr]]
\end{lstlisting}
This would return the following suggestions: (Note that \verb|f| and \verb|Type| are not suggested, since they cannot have type \verb|Bool|)
 \begin{itemize}
 	\item \verb|[[Expr]] [[Expr]]|
 	\item \verb|let [[ID]] = [[Expr]]; [[Expr]]|
 	\item \verb|true|
 	\item \verb|false|
 	\item \verb|if [[Expr]] then [[Expr]] else [[Expr]] end|
 \end{itemize}

### 2023-01-18 paper-thesis/chapters/22-adding-datatypes.tex
Finish data types section

@@
 To eliminate a constructor of the form \verb|C : args -> N params indices| it generates a function \verb|args -> P indices (C pars args)|.
+For recursive datatypes, a previous case is generated, such as the \verb|P x| in the eliminator of \verb|Nat| above.
 	\item \verb|indices| are the indices of the value that is to be eliminated.
@@
 If this fails (because there are not enough arguments), the eliminator cannot be reduced.
-	\item Beta head reduce \verb|v|, if this does not result in one of the constructors at the head then the eliminator cannot be reduced. 
-	\item 
+	\item Beta head reduce \verb|v|, if this does not result in one of the constructors of the datatype at the head then the eliminator cannot be reduced. 
+	\item Apply the relevant function that eliminates the constructor to the arguments of the datatype. Generate recursive calls to the eliminator for recursive datatypes.
 \end{enumerate}
@@
 \label{ch:datatypes:positivity}
+
+When defining a recursive data type, we would like our datatype to be \emph{strictly positive}, meaning that it can only be recursive on positive positions.
+If this rule would not exist, we would be able to create non-terminating functions using the datatype.
+Since under the curry-howard correspondence this allows to prove false, this is undesirable.
+
+\begin{lstlisting}
+data Bad : Set where
+	bad : (Bad -> Bad) -> Bad
+	--     A       B       C
+	-- A is in a negative position, B and C are OK
+\end{lstlisting}
+
+The implementation is simply an extra check during the type checking of data type declarations. It checks that the datatype only refers to itself in positive positions.

### 2023-01-18 paper-thesis/chapters/12-editor-services.tex
Minor fix

@@
 
-\section{How well it works}
+\section{Quality of suggestions}
 In order to get suggestions, we can now insert a placeholder \verb|[[Expr]]| and press \verb|ctrl + space| in the editor to get semantic suggestions.

### 2023-01-20 paper-thesis/chapters/10-coc.tex
Reindex chapters + writing
Write the chapter about universes + update other chapters

--- 2022-11-07 paper-thesis/chapters/02-coc.tex
@@
 
-The base language that has been implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}.
+The base language that has been implemented is the Calculus of Constructions~\cite{Coquand_Huet_1988}.
 One extra feature was added that is not present in the Calculus of Constructions: let bindings.
 Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
-The abstract syntax of the language is available in figure \ref{fig:syntax}.
+The concrete syntax (written in SDF3~\cite{sdf3}) of the language is available in figure \ref{fig:syntax}.
 
@@
 \begin{lstlisting}
-Type        : Expr
-Let         : ID * Expr * Expr -> Expr
-Var         : ID -> Expr
-FnType      : ID * Expr * Expr -> Expr
-FnConstruct : ID * Expr * Expr -> Expr
-FnDestruct  : Expr * Expr -> Expr
+Expr.Let = "let" ID "=" Expr ";" Expr
+Expr.Type = "Type"
+Expr.Var = ID
+Expr.FnType = ID ":" Expr "->" Expr {right}
+Expr.FnConstruct = "\\" ID ":" Expr "." Expr
+Expr.FnDestruct = Expr Expr {left}
 \end{lstlisting}
-\caption{The syntax for the base language. FnConstruct is a lambda function, FnDestruct is application of a lambda function.}
+\caption{The concrete syntax for the base language. $\co{FnConstruct}$ is a lambda function, $\co{FnDestruct}$ is application of a lambda function.}
 \label{fig:syntax}
 \end{figure}
+
+Please observe that the syntax definition does not have a separate sort for types, as types may be arbitrary expressions in a dependently typed language.
+Furthermore, \verb|FnType| assigns a name to its first argument to allow the return type of the function to depend on the argument type.
 
@@
 let f = \T: Type. \x: T. x;
-f (_: Type -> Type) (\x: Type. x)
-\end{lstlisting}
-The AST of this program in Aterm format\cite{aterm} would be:
-\begin{lstlisting}
-Let(
-  "f",
-  FnConstruct("T", Type(), FnConstruct("x", Var("T"), Var("x"))),
-  FnDestruct(
-	FnDestruct(Var("f"), FnConstruct("_", Type(), Type())),
-	FnConstruct("x", Type(), Var("x"))
-  )
-)
+f (T: Type -> Type) (\y: Type. y)
 \end{lstlisting}
@@
 To type check the base language, we need to store information about the names that are in scope at each point in the program.
-There are two different cases, names that do not have a known value (only a type), such as function arguments, and names that do have a known value, such as let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need to value of a binding to compare types, this is needed in dependently typed languages.}
+There are two different cases, names that do not have a known value (only a type), which are names created by function arguments and dependent function types, and names that do have a known value, which are names created by let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need \emph{the value} of a binding to compare types, this is needed in dependently typed languages.}
 
-In Statix, all this information can be stored in a \emph{scope graph} \cite{scope_graphs}, which is a feature of Statix.
+In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}.
 It is a graph consisting of nodes for scopes, labeled edges for visibility relations, scoped declarations for a relation, and queries for references.
@@
 A unique requirement for dependently typed languages is beta reduction during type checking, since types may require evaluation to compare.
+Beta reduction is the process of reduction a term to its beta normal form, which is the state where no further beta reductions are possible~\cite{tapl}.
+It works by matching a term of the form \verb|(\x.
+b) e|.
+and substituting \verb|x| in \verb|b| with \verb|e|.
+Beta reduction applies this rule anywhere in the term, whereas beta head-reduction only applies this rule at the head (outermost expression) of the term, and produces a term in beta-head normal form.
 
-We implemented beta reduction using a Krivine abstract machine\cite{krivine}.
+We implemented beta reduction using a Krivine abstract machine~\cite{krivine}.
 The machine can head evaluate lambda expressions with a call-by-name semantics.
 It works by keeping a stack of all arguments that have not been applied yet.
-This turned out to be the more natural way of expressing this over substitution-based evaluation relation.
+This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation.
 We originally tried to implement the latter, which works fine for the base language.
-However, it runs into trouble when implementing inductive data types; more information about this will be in the full master thesis.
+However, it runs into trouble when implementing inductive data types; more information about this will be in the full master's thesis~\cite{thesis}.
 An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.
@@
 First, the primary predicate is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression.
-The scope acts as the environment from \cite{krivine}, using \verb|NSubst| to store substitutions.
+The scope acts as the environment from~\cite{krivine}, using \verb|NSubst| to store substitutions.
 All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
-We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|.
+We use the syntax $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), ps) == (s2, e2)|.
+The \verb|p| in this definition is the argument stack of the Krivine machine.
 Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
-One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.
+One predicate that is used for this is the \verb|rebuild| predicate, which takes a scoped expression and the stack of arguments (of the Krivine machine state) and converts it to an expression by adding \verb|FnDestruct|s.
+Its signature is:
+\begin{lstlisting}
+rebuild : (scope * Expr) * list((scope * Expr)) -> (scope * Expr)
+\end{lstlisting}
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[h]
+	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$}]{Beta head-reduction rules}
+
 	\begin{mathpar}
@@
 			\bhr
-			{ \scope{ \co{sPutSubst}(s, n, (s, v))}{b} }
-			{ t }
-			{ \scope{ s' } { e' } }
+			{ \scope{ \co{sPutSubst}(s, x, (s, e))}{b} }
+			{ \overline{p} }
+			{ \scope{ s' } { b' } }
 		}{
 			\bhr
-			{ \scope{ s }{ \co{Let}(n, v, b) } }
-			{ t }
-			{ \scope { s' } { e' } }
+			{ \scope{ s }{ \co{Let}(x, e, b) } }
+			{ \overline{p} }
+			{ \scope { s' } { b' } }
 		}
@@
 		\inferrule{
-			\co{sGetName}(s, n) = \co{NSubst}(se, e)
+			\co{sGetName}(s, x) = \co{NSubst}(s_e, e)
 			\\ \bhr
-			{\scope{se}{e}}
-			{t}
-			{\scope{se'}{e'}}
+			{\scope{s_e}{e}}
+			{\overline{p}}
+			{\scope{s_{e'}}{e'}}
 		}{
 			\bhr
-			{\scope{s}{\co{Var}(n)}}
-			{t}
-			{\scope{se'}{e'}}
+			{\scope{s}{\co{Var}(x)}}
+			{\overline{p}}
+			{\scope{s_{e'}}{e'}}
 		}
@@
 		\inferrule{
-			\co{sGetName}(s, n) = \co{NType}(t)
+			\co{sGetName}(s, x) = \co{NType}(t)
 		}{
 			\bhr
-			{\scope{s}{\co{Var}(n)}}
-			{t}
-			{\co{rebuild}(s, \co{Var}(n), t)}
+			{\scope{s}{\co{Var}(x)}}
+			{\overline{p}}
+			{\co{rebuild}(s, \co{Var}(x), \overline{p})}
 		}
@@
 			\bhr
-			{\scope{s}{\co{FnType}(n, a, b)}}
+			{\scope{s}{\co{FnType}(x, a, b)}}
 			{[]}
-			{\scope{s}{\co{FnType}(n, a, b)}}
+			{\scope{s}{\co{FnType}(x, a, b)}}
 		}
@@
 		\inferrule{
-		} { relation
+		} {
 			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
+			{\scope{s}{\co{FnConstruct}(x, a, b)}}
 			{[]}
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
+			{\scope{s}{\co{FnConstruct}(x, a, b)}}
 		}
@@
 			\bhr
-			{\scope{\co{sPutSubst}(s, n, a)}{b}}
-			{ts}
+			{\scope{\co{sPutSubst}(s, x, p)}{b}}
+			{\overline{p}}
 			{\scope{s'}{e'}}
@@
 			\bhr
-			{\scope{s}{\co{FnConstruct}(n, a, b)}}
-			{(t::ts)}
+			{\scope{s}{\co{FnConstruct}(x, \_, b)}}
+			{(p::\overline{p})}
 			{\scope{s'}{e'}}
@@
 			{\scope{s}{f}}
-			{(a::ts)}
+			{(a::\overline{p})}
 			{\scope{s'}{e'}}
@@
 			{\scope{s}{\co{FnDestruct}(f, a)}}
-			{ts}
+			{\overline{p}}
 			{\scope{s'}{e'}}
@@
 	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
+	\item $\bhr{\scope{s1}{e1}}{\overline{p}}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1), ps) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
 	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
@@
 
-\begin{figure}[ht]
+\begin{figure}[h]
+	\figuresection[\fbox{$\toe{s}{e}{t}$}]{Type checking rules}
 	\begin{mathpar}
+		\mprset{vskip=0.7ex}
 		% Type rule
@@
 		\inferrule{
-			\toe{s}{v}{vt}
-			\\ \toe{\co{sPutSubst}(s, n, (s, v))}{b}{t}
+			\toe{s}{e}{t_e}
+			\\ \toe{\co{sPutSubst}(s, x, (s, e))}{b}{t_b}
 		}{
-			\toe{s}{\co{Let}(n, v, b)}{t}
+			\toe{s}{\co{Let}(x, e, b)}{t_b}
 		}
@@
 		\inferrule{
-			\co{sGetName}(s, n) = \co{NType}(t)
+			\co{sGetName}(s, x) = \co{NType}(t)
 		}{
-			\toe{s}{\co{Var}(n)}{t}
+			\toe{s}{\co{Var}(x)}{t}
 		}
@@
 		\inferrule{
-			\co{sGetName}(s, n) = \co{NSubst}(se, e)
-			\\ \toe{se}{e}{t}
+			\co{sGetName}(s, x) = \co{NSubst}(s_e, e)
+			\\ \toe{s_e}{e}{t}
 		}{
-		 	\toe{s}{\co{Var}(n)}{t}
+		 	\toe{s}{\co{Var}(x)}{t}
 		}
@@
 		% FnType rule
+
 		\inferrule{
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()}
+			\toe{s}{a}{t_a}
+			\\ \beq{t_a}{\co{Type}()}
 			\\ \bred{s}{a}{a'}
-			\\\\  \toe{\co{sPutType}(s, n, a')}{b}{bt}
-			\\ \beq{bt}{\co{Type}()}
+			\\\\ \toe{\co{sPutType}(s, x, a')}{b}{t_b}
+			\\ \beq{t_b}{\co{Type}()}
 		} {
-			\toe{s}{\co{FnType}(n, a, b)}{\co{Type}()}
+			\toe{s}{\co{FnType}(x, a, b)}{\co{Type}()}
 		}
@@
 		\inferrule{
-			\toe{s}{a}{at}
-			\\ \beq{at}{\co{Type}()}
+			\toe{s}{a}{t_a}
+			\\ \beq{t_a}{\co{Type}()}
 			\\ \bred{s}{a}{a'}
-			\\\\ \toe{\co{sPutType}(s, n, a')}{b}{bt}
+			\\\\ \toe{\co{sPutType}(s, x, a')}{b}{t_b}
 		} {
-			\toe{s}{\co{FnConstruct}(n, a, b)}{\co{FnType}(n, a', bt)}
+			\toe{s}{\co{FnConstruct}(x, a, b)}{\co{FnType}(x, a', t_b)}
 		}
@@
 		\inferrule{
-			\toe{s}{f}{ft}
+			\toe{s}{f}{t_f}
 			\\ \bhr
-				{\scope{s}{ft}}
+				{\scope{s}{t_f}}
 				{[]}
-				{\scope{s'}{\co{FnType}(da, dt, db)}}
-			\\\\ \toe{s}{a}{at}
-			\\ \beq{at}{\scope{sf}{dt}}
-			\\ \bred{s'}{db}{db'}
+				{\scope{s_f}{\co{FnType}(x, t_{da}, t_b)}}
+			\\\\ \toe{s}{a}{t_a}
+			\\ \beq{t_a}{\scope{s_f}{t_{da}}}
+			\\ \bred{\co{sPutSubst}(s_f, x, (s, a))}{t_b}{t_b'}
 		} {
-			\toe{s}{\co{FnDestruct}(f, a)}{db'}
+			\toe{s}{\co{FnDestruct}(f, a)}{t_b'}
 	 	}
 
+		\mprset{vskip=0ex}
 	\end{mathpar}

### 2023-01-20 paper-thesis/chapters/21-adding-inference.tex
Write part of the datatypes chapter
Meeting results
Write the chapter about universes + update other chapters

@@
 For example, it can infer both programs in the introduction of this chapter, but it fails to infer the following program:
+
+\todo{higher-order pattern unification / miller pattern unification, what agda does}
 
@@
 \todo{Find program where this fails.}
-
-\section{Analysis of the implementation}
-\label{analysing-inference-impl}
-
-This process of defining a rule for each constructor in the language is a bit tedious, and it easy to accidentally get wrong.
-Stratego (the transformation language of Spoofax) allows for \emph{generic term deconstruction} \cite{generictermdestr}, if Statix would also have this feature this would make writing the rules above in a generic way possible.
-Borrowing some of the syntax from Stratego, this would look like this:
-
-\begin{lstlisting}
-expectBetaEq((s1, e1@c#(xs)), (_, Infer(e2))) :-
-	e2 == c#(<all(Infer)>(xs)),
-	expectBetaEq((s1, e1), (sEmpty(), e2)).
-\end{lstlisting}

### 2023-01-20 paper-thesis/chapters/22-adding-datatypes.tex
Write the chapter about universes + update other chapters

@@
 	\item Create a new scope \verb|s1| whose parent is the scope the declaration was in \verb|s0|.
-	\item Declare \verb|N : Params -> Indices -> Type| in \verb|s1|.
+	\item \label{name-decl}Declare \verb|N : Params -> Indices -> Type| in \verb|s1|.
 	\item \label{datatype-decl}We use a new scope-graph relation \verb|datatype : ID -> Expr| and declare \\ \verb|N : DataType(name, params, indices, constructors)| in \verb|s1| so that we can access information about the data type later.
@@
 
-	\item Create a new scope \verb|s4| with \verb|s1| as the parent which will be the scope that the rest of the program will be type-checked with.
+	\item Create a new scope \verb|s4| with \verb|s1| as the parent.
 Then type-check each constructor with \verb|s2| as scope (so that the constructors can depend on parameters, but cannot depend on indices), and declare the type of that constructor in \verb|s4|.
 The parameters are automatically added to the type of the constructor.
+
+	\item Type-check the rest of the program with scope \verb|s4|.
 \end{enumerate}

### 2023-01-20 paper-thesis/chapters/23-adding-universes.tex
Write the chapter about universes + update other chapters

\chapter{Universes}
\label{ch:universes}

Our language so far has the property that the type of \verb|Type| is \verb|Type|, this is called \emph{type in type}. This allows for \emph{Girard's paradox}, which means that the logic created by this language is inconsistent, we can prove false.

\section{Universes}

The solution to this problem is introducing universes into the language. We now have the following hierarchy of types:
\begin{lstlisting}
true : Bool : Type 0 : Type 1 : Type 2 : ...
\end{lstlisting}

In the rest of this chapter we will show how universes have been implemented. It turns out to be relatively easy to add this feature to the language, only requiring a few changes.

\section{Implementing universes in the CoC}

The constructor of Type has been changed to \verb|Type : int -> Expr|. We need to update all occurrences of \verb|Type| in figure \ref{fig:type-check-rules}. Figure \ref{fig:type-check-rules-universes} contains all the rules that require a change.

The three rules that were changed are:
\begin{enumerate}
	\item The rule for type-checking a \verb|Type| expression, which now is explicit about the universes of each \verb|Type|.
	\item The rule for type-checking \verb|FnType|, which takes the maximum of the two arguments.
	\item The rule for type-checking \verb|FnConstruct|, which ignores the universe of its argument type, and only verifies that it is a type.
\end{enumerate}

\begin{figure}[h]
	\figuresection[\fbox{$\toe{s}{e}{t}$}]{Type checking rules with universes}
	\begin{mathpar}
		\mprset{vskip=0.7ex}
		% Type rule
		\inferrule{
		} {
			\toe{s}{\co{Type}(u)}{\co{Type}(u+1)}
		}

		% FnType rule
		\inferrule{
			\toe{s}{a}{t_a}
			\\ \beq{t_a}{\co{Type}(u_a)}
			\\ \bred{s}{a}{a'}
			\\\\ \toe{\co{sPutType}(s, x, a')}{b}{t_b}
			\\ \beq{t_b}{\co{Type}(u_b)}
		} {
			\toe{s}{\co{FnType}(x, a, b)}{\co{Type}(max(u_a, u_b))}
		}

		% FnConstruct rule
		\inferrule{
			\toe{s}{a}{t_a}
			\\ \beq{t_a}{\co{Type}(u_a)}
			\\ \bred{s}{a}{a'}
			\\\\ \toe{\co{sPutType}(s, x, a')}{b}{t_b}
		} {
			\toe{s}{\co{FnConstruct}(x, a, b)}{\co{FnType}(x, a', t_b)}
		}

		\mprset{vskip=0ex}
	\end{mathpar}
	\caption{Rules for type checking the Calculus of Constructions with universes}
	\label{fig:type-check-rules-universes}
\end{figure}

\newpage
\section{Implementing universes with inductive data types}

We change data type declarations such that the universe the datatype lives in is now explicit. For example, the \verb|Nat| type now looks like this, explicitly stating that the type of \verb|Nat| is \verb|Type 0|.
\begin{lstlisting}
data Nat : -> Type 0 where
	zero : Nat,
	suc : Nat -> Nat;
\end{lstlisting}

We need to make the following changes to the type-checking algorithm:
\begin{enumerate}
	\item In point \ref{name-decl} of type-checking data type declarations in section \ref{ch:datatypes:typechecking}, we need to return \verb|Type u| instead of \verb|Type|, where \verb|u| is the declared universe level.
	\item For all constructors, we need to check that the universe of the constructor arguments is smaller than or equal to \verb|u|. This ensures that the datatype does actually live in the universe \verb|u|.
\end{enumerate}

### 2023-01-23 paper-thesis/chapters/30-comparing-with-haskell.tex
Compare with Haskell

@@
 \end{lstlisting}
+
+\section{Comparison}
+
+The base implementations of the language are quite similar.
+One difference is the way the definitions are clustered\todo{not right word?}.
+In Statix, one could put each language construct in a separate file, keeping the definitions for that construct together.
+In the Haskell implementation this is not easily possible\footnote{We could still define a function that then calls the actual implementation in the separate files, but this is still inferior to the Statix implementation.}, since function definitions can not be split over a file.
+
+Another advantage that Statix has is that it has first-order inference built-in, which makes implementing a basic form of inference as described in chapter \ref{ch:inference} way easier.
+However, if we want a more complex form of inference then an implementation in Haskell would be better, since it is a more expressive language and it has more libraries available.
+
+Finally, creating the language in Spoofax automatically gives us editor services such as code highlighting and semantic autocompletion, as dicusssed in chapter \ref{chap:editor-services}.
+Implementing these in Haskell would require some work.

### 2023-01-23 paper-thesis/chapters/10-coc.tex
Compare with Haskell
Work on comparison with lambdapi

@@
 We originally tried to implement the latter, which works fine for the base language.
-However, it runs into trouble when implementing inductive data types; more information about this will be in the full master's thesis~\cite{thesis}.
+However, it runs into trouble when implementing inductive data types.
 An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.
@@
 
+One thing to note is that some rules use \verb|betaReduce|.
+The goal of this beta reduce is to make the term into a term that does not need an environment (by substituting all let bindings).
+A full beta reduce is not necessary, but this is merely a performance optimization.
+
 The inference rules in figure \ref{fig:type-check-rules} can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
@@
 \end{figure}
+
+\section{Discussion of a substitution-based approach}
+\todo{Do this}

### 2023-01-23 paper-thesis/chapters/31-comparing-with-lambdapi.tex
Work on comparison with lambdapi

@@
 \label{ch:comp-lambdapi}
+
+There exist several \emph{logical frameworks}, tools designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}.
+Since these tools are designed specifically for the task, implementing the type system takes less effort in them compared to Spoofax, but for other tasks such as defining a parser or editor services they are not as well equipped.
+
+In this chapter we will be implementing the language defined in chapter \ref{chap:baselang} in lambdapi, a proof assistant based on the $\lambda \Pi$-calculus modulo rewriting~\cite{BoespflugCH12}.
+
+\section{Defining symbols}
+
+We will define a symbol in the meta language (lambdapi) for each construct in the object language.
+The result is visible in figure \ref{fig:lp-symbols}.
+We will leave \verb|Let| out of the language for now, it will be discussed separately in section \ref{ch:comp-lambdapi:let}.
+
+\begin{figure}[h]
+\begin{lstlisting}
+constant symbol TmSort : TYPE;
+symbol TmType : Π (a : TmSort), TYPE;
+
+constant symbol Type : TmSort;
+
+constant symbol FnType :
+  Π (A : TmSort), 
+  Π (B : TmType A → TmSort), 
+TmSort;
+
+symbol FnConstruct :
+  Π (A : TmSort), 
+  Π (B : TmType A → TmSort), 
+  Π (f : Π (x : TmType A), TmType (B x)), 
+  TmType (FnType A B);
+
+symbol FnDestruct :
+  Π (A : TmSort), 
+  Π (B : TmType A → TmSort), 
+  Π (f : TmType (FnType A B)),
+  Π (a : TmType A),
+  TmType (B a);
+\end{lstlisting}
+	\caption{Symbols of the Calculus of Constructions}
+\label{fig:lp-symbols}
+\end{figure}
+
+First, \verb|TmSort : TYPE| is the meta-language type of a type in the object language.
+So in any place where we say \verb|A : TmSort|, this means \verb|A| is a type in the object language.
+Next, \verb|TmType : (a : TmSort) -> TYPE| is the meta-language type of a term of type \verb|a| in the object language.
+So if we have \verb|x : TmType Bool| then \verb|x| is a boolean in the object language.
+
+Now we will define a symbol for each construct in the language we are defining.
+\begin{enumerate}
+	\item \verb|Type| is a type.
+	\item \verb|FnType| is a type, but it takes two arguments: \verb|A| is the argument type and \verb|B| is the return type, which is allowed to depend on a value of the argument type.
+	\item \verb|FnConstruct| takes three arguments: \verb|A| is the argument type, \verb|B| is the return type (allowed to depend on \verb|A| again), and \verb|f| is a function in the meta language of type \verb|x : A -> B x|.
+	\item \verb|FnDestruct| takes four arguments: \verb|A| is the argument type, \verb|B| is the return type (allowed to depend on \verb|A| again), a term of type \verb|A -> B| and an argument of type \verb|A|.
+\end{enumerate}
+
+Note that these types can only represent type-correct terms, \emph{intrinsically typed} terms.
+This is useful because it means the meta language does the type checking for us.
+The disadvantage is that it requires extra information: We need to give \verb|B| for \verb|FnConstruct| and \verb|A| and \verb|B| for \verb|FnDestruct|, which is information we don't have to provide to type-check terms in Statix.
+These could easily be automatically generated by a type checker but are tedious to specify manually.
+
+It seems like it should be possible to infer these, but lambdapi fails to infer even the simplest ones. \todo{not sure why}.
+
+\section{Reduction rules}
+
+We define reduction rules to reduce the symbols we defined in the previous section into lambdapi.
+Lambdapi can then evaluate the rules using its own semantics, not requiring any additional rules similar to the ones in figure \ref{fig:beta-head-reduce-rules}.
+The reduction rules are given in figure \ref{fig:lp-rules}.
+Each of the constructs is reduced to the corresponding construct in the meta language.
+
+	\todo{invisible chars}
+\begin{figure}[h]
+	\begin{lstlisting}
+rule TmType Type ↪ TmSort;
+rule TmType (FnType $A $B) ↪ Π (x : TmType $A), TmType ($B x);
+rule FnConstruct _ _ $f ↪ $f;
+rule FnDestruct _ _ $f $a ↪ $f $a;
+	\end{lstlisting}
+	\caption{Reduction Rules for the Calculus of Constructions}
+	\label{fig:lp-rules}
+\end{figure}
+
+\section{\label{ch:comp-lambdapi:let}Defining let bindings}
+
+Let bindings require substitution, which is not possible to encode in lambdapi.
+We can encode a less powerful version of let bindings, which are not substituted but evaluated via functions.
+The definition of this is given in figure \ref{fig:lp-let}.
+The body of the let binding is allowed to depend on a value of type \verb|A|, but it is not aware of the exact value \verb|v| of the let binding.
+
+	\todo{invisible chars}
+\begin{figure}[h]
+	\begin{lstlisting}
+symbol Let : 
+	Π (A : TmSort), 
+	Π (B : TmType A → TmSort), 
+	Π (v : TmType A),
+	Π (b : TmType A → TmType (B v)),
+	TmType (B v);
+rule Let _ _ $v $b ↪ $b $v;
+	\end{lstlisting}
+	\caption{Definition of less powerful Let in lampdapi}
+	\label{fig:lp-let}
+\end{figure}
+
+One program which would not type-check with this approach is the following.
+It fails to compile since it cannot know that \verb|b| is a boolean, which is required by the definition of f.
+\footnote{Assuming we introduce booleans into the language, there are examples that don't require booleans but they are a bit more difficult to understand}.
+\begin{lstlisting}
+let T = Bool;
+\f: Bool -> Bool;
+\b: T. f b
+\end{lstlisting}

### 2023-01-23 paper-thesis/chapters/98-relatedwork.tex
Work on comparison with lambdapi

@@
 The primary difference is that they use a bidirectional type system, whereas this paper does not.
-
-There exist several so-called \emph{logical frameworks}, tools designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}.
-Since these tools are designed specifically for the task, implementing the type system takes less effort in them compared to Spoofax, but for other tasks such as defining a parser or editor services they are not as well equipped.
-Some logical frameworks such as Twelf and Dedukti support Miller's \emph{higher-order pattern unification}~\cite{Miller89}, which can be used as a more powerful way of inferring implicit arguments than the first-order unification built into Statix.
-Andromeda 2 also supports \emph{extensionality rules} that can match on the type of an equality.
-We expect that adding extensionality rules to our implementation would be possible to do in Statix, but we leave an actual implementation to future work.

### 2023-01-25 paper-thesis/chapters/10-coc.tex
Work from yesterday

@@
 
-Finally, we define \verb|expectBetaEq|.
-This rule first beta reduces the heads of both sides, and then compares them.
-If the head is not the same, the rule fails.
-Otherwise, it recurses on the sub-expressions.
-One special case is when comparing two \verb|FnConstruct|s.
-Here we need to take into account alpha equality: two expressions which only differ in the names that they use should be considered equal.
-We implement this by substituting in the body of the functions, replacing their argument names with placeholders.
-
 \begin{figure}[h]
@@
 	\label{fig:beta-head-reduce-rules}
+\end{figure}
+
+\section{Beta equality}
+\label{sec:coc-betaeq}
+
+Finally, we define \verb|expectBetaEq|.
+This rule first beta reduces the heads of both sides, and then compares them.
+If the head is not the same, the rule fails.
+Otherwise, it recurses on the sub-expressions.
+One special case is when comparing two \verb|FnConstruct|s.
+Here we need to take into account alpha equality: two expressions which only differ in the names that they use should be considered equal.
+We implement this by substituting in the body of the functions, replacing their argument names with a unique placeholder.
+
+This substitution is called \verb|AlphaEqVars : ID * ID -> Expr|.
+The combination of the ID on the left and right hand side of the equality guarantees that the substitution is unique.
+In figure \ref{fig:beta-eq-rules} we show how \verb|AlphaEqVars| is used to determine equality of functions.
+
+\begin{figure}[h]
+	\figuresection[\fbox{$\beq{\scope{s_1}{e_1}}{\scope{s_2}{e_2}}$}]{Beta equality rules}
+
+	\begin{mathpar}
+		\inferrule{
+		} {
+			\beq
+			{ \co{AlphaEqVars}(x_1, x_2) }
+			{ \co{AlphaEqVars}(x_1, x_2) }
+		}
+
+		\inferrule{
+			\beq{\scope{s_1}{a_1}}{\scope{s_2}{a_2}}
+			\\\\ s_1' = \co{sPutSubst}(s_1, x_1, (\co{sEmpty}(), \co{AlphaEqVars}(x_1, x_2)))
+			\\\\ s_2' = \co{sPutSubst}(s_2, x_2, (\co{sEmpty}(), \co{AlphaEqVars}(x_1, x_2)))
+			\\\\ \beq{
+				\scope{s_1'}{b_1}
+				}{ 
+				\scope{s_2'}{b_2}
+			}
+		} {
+			\beq
+			{ \scope{s_1}{\co{FnType}(x_1, a_1, b_1)} }
+			{ \scope{s_2}{\co{FnType}(x_1, a_2, b_2)} }
+		}
+
+		\inferrule{
+			\beq{\scope{s_1}{a_1}}{\scope{s_2}{a_2}}
+			\\\\ s_1' = \co{sPutSubst}(s_1, x_1, (\co{sEmpty}(), \co{AlphaEqVars}(x_1, x_2)))
+			\\\\ s_2' = \co{sPutSubst}(s_2, x_2, (\co{sEmpty}(), \co{AlphaEqVars}(x_1, x_2)))
+			\\\\ \beq{
+				\scope{s_1'}{b_1}
+			}{ 
+				\scope{s_2'}{b_2}
+			}
+		} {
+			\beq
+			{ \scope{s_1}{\co{FnConstruct}(x_1, a_1, b_1)} }
+			{ \scope{s_2}{\co{FnConstruct}(x_1, a_2, b_2)} }
+		}
+	\end{mathpar}
+
+	\caption{Rules for beta equality in the Calculus of Constructions}
+	\label{fig:beta-eq-rules}
 \end{figure}

### 2023-01-25 paper-thesis/chapters/21-adding-inference.tex
Work from yesterday

@@
 
-\todo{higher-order pattern unification / miller pattern unification, what agda does}
-
 \begin{lstlisting}
@@
 
+A more powerful algorithm that could solve this is \emph{higher-order pattern unification} as defined by Miller~\cite{Miller89}, which is implemented in Agda.
+Implementing this in Statix is theoretically possible\footnote{since Statix is turing-complete}, but would be quite difficult to do in practice.
+
 \section{Inference in Statix}
@@
 This will work by explicitly denoting which variables \emph{could be} free, and explicitly handling these cases in a way that approximates first-order unification.
-We will denote this algorithm as \emph{approximated first-order unification (AFOU)}.
+We will name this algorithm \emph{approximated first-order unification (AFOU)}.
 
@@
 \todo{Find program where this fails.}
+
+\begin{lstlisting}
+postulate f : (A : Type -> _: A -> _: A -> A);
+postulate u : X : Bool -> Bool;
+postulate v : Y : Bool -> Bool;
+f _ u v
+\end{lstlisting}

### 2023-01-25 paper-thesis/chapters/40-discussing-statix.tex
Meeting results
Work from yesterday

@@
-\chapter{The usability of Statix for defining dependently typed languages}
+\chapter{Ergonomics of Spoofax}
+
+Now that we've implemented a feature-rich dependently typed language in Statix, this chapter discusses the experience.
+
+\section{Statix}
+
+- expressive language
+- more control over metavariables
+- nested constraints (datatypes reduce)
+- generic functions? reverse/concat for lists
+- debugging (statix tests are not great)
+
+\section{Spoofax}
+
+- errors
+- performance
+- testing works well
+-

### 2023-01-25 paper-thesis/chapters/31-comparing-with-lambdapi.tex
Work from yesterday
Result of meeting

@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{lstlisting}
@@
 	\todo{invisible chars}
-\begin{figure}[h]
+\begin{figure}[ht]
 	\begin{lstlisting}
@@
 	\todo{invisible chars}
-\begin{figure}[h]
+\begin{figure}[ht]
 	\begin{lstlisting}
 symbol Let : 
-	Π (A : TmSort), 
-	Π (B : TmType A → TmSort), 
-	Π (v : TmType A),
-	Π (b : TmType A → TmType (B v)),
+	`Π` (A : TmSort), 
+	`Π` (B : TmType A → TmSort), 
+	`Π` (v : TmType A),
+	`Π` (b : TmType A → TmType (B v)),
 	TmType (B v);

### 2023-01-25 paper-thesis/chapters/40-discussing-statix.tex
Result of meeting

--- 2023-01-14 paper-thesis/chapters/40-discussing-statix.tex
@@
-\chapter{The usability of Statix for defining dependently typed languages}
+\chapter{Ergonomics of Spoofax}
+
+Now that we've implemented a feature-rich dependently typed language in Statix, this chapter discusses the experience, and recommends some features which might improve the experience.
+
+\section{Statix}
+
+Statix is a simple language, not supporting too many complex features.
+This works well in some ways, but having some more features available could improve the experience drastically.
+Some features that would've helped this project create better code are:
+
+\begin{enumerate}
+	\item Nested constraints: Defining a constraint that can use the metavariables of a parent constraint. This works similar to nested function definitions. For example, for type checking datatypes:
+	\begin{lstlisting}
+typeOfExpr_(s, DataTypeDecl(n, ps, is, ue, cs, b)) = ... :-
+
+	isStrictlyPositive : (scope * Expr)
+	isStrictlyPositive(s, FnType(...)) = ...
+	\end{lstlisting}
+	The definition of \verb|isStrictlyPositive| can then use all the metavariables of \verb|DataTypeDecl| without having to explicitly pass them.
+
+	\item In order to implement more powerful inference algorithms, it would be useful to have a mechanism to check if metavariables have been instantiated or not.
+Since the order in which Statix evaluates constraints is so dynamic, we're not entirely sure how this mechanism would be implemented.
+More research is needed on this.
+
+	\item Generic constraints.
+For example, we created two separate relations for reversing lists of different types, making these one generic function would've been better.
+This was the only case where this feature would've been useful for this thesis, but it may be more useful for other work.
+
+	\item A better way to control the errors generated by Statix, such that users of the language get good errors.
+Statix does support \texttt{expr | error "Error text here"} annotations, but these don't allow constraints that instantiate metavariables on the left hand side whereas this would be needed for good error messages for dependently typed languages, since \verb|expectBetaEq| may infer variables.
+\end{enumerate}
+
+\section{Spoofax}
+
+- performance
+- testing works well
+- specifying grammars is still difficult
+- debugging (statix tests are not great)

### 2023-01-26 paper-thesis/chapters/40-discussing-statix.tex
Work on "Ergonomics of Spoofax"

@@
 Statix does support \texttt{expr | error "Error text here"} annotations, but these don't allow constraints that instantiate metavariables on the left hand side whereas this would be needed for good error messages for dependently typed languages, since \verb|expectBetaEq| may infer variables.
+\todo{This is not true}
 \end{enumerate}
@@
 
-- performance
-- testing works well
-- specifying grammars is still difficult
-- debugging (statix tests are not great)
+Next we will describe our experience with Spoofax as a whole.
+
+\begin{enumerate}
+	\item Spoofax is still slow, even though its performance has improved over the past years.
+On a high-end computer\footnote{Intel i7 6700k @ 4.2GHz, 16 GB Ram @ 2133MHz, Arch Linux @ Jan 2023, Spoofax 3 v0.19.2}, we took the following measurements.
+	\begin{itemize}
+		\item Full build (non-incremental) takes 55 seconds
+		\item Incremental build changing only one Statix file takes 5 seconds
+		\item Running all 171 tests takes 18 seconds
+	\end{itemize}
+	While these numbers are not unworkably high, they are still magnitudes higher than the implementation in Haskell (chapter \ref{ch:comp-haskell}) and other general-purpose languages.
+Improving these numbers would make the experience better.
+
+	\item The SPT testing DSL works very well, it makes writing integration tests very easy.
+However, support for unit tests is still lacking.
+The to create unit tests is using Statix tests, allowing assertions on specific relations.
+But these tests are not integrated with SPT tests (they don't even reside in the test folder) and there is no way to run all Statix tests, to ensure no regressions happened.
+Ideally there would be one button that runs all SPT and Statix tests.
+
+	\item Debugging Statix is still quite tedious.
+When we have a constraint that does not do what we expect it to do (ie.
+we have a failing Statix test), one either has to stare at its definition until they figure out the problem, or start unfolding the definitions of the constraints into the Statix test, finding out which part of the definition is misbehaving and repeating the unfolding.
+Having support for a debugger would help tremendously with this.
+\end{enumerate}

### 2023-01-26 paper-thesis/chapters/000-abstract.tex
Cleanup work

@@
 However, so far Statix has not been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking.
-In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
+In this thesis, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
 While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.

### 2023-01-26 paper-thesis/chapters/01-introduction.tex
Cleanup work

@@
 
-Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages \cite{spoofax}.
+Spoofax is a language workbench: a collection of tools that enable the development of textual languages~\cite{spoofax}.
 When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
-To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type systems.
-However, no attempts have been made to express dependently typed languages in Statix.
+It is a declarative language that uses inference rules to define static semantics.
+From these rules a type checker and other editor tools can be derived.
+Statix aims to cover a broad range of languages and type systems.
+However, no attempts have been made to express dependently typed languages in Statix until now.
 
@@
 This expressiveness also makes dependent type systems more complicated to implement.
-Especially, deciding equality of types requires evaluation of the terms they are parameterized by.
+Especially, deciding equality of types requires normalization of the terms they are parameterized by.
 
-\section{Contributions}
-This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language.
+\section*{Contributions}
+This goal of this thesis is to investigate how well Statix is fit for the task of defining a simple dependently-typed language.
 We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix.
-The goal is not only to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.
+The goal is not only to show that Statix can implement it, but also that the implementation is concise.
 
@@
 \begin{itemize}
-	\item 
+	\item We present a implementation of the Calculus of Constructions in Statix (Chapter \ref{chap:baselang})
+	\item We show the implementation is easily extendable (Chapter \ref{chap:bools})
+	\item We show how to add inference to the implementation (Chapter \ref{ch:inference})
+	\item We show how to add support for inductive datatypes to the implementation (Chapter \ref{ch:datatypes})
+	\item We show how to add support for universes to the implementation (Chapter \ref{ch:universes})
 \end{itemize}
 
-\section{Outline}
+\section*{Discussions}
 
-TODO navigation
+Furthermore, the thesis contains some discussions:
+\begin{itemize}
+	\item We discuss how well semantic code completion works for our language (Chapter \ref{chap:editor-services})
+	\item We compare our implementation with an implementation of the same language in Haskell (Chapter \ref{ch:comp-haskell})
+	\item We compare our implementation with an implementation of the same language in LambdaPi (Chapter \ref{ch:comp-lambdapi})
+	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (Chapter \ref{ch:ergonomics})
+	\item We discuss related work (Chapter \ref{ch:relatedwork}) and future work (Chapter \ref{ch:futurework})
+
+\end{itemize}

### 2023-01-26 paper-thesis/chapters/02-background-spoofax.tex
Cleanup work

@@
 
-The Spoofax Language Workbench \cite{spoofax} is a platform to develop domain-specific and
+The Spoofax Language Workbench~\cite{spoofax} is a platform to develop domain-specific and
 general-purpose programming languages.
@@
 
-The metalanguage SDF3 is used to define the syntax of the language \cite{sdf3}, which is then parsed using Spoofax's SGLR parser \cite{sdf3_parser}.
-The parser produces a abstract syntax tree in the ATerm format \cite{aterm}.
+The metalanguage SDF3 is used to define the syntax of the language~\cite{sdf3}, which is then parsed using Spoofax's SGLR parser~\cite{sdf3_parser}.
+The parser produces a abstract syntax tree in the ATerm format~\cite{aterm}.
 A pretty printer is also generated based on this syntax definition.
 
-Next, the metalanguage Statix \cite{scopes_as_types} is used to define the static semantics of the language.
+Next, the metalanguage Statix~\cite{scopes_as_types} is used to define the static semantics of the language.
 It uses scope graphs to represent binding and typing information of the language.
@@
 
-Finally, Stratego \cite{stratego} is used to define program transformations, using rewrite rules.
+Finally, Stratego~\cite{stratego} is used to define program transformations, using rewrite rules.
 This can be used for modifications to the AST, but also to build transformations to a different language (such as compiling to assembly).
@@
 \textit{Note: This section is adapted from the Explanation in "Composable Type System Specification
-	using Heterogeneous Scope Graphs". \cite[sect. 4.1.2]{Zwaan21}}
+	using Heterogeneous Scope Graphs".~\cite[sect. 4.1.2]{Zwaan21}}
 
-In Statix, scope graphs are used to represent a programs name and type information \cite{scopes_as_types}. A scope graph is a graph that consists of the following components:
+In Statix, scope graphs are used to represent a programs name and type information~\cite{scopes_as_types}. A scope graph is a graph that consists of the following components:
 

### 2023-01-26 paper-thesis/chapters/03-background-dependent-types.tex
Update 03-background-dependent-types.tex
Cleanup work

@@
 
-The Curry-Howard correspondence is a relationship that can be used to interpret typed programs as mathematical proofs \cite{chc}.
+The Curry-Howard correspondence is a relationship that can be used to interpret typed programs as mathematical proofs~\cite{chc}.
 This is done by representing false statements as empty types, and true statements as non-empty (inhabited) types.
@@
 
-The lambda cube is a framework to categorize programming languages based on whether terms and types can depend on each other \cite{lambda_cube}.
+The lambda cube is a framework to categorize programming languages based on whether terms and types can depend on each other~\cite{lambda_cube}.
 Specifically, it categorizes languages on three axis: \footnote{Note that the "Terms can depend on terms" axis is missing, this is because a language where terms cannot depend on terms cannot compute anything and is thus pretty useless.}

### 2023-01-27 paper-thesis/chapters/10-coc.tex
A substitution based approach

@@
 
-In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}.
-It is a graph consisting of nodes for scopes, labeled edges for visibility relations, scoped declarations for a relation, and queries for references.
+In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}, as explained in chapter \ref{chap:bg-spoofax}.
 We only use a single type of edge, called \verb|P| (parent) edges.
-It also only has a single relation, called \verb|name|.
+We also only have a single relation, called \verb|name|.
 This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.
@@
 The \verb|sPutType| and \verb|sPutSubst| predicates generate a new scope given a parent scope and a type or a substitution respectively.
-To query the scope graph use \verb|sGetName|, which will return a \verb|NameEntry| that the query found.
+To query the scope graph we use \verb|sGetName|, which will return the closest \verb|NameEntry| with a matching same.
 Finally, \verb|sEmpty| returns a fresh empty scope.
@@
 It works by keeping a stack of all arguments that have not been applied yet.
-This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation.
-We originally tried to implement the latter, which works fine for the base language.
-However, it runs into trouble when implementing inductive data types.
-An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.
+This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation, which is an alternative we will discuss in section \ref{sec:coc-subst}.
 
@@
 
-Finally, we define \verb|expectBetaEq|.
+We need to define \verb|expectBetaEq|, which asserts that two scoped expressions are equal under beta reduction.
 This rule first beta reduces the heads of both sides, and then compares them.
@@
 \section{Discussion of a substitution-based approach}
-\todo{Do this}
+\label{sec:coc-subst}
+
+An alternative for a Krivine machine, which keeps a stack of arguments it has encountered, is a substitution-based relation.
+This beta-reduces a \verb|FnDestruct| by doing a nested beta-reduction of the function, and substituting into that, as is shown in figure \ref{fig:subst-approach}.
+
+Although these rules look cleaner, they are more complicated to implement, requiring a separate relation to check if \verb|f| is a \verb|FnConstruct| or something else.
+For the pure calculus of constructions, this still works quite well but when adding inductive datatypes (chapter \ref{ch:datatypes}) this quickly becomes a mess.
+
+An additional benefit of Krivine machines is that the run-time performance of them tends to be a bit higher, though this is only a constant factor and not a time-complexity improvement.
+
+\begin{figure}[h]
+	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{}{\scope{s_2}{e_2}}$}]{A substitution-based approach for beta reduction}
+	\begin{mathpar}
+		\mprset{vskip=0.7ex}
+
+		\inferrule{
+			\bhr
+			{\scope{s}{f}}{}
+			{\scope{s_{f'}}{\co{FnConstruct}(x, \_, b)}}
+			\\ \bhr
+			{\scope{\co{sPutSubst}(s_{f'}, x, (s, a))}{b}}{}
+			{\scope{s_{b'}}{b'}}
+		} {
+			\bhr
+			{\scope{s}{\co{FnDestruct}(f, a)}}{}
+			{\scope{s_{b'}}{b'}}
+		}
+
+		\inferrule{
+			\bhr
+			{\scope{s}{f}}{}
+			{\scope{s_{f'}}{f'}}
+			\\ \nexists x e_1 e_2. f' = \co{FnConstruct}(x, e_1, e_2)
+		} {
+			\bhr
+			{\scope{s}{\co{FnDestruct}(f, a)}}{}
+			{\scope{s}{\co{FnDestruct}(f, a)}}{}
+		}
+
+		\inferrule{
+		} {
+			\bhr
+			{\scope{s}{\co{FnConstruct}(x, a, b)}}{}
+			{\scope{s}{\co{FnConstruct}(x, a, b)}}{}
+		}
+
+		\mprset{vskip=0ex}
+	\end{mathpar}
+	\caption{A substitution-based approach for beta reduction}
+	\label{fig:subst-approach}
+\end{figure}

### 2023-01-27 paper-thesis/chapters/11-variable-capture.tex
Cleanup work
Improve the variable capture chapter

@@
 The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is variable capture.
-Variable capture is the phenomenon of free variables in a term becoming bound when a naive substitution happens \cite{tapl}.
-This section will explore several ways of solving this.
+Variable capture is the phenomenon of free variables in a term becoming bound when a naive substitution happens~\cite{tapl}.
+This chapter will explore several ways of solving this.
 
@@
 \end{lstlisting}
-The algorithm so far would tell you it is \verb|T : Type -> T : T -> T|.
+The implementation so far would tell you it is \verb|T : Type -> T : T -> T|.
 Given the scoping rules of the language, that is equivalent to \verb|T : Type -> x : T -> x|.
@@
 Using de Bruijn indices in statix is possible, but sacrifices a lot.
-It would require a transformation on the AST before typechecking.
-Modifying the AST like this causes problems, because it changes AST nodes.
-All editor services that rely on \verb|.ref| annotations, such as renaming, can no longer be used.
-It also loses a lot of the benefits of using Spoofax, since using scope graphs relies on using names.
+Many editor services (renaming, go to definition) require \verb|.ref| annotations to be set on names, and this is not possible if the names are no longer a part of the AST.
 
@@
 This doesn't work for a variety of reasons.
-The simplest being, it doesn't actually solve the problem.
+First of all, it doesn't actually solve the problem.
 Names can be duplicated during beta reduction of terms, so we still don't have the guarantee that each variable has a unique name.
-Furthermore, this is a pre-analysis transformation, so similarly to using de Bruijn indices, it breaks editor services such as renaming.
+Furthermore, this changes the names in the AST, so it breaks editor services such as renaming.
 
@@
 It requires a new relation to traverse through the type and rename.
-It also increases the time complexity, as constant traversals of the type are needed.
+It also increases the time complexity, as many traversals of the type are needed.
+This was successfully implemented, but we chose against using it in favor of the solution presented in the next section.
 
@@
 To be precise, at the grammar level we have two sorts, \verb|RID| is a "Raw ID", being just a string.
-\verb|ID| will have two constructors, one being \verb|Syn|, a syntactical \verb|ID|.
-The second one is \verb|ScopedName|, it is defined in statix, so there is no syntax for it.
-It will be generated by \verb|typeOfExpr|.
+\verb|ID| will have two constructors, one being \verb|Syn|, a syntactical \verb|ID|, referring to the lexically closest match.
+The second one is \verb|ScopedName|, which is defined below.
 
@@
 The scope is used to uniquely identify the name.
-The main idea is that whenever we encounter a syntactical name, we replace it with a scoped name, so it unambiguous.
+The main idea is that whenever we encounter a syntactical name during type checking, we replace it with a scoped name, so it unambiguous.
 The scope graph will never have a syntactical name in it.
@@
 typeOfExpr(s1, FnConstruct(Syn("T"), Var(Syn("T")), Var(Syn("T"))))
-
-(ScopedName(s1, T) : Type)
-	       [s1]
 \end{lstlisting}
+\begin{tikzpicture}[scopegraph, node distance = 2.5em and 3.0em]
+	\node[scope] (s1) {$s1$};
+	\node[type, above of = s1] (s1l) {$\texttt{ScopedName(s1, T) : Type()}$};
+	\draw (s1) edge[type] (s1l);
+\end{tikzpicture}
 
@@
 typeOfExpr(s2, Var(Syn("T")))
+\end{lstlisting}
+\begin{tikzpicture}[scopegraph, node distance = 2.5em and 2.5em]
+	\node[scope] (s1) {$s1$};
+	\node[type, above of = s1] (s1l) {$\texttt{ScopedName(s1, T) : Type()}$};
+	\draw (s1) edge[type] (s1l);
 
-(ScopedName(s1, T) : Type)     (ScopedName(s2, T) : Var(ScopedName(s1, T)))
-	       [s1]<----------------------------------[s2]
-\end{lstlisting}
+	\node[scope, right of = s1, xshift=18em] (s2) {$s2$};
+	\node[type, above of = s2] (s2l) {$\texttt{ScopedName(s2, T) : Var(ScopedName(s1, T))}$};
+	\draw (s2) edge[type] (s2l);
+
+	\draw (s2) edge[] (s1);
+\end{tikzpicture}
 
@@
 FnType(ScopedName(s1, T), Type(),
-	FnType(ScopedName(s2, T), Var(ScopedName(s1, T)), Var(ScopedName(s1, T))))
+	FnType(ScopedName(s2, T), Var(ScopedName(s1, T)), 
+		Var(ScopedName(s1, T))))
 \end{lstlisting}
@@
 
-Because the expression above, with \verb|ScopedName|s, is not particularly readable, we add a new post-analysis Stratego pass that converts the \verb|ScopedName|s to ticked named.
+Because the expression above, with \verb|ScopedName|s, is not particularly readable, we add a new post-analysis Stratego pass that converts the \verb|ScopedName|s to ticked names.
 For example, the above would be transformed to:
@@
 \begin{enumerate}
-	\item When we encounter a \verb|ScopedName| for the first time, we keep adding ticks to the name until we find a name that has not been used before. 
+	\item When we encounter a \verb|ScopedName| in a definition, we keep adding ticks to the name until we find a name that has not been used before. 
 	\item We define a dynamic rule \verb|Rename :: string -> string| and we store the new name we generated using this rule.
-	\item When we encounter a \verb|ScopedName|, use the \verb|Rename| rule to find what the name was transformed to.
+	\item When we encounter a \verb|ScopedName| in a variable, use the \verb|Rename| rule to find what the name was transformed to.
 \end{enumerate}

### 2023-01-27 paper-thesis/chapters/21-adding-inference.tex
Improve inference chapter

@@
 A more powerful algorithm that could solve this is \emph{higher-order pattern unification} as defined by Miller~\cite{Miller89}, which is implemented in Agda.
-Implementing this in Statix is theoretically possible\footnote{since Statix is turing-complete}, but would be quite difficult to do in practice.
+Implementing this in Statix is theoretically possible\footnote{Since Statix is turing-complete}, but would be quite difficult to do in practice.
 
@@
 
-\section{Implementing AFOU}
+\section{Implementing AFOU} \todo{Do we want all these examples in Statix or math notation?}
 \label{implementing-inference}
@@
 expectBetaEq((s1, e1@BoolIf(c1, t1, b1)), (_, Infer(e2))) :-
-	{c2 t2 b2}
 	e2 == BoolIf(Infer(c2), Infer(t2), Infer(b2)),
@@
 	\end{lstlisting}
-	\verb|FnType| and \verb|FnConstruct| work similarly,
+	\verb|FnType| and \verb|FnConstruct| work similarly, but we don't only have sub-expressions but also a name to consider. Sadly, this is the first case where we have to approximate. 
+	\begin{lstlisting}
+expectBetaEq((s1, e1@FnConstruct(arg_name1, arg_type1, body1)), 
+	(_, Infer(e2))) :-
+	e2 == FnConstruct(arg_name1, Infer(arg_type2), Infer(body2)),
+	expectBetaEq_((s1, e1), (sEmpty(), e2)).
+	\end{lstlisting}
+	\label{inf-case2}
+	Ideally, we would generate a new name iff \verb|e2| is a free variable, otherwise using the already generated name.
+We can't query whether \verb|e2| is free, so as a best-attempt we always assume that \verb|e2| is free and generate a new name, this will fail in some cases.
+We will discuss this in the next section.
 
@@
 	\item \label{inf-case4} An infer expression on both sides.
-This is where the approximation comes in.
+Here, more approximation is required.
 In normal first-order unification, we would see if either side is known, and possible apply one of the rules above depending on the result.
@@
 
-The only difference between AFOU and FOU is case \ref{inf-case4}.
-Instead of asserting that both sides are equal under beta equality, we assert that both sides are identical.
+The only difference between AFOU and FOU is case \ref{inf-case2} (for functions) and \ref{inf-case4}.
+
+In case \ref{inf-case2}, this will only infer the value if it has not been inferred yet.
+This means that if the same value is inferred from two different places, this will infer two different values and fail.
+
+In case \ref{inf-case4}, instead of asserting that both sides are equal under beta equality, we assert that both sides are identical.
 Both sides being identical implies that they are beta-equal, so this approximation is sound, meaning there is no program that AFOU can infer but FOU cannot.
 
-The approximation is not complete: There are situations where AFOU fails to infer a program that FOU can infer.
-However, these programs are rare.
-This happens when a program contains multiple infer points, two of these are asserted to be beta-equal, and they are not identical but they are beta-equal.
-\todo{Find program where this fails.}
+The approximations are not complete: There are situations where AFOU fails to infer a program that FOU can infer.
+However, these programs are surprisingly uncommon considering how rough the approximations are.
+They mostly fail when we attempt to infer function values, and two functions are inferred that are beta-equal, and they are beta-equal but not identical.
+An example of a program where these approximation fails is:
 

### 2023-01-27 paper-thesis/chapters/22-adding-datatypes.tex
Work from yesterday
Minor writing changes

@@
 
-A difference between Agda and our implementation is that Agda has case matching as a native construct, whereas we chose to use eliminators \todo{Citation}.
+A difference between Agda and our implementation is that Agda has case matching as a native construct, whereas we chose to use eliminators~\cite{eliminators}.
 An eliminator is a function that can be used to case match on a data type.
@@
 Figure \ref{scope-graph-eq} shows the scope graph generated by this data type.
-\begin{figure}[!h]
+\begin{figure}[!ht]
 	\begin{framed}

### 2023-01-30 paper-thesis/chapters/31-comparing-with-lambdapi.tex
Add error annotations

@@
 constant symbol TmSort : TYPE;
-symbol TmType : Π (a : TmSort), TYPE;
+symbol TmType : `Π` (a : TmSort), TYPE;
 
@@
 constant symbol FnType :
-  Π (A : TmSort), 
-  Π (B : TmType A → TmSort), 
+  `Π` (A : TmSort), 
+  `Π` (B : TmType A → TmSort), 
 TmSort;
@@
 symbol FnConstruct :
-  Π (A : TmSort), 
-  Π (B : TmType A → TmSort), 
-  Π (f : Π (x : TmType A), TmType (B x)), 
+  `Π` (A : TmSort), 
+  `Π` (B : TmType A → TmSort), 
+  `Π` (f : `Π` (x : TmType A), TmType (B x)), 
   TmType (FnType A B);
@@
 symbol FnDestruct :
-  Π (A : TmSort), 
-  Π (B : TmType A → TmSort), 
-  Π (f : TmType (FnType A B)),
-  Π (a : TmType A),
+  `Π` (A : TmSort), 
+  `Π` (B : TmType A → TmSort), 
+  `Π` (f : TmType (FnType A B)),
+  `Π` (a : TmType A),
   TmType (B a);
@@
 
-	\todo{invisible chars}
 \begin{figure}[ht]
 	\begin{lstlisting}
-rule TmType Type ↪ TmSort;
-rule TmType (FnType $A $B) ↪ Π (x : TmType $A), TmType ($B x);
-rule FnConstruct _ _ $f ↪ $f;
-rule FnDestruct _ _ $f $a ↪ $f $a;
+rule TmType Type `↪` TmSort;
+rule TmType (FnType $A $B) `↪` `Π` (x : TmType $A), TmType ($B x);
+rule FnConstruct _ _ $f `↪` $f;
+rule FnDestruct _ _ $f $a `↪` $f $a;
 	\end{lstlisting}
@@
 	TmType (B v);
-rule Let _ _ $v $b ↪ $b $v;
+rule Let _ _ $v $b `↪` $b $v;
 	\end{lstlisting}

### 2023-01-30 paper-thesis/chapters/40-discussing-statix.tex
Cleanup work
Ok this is already a thing

@@
 \chapter{Ergonomics of Spoofax}
+\label{ch:ergonomics}
 
@@
 This was the only case where this feature would've been useful for this thesis, but it may be more useful for other work.
-
-	\item A better way to control the errors generated by Statix, such that users of the language get good errors.
-Statix does support \texttt{expr | error "Error text here"} annotations, but these don't allow constraints that instantiate metavariables on the left hand side whereas this would be needed for good error messages for dependently typed languages, since \verb|expectBetaEq| may infer variables.
-\todo{This is not true}
 \end{enumerate}

### 2023-01-31 paper-thesis/chapters/99-conclusion.tex
Clean up citations + other work

@@
 
-We have demonstrated that the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph.
-We have also presented a few extensions to the Calculus of Constructions and discussed how they could be implemented.
+This thesis presents an implementation of a dependently typed language in Statix. Our aim was to anther the following research questions:
+
+\paragraph{Can the Calculus of Constructions be implemented in Statix? (Chapter \ref{chap:baselang})}
+
+\paragraph{We show the implementation is easily extendable (Chapter \ref{chap:bools})}
+
+\paragraph{We show how to add inference to the implementation (Chapter \ref{ch:inference})}
+
+\paragraph{We show how to add support for inductive datatypes to the implementation (Chapter \ref{ch:datatypes})}
+
+\paragraph{We show how to add support for universes to the implementation (Chapter \ref{ch:universes})}
+
+\section{Future Work}

### 2023-01-31 paper-thesis/chapters/99-conclusion.tex
Work on conclusion

--- 2022-11-07 paper-thesis/chapters/99-conclusion.tex
@@
 
-We have demonstrated that the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph.
-We have also presented a few extensions to the Calculus of Constructions and discussed how they could be implemented.
+This thesis presents an implementation of a dependently typed language in Statix. Our aim was to anther the following research questions:
+
+\paragraph{RQ1: Can the Calculus of Constructions be implemented in Statix?}
+As we demonstrated in chapter \ref{chap:baselang}, the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph.
+We defined the beta-reduction, beta-equality and type checking rules and then converted them to Statix code.
+Beta reduction was defined using a Krivine machine.
+In chapter \ref{chap:namecolls} we solved the variable capture problem which the naive implementation from chapter \ref{chap:baselang} suffered from, by using scoped names.
+
+\paragraph{RQ2: Is the implementation is easily extendable?}
+We showed in chapter \ref{chap:bools} that the implementation is easily extendable, by extending it with booleans, postulate, and type assertions.
+We define a four step process that can be used to extend the language, which we also use to answer the remaining research questions.
+
+\paragraph{RQ3: Can we add inference to the implementation?}
+In chapter \ref{ch:inference} we discuss how to add inference to the implementation.
+If we want to keep the implementation clean and concise, we need to compromise on how powerful the inference algorithm is, by defining an approximated version of first-order inference.
+We implemented this algorithm in a concise way.
+
+\paragraph{RQ4: Can we add support for inductive data types to the implementation?}
+Chapter \ref{ch:datatypes} adds support for inductive data types with parameters and indices.
+We give steps for type-checking data type declarations.
+Additionally, we give each data type an eliminator and show how to type-check and beta reduce eliminators.
+Finally, we show that positivity checking can be implemented concisely.
+
+\paragraph{RQ5: Can we add support for universes to the implementation?}
+In chapter \ref{ch:universes} we showed that we can add universes to the language. This can be done easily and concisely.
+
+\section{Future Work}
+While the language as implemented currently is fully usable, there are still some open questions.

### 2023-01-31 paper-thesis/chapters/31-comparing-with-lambdapi.tex
Fix the arrrows

@@
 	\begin{lstlisting}
-rule TmType Type `↪` TmSort;
-rule TmType (FnType $A $B) `↪` `Π` (x : TmType $A), TmType ($B x);
-rule FnConstruct _ _ $f `↪` $f;
-rule FnDestruct _ _ $f $a `↪` $f $a;
+rule TmType Type `$\hookrightarrow$` TmSort;
+rule TmType (FnType $A $B) `$\hookrightarrow$` `Π` (x : TmType $A), TmType ($B x);
+rule FnConstruct _ _ $f `$\hookrightarrow$` $f;
+rule FnDestruct _ _ $f $a `$\hookrightarrow$` $f $a;
 	\end{lstlisting}
@@
 
-	\todo{invisible chars}
 \begin{figure}[ht]
@@
 	TmType (B v);
-rule Let _ _ $v $b `↪` $b $v;
+rule Let _ _ $v $b `$\hookrightarrow$` $b $v;
 	\end{lstlisting}

### 2023-02-02 paper-thesis/chapters/98-relatedwork.tex
Add the first two sections to related work

@@
 \chapter{Related Work}
+\label{ch:relatedwork}
+
+\paragraph{Scopes as Types}
+
 The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
-Van Antwerpen et al.
-\cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types.
-It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.
+Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types.
+It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can be arbitrary terms.
 
-Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language.
-The pi-forall language\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper.
+\paragraph{Bidirectional Typing}
+
+In chapter \ref{ch:comp-haskell} we compared the implementation in Statix with an implementation in Haskell.
+The pi-forall language~\cite{pi_forall} is a language with a similar complexity to the language presented in this paper.
 In principle, the implementations are very similar.
-For example, the inference rules presented in \cite{pi_forall} are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper.
-The primary difference is that they use a bidirectional type system, whereas this paper does not.
+For example, the inference rules of pi-forall are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper.
+The primary difference is that they use a bidirectional type system~\cite{bidirectional}, whereas this paper uses Statix' unification.
+
+\paragraph{}

### 2023-02-02 lody/src/main/java/mb/helloworld/task/HelloWorldShowReplaceWorlds.java
Follow example

@@ -1,0 +1,73 @@
+package mb.helloworld.task;
+
+import java.io.Serializable;
+import java.util.Objects;
+
+import javax.inject.Inject;
+
+import org.checkerframework.checker.nullness.qual.Nullable;
+
+import mb.lody.LodyClassLoaderResources;
+import mb.lody.LodyScope;
+import mb.lody.task.LodyParse;
+import mb.pie.api.ExecContext;
+import mb.pie.api.TaskDef;
+import mb.pie.api.stamp.resource.ResourceStampers;
+import mb.resource.ResourceKey;
+import mb.spoofax.core.language.command.CommandFeedback;
+import mb.spoofax.core.language.command.ShowFeedback;
+import mb.aterm.common.TermToString;
+
+@LodyScope
+public class HelloWorldShowReplaceWorlds implements TaskDef<HelloWorldShowReplaceWorlds.Args, CommandFeedback> {
+    public static class Args implements Serializable {
+        private static final long serialVersionUID = 1L;
+
+        public final ResourceKey file;
+
+        public Args(ResourceKey file) {
+            this.file = file;
+        }
+
+        @Override
+        public boolean equals(@Nullable Object o) {
+            if (this == o) return true;
+            if (o == null || getClass() != o.getClass()) return false;
+            final Args args = (Args) o;
+            return file.equals(args.file);
+        }
+
+        @Override
+        public int hashCode() {
+            return Objects.hash(file);
+        }
+
+        @Override
+        public String toString() {
+            return "Args{" + "file=" + file + '}';
+        }
+    }
+
+
+    private final LodyClassLoaderResources classloaderResources;
+    private final LodyParse parse;
+
+    @Inject
+    public HelloWorldShowReplaceWorlds(LodyClassLoaderResources classloaderResources, LodyParse parse) {
+        this.classloaderResources = classloaderResources;
+        this.parse = parse;
+    }
+
+
+    @Override
+    public CommandFeedback exec(ExecContext context, Args args) throws Exception {
+        context.require(classloaderResources.tryGetAsNativeResource(getClass()), ResourceStampers.hashFile());
+        final ResourceKey file = args.file;
+        return CommandFeedback.of(ShowFeedback.showText("A", "B"));
+    }
+
+    @Override
+    public String getId() {
+        return getClass().getName();
+    }
+}

### 2023-02-02 lody/src/main/java/mb/lody/task/Compile.java
Change the names

@@ -1,4 +1,4 @@
-package mb.helloworld.task;
+package mb.lody.task;
 
 import java.io.Serializable;
 import java.util.Objects;
@@ -19,7 +19,7 @@ import mb.spoofax.core.language.command.ShowFeedback;
 import mb.aterm.common.TermToString;
 
 @LodyScope
-public class HelloWorldShowReplaceWorlds implements TaskDef<HelloWorldShowReplaceWorlds.Args, CommandFeedback> {
+public class Compile implements TaskDef<Compile.Args, CommandFeedback> {
     public static class Args implements Serializable {
         private static final long serialVersionUID = 1L;
 
@@ -53,7 +53,7 @@ public class HelloWorldShowReplaceWorlds implements TaskDef<HelloWorldShowReplac
     private final LodyParse parse;
 
     @Inject
-    public HelloWorldShowReplaceWorlds(LodyClassLoaderResources classloaderResources, LodyParse parse) {
+    public Compile(LodyClassLoaderResources classloaderResources, LodyParse parse) {
         this.classloaderResources = classloaderResources;
         this.parse = parse;
     }

### 2023-02-02 lody/src/main/java/mb/lody/task/Compile.java
Add transform step

@@ -51,11 +51,13 @@ public class Compile implements TaskDef<Compile.Args, CommandFeedback> {
 
     private final LodyClassLoaderResources classloaderResources;
     private final LodyParse parse;
+    private final CompileTransform transformer;
 
     @Inject
-    public Compile(LodyClassLoaderResources classloaderResources, LodyParse parse) {
+    public Compile(LodyClassLoaderResources classloaderResources, LodyParse parse, CompileTransform transformer) {
         this.classloaderResources = classloaderResources;
         this.parse = parse;
+        this.transformer = transformer;
     }
 
 
@@ -63,7 +65,10 @@ public class Compile implements TaskDef<Compile.Args, CommandFeedback> {
     public CommandFeedback exec(ExecContext context, Args args) throws Exception {
         context.require(classloaderResources.tryGetAsNativeResource(getClass()), ResourceStampers.hashFile());
         final ResourceKey file = args.file;
-        return CommandFeedback.of(ShowFeedback.showText("A", "B"));
+        return context.require(transformer, parse.inputBuilder().withFile(file).buildAstSupplier()).mapOrElse(
+            ast -> CommandFeedback.of(ShowFeedback.showText(TermToString.toString(ast), "Title")),
+            e -> CommandFeedback.ofTryExtractMessagesFrom(e, file)
+        );
     }
 
     @Override

### 2023-02-02 lody/src/main/java/mb/lody/task/CompileTransform.java

@@ -1,0 +1,32 @@
+package mb.lody.task;
+
+import mb.lody.LodyClassLoaderResources;
+import mb.lody.LodyScope;
+import mb.pie.api.ExecContext;
+import mb.pie.api.stamp.resource.ResourceStampers;
+import mb.stratego.pie.AstStrategoTransformTaskDef;
+
+import javax.inject.Inject;
+import java.io.IOException;
+
+@LodyScope
+public class CompileTransform extends AstStrategoTransformTaskDef {
+  private final LodyClassLoaderResources classloaderResources;
+
+  @Inject
+  public CompileTransform( // 1
+    LodyClassLoaderResources classloaderResources,
+    LodyGetStrategoRuntimeProvider getStrategoRuntimeProvider
+  ) {
+    super(getStrategoRuntimeProvider, "compile-program"); // 2
+    this.classloaderResources = classloaderResources;
+  }
+
+  @Override public String getId() { // 3
+    return getClass().getName();
+  }
+
+  @Override protected void createDependencies(ExecContext context) throws IOException { // 4
+    context.require(classloaderResources.tryGetAsNativeResource(getClass()), ResourceStampers.hashFile());
+  }
+}

### 2023-02-03 paper-thesis/chapters/000-abstract.tex
Improve the introduction

--- 2022-11-14 paper-thesis/chapters/000-abstract.tex
@@
-Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone.
+\noindent Static type systems can greatly enhance the quality of programs, but implementing a type checker is challenging and error-prone.
 The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system.
-However, so far Statix has not been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking.
-In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
+However, so far Statix has not been used to implement a type system with dependent types, an expressive class of type systems which require evaluation of terms during type checking.
+
+In this thesis, we present an specification of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
 While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.

### 2023-02-03 paper-thesis/chapters/01-introduction.tex
Improve the introduction

@@
 
-Spoofax is a language workbench: a collection of tools that enable the development of textual languages~\cite{spoofax}.
-When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics.
+Type checking programming languages is commonly done by writing a type checker in a general purpose language.
+This requires a lot of effort, and it is easy to make mistakes.
+In this thesis we will specifically focus on dependently typed languages, which differ from other languages because they allow types to be parameterized by values.
+This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree.
+This expressiveness also makes dependent type systems more complicated to type check, since deciding equality of types requires normalization of the terms they are parameterized by.
+
+Instead of writing the type checker in a general purpose language, in this thesis we will systematically derive a type checker from a high-level, declarative specification.
+This allows for a cleaner implementation that is easy to extend and maintain.
+
+We will do this using the Spoofax language workbench, which is a collection of tools that can derive a parser and type checker from a high level specification of the language~\cite{spoofax}.
+When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics~\cite{scopes_as_types}.
 It is a declarative language that uses inference rules to define static semantics.
-From these rules a type checker and other editor tools can be derived.
 Statix aims to cover a broad range of languages and type systems.
-However, no attempts have been made to express dependently typed languages in Statix until now.
-
-Dependently typed languages are different from other languages, because they allow types to be parameterized by values.
-This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree.
-This expressiveness also makes dependent type systems more complicated to implement.
-Especially, deciding equality of types requires normalization of the terms they are parameterized by.
+However, no attempts have been made yet to express a dependently typed language in Statix until now.
 
@@
 This goal of this thesis is to investigate how well Statix is fit for the task of defining a simple dependently-typed language.
-We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix.
+We want to investigate whether typical features of dependently typed languages can be encoded in Statix.
 The goal is not only to show that Statix can implement it, but also that the implementation is concise.
@@
 \begin{itemize}
-	\item We present a implementation of the Calculus of Constructions in Statix (Chapter \ref{chap:baselang})
-	\item We show the implementation is easily extendable (Chapter \ref{chap:bools})
-	\item We show how to add inference to the implementation (Chapter \ref{ch:inference})
-	\item We show how to add support for inductive datatypes to the implementation (Chapter \ref{ch:datatypes})
-	\item We show how to add support for universes to the implementation (Chapter \ref{ch:universes})
+	\item We present a implementation of the Calculus of Constructions, a lambda calculus with dependent types, in Statix (Chapter~\ref{chap:baselang})
+	\item We show the language is easily extendable, by extending it with booleans among other features (Chapter~\ref{chap:bools})
+	\item We show how to add inference of implicit arguments to the implementation (Chapter~\ref{ch:inference})
+	\item We show how to add support for inductive datatypes to the implementation (Chapter~\ref{ch:datatypes})
+	\item We show how to add support for universes to the implementation (Chapter~\ref{ch:universes})
 \end{itemize}
@@
 	\item We discuss how well semantic code completion works for our language (Chapter \ref{chap:editor-services})
-	\item We compare our implementation with an implementation of the same language in Haskell (Chapter \ref{ch:comp-haskell})
-	\item We compare our implementation with an implementation of the same language in LambdaPi (Chapter \ref{ch:comp-lambdapi})
+	\item We compare our implementation with an implementation of the same language in Haskell (Chapter \ref{ch:comp-haskell}) and LambdaPi (Chapter \ref{ch:comp-lambdapi})
 	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (Chapter \ref{ch:ergonomics})
-	\item We discuss related work (Chapter \ref{ch:relatedwork}) and future work (Chapter \ref{ch:futurework})
+	\item We discuss related work (Chapter \ref{ch:relatedwork}) and future work (Chapter \ref{ch:conclusion})
+\end{itemize}
 
-\end{itemize}
+Before explaining these contributions, we provide background information on Spoofax and Statix (Chapter \ref{chap:bg-dp}) and the Calculus of Constructions (Chapter \ref{chap:bg-spoofax}).

### 2023-02-03 paper-thesis/chapters/01-introduction.tex
Start reworking spoofax chapter

@@
 
-Before explaining these contributions, we provide background information on Spoofax and Statix (Chapter \ref{chap:bg-dp}) and the Calculus of Constructions (Chapter \ref{chap:bg-spoofax}).
+\noindent Before explaining these contributions, we provide background information on Spoofax and Statix (Chapter \ref{chap:bg-dp}) and the Calculus of Constructions (Chapter \ref{chap:bg-spoofax}).
+
+The source code for this thesis can be found online: \url{https://github.com/JonathanBrouwer/master-thesis}

### 2023-02-03 paper-thesis/chapters/02-background-spoofax.tex
A substitution based approach
Start reworking spoofax chapter

@@
 
-This chapter will explain the concepts behind the Spoofax Language Workbench. If you are already familiar with Spoofax, you can skip this chapter.
-
-\section{Spoofax}
+This chapter will explain the concepts behind the Spoofax Language Workbench. Readers already familiar with Spoofax are recommended to only skim through this chapter.
 
@@
 general-purpose programming languages.
-Each aspect of the language is defined in one of three metalanguages, each with their own purpose.
-From these languages Spoofax generates a parser, type checker and other useful tools.
+Each aspect of the language is defined in one the meta-languages, each with their own purpose.
+From these specifications Spoofax generates a parser, type checker and other useful tools.
+The meta-languages that are used in this thesis are discussed in the sections below:
 
-The metalanguage SDF3 is used to define the syntax of the language~\cite{sdf3}, which is then parsed using Spoofax's SGLR parser~\cite{sdf3_parser}.
+\section{SDF3}
+
+The meta-language SDF3 is used to define the syntax of the language~\cite{sdf3}, which is then parsed using Spoofax's SGLR parser~\cite{sdf3_parser}.
 The parser produces a abstract syntax tree in the ATerm format~\cite{aterm}.
@@
 
-Next, the metalanguage Statix~\cite{scopes_as_types} is used to define the static semantics of the language.
-It uses scope graphs to represent binding and typing information of the language.
-How this works is discussed in section \ref{sec:statix}.
+\paragraph{Syntax definitions}
+The syntax of
 
-Finally, Stratego~\cite{stratego} is used to define program transformations, using rewrite rules.
-This can be used for modifications to the AST, but also to build transformations to a different language (such as compiling to assembly).
+\paragraph{ATerm format}
 
-\section{\label{sec:statix}Statix and Scope graphs}
+\section[Statix and Scope graphs]{\label{sec:statix}Statix and Scope graphs\protect\footnote{Note: This section is adapted from "Composable Type System Specification using Heterogeneous Scope Graphs".~\cite[sect.
+4.1.2]{Zwaan21}}}
 
-\textit{Note: This section is adapted from the Explanation in "Composable Type System Specification
-	using Heterogeneous Scope Graphs".~\cite[sect. 4.1.2]{Zwaan21}}
-
-In Statix, scope graphs are used to represent a programs name and type information~\cite{scopes_as_types}. A scope graph is a graph that consists of the following components:
+In Statix, scope graphs are used to represent a programs name and type information~\cite{scopes_as_types}~\cite{nameres}. A scope graph is a graph that consists of the following components:
 
@@
 Its default value is \verb|true|, meaning that all datums under the specified relation that satisfy the path well-formedness condition are returned.
-	\item Result comparison parameters:
-	\begin{itemize}
-		\item A strict partial order on paths, described by less-than relations on labels. This relation defines a prefix-order on paths.
-		\item A `shadow' predicate, which takes two datums as input. Its default value is \verb|false|, meaning that no shadowing is performed.
 \end{itemize}
-	When the result set contains multiple datums, and it holds for a datum $d$ that there exists another datum $d'$ which path is strictly smaller that $d$, and the shadow predicate holds, then $d$ is removed from the result set.
-\end{itemize}

### 2023-02-07 paper-thesis/chapters/02-background-spoofax.tex
Rewrite spoofax background

@@
 The meta-language SDF3 is used to define the syntax of the language~\cite{sdf3}, which is then parsed using Spoofax's SGLR parser~\cite{sdf3_parser}.
-The parser produces a abstract syntax tree in the ATerm format~\cite{aterm}.
-A pretty printer is also generated based on this syntax definition.
 
 \paragraph{Syntax definitions}
-The syntax of
+The syntax of a language is defined through rules of the form \verb|A.C = a| with \verb|A| being a \emph{Non-Terminal}, \verb|C| being the name of a constructor and \verb|a| a list of symbols.
+To remove ambiguity, one can specify a relative priority for each constructor.
+For example, below is the syntax of a simple language:
+\begin{lstlisting}
+context-free sorts
+	Expr
+context-free syntax
+	Expr.Add = Expr "+" Expr
+	Expr.Mul = Expr "*" Expr
+	Expr.X = "x"
+context-free priorities
+	Expr.X > Expr.Mul > Expr.Add
+\end{lstlisting}
 
 \paragraph{ATerm format}
+Using the Syntax, Spoofax will generate a parser.
+This parser takes in a program in textual format, and outputs ATerms.
+ATerms are a way of encoding an abstract syntax tree~\cite{aterm}.
+For example, for the program \verb|x + x * x| the following ATerm is produced:
+\begin{lstlisting}
+Add(X(), Mul(X(), X()))
+\end{lstlisting}
 
-\section[Statix and Scope graphs]{\label{sec:statix}Statix and Scope graphs\protect\footnote{Note: This section is adapted from "Composable Type System Specification using Heterogeneous Scope Graphs".~\cite[sect.
+\section{Statix}
+Now that the syntax has been determined, we move on to the static semantics.
+Statix is a declarative language for describing the static semantics of a language~\cite{scopes_as_types}.
+Using these rules, Statix automatically derives a type checker.
+The rules are based on \emph{equality constraints}.
+The basic constraints
+are equality constraints \verb|t1 == t2| and inequality constraints \verb|t1 != t2| on terms.
+An equality constraint is states that two terms are equal or can be unified.
+An inequality constraint states that two terms are not equal.
+
+The constraints are generated by predicates.
+A predicate defines a relation on terms and scopes, it is of the form $\mathtt{p(t_1, ..., t_n) :- c_1, ..., c_n}$.
+Each constraint $c_1...c_n$ is another predicate or an equality constraint.
+Below is an example that uses these mechanics to check that an expression does not contain any \verb|Muls|.
+Note that \verb|noMuls| is not defined for the \verb|Mul| constructor, so it will fail if it encounters one.
+\begin{lstlisting}
+noMuls(Add(e1, e2)) :- noMuls(e1), noMuls(e2).
+noMuls(X()).
+\end{lstlisting}
+
+Statix also has functional predicates, these define a one-to-one relationship between input terms and the output term. For example below we use this to associate a type with each expression.
+\begin{lstlisting}
+typeOfExpr(X()) = INT().
+typeOfExpr(Add(e1, e2)) = INT() :-
+	typeOfExpr(e1) == INT(),
+	typeOfExpr(e2) == INT().
+\end{lstlisting}
+
+In Statix, the language designer writes down the declarative semantics. The Statix solver decides the order in which constraints are considered, the language designer has no control over that.
+
+\section[Scope graphs]{\label{sec:statix}Scope graphs\protect\footnote{Note: This section is adapted from "Composable Type System Specification using Heterogeneous Scope Graphs".~\cite[sect.
 4.1.2]{Zwaan21}}}
 
-In Statix, scope graphs are used to represent a programs name and type information~\cite{scopes_as_types}~\cite{nameres}. A scope graph is a graph that consists of the following components:
+Statix also has scope graphs.
+Scope graphs are used to represent a programs' name and type information~\cite{scopes_as_types}~\cite{nameres}.
+A scope graph is a graph that consists of the following components:
 

### 2023-02-07 paper-thesis/chapters/03-background-dependent-types.tex
Start reworking spoofax chapter
Improve chapter 3

@@
 
-This chapter will explain the concepts behind dependent types. If you are already familiar with Dependent Types, you can skip this chapter.
+This chapter will explain the concepts behind dependent types. Readers already familiar with Dependent Types are recommended to only skim through this chapter.
 
-\section{What are dependent types?}
+\section{What are Dependent Types?}
 
@@
 
-The biggest advantage of dependent types is that they increase the expressiveness of a type system a lot, for example it is now possible to express a function that adds two vectors of the same length and returns a new vector of that length.
+The biggest advantage of dependent types is that they increase the expressiveness of a type system a lot, for example:
 \begin{lstlisting}
-add_vec : (n : Nat) -> Vec n -> Vec n -> Vec n
+append : (A: Set) -> (n : Nat) -> Vec A n -> Vec A n -> Vec A (n + n)
 \end{lstlisting}
+A \verb|Vec A n| is a list with elements of type \verb|A| that is exactly length \verb|n| (an integer, which is a value!).
+This append function appends two \verb|Vecs| of equal length, returning a \verb|Vec| of length \verb|n + n|.
+This is an example of a type-level computation.
 
-\section{The Curry-Howard Isomorphism}
+Dependent types are useful because dependent type systems allow for the Curry-Howard Correspondence~\cite{chc}.
+Using this correspondence, types correspond to propositions and a term of a type correspond to a proof of a proposition.
+Thus if our type system is sound\footnote{The type system we will define in chapter \ref{chap:baselang} is not sound, we will fix this in chapter \ref{ch:universes}.}, that is, we cannot create a term of an empty type, then we can use our type systems to prove things.
 
-The Curry-Howard correspondence is a relationship that can be used to interpret typed programs as mathematical proofs~\cite{chc}.
-This is done by representing false statements as empty types, and true statements as non-empty (inhabited) types.
-
-For example, it is possible to define a type \verb|IsEven n|, that represent the statement "\verb|n| is even", and declare it in such a way that it is only inhabited if \verb|n| is even.
-
-In order to use a type system to prove things it is important that the type system is sound, that is, it is impossible to prove a false statement.
-Using the Curry-Howard correspondence, that means it should be impossible to obtain an element of the empty type.
-We will explore this further in chapter \ref{ch:universes}.
-
-\section{The Lambda Cube}
+\section{The Calculus of Constructions}
 
 The lambda cube is a framework to categorize programming languages based on whether terms and types can depend on each other~\cite{lambda_cube}.
-Specifically, it categorizes languages on three axis: \footnote{Note that the "Terms can depend on terms" axis is missing, this is because a language where terms cannot depend on terms cannot compute anything and is thus pretty useless.}
+Specifically, it categorizes languages on three axes: \footnote{Note that the "Terms can depend on terms" axis is missing, this is because a language where terms cannot depend on terms cannot compute anything and is thus pretty useless.}
 
@@
 
-The \textit{Calculus of Constructions} is the language that has all three of these features, and it is what we will implement in this paper.
-If you want to learn more, I recommend playing around with the \textit{\href{https://agda.readthedocs.io/en/latest/}{Agda}} language, which is a real-world language that satisfies all three axis.
+The \textit{Calculus of Constructions} is the language that has all three of these features, it will be implemented in chapter \ref{chap:baselang}.

### 2023-02-08 paper-thesis/chapters/02-background-spoofax.tex
Work on CoC chapter

@@
 
-\section[Scope graphs]{\label{sec:statix}Scope graphs\protect\footnote{Note: This section is adapted from "Composable Type System Specification using Heterogeneous Scope Graphs".~\cite[sect.
+\section[Scope graphs]{\label{sec:statix}Scope Graphs\protect\footnote{Note: This section is adapted from "Composable Type System Specification using Heterogeneous Scope Graphs".~\cite[sect.
 4.1.2]{Zwaan21}}}

### 2023-02-08 paper-thesis/chapters/10-coc.tex
Improve the variable capture chapter
Improve booleans chapter
Work on CoC chapter

@@
-\chapter{Calculus of Constructions}
+\chapter{Calculus of Constructions in Statix}
 \label{chap:baselang}
@@
 Section \ref{sec:coc-dynsyms} describes the dynamic semantics of the language, and finally \ref{sec:coc-typecheck} how to type check the language.
+This chapter is the main contribution of this thesis.
 
-\section{The language}
+\section{The Language}
 \label{sec:coc-syntax}
@@
 \begin{lstlisting}
-Expr.Let = "let" ID "=" Expr ";" Expr
 Expr.Type = "Type"
@@
 Expr.FnDestruct = Expr Expr {left}
+Expr.Let = "let" ID "=" Expr ";" Expr
 \end{lstlisting}
@@
 
-Please observe that the syntax definition does not have a separate sort for types, as types may be arbitrary expressions in a dependently typed language.
-Furthermore, \verb|FnType| assigns a name to its first argument to allow the return type of the function to depend on the argument type.
+There is only one sort: Expr. The syntax definition does not have a separate sort for types, as types may be arbitrary expressions in a dependently typed language. The following constructors exist:
+\begin{itemize}
+	\item \verb|Type| is the Type of Types.
+	\item \verb|Var| is a variable, it uses the lexical sort \verb|ID|, which is defined as \verb|[a-zA-Z\_][a-zA-Z0-9\_]*|.
+	\item \verb|FnType| is the type of a function.
+It assigns a name to its first argument to allow the return type of the function to depend on the argument type.
+It is right associative, meaning \verb|A -> B -> C| is interpreted as \verb|A -> (B -> C)|.
+	\item \verb|FnConstruct| creates an anonymous function (lambda function).
+	\item \verb|FnDestruct| applies a function to an argument. It is left associative, meaning \verb|a b c| is interpreted as \verb|(a b) c|.
+	\item \verb|Let| is a let binding. It introduces a substitutable variable.
+\end{itemize}
 
@@
 
-To type check the base language, we need to store information about the names that are in scope at each point in the program.
-There are two different cases, names that do not have a known value (only a type), which are names created by function arguments and dependent function types, and names that do have a known value, which are names created by let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need \emph{the value} of a binding to compare types, this is needed in dependently typed languages.}
+To type check the base language, we need an environment to store information about the names that are in scope at each point in the program.
+There are two different types of names that we may want to store, names that do not have a known value (only a type), which are names created by function arguments and dependent function types, and names that do have a known value, which are names created by let bindings.\footnote{In non-dependent languages there is no such distinction, but because we may need \emph{the value} of a binding to compare types, this is needed in dependently typed languages.}
 
@@
 The \verb|sPutType| and \verb|sPutSubst| predicates generate a new scope given a parent scope and a type or a substitution respectively.
-To query the scope graph we use \verb|sGetName|, which will return the closest \verb|NameEntry| with a matching same.
+These return a scope that represent an environment that has been extended with the new name.
+To query the scope graph we use \verb|sGetName|, which will return the closest \verb|NameEntry| with a matching name.
 Finally, \verb|sEmpty| returns a fresh empty scope.
@@
 
-\section{Beta equality}
+\section{Beta Equality}
 \label{sec:coc-betaeq}
@@
 
-\section{Type checking the Calculus of Constructions}
+\section{Type Checking}
 \label{sec:coc-typecheck}
@@
 
-\section{Discussion of a substitution-based approach}
+\section{Discussion of a Substitution-Based Approach}
 \label{sec:coc-subst}
@@
 			{\scope{s_{f'}}{f'}}
-			\\ \nexists x e_1 e_2. f' = \co{FnConstruct}(x, e_1, e_2)
+			\\ \nexists x \, e_1 \, e_2. \, \, f' = \co{FnConstruct}(x, e_1, e_2)
 		} {

### 2023-02-08 paper-thesis/chapters/23-adding-universes.tex
Work from yesterday
Minor writing changes
Work on CoC chapter

@@
 Our language so far has the property that the type of \verb|Type| is \verb|Type|, this is called \emph{type in type}.
-This allows for \emph{Girard's paradox}, which means that the logic created by this language is inconsistent, we can prove false.
+This allows for \emph{Girard's paradox}, which means that the logic created by this language is inconsistent: we can prove false.
 
@@
 
-\section{Implementing universes in the CoC}
+\section{Implementing Universes in the CoC}
 
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 	\figuresection[\fbox{$\toe{s}{e}{t}$}]{Type checking rules with universes}
@@
 \newpage
-\section{Implementing universes with inductive data types}
+\section{Implementing Universes with Inductive Data Types}
 

### 2023-02-08 paper-thesis/chapters/10-coc.tex
More writing on coc

@@
 
-An example program is the following, which defines a polymorphic identity function and applies it to a function:
+An example program is the following, which defines a polymorphic identity function (an identity function that is generic over its type) and applies it to a function:
 
@@
 \end{lstlisting}
+
+The type that the function is generic over needs to be explicitly specified.
+In most languages, generics are inferred, this inference will also be possible in this language after implementing inference in chapter~\ref{ch:inference}.
 
@@
 We also only have a single relation, called \verb|name|.
-This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.
+The relation associates a \verb|NameEntry| with each name in the scope graph.
+The \verb|NameEntry| can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.
 The Statix definition of these concepts is given below:
@@
 constructors
-    NameType : Expr -> NameEntry
-    NameSubst : scope * Expr -> NameEntry
+    NType : Expr -> NameEntry
+    NSubst : scope * Expr -> NameEntry
 relations
@@
 
-We implemented beta reduction using a Krivine abstract machine~\cite{krivine}.
+We implemented beta-head reduction using a Krivine abstract machine~\cite{krivine}.
 The machine can head evaluate lambda expressions with a call-by-name semantics.
+This is a strategy under which the leftmost, outermost term is always reduced first~\cite{tapl}.
 It works by keeping a stack of all arguments that have not been applied yet.
@@
 
-In conventional dependently typed languages, evaluation is often done using De Bruijn indices.
+In conventional dependently typed languages, evaluation is often done using De Bruijn indices, which use numbers to identify variables.
 However, we chose to use names rather than De Bruijn indices, because scope graphs work based on names.
-Using De Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations (which are Spoofax annotations that declare one name as being a use of another name that is a definition), such as renaming.
+Using De Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations (which are Spoofax annotations that declare one name as being a use of another name that is a definition).
 
@@
 We use the syntax $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), ps) == (s2, e2)|.
-The \verb|p| in this definition is the argument stack of the Krivine machine.
+The $\overline{p}$ in this definition is the argument stack of the Krivine machine.
+The argument stack is a stack of scoped expressions, which are the arguments that are not yet paired with a matching function.
 Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language.
@@
 
-The inference rules in figure \ref{fig:type-check-rules} can be directly translated to Statix rules. For example, the rule for \verb|Let| bindings is expressed like this in Statix:
+The inference rules in figures \ref{fig:beta-head-reduce-rules}, \ref{fig:beta-eq-rules}, and \ref{fig:type-check-rules} can be directly translated to Statix rules.
+For example, the rule for \verb|Let| bindings in figure \ref{fig:type-check-rules} is expressed like this in Statix:
 \begin{lstlisting}
@@
 Although these rules look cleaner, they are more complicated to implement, requiring a separate relation to check if \verb|f| is a \verb|FnConstruct| or something else.
-For the pure calculus of constructions, this still works quite well but when adding inductive datatypes (chapter \ref{ch:datatypes}) this quickly becomes a mess.
+For the pure calculus of constructions this still works quite well, but when adding inductive datatypes (chapter \ref{ch:datatypes}) the rules required become a lot more complex than those for a Krivine machine.
 

### 2023-02-08 paper-thesis/chapters/01-introduction.tex
Fix chapter references

@@
 \begin{itemize}
-	\item We discuss how well semantic code completion works for our language (Chapter \ref{chap:editor-services})
-	\item We compare our implementation with an implementation of the same language in Haskell (Chapter \ref{ch:comp-haskell}) and LambdaPi (Chapter \ref{ch:comp-lambdapi})
-	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (Chapter \ref{ch:ergonomics})
-	\item We discuss related work (Chapter \ref{ch:relatedwork}) and future work (Chapter \ref{ch:conclusion})
+	\item We discuss how well semantic code completion works for our language (chapter~\ref{chap:editor-services})
+	\item We compare our implementation with an implementation of the same language in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi})
+	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (chapter~\ref{ch:ergonomics})
+	\item We discuss related work (chapter~\ref{ch:relatedwork}) and future work (chapter~\ref{ch:conclusion})
 \end{itemize}
 
-\noindent Before explaining these contributions, we provide background information on Spoofax and Statix (Chapter \ref{chap:bg-dp}) and the Calculus of Constructions (Chapter \ref{chap:bg-spoofax}).
+\noindent Before explaining these contributions, we provide background information on Spoofax and Statix (chapter~\ref{chap:bg-dp}) and the Calculus of Constructions (chapter~\ref{chap:bg-spoofax}).
 

### 2023-02-08 paper-thesis/chapters/03-background-dependent-types.tex
Fix chapter references

@@
 Using this correspondence, types correspond to propositions and a term of a type correspond to a proof of a proposition.
-Thus if our type system is sound\footnote{The type system we will define in chapter \ref{chap:baselang} is not sound, we will fix this in chapter \ref{ch:universes}.}, that is, we cannot create a term of an empty type, then we can use our type systems to prove things.
+Thus if our type system is sound\footnote{The type system we will define in chapter~\ref{chap:baselang} is not sound, we will fix this in chapter~\ref{ch:universes}.}, that is, we cannot create a term of an empty type, then we can use our type systems to prove things.
 
@@
 
-The \textit{Calculus of Constructions} is the language that has all three of these features, it will be implemented in chapter \ref{chap:baselang}.
+The \textit{Calculus of Constructions} is the language that has all three of these features, it will be implemented in chapter~\ref{chap:baselang}.

### 2023-02-08 paper-thesis/chapters/10-coc.tex
Fix chapter references

@@
 
-In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}, as explained in chapter \ref{chap:bg-spoofax}.
+In Statix, all this information can be stored in a \emph{scope graph}~\cite{scope_graphs}, as explained in chapter~\ref{chap:bg-spoofax}.
 We only use a single type of edge, called \verb|P| (parent) edges.
@@
 Although these rules look cleaner, they are more complicated to implement, requiring a separate relation to check if \verb|f| is a \verb|FnConstruct| or something else.
-For the pure calculus of constructions this still works quite well, but when adding inductive datatypes (chapter \ref{ch:datatypes}) the rules required become a lot more complex than those for a Krivine machine.
+For the pure calculus of constructions this still works quite well, but when adding inductive datatypes (chapter~\ref{ch:datatypes}) the rules required become a lot more complex than those for a Krivine machine.
 

### 2023-02-08 paper-thesis/chapters/20-adding-booleans.tex
Improve booleans chapter
Work on CoC chapter
Fix chapter references

@@
 One for if \verb|c| evaluates to true, one for if \verb|c| evaluates to false, and finally a third case for other cases (such as when \verb|c| is a variable that does not have a substitution).
-These rules are stated below: (Remember that \verb|rebuild| is the rule introduced in chapter \ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s)
+These rules are stated below: (Remember that \verb|rebuild| is the rule introduced in chapter~\ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s)
 
 \begin{lstlisting}
-betaReduceHead((s, BoolIf(c, b1, b2)), t) = 
-  betaReduceHeadIf(s, betaReduceHead((s, c), []), c, b1, b2, t).
-betaReduceHeadIf(s, (_, BoolTrue()), _, b1, _, t) = 
-  betaReduceHead((s, b1), t).
-betaReduceHeadIf(s, (_, BoolFalse()), _, _, b2, t) = 
-  betaReduceHead((s, b2), t).
-betaReduceHeadIf(s, _, c, b1, b2, t) = 
-  rebuild((s, BoolIf(c, b1, b2)), t).
+betaReduceHead((s, BoolIf(c, b1, b2)), ps) = 
+  betaReduceHeadIf(s, betaReduceHead((s, c), []), c, b1, b2, ps).
+betaReduceHeadIf(s, (_, BoolTrue()), _, b1, _, ps) = 
+  betaReduceHead((s, b1), ps).
+betaReduceHeadIf(s, (_, BoolFalse()), _, _, b2, ps) = 
+  betaReduceHead((s, b2), ps).
+betaReduceHeadIf(s, _, c, b1, b2, ps) = 
+  rebuild((s, BoolIf(c, b1, b2)), ps).
 \end{lstlisting}
@@
 
+	\\
+
 	\inferrule{
@@
 		{ \scope{s'}{\co{BoolTrue}()} }
-		\\\bhr
+		\\\\\bhr
 		{ \scope{s}{b1} }
-		{ t }
+		{ \overline{p} }
 		{ \scope{s''}{b1'} }
@@
 		{ \scope{s}{\co{BoolIf}(c, b1, b2)} }
-		{ t }
+		{ \overline{p} }
 		{ \scope{s}{b1'}}
@@
 		{ \scope{s'}{\co{BoolFalse}()} }
-		\\\bhr
+		\\\\\bhr
 		{ \scope{s}{b2} }
-		{ t }
+		{ \overline{p} }
 		{ \scope{s''}{b2'} }
@@
 		{ \scope{s}{\co{BoolIf}(c, b1, b2)} }
-		{ t }
+		{ \overline{p} }
 		{ \scope{s}{b2'}}
+	}
+
+	\inferrule{
+	\bhr
+	{ \scope{s}{c} }
+	{ [] }
+	{ \scope{s'}{c'} }
+	\\\\ c' \neq \co{BoolTrue}()
+	\\ c' \neq \co{BoolFalse}()
+} {
+	\bhr
+	{ \scope{s}{\co{BoolIf}(c, b1, b2)} }
+	{ \overline{p} }
+	{ \co{rebuild}((s, \co{BoolIf}(c, b1, b2)), \overline{p}) }
 }
@@
 	\inferrule{
-		\toe{s}{c}{ct}
-		\\\beq{ct}{\co{BoolType}()}
+		\toe{s}{c}{t_c}
+		\\\beq{t_c}{\co{BoolType}()}
 		\\\\
-		\\\toe{s}{b1}{tb1}
-		\\\toe{s}{b2}{tb2}
-		\\\beq{tb1}{tb2}
+		\\\toe{s}{b1}{t_{b1}}
+		\\\toe{s}{b2}{t_{b2}}
+		\\\beq{t_{b1}}{t_{b2}}
 	} {
-		\toe{s}{\co{BoolIf}(c, b1, b2)}{tb1}
+		\toe{s}{\co{BoolIf}(c, b1, b2)}{t_{b1}}
 	}
@@
 			{\scope{s}{b}}
-			{a}
+			{\overline{p}}
 			{\scope{s'}{b'}}
@@
 			{\scope{s}{\co{Postulate}(n, t, b)}}
-			{a}
+			{\overline{p}}
 			{\scope{s'}{b'}}
@@
 		\inferrule{
-			\toe{s}{t}{tt}
-			\\\beq{tt}{\co{Type}()}
+			\toe{s}{t}{t_t}
+			\\\beq{t_t}{\co{Type}()}
 			\\\\\bred{s}{t}{t'}
-			\\\toe{\co{sPutType}(s, n, t')}{b}{bt}
+			\\\toe{\co{sPutType}(s, n, t')}{b}{t_b}
 		} {
-			\toe{s}{\co{Postulate}(n, t, b)}{bt}
+			\toe{s}{\co{Postulate}(n, t, b)}{t_b}
 		}
@@
 
-\section{Extensibility of the approach}
+\section{Extensibility of the Approach}
 
@@
 \begin{itemize}
-	\item Inference (chapter \ref{ch:inference})
-	\item Inductive Datatypes (chapter \ref{ch:datatypes})
-	\item Universes (chapter \ref{ch:universes})
+	\item Inference (chapter~\ref{ch:inference})
+	\item Inductive Datatypes (chapter~\ref{ch:datatypes})
+	\item Universes (chapter~\ref{ch:universes})
 \end{itemize}

### 2023-02-08 paper-thesis/chapters/30-comparing-with-haskell.tex
Add error annotations
Work on CoC chapter
Fix chapter references

@@
 
-In this chapter, we implement the language defined in chapter \ref{chap:baselang} in Haskell, and then compare the implementation with the implementation in Statix.
+In this chapter, we implement the language defined in chapter~\ref{chap:baselang} in Haskell, and then compare the implementation with the implementation in Statix.
 We want the design of the implementation in Haskell to be similar to conventional implementations of dependently typed languages, so we can compare the Statix implementation with them.
@@
 
-\section{Defining environments}
+\section{Defining Environments}
 
 Next, we need to decide how we want to store the environment.
-We still need to store both function arguments and substitutions, which we called \verb|NType| and \verb|NSubst| in chapter \ref{chap:baselang}.
+We still need to store both function arguments and substitutions, which we called \verb|NType| and \verb|NSubst| in chapter~\ref{chap:baselang}.
 We will use the same names, and store the environment as a list, with the head of the list representing De Bruijn index 0.
@@
 
-\section{Defining beta reduction}
+\section{Defining Beta Reduction}
 
@@
 The base implementations of the language are quite similar.
-One difference is the way the definitions are clustered\todo{not right word?}.
+One difference is the way the definitions are spread.
 In Statix, one could put each language construct in a separate file, keeping the definitions for that construct together.
@@
 
-Another advantage that Statix has is that it has first-order inference built-in, which makes implementing a basic form of inference as described in chapter \ref{ch:inference} way easier.
+Another advantage that Statix has is that it has first-order inference built-in, which makes implementing a basic form of inference as described in chapter~\ref{ch:inference} way easier.
 However, if we want a more complex form of inference then an implementation in Haskell would be better, since it is a more expressive language and it has more libraries available.
 
-Finally, creating the language in Spoofax automatically gives us editor services such as code highlighting and semantic autocompletion, as dicusssed in chapter \ref{chap:editor-services}.
+Finally, creating the language in Spoofax automatically gives us editor services such as code highlighting and semantic autocompletion, as dicusssed in chapter~\ref{chap:editor-services}.
 Implementing these in Haskell would require some work.

### 2023-02-08 paper-thesis/chapters/31-comparing-with-lambdapi.tex
Work on CoC chapter
Fix chapter references

@@
 
-In this chapter we will be implementing the language defined in chapter \ref{chap:baselang} in lambdapi, a proof assistant based on the $\lambda \Pi$-calculus modulo rewriting~\cite{BoespflugCH12}.
+In this chapter we will be implementing the language defined in chapter~\ref{chap:baselang} in lambdapi, a proof assistant based on the $\lambda \Pi$-calculus modulo rewriting~\cite{BoespflugCH12}.
 
-\section{Defining symbols}
+\section{Defining Symbols}
 
@@
 
-\section{Reduction rules}
+\section{Reduction Rules}
 
@@
 
-\section{\label{ch:comp-lambdapi:let}Defining let bindings}
+\section{\label{ch:comp-lambdapi:let}Defining Let Bindings}
 

### 2023-02-08 paper-thesis/chapters/40-discussing-statix.tex
Fix chapter references

@@
 	\end{itemize}
-	While these numbers are not unworkably high, they are still magnitudes higher than the implementation in Haskell (chapter \ref{ch:comp-haskell}) and other general-purpose languages.
+	While these numbers are not unworkably high, they are still magnitudes higher than the implementation in Haskell (chapter~\ref{ch:comp-haskell}) and other general-purpose languages.
 Improving these numbers would make the experience better.

### 2023-02-08 paper-thesis/chapters/99-conclusion.tex
Improve the introduction
Fix chapter references

@@
 \chapter{Conclusion}
+\label{ch:conclusion}
 
@@
 \paragraph{RQ1: Can the Calculus of Constructions be implemented in Statix?}
-As we demonstrated in chapter \ref{chap:baselang}, the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph.
+As we demonstrated in chapter~\ref{chap:baselang}, the Calculus of Constructions can be implemented concisely in Statix, by storing substitutions in the scope graph.
 We defined the beta-reduction, beta-equality and type checking rules and then converted them to Statix code.
 Beta reduction was defined using a Krivine machine.
-In chapter \ref{chap:namecolls} we solved the variable capture problem which the naive implementation from chapter \ref{chap:baselang} suffered from, by using scoped names.
+In chapter~\ref{chap:namecolls} we solved the variable capture problem which the naive implementation from chapter~\ref{chap:baselang} suffered from, by using scoped names.
 
 \paragraph{RQ2: Is the implementation is easily extendable?}
-We showed in chapter \ref{chap:bools} that the implementation is easily extendable, by extending it with booleans, postulate, and type assertions.
+We showed in chapter~\ref{chap:bools} that the implementation is easily extendable, by extending it with booleans, postulate, and type assertions.
 We define a four step process that can be used to extend the language, which we also use to answer the remaining research questions.
@@
 \paragraph{RQ3: Can we add inference to the implementation?}
-In chapter \ref{ch:inference} we discuss how to add inference to the implementation.
+In chapter~\ref{ch:inference} we discuss how to add inference to the implementation.
 If we want to keep the implementation clean and concise, we need to compromise on how powerful the inference algorithm is, by defining an approximated version of first-order inference.
@@
 \paragraph{RQ4: Can we add support for inductive data types to the implementation?}
-Chapter \ref{ch:datatypes} adds support for inductive data types with parameters and indices.
+chapter~\ref{ch:datatypes} adds support for inductive data types with parameters and indices.
 We give steps for type-checking data type declarations.
@@
 \paragraph{RQ5: Can we add support for universes to the implementation?}
-In chapter \ref{ch:universes} we showed that we can add universes to the language. This can be done easily and concisely.
+In chapter~\ref{ch:universes} we showed that we can add universes to the language. This can be done easily and concisely.
 

### 2023-02-08 paper-thesis/chapters/10-coc.tex
More feedback processed

@@
 These return a scope that represent an environment that has been extended with the new name.
-To query the scope graph we use \verb|sGetName|, which will return the closest \verb|NameEntry| with a matching name.
+To query the scope graph, we use \verb|sGetName|, which will return the closest \verb|NameEntry| with a matching name.
 Finally, \verb|sEmpty| returns a fresh empty scope.
 
-We will define a \emph{scoped expression}, as a pair of a scope and an expression. The scope acts as the environment of the expression, containing all of the context needed to evaluate the expression.
+We define a \emph{scoped expression}, as a pair of a scope and an expression. The scope acts as the environment of the expression, containing all of the context needed to evaluate the expression.
 
@@
 All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}.
-We use the syntax $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ to express \verb|betaReduceHead((s1, e1), ps) == (s2, e2)|.
+We use the syntax $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ to express \textit{betaReduceHead((s1, e1), ps) == (s2, e2)}.
 The $\overline{p}$ in this definition is the argument stack of the Krivine machine.
@@
 	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
-	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $\bhr{\scope{s1}{e1}}{\overline{p}}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1), ps) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
-	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
-	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
+	\item $\beq{\scope{s_1}{e_1}}{\scope{s_2}{e_2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
+	\item $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ is the same as \verb|betaReduceHead((s1, e1), ps) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
+	\item $\bred{s_1}{e_1}{e_2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
+	\item $\scope{\co{sEmpty}}{e}$ is the same as $e$ (empty scopes can be left out)
 \end{enumerate}

### 2023-02-08 paper-thesis/chapters/10-coc.tex
First working compilation

@@
 
-In conventional dependently typed languages, evaluation is often done using De Bruijn indices, which use numbers to identify variables.
+In conventional dependently typed languages, evaluation is often done using De Bruijn indices.
+De Bruijn representation~\cite[Section 6.1]{tapl} uses the distance from a binder to identify a variable.
+In this representation, alpha equivalence is the same as syntactic equivalence, which can simplify the manipulation of terms.
 However, we chose to use names rather than De Bruijn indices, because scope graphs work based on names.

### 2023-02-09 paper-thesis/chapters/02-background-spoofax.tex
Minor fixes

@@
 Using these rules, Statix automatically derives a type checker.
-The rules are based on \emph{equality constraints}.
-The basic constraints
-are equality constraints \verb|t1 == t2| and inequality constraints \verb|t1 != t2| on terms.
+The two basic constraints that can be used in the rules are equality constraints \verb|t1 == t2| and inequality constraints \verb|t1 != t2|.
 An equality constraint is states that two terms are equal or can be unified.
@@
 The constraints are generated by predicates.
-A predicate defines a relation on terms and scopes, it is of the form $\mathtt{p(t_1, ..., t_n) :- c_1, ..., c_n}$.
+A predicate defines a relation on terms and scopes, it is of the form $\mathtt{p(t_1, ..., t_n) :-\,\,c_1, ..., c_n}$.
 Each constraint $c_1...c_n$ is another predicate or an equality constraint.

### 2023-02-14 paper-thesis/chapters/11-variable-capture.tex
Work on CoC chapter
Fix chapter references
Fix a few details

@@
 
-The implementation of the base language shown in chapter \ref{chap:baselang} has one big problem, that is variable capture.
+We have now implemented the Calculus of Constructions in Statix.
+The implementation has one big problem, that is variable capture.
 Variable capture is the phenomenon of free variables in a term becoming bound when a naive substitution happens~\cite{tapl}.
@@
 
-Finally, we need to find the type of \verb|Var("T")|.
+Finally, we need to find the type of \verb|Var("T")| in \verb|s2|.
 This finds the lexically closest definition of \verb|T| (the one in s2), which is correct.
@@
 \section{Alternative Solutions}
+
+Now that the problem is clear, we will explore several attempts at a solution that failed, before settling on the final solution in section \ref{scopes_for_names}.
 
@@
 
-The first solution that was attempted was having a pre-analysis transformation that gives each variable a unique name.
+Another solution that was attempted was having a pre-analysis transformation that gives each variable a unique name.
 This doesn't work for a variety of reasons.
@@
 It requires a new relation to traverse through the type and rename.
-It also increases the time complexity, as many traversals of the type are needed.
+This is also an inefficient solution, as many traversals of the type are needed.
 This was successfully implemented, but we chose against using it in favor of the solution presented in the next section.
 
-\section{Using scopes to distinguish names}
+\section{Using Scopes to Distinguish Names}
+\label{scopes_for_names}
 
@@
 
-The \verb|ScopedName| constructor has a scope and a raw ID.
+The \verb|ScopedName| constructor takes a scope and a raw ID.
 The scope is used to uniquely identify the name.
@@
 
-\section{Improving the readability of types}
+\section{Improving the Readability of Types}
 

### 2023-02-14 paper-thesis/chapters/03-background-dependent-types.tex
Chapter transitions

@@
 
-The \textit{Calculus of Constructions} is the language that has all three of these features, it will be implemented in chapter~\ref{chap:baselang}.
+The \textit{Calculus of Constructions} is the language that has all three of these features.
+Now that we have explained the concepts behind Spoofax and Dependent Types, we will further explain the Calculus of Constructions and implement it in Spoofax in chapter~\ref{chap:baselang}.
+Then in the chapters after that, we will extend the language with more features.

### 2023-02-14 paper-thesis/chapters/10-coc.tex
Syntax highlighting
Fix a few details
Chapter transitions

@@
 \begin{figure}[h]
+\lstset{language=SDF3}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\lstset{language=base}
 \caption{The concrete syntax for the base language. $\co{FnConstruct}$ is a lambda function, $\co{FnDestruct}$ is application of a lambda function.}
@@
 
-An additional benefit of Krivine machines is that the run-time performance of them tends to be a bit higher, though this is only a constant factor and not a time-complexity improvement.
+An additional benefit of Krivine machines is that the run-time performance of them tends to be a bit better, though this is only a constant factor and not a time-complexity improvement.
 
@@
 \end{figure}
+
+We now have an implementation of the Calculus of Constructions in Spoofax.
+This implementation still has the issue of variable capture, which we will discuss and solve on chapter \ref{chap:namecolls}.
+Next, from chapter \ref{chap:bools} onward we will show how to extend the language.

### 2023-02-15 paper-thesis/chapters/12-editor-services.tex
Meeting results
Minor writing changes
Add error annotations
Add chapter transitions

@@
 \section{Setup required}
-To set up editor services, we followed the steps at \todo{cite} \url{https://spoofax.dev/spoofax-pie/develop/guide/static-semantics/code-completion/}. To be precise, we followed the following steps:
+To set up editor services, we followed the steps in the Spoofax documentation \cite{semcomplet}. To be precise, we followed the following steps:
 \begin{enumerate}
@@
  \end{itemize}
+
+We have now discussed semantic code completion. Next, we will compare the implementation with one in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi}).

### 2023-02-15 paper-thesis/chapters/22-adding-datatypes.tex
Work on CoC chapter
Fix chapter references
Add chapter transitions

@@
 \item Data types must be strictly positive.
-This ensures that we cannot make non-terminating programs\footnote{After we add support for universes in chapter \ref{ch:universes}}.
+This ensures that we cannot make non-terminating programs\footnote{After we add support for universes in chapter~\ref{ch:universes}}.
 For example, the following data type is forbidden, because it refers to itself in a negative position.
@@
 
-\section{Type-checking data type declarations}
+\section{Type-checking Data Type Declarations}
 \label{ch:datatypes:typechecking}
@@
 
-\section{Type-checking eliminators}
+\section{Type-checking Eliminators}
 
@@
 The implementation is simply an extra check during the type checking of data type declarations. It checks that the datatype only refers to itself in positive positions.
+
+We have now shown how to add inductive datatypes to the implementation. Next, we will Universes (chapter~\ref{ch:universes}) to the implementation.

### 2023-02-15 paper-thesis/chapters/23-adding-universes.tex
Add chapter transitions

@@
 \end{enumerate}
+
+We have now finished the implementation of the language.
+Next, we will discuss how well semantic code completion works with this implementation in chapter~\ref{chap:editor-services}.
+Then we will compare the implementation with one in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi}).

### 2023-02-15 paper-thesis/chapters/01-introduction.tex
Minor fixes
Some layout fixes

@@
 \section*{Contributions}
-This goal of this thesis is to investigate how well Statix is fit for the task of defining a simple dependently-typed language.
+This goal of this thesis is to investigate how well Statix is fit for the task of defining a dependently-typed language.
 We want to investigate whether typical features of dependently typed languages can be encoded in Statix.
@@
 
-The source code for this thesis can be found online: \url{https://github.com/JonathanBrouwer/master-thesis}
+The source code for this thesis can be found online \\ \url{https://github.com/JonathanBrouwer/master-thesis}

### 2023-02-15 paper-thesis/chapters/20-adding-booleans.tex
Add chapter transitions
Some layout fixes

@@
 Now that a few features have been implemented, we can discuss how easy the language is to extend.
-The conclusion so far is that it is doable in a clean way.
-Specifically, the approach for each of the features above was:
+From the three examples above and the following chapters we can conclude that extending the language in a clean way is possible.
+To extend the language, we used the following approach:
 
@@
 	\item Define the parsing rules for the new feature
-	\item Create a new file, \verb|tp_bools.stx| and import this file in the main \verb|type_check.stx| file.
+	\item Create a new file, \verb|tp_[feature].stx| and import this file in the main \verb|type_check.stx| file.
 	\item In the new file, define a case for the \verb|betaReduceHead| rule for the constructors that were added.
 Also define cases for \verb|expectBetaEq| and \verb|betaReduce| if necessary.
-It is necessary iff the new constructors are not always eliminated by \verb|betaReduceHead|.
+This is only necessary iff the new constructors are not always eliminated by \verb|betaReduceHead|.
 	\item In the new file, define a case for the \verb|typeOfExpr| rule for the constructors that were added.
@@
 If we decide that we don't like a feature after all, we can remove it simply by unimporting the file.
-This problem is similar to the expression problem\cite{expression_problem}, which Statix solves very well.
+This problem is similar to the expression problem\cite{expression_problem}, where we want to extend datatypes and their associated behaviour, which Statix solves very well.
 
-In the following chapters, we're going to be extending the language further with these more interesting features using this approach:
+In the following chapters, we're going to be extending the language further with these more interesting features using the approach defined above:
 \begin{itemize}

### 2023-02-15 paper-thesis/chapters/40-discussing-statix.tex
Some layout fixes

@@
 
-Now that we've implemented a feature-rich dependently typed language in Statix, this chapter discusses the experience, and recommends some features which might improve the experience.
+Now that we've implemented a feature-rich dependently typed language in Statix, this chapter discusses the experience, and recommends some changes to Statix and Spoofax which might improve the experience.
 
@@
 	\end{lstlisting}
-	The definition of \verb|isStrictlyPositive| can then use all the metavariables of \verb|DataTypeDecl| without having to explicitly pass them.
+	The definition of \texttt{isStrictlyPositive} can then use all the metavariables of \\ \texttt{DataTypeDecl} without having to explicitly pass them.
 

### 2023-02-15 paper-thesis/chapters/98-relatedwork.tex
Fix chapter references
Some layout fixes

@@
 \label{ch:relatedwork}
+
+In chapter \ref{ch:comp-haskell} and \ref{ch:comp-lambdapi} we already compared our implementation with two different implementations.
 
@@
 
-In chapter \ref{ch:comp-haskell} we compared the implementation in Statix with an implementation in Haskell.
+In chapter~\ref{ch:comp-haskell} we compared the implementation in Statix with an implementation in Haskell.
 The pi-forall language~\cite{pi_forall} is a language with a similar complexity to the language presented in this paper.

### 2023-02-15 paper-thesis/chapters/11-variable-capture.tex
Add chapter transitions
Meeting results

@@
 \end{lstlisting}
-The implementation so far would tell you it is \verb|T : Type -> T : T -> T|.
+The implementation so far would tell you it is \verb|T : Type -> T : T -> T| \todo{why is this wrong}.
 Given the scoping rules of the language, that is equivalent to \verb|T : Type -> x : T -> x|.
@@
 \end{enumerate}
+
+We have now solved the problem of variable capture.
+In chapter \ref{chap:bools}, we explain how to extend the language further, by extending it with booleans.
+Then, from chapter \ref{ch:inference} onward we use this method to extend the language with various features.

### 2023-02-19 paper-thesis/chapters/10-coc.tex
Some layout fixes
Type checking example

@@
 	\item \verb|Type| is the Type of Types.
-	\item \verb|Var| is a variable, it uses the lexical sort \verb|ID|, which is defined as \verb|[a-zA-Z\_][a-zA-Z0-9\_]*|.
+	\item \verb|Var| is a variable, it uses the lexical sort \verb|ID|, which is defined as \\ \verb|[a-zA-Z\_][a-zA-Z0-9\_]*|.
 	\item \verb|FnType| is the type of a function.
@@
 
+\section{An example}
+\todo{Example}
+
+\begin{figure}[h]
+	\figuresection{Type checking let}
+	\begin{mathpar}
+	\mprset{vskip=0.3ex}
+		\inferrule{
+			\inferrule{
+			}{
+			\toe{s_0}{\co{Type}()}{\co{Type}()}
+			}
+			\\ \toe{s_1}{b}{\co{Type}()}
+		}{
+			\toe{s_0}{\co{Let}("T", \co{Type}(), b)}{\co{Type}()}
+		}
+	\mprset{vskip=0ex}
+	\end{mathpar}
+
+	\figuresection{Type checking let body}
+	\begin{mathpar}
+		\mprset{vskip=0.7ex}
+		\inferrule{
+			\toe{s_1}{\co{FnConstruct}("v", \co{Type}(), \co{Var}("v"))}{\co{FnType}("v", \co{Type}(), \co{Type}())}
+			\\\\ \bhr{\scope{s_1}{\co{FnType}("v", \co{Type}(), \co{Type}())}}{[]}{\scope{s_1}{\co{FnType}("v", \co{Type}(), \co{Type}())}}
+			\\\\ \toe{s_1}{\co{Var}("T")}{\co{Type}()}
+			\\ \beq{\co{Type}()}{\co{Type}()}
+			\\ \bred{s_2}{\co{Type}()}{\co{Type}()}
+		}{
+			\toe{s_1}{\co{FnDestruct}(\co{FnConstruct}("v", \co{Type}(), \co{Var}("v")), \co{Var}("T")))}{\co{Type}()}
+		}
+		\mprset{vskip=0ex}
+	\end{mathpar}
+
+	\figuresection{Type checking function}
+	\begin{mathpar}
+		\mprset{vskip=0.7ex}
+		\inferrule{
+			\toe{s_1}{\co{Type}()}{\co{Type}()}
+			\\ \beq{\co{Type}()}{\co{Type}()}
+			\\\\ \bred{s_1}{\co{Type}()}{\co{Type}()}
+			\\ \toe{s_3}{\co{Var}("v")}{\co{Type}()}
+		}{
+			\toe{s_1}{\co{FnConstruct}("v", \co{Type}(), \co{Var}("v"))}{\co{FnType}("v", \co{Type}(), \co{Type}())}
+		}
+		\mprset{vskip=0ex}
+	\end{mathpar}
+
+	\figuresection{Corresponding Scope Graph} \\
+	\begin{tikzpicture}[scopegraph, node distance = 2.5em and 2.5em]
+		\node[scope] (s0) {$s0$};
+
+		\node[scope, right of = s0, xshift=5em] (s1) {$s1$};
+		\draw (s1) edge[] (s0);
+		\node[type, above of = s1] (s1l) {$\texttt{T = Type()}$};
+		\draw (s1) edge[type] (s1l);
+
+		\node[scope, right of = s1, xshift=6em] (s2) {$s2$};
+		\draw (s2) edge[] (s1);
+		\node[type, above of = s2] (s2l) {$\texttt{v = Type()}$};
+		\draw (s2) edge[type] (s2l);
+
+		\node[scope, below right of = s1, xshift=7em, yshift=-3em] (s3) {$s3$};
+		\draw (s3) edge[] (s1);
+		\node[type, above of = s3] (s3l) {$\texttt{v : Type()}$};
+		\draw (s3) edge[type] (s3l);
+	\end{tikzpicture}
+	\caption{Type checking example}
+	\label{fig:type-check-example}
+\end{figure}
+
 \section{Discussion of a Substitution-Based Approach}

### 2023-02-19 paper-thesis/chapters/98-relatedwork.tex
Meeting results
Type checking example

@@
 
-In chapter \ref{ch:comp-haskell} and \ref{ch:comp-lambdapi} we already compared our implementation with two different implementations.
+In chapter \ref{ch:comp-haskell} and \ref{ch:comp-lambdapi} we already compared our implementation with two different implementations. In this chapter we discuss remaining related work.
+TODO TODO TODO
+
+\section{Other languages implemented in Statix}
+
+In this thesis, we implemented a dependently typed language in Statix. In the following work, other languages are implemented in Statix:
 
@@
 The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
-Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types} present an implementation of System F that does lazy substitutions, by using scopes as types.
+Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types}~\cite{nameres} present an implementation of System F in Statix that does lazy substitutions, by using scopes as types.
 It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can be arbitrary terms.
+
+\paragraph{Correct by Construction Language Implementations}
+
+Construction Language Implementations~\cite{Rouvoet21} describes how to create a declarative specification of a language in Statix, and then to use a dependently typed language to integrate the specification of well-typing in the representation of the program that is being interpreted or transformed.
+The work from this thesis and ours could be combined, by using our dependently typed languages to do this verification.
+
+\section{Other dependently typed languages}
+
+We already implemented the calculus of constructions in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi}), and compared the implementation in Statix with these implementations.
+
+Below are some dependently typed languages, we discuss what make these languages unique and if these languages could be implemented in Statix:
 
@@
 
-\paragraph{}
+\paragraph{Other}

### 2023-02-20 paper-thesis/chapters/01-introduction.tex
Add example to background

@@
 
-Type checking programming languages is commonly done by writing a type checker in a general purpose language.
+Type checking is commonly done by writing a type checker in a general purpose language.
 This requires a lot of effort, and it is easy to make mistakes.
-In this thesis we will specifically focus on dependently typed languages, which differ from other languages because they allow types to be parameterized by values.
+In this thesis we will specifically focus on dependently typed languages, which differ from other languages because they allow types to be parameterized by values~\cite{lambda_cube}.
 This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree.
-This expressiveness also makes dependent type systems more complicated to type check, since deciding equality of types requires normalization of the terms they are parameterized by.
+This expressiveness also makes dependent type systems more complicated to type check, since deciding equality of types requires normalization of the terms they are parameterized by~\cite{Coquand_Huet_1988}.
 

### 2023-02-20 paper-thesis/chapters/02-background-spoofax.tex
Syntax highlighting
Chapter transitions
Add example to background

@@
 For example, below is the syntax of a simple language:
+
+\lstset{language=SDF3}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\lstset{language=base}
 
@@
 \end{itemize}
+
+\subsection*{Example}
+To get a better understanding of scope graphs, we consider an example in a non-dependent language. 
+\lstset{language=java}
+\begin{lstlisting}
+	int x = 5;
+	int y = x + 3;
+	return x + y;
+\end{lstlisting}
+Remember that you can think of scopes as a type-checking environment. During type checking, the following steps happen:
+\begin{enumerate}
+	\item \verb|5| is type-checked in the root scope \verb|s0|. 
+	\item The definition of \verb|x| creates a new scope \verb|s1| that contains the type of \verb|x|, which is used to type-check the rest of the program.
+	\item \verb|x + 3| is type-checked in the new scope \verb|s1|, the \verb|x| is resolved to the declaration in \verb|s1|.
+	\item The definition of \verb|y| creates a new scope \verb|s2| that contains the type of \verb|y|, which is used to type-check the rest of the program.
+	\item \verb|x + y| is type-checked in the new scope \verb|s2|, to resolve \verb|x| we need to follow a \verb|P| edge.
+\end{enumerate}
+
+\lstset{language=}
+\begin{tikzpicture}[scopegraph, node distance = 2.5em and 2.5em]
+	\node[scope] (s0) {$s0$};
+
+	\node[scope, right of = s0, xshift=5em] (s1) {$s1$};
+	\draw (s1) edge[lbl={P}] (s0);
+	\node[type, above of = s1] (s1l) {$\texttt{x : Int()}$};
+	\draw (s1) edge[type] (s1l);
+
+	\node[scope, right of = s1, xshift=6em] (s2) {$s2$};
+	\draw (s2) edge[lbl={P}] (s1);
+	\node[type, above of = s2] (s2l) {$\texttt{y : Int()}$};
+	\draw (s2) edge[type] (s2l);
+\end{tikzpicture}

### 2023-02-22 lody/src/main/java/mb/lody/task/Compile.java
Basic compiler working

@@ -66,7 +66,7 @@ public class Compile implements TaskDef<Compile.Args, CommandFeedback> {
         context.require(classloaderResources.tryGetAsNativeResource(getClass()), ResourceStampers.hashFile());
         final ResourceKey file = args.file;
         return context.require(transformer, parse.inputBuilder().withFile(file).buildAstSupplier()).mapOrElse(
-            ast -> CommandFeedback.of(ShowFeedback.showText(TermToString.toString(ast), "Title")),
+            ast -> CommandFeedback.of(ShowFeedback.showText(TermToString.toString(ast), "test.clj")),
             e -> CommandFeedback.ofTryExtractMessagesFrom(e, file)
         );
     }

### 2023-02-22 paper-thesis/chapters/01-introduction.tex
Improve the introduction a bit

@@
 
-Type checking is commonly done by writing a type checker in a general purpose language.
-This requires a lot of effort, and it is easy to make mistakes.
-In this thesis we will specifically focus on dependently typed languages, which differ from other languages because they allow types to be parameterized by values~\cite{lambda_cube}.
-This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree.
-This expressiveness also makes dependent type systems more complicated to type check, since deciding equality of types requires normalization of the terms they are parameterized by~\cite{Coquand_Huet_1988}.
-
+While we keep building more and more complex programming languages, their type checkers are still often written in general purpose languages.
+This takes a lot of effort, and it is easier to make mistakes.
 Instead of writing the type checker in a general purpose language, in this thesis we will systematically derive a type checker from a high-level, declarative specification.
@@
 
-We will do this using the Spoofax language workbench, which is a collection of tools that can derive a parser and type checker from a high level specification of the language~\cite{spoofax}.
+In this thesis we will specifically focus on dependently typed languages, which differ from other languages because they allow types to be parameterized by values~\cite{lambda_cube}.
+This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree.
+This expressiveness also makes dependent type systems more complicated to type check, since deciding equality of types requires evaluation of the terms they are parameterized by~\cite{Coquand_Huet_1988}.
+
+We will write the declarative specification in the Spoofax language workbench, which is a collection of tools that can derive a parser and type checker from a high level specification of the language~\cite{spoofax}.
 When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics~\cite{scopes_as_types}.
-It is a declarative language that uses inference rules to define static semantics.
+It is a declarative language that uses inference rules and scope graphs~\cite{scope_graphs} to define static semantics.
 Statix aims to cover a broad range of languages and type systems.
@@
 This goal of this thesis is to investigate how well Statix is fit for the task of defining a dependently-typed language.
-We want to investigate whether typical features of dependently typed languages can be encoded in Statix.
-The goal is not only to show that Statix can implement it, but also that the implementation is concise.
+The goal is not only to show that Statix can implement it, but also that the implementation is more concise than in a general purpose language.
+We will start by implementing the Calculus of Constructions~\cite{Coquand_Huet_1988}, a lambda calculus with dependent types, as this is one of the most simple dependently typed languages (chapter~\ref{chap:baselang}).
 
-The technical contributions of this thesis are the following:
+However, this language is not very practical to write actual complex programs in, so to improve this situation we will extend the language with typical features of dependently typed languages.
+We will do this in the following steps:
 \begin{itemize}
-	\item We present a implementation of the Calculus of Constructions, a lambda calculus with dependent types, in Statix (Chapter~\ref{chap:baselang})
 	\item We show the language is easily extendable, by extending it with booleans among other features (Chapter~\ref{chap:bools})

### 2023-02-22 paper-thesis/chapters/10-coc.tex
Work
Bit of cleanup on coc chapter

@@
 The base language that has been implemented is the Calculus of Constructions~\cite{Coquand_Huet_1988}.
-One extra feature was added that is not present in the Calculus of Constructions: let bindings.
+We chose this language as it is the core language of many dependently typed languages~\cite{core_coq}\cite{lean}, so it is representative of dependently typed languages as a whole.
+
+One extra feature has been added that is not present in the pure Calculus of Constructions: let bindings.
 Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful.
@@
 \begin{figure}[h]
-	\figuresection{Type checking let}
+	\figuresection[\fbox{$b_0 = \co{FnDestruct}(b_1, \co{Var}("T")))$}]{Type checking let}
 	\begin{mathpar}
@@
 			}
-			\\ \toe{s_1}{b}{\co{Type}()}
+			\\ \toe{s_1}{b_0}{\co{Type}()}
 		}{
-			\toe{s_0}{\co{Let}("T", \co{Type}(), b)}{\co{Type}()}
+			\toe{s_0}{\co{Let}("T", \co{Type}(), b_0)}{\co{Type}()}
 		}
@@
 
-	\figuresection{Type checking let body}
+	\figuresection[\fbox{$b_1 = \co{FnConstruct}("v", \co{Type}(), \co{Var}("v"))$}]{Type checking let body}
 	\begin{mathpar}
@@
 		\inferrule{
-			\toe{s_1}{\co{FnConstruct}("v", \co{Type}(), \co{Var}("v"))}{\co{FnType}("v", \co{Type}(), \co{Type}())}
+			\toe{s_1}{b_1}{\co{FnType}("v", \co{Type}(), \co{Type}())}
 			\\\\ \bhr{\scope{s_1}{\co{FnType}("v", \co{Type}(), \co{Type}())}}{[]}{\scope{s_1}{\co{FnType}("v", \co{Type}(), \co{Type}())}}
@@
 		}{
-			\toe{s_1}{\co{FnDestruct}(\co{FnConstruct}("v", \co{Type}(), \co{Var}("v")), \co{Var}("T")))}{\co{Type}()}
+			\toe{s_1}{\co{FnDestruct}(b_1, \co{Var}("T")))}{\co{Type}()}
 		}

### 2023-02-22 paper-thesis/chapters/10-coc.tex
Start example

@@
 \section{An example}
-\todo{Example}
+In this section we will give an example of how to type check a program. The program we will type check is:
+\begin{lstlisting}
+let T = Type;
+(\v: Type. v) T
+\end{lstlisting}
+
+The process that happens during type-checking is shown in figure \ref{fig:type-check-example}.
+\begin{enumerate}
+	\item First, we type-check the outermost expression \verb|Let| in an empty scope $s_0$.
+That rule states we have to type-check the let-bound value using the same scope, then it creates a \verb|NSubst| declaration in a new scope $s_1$, with which it type checks the body (which is here abbreviated as $b_0$).
+
+	\item \todo{tododododood}
+\end{enumerate}
 

### 2023-02-27 paper-thesis/chapters/10-coc.tex
Finish CoC example

@@
 \end{lstlisting}
-
 The process that happens during type-checking is shown in figure \ref{fig:type-check-example}.
@@
 That rule states we have to type-check the let-bound value using the same scope, then it creates a \verb|NSubst| declaration in a new scope $s_1$, with which it type checks the body (which is here abbreviated as $b_0$).
-
-	\item \todo{tododododood}
+	\item To type-check the \verb|FnDestruct| in the let body, we apply the corresponding rule.
+This requires type checking the given function, checking if the argument type matches, then substituting the argument in the return type creating scope $s_2$ and returning this.
+The only non-trivial step is type-checking the function.
+	\item To type check the function, we type check the argument and assert that it is a type.
+Then we type check the body with the argument in scope, creating scope $s_3$, and finally we construct the type.
 \end{enumerate}

### 2023-02-28 paper-thesis/chapters/24-backend.tex
Start backend

\chapter{Back-end}
\label{ch:backend}

We have now fully built the front-end (parser and type checker) of our language. However, the language also needs a back-end to be usable.

### 2023-03-01 paper-thesis/chapters/24-backend.tex
Work on backend chapter

@@
 
-We have now fully built the front-end (parser and type checker) of our language. However, the language also needs a back-end to be usable.
+We have now fully built the front-end (parser and type checker) of our language.
+However, the language also needs a back-end to be usable.
+A back-end can be either an interpreter or a compiler, we will explore both options in this chapter.
+
+\section{Interpreter}
+It is valuable to define an interpreter for our language, because it is a way to run a program without relying on the definition of the language you are compiling to.
+For a dependently typed language specifically it is easy to add on, as we already needed to define the dynamic semantics of our language for type checking anyways.
+
+However, this big advantage is also a disadvantage: Dynamic semantics would usually be defined in Stratego.
+We need to find a way to re-use the definition from Statix, since we don't want to repeat the definition.
+Luckily, Statego has a way to evaluate Statix relations on terms.
+
+\todo{finish this after implementing}
+
+\section{Compiler}
+The alternative to writing an interpreter is writing a compiler to another language.
+In order to explore this avenue, we wrote a compiler to Clojure, a dynamically typed dialect of Lisp that runs on the JVM~\cite{clojure}.
+The advantage of a compiling to such a high-level language is that the actual compilation steps are relatively straight-forward, but it still shows that we can write a compiler for a dependently typed language in Spoofax.
+
+Both languages support first-class functions, so we can map most constructs of our language directly onto constructs of Clojure.
+Therefore, most of our compiler has rules that look like the example below, where the constructors are compiled recursively using the \verb|compile-expr| strategy, ignoring the specified types.
+The constructors starting with \verb|C| are constructors from Clojure.
+\begin{lstlisting}
+compile-expr : FnConstruct(x, _, b) -> 
+  CFn(<compile-id> x, <compile-expr> b)
+\end{lstlisting}
+
+Types require a bit more attention.
+Since our language has no way to match on types, types can actually be completely eliminated during type checking.
+The easiest way to do this is to compile all types to an arbitrary value (we chose the integer 0), since the value is irrelevant anyways.
+A more efficient solution could be to implement type erasure~\cite[Section 23.6]{tapl}, where type-level functions are completely removed.
+We decided against this as the goal of this section is to show that we can compile the language with ease, not to write an efficient compiler.
+
+Finally, inductive data types are a bit harder to compile, as Clojure does not have an equivalent to sum types on which we can match. \todo{finish this after implementing}

### 2023-03-01 paper-thesis/chapters/24-backend.tex
Add meeting notes

@@
 \todo{finish this after implementing}
+- discuss efficiency (reference "Optimizing first class pattern matching")
 
@@
 
+-analysis is hard, agda doesn't do it
+- if compiled language doesn't matter
+- erasure annotations (idris/agda)
+
 Finally, inductive data types are a bit harder to compile, as Clojure does not have an equivalent to sum types on which we can match. \todo{finish this after implementing}
+- agda to scheme compiler
+- atoms voor constructors

### 2023-03-06 lody/src/main/java/mb/lody/task/Interpret.java
Basic interpeter working

@@ -18,7 +18,7 @@ import mb.spoofax.core.language.command.ShowFeedback;
 import mb.aterm.common.TermToString;
 
 @LodyScope
-public class Compile implements TaskDef<Compile.Args, CommandFeedback> {
+public class Interpret implements TaskDef<Interpret.Args, CommandFeedback> {
     public static class Args implements Serializable {
         private static final long serialVersionUID = 1L;
 
@@ -50,10 +50,10 @@ public class Compile implements TaskDef<Compile.Args, CommandFeedback> {
 
     private final LodyClassLoaderResources classloaderResources;
     private final LodyParse parse;
-    private final CompileTransform transformer;
+    private final InterpretTransform transformer;
 
     @Inject
-    public Compile(LodyClassLoaderResources classloaderResources, LodyParse parse, CompileTransform transformer) {
+    public Interpret(LodyClassLoaderResources classloaderResources, LodyParse parse, InterpretTransform transformer) {
         this.classloaderResources = classloaderResources;
         this.parse = parse;
         this.transformer = transformer;
@@ -65,7 +65,7 @@ public class Compile implements TaskDef<Compile.Args, CommandFeedback> {
         context.require(classloaderResources.tryGetAsNativeResource(getClass()), ResourceStampers.hashFile());
         final ResourceKey file = args.file;
         return context.require(transformer, parse.inputBuilder().withFile(file).buildAstSupplier()).mapOrElse(
-            ast -> CommandFeedback.of(ShowFeedback.showText(TermToString.toString(ast), "test.clj")),
+            ast -> CommandFeedback.of(ShowFeedback.showText(TermToString.toString(ast), "output.aterm")),
             e -> CommandFeedback.ofTryExtractMessagesFrom(e, file)
         );
     }

### 2023-03-06 lody/src/main/java/mb/lody/task/InterpretTransform.java

@@ -1,0 +1,32 @@
+package mb.lody.task;
+
+import mb.lody.LodyClassLoaderResources;
+import mb.lody.LodyScope;
+import mb.pie.api.ExecContext;
+import mb.pie.api.stamp.resource.ResourceStampers;
+import mb.stratego.pie.AstStrategoTransformTaskDef;
+
+import javax.inject.Inject;
+import java.io.IOException;
+
+@LodyScope
+public class InterpretTransform extends AstStrategoTransformTaskDef {
+  private final LodyClassLoaderResources classloaderResources;
+
+  @Inject
+  public InterpretTransform( // 1
+    LodyClassLoaderResources classloaderResources,
+    LodyGetStrategoRuntimeProvider getStrategoRuntimeProvider
+  ) {
+    super(getStrategoRuntimeProvider, "interpret-program"); // 2
+    this.classloaderResources = classloaderResources;
+  }
+
+  @Override public String getId() { // 3
+    return getClass().getName();
+  }
+
+  @Override protected void createDependencies(ExecContext context) throws IOException { // 4
+    context.require(classloaderResources.tryGetAsNativeResource(getClass()), ResourceStampers.hashFile());
+  }
+}

### 2023-03-06 lody/src/main/java/mb/lody/task/Interpret.java
Fix data type declarations

@@ -65,7 +65,7 @@ public class Interpret implements TaskDef<Interpret.Args, CommandFeedback> {
         context.require(classloaderResources.tryGetAsNativeResource(getClass()), ResourceStampers.hashFile());
         final ResourceKey file = args.file;
         return context.require(transformer, parse.inputBuilder().withFile(file).buildAstSupplier()).mapOrElse(
-            ast -> CommandFeedback.of(ShowFeedback.showText(TermToString.toString(ast), "output.aterm")),
+            ast -> CommandFeedback.of(ShowFeedback.showText(TermToString.toString(ast), "output.lody")),
             e -> CommandFeedback.ofTryExtractMessagesFrom(e, file)
         );
     }

### 2023-03-08 paper-thesis/chapters/11-variable-capture.tex
Work on interpreter text + other work on text

@@
 
-Almost all compilers that typecheck dependently typed languages use de Bruijn representation for variables.
+Almost all compilers that typecheck dependently typed languages use de Bruijn representation for variables~\cite{lean}.
 Using de Bruijn indices in statix is possible, but sacrifices a lot.
-Many editor services (renaming, go to definition) require \verb|.ref| annotations to be set on names, and this is not possible if the names are no longer a part of the AST.
+Many editor services (renaming, go to definition) require \verb|.ref| annotations (which specify which other name a certain name refers to) to be set on names, and this is not possible if the names are no longer a part of the AST and the scope graph.
 
@@
 Names can be duplicated during beta reduction of terms, so we still don't have the guarantee that each variable has a unique name.
-Furthermore, this changes the names in the AST, so it breaks editor services such as renaming.
+Furthermore, this changes the names in the AST, so it breaks editor services in the same was as De Bruijn Indices did.
 
-\subsection{Renaming terms dynamically}
+\subsection{Capture-avoiding substitution}
 
 Anytime that we introduce a new name in a type, we could check if the name already exists in the environment, and if it does, choose a different unique name.
-This approach is possible but tedious to implement in Statix.
+This approach is called capture-avoiding substitution by renaming~\cite{capture_avoiding_sub}.
+This is possible but tedious to implement in Statix.
 It requires a new relation to traverse through the type and rename.
 This is also an inefficient solution, as many traversals of the type are needed.
-This was successfully implemented, but we chose against using it in favor of the solution presented in the next section.
+This was successfully implemented, but we chose against using it since we found a better solution.
 
@@
 
-Because the expression above, with \verb|ScopedName|s, is not particularly readable, we add a new post-analysis Stratego pass that converts the \verb|ScopedName|s to ticked names.
+Because the expression above, with \verb|ScopedName|s, is not particularly readable, we add a new post-analysis Stratego pass (an AST transformation that runs directly after type-checking) that converts the \verb|ScopedName|s to ticked names.
 For example, the above would be transformed to:

### 2023-03-08 paper-thesis/chapters/20-adding-booleans.tex
Work on interpreter text + other work on text

@@
 The goal of this is to show that this language is easy to extend, and to add some features that make testing the language easier.
+
+It is important that our implementation is easily extensible since real-world dependently typed languages have complex features such as inductive data types and term inference, that make the language a lot easier to use.
+We will first show how to add support for these relatively simple features, and then use the same techniques to add support for more complex features in the following chapters.
 

### 2023-03-08 paper-thesis/chapters/24-backend.tex
Work on interpreter text + other work on text

@@
 However, this big advantage is also a disadvantage: Dynamic semantics would usually be defined in Stratego.
-We need to find a way to re-use the definition from Statix, since we don't want to repeat the definition.
-Luckily, Statego has a way to evaluate Statix relations on terms.
+We need to find a way to re-use the definition from Statix, since we don't want to repeat the definition of the dynamic semantics.
 
-\todo{finish this after implementing}
-- discuss efficiency (reference "Optimizing first class pattern matching")
+\subsection{Implementation}
+
+Luckily, Stratego has a strategy built in that runs a Statix predicate on a term: \verb|stx-evaluate|.
+Using this strategy, the actual implementation of our interpreter in Stratego is just one line: Match an expression \verb|e| and run the functional predicate \verb|interpret| from the \verb|main.stx| file on it.
+
+\begin{lstlisting}
+interpret : e -> <stx-evaluate(|"main", "main!interpret")> [e]
+\end{lstlisting}
+
+Then, in the \verb|main.stx| file we define \verb|interpret| to be beta reducing the expression in an empty scope.
+
+\begin{lstlisting}
+interpret(e) = betaReduce((sEmpty(), e)).
+\end{lstlisting}
+
+These two lines are the only required changes to implement an interpreter\footnote{Other than changing the project config to define a button in the Spoofax UI that when pressed runs the interpret strategy on a given term}.
+This shows that it is very straight-forward to implement an interpreter.
+
+\subsection{Discussion of defining in Statix versus Stratego}
+
+An advantage of implementing the dynamic semantics in Statix instead of Stratego is that we can use scope graphs as a part of our dynamic semantics rules.
+
+A disadvantage is that Stratego is specifically optimized for these transformations~\cite{fcpmc}. Therefore, the resulting interpreter will be slower when implemented in Statix. \todo{comparison}
 

### 2023-03-13 pres-evcs/presentation.tex
Work on EVCS presentation

% !TeX document-id = {2870843d-1baa-4f6a-bd0a-a5c796104a32}
% !TeX encoding = UTF-8
% TU Delft beamer template

\documentclass[aspectratio=43]{beamer}
\usepackage{csquotes}
\usepackage{calc}
\usepackage[absolute,overlay]{textpos}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{mathtools}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{comment}
\usepackage{siunitx}
\usepackage{MnSymbol,wasysym}
\usepackage{array}
\usepackage{qrcode}

\setbeamertemplate{navigation symbols}{} % remove navigation symbols
\mode<presentation>{\usetheme[verticalbar=false]{tud}}

\newcommand{\absimage}[4][0.5,0.5]{%
	\begin{textblock}{#3}%width
		[#1]% alignment anchor within image (centered by default)
		(#2)% position on the page (origin is top left)
		\includegraphics[width=#3\paperwidth]{#4}%
\end{textblock}}

\newcommand{\mininomen}[2][1]{{\let\thefootnote\relax%
	\footnotetext{\begin{tabular}{*{#1}{@{\!}>{\centering\arraybackslash}p{1em}@{\;}p{\textwidth/#1-2em}}}%
	#2\end{tabular}}}}

\title[]{Dependently Typed Languages in Statix}
\institute[]{Delft University of Technology, The Netherlands}
\author{Jonathan Brouwer \and Jesper Cockx \and Aron Zwaan}

\begin{document}
\section{Introduction}
{
\setbeamertemplate{footline}{\usebeamertemplate*{minimal footline}}
\frame{\titlepage}
}

\begin{frame}[fragile]{Background: What are Dependent Types?}
\begin{itemize}
	\item Types may depend on values!
	\begin{exampleblock}{Example}
		\texttt{concat : (A: Set) -> (n : Nat) -> Vec A n -> Vec A n
			\\ \hspace*{48pt} -> Vec A (n + n)}
	\end{exampleblock}
% - n is a RUNTIME value, it does not have to be known at compile time
% - n occurs insie a type `Vec`!
% - The type system enforces these lengths are always correct
% - Challenge: Type checking may require evaluating arbitrary terms to decide equality
	\item Curry-Howard correspondence
% Can be used to write proofs about code
\end{itemize}

\end{frame}

\begin{frame}[fragile]{Research Question}
How well Statix is fit for the task of defining a dependently-typed language.
% Statix is turing complete, so it's going to be possible
% But is it going to be easier or harder than a general purpose language like haskell?
\end{frame}

\begin{frame}[fragile]{Why is this important?}
	\begin{block}{From the perspective of Spoofax research}
		Developing a language with a complex type system tests the boundaries of what Spoofax can do.
	\end{block}
	% - What can we improve about Statix?

	\begin{block}{From the perspective of Dependent Types research}
		A rapid prototyping platform.
	\end{block}
	% - Try a new feature without having to implement in general purpose
	% - Spoofax provides an easy way to get a parser, type checker, etc
\end{frame}

\begin{frame}[fragile]{Calculus of Constructions}
	A lambda calculus with dependent types.

	\begin{exampleblock}{Example 1}
		\texttt{($\backslash$v: Type. v) T}
	\end{exampleblock}

	\begin{exampleblock}{Example 2}
		\texttt{let f = $\backslash$T: Type. $\backslash$x: T. x; \\
f (T: Type -> Type) ($\backslash$y: Type. y)
		}
	\end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Extra contributions}
\begin{enumerate}
	\item Implemented Inference
	\item Implemented Inductive Data Types
	\item Implemented Universes
	\item Interpreter
	\item Compiler to Clojure
	\item Comparison with implementation in Haskell
	\item Comparison with implementation in LambdaPi
	\item Evaluation of Spoofax
\end{enumerate}
\end{frame}

% block
% exampleblock
% alertblock

\end{document}

### 2023-03-15 pres-evcs/presentation.tex
Add support for types to compiler
Work on presentation

@@
 
+\begin{frame}[fragile]{Type Checking}
+	\begin{block}{Type checking relation}
+		\texttt{typeOfExpr : scope * Expr -> Expr}
+	\end{block}
+	% Returns an Expr, not a Type
+	% Scope -> Two different types of nodes
+	\begin{block}{How do we use scopes?}
+		One relation \texttt{name $\rightarrow$ NameEntry}, NameEntry is either:
+		\begin{itemize}
+			\item NType: Stores a type
+			\item NSubst: Stores a substitution
+		\end{itemize}
+	\end{block}
+\end{frame}
+
+\begin{frame}[fragile]{Type Checking: Requires Evaluation}
+	\begin{exampleblock}{Example 1}
+		\texttt{let T = Bool;\\let b: T = true;}
+	\end{exampleblock}
+	\begin{block}{Evaluation relation} 
+		\texttt{betaHeadReduce : scope * Expr -> scope * Expr} \\
+		\texttt{betaReduce : scope * Expr -> Expr} \\
+		\texttt{exectBetaEq : (scope * Expr) * (scope * Expr)}
+	\end{block}
+\end{frame}
+
+\begin{frame}[fragile]{Type Checking: Rules}
+	\begin{figure}
+		\centering
+		\includegraphics[width=0.5\linewidth]{img/screenshot001}
+		\label{fig:screenshot001}
+	\end{figure}
+	\begin{figure}
+		\centering
+		\includegraphics[width=0.5\linewidth]{img/screenshot002}
+		\label{fig:screenshot002}
+	\end{figure}
+
+\end{frame}
+
 \begin{frame}[fragile]{Extra contributions}
@@
 
+\begin{frame}[fragile]{Conclusions}
+	Spoofax is a great tool for developing dependently typed languages!\footnote{But there is still room for improvement}
+\end{frame}
+
 % block

### 2023-03-15 pres-evcs/presentation.tex
Presentation feedback

@@
 	\begin{exampleblock}{Example}
-		\texttt{concat : (A: Set) -> (n : Nat) -> Vec A n -> Vec A n
-			\\ \hspace*{48pt} -> Vec A (n + n)}
+		\texttt{concat : (A: Set) -> (n m : Nat) -> Vec A n -> Vec A m
+			\\ \hspace*{48pt} -> Vec A (n + m)}
 	\end{exampleblock}
@@
 % - Challenge: Type checking may require evaluating arbitrary terms to decide equality
-	\item Curry-Howard correspondence
+	\item Proof assistants
+	\item For example: Agda, Coq, Lean, ...
 % Can be used to write proofs about code
@@
 \begin{frame}[fragile]{Research Question}
-How well Statix is fit for the task of defining a dependently-typed language.
+\large{How suitable is Statix for defining a dependently-typed language?}
 % Statix is turing complete, so it's going to be possible
@@
 \begin{frame}[fragile]{Why is this important?}
-	\begin{block}{From the perspective of Spoofax research}
+	\begin{block}{Spoofax perspective}
 		Developing a language with a complex type system tests the boundaries of what Spoofax can do.
@@
 
-	\begin{block}{From the perspective of Dependent Types research}
-		A rapid prototyping platform.
+	\begin{block}{Dependent Types perspective}
+		Using a language workbench helps with rapid prototyping.
 	\end{block}
@@
 	% - Spoofax provides an easy way to get a parser, type checker, etc
+\end{frame}
+
+\begin{frame}{Primary Contribution}
+	todododododo
 \end{frame}
@@
 	\begin{block}{How do we use scopes?}
+		A scope is used as a combination of an environment and a context.\\
 		One relation \texttt{name $\rightarrow$ NameEntry}, NameEntry is either:
 		\begin{itemize}
-			\item NType: Stores a type
-			\item NSubst: Stores a substitution
+			\item NType: Stores a type -> Corresponds with context
+			\item NSubst: Stores a substitution -> Corresponds with environment
 		\end{itemize}
@@
 	\begin{exampleblock}{Example 1}
-		\texttt{let T = Bool;\\let b: T = true;}
+		\texttt{let T = if false then Int else Bool end;\\let b: T = true;}
 	\end{exampleblock}
@@
 
-\begin{frame}[fragile]{Type Checking: Rules}
+\begin{frame}[fragile]{Type Checking: From inference rules to Statix code}
+	todododododod From inference rules to Statix code
+
 	\begin{figure}
@@
 \begin{frame}[fragile]{Extra contributions}
+	% Refer to thesis for this
+Features
 \begin{enumerate}
@@
 	\item Compiler to Clojure
+\end{enumerate}
+Evaluation
+\begin{enumerate}
 	\item Comparison with implementation in Haskell
@@
 	Spoofax is a great tool for developing dependently typed languages!\footnote{But there is still room for improvement}
+
+	- using scopes as env+context 
+	todo beperkingen noemen 
+	- statix enforces syntactic uniqueness of metavariable solutions but dependently typed languages use a weaker notion of equality 
+	- full thesis coming soonTM
 \end{frame}

### 2023-03-17 pres-evcs/presentation.tex
add note
Update presentation

@@
 \large{How suitable is Statix for defining a dependently-typed language?}
+\begin{itemize}
+	\item Defining system F was a challenge\footnote{Hendrik van Antwerpen et al. Scopes as types.}, will it even be possible?
+	\item Will it be easier than doing it in Haskell?
+\end{itemize}
+
+% Mention system f (polymorphic lambda calculus) being hard (scopes as types)
 % Statix is turing complete, so it's going to be possible
@@
 
-\begin{frame}{Primary Contribution}
-	todododododo
-\end{frame}
-
-\begin{frame}[fragile]{Calculus of Constructions}
+\begin{frame}[fragile]{Primary Contribution: Calculus of Constructions in Statix}
 	A lambda calculus with dependent types.
@@
 	\begin{block}{How do we use scopes?}
-		A scope is used as a combination of an environment and a context.\\
+		A scope is used as a replacement for an environment and a context..\\
 		One relation \texttt{name $\rightarrow$ NameEntry}, NameEntry is either:
 		\begin{itemize}
-			\item NType: Stores a type -> Corresponds with context
-			\item NSubst: Stores a substitution -> Corresponds with environment
+			\item NType: Stores a type (Corresponds to a context)
+			\item NSubst: Stores a substitution (Corresponds to an environment)
 		\end{itemize}
@@
 \begin{frame}[fragile]{Type Checking: From inference rules to Statix code}
-	todododododod From inference rules to Statix code
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\linewidth]{img/screenshot004}
+\end{figure}
 
@@
 	\centering
-		\includegraphics[width=0.5\linewidth]{img/screenshot001}
-		\label{fig:screenshot001}
-	\end{figure}
-	\begin{figure}
-		\centering
-		\includegraphics[width=0.5\linewidth]{img/screenshot002}
-		\label{fig:screenshot002}
+	\begin{exampleblock}{Equivalent Statix code} 
+		\texttt{typeOfExpr (s, Let(x, e, b)) = bt :-\\
+			\,\, typeOfExpr (s, e) == et, \\
+			\,\, typeOfExpr (sPutSubst (s, x, (s, e)), b) == bt}
+	\end{exampleblock}
 \end{figure}
@@
 \begin{frame}[fragile]{Conclusions}
-	Spoofax is a great tool for developing dependently typed languages!\footnote{But there is still room for improvement}
+	Spoofax is a great tool for developing dependently typed languages!
+	\begin{itemize}
+		\item We can use scopes to represent environments and contexts
+		\item Statix can still use improvements
+	\end{itemize}
 
-	- using scopes as env+context 
-	todo beperkingen noemen 
-	- statix enforces syntactic uniqueness of metavariable solutions but dependently typed languages use a weaker notion of equality 
-	- full thesis coming soonTM
+	%- using scopes as env+context  
+	%- statix enforces syntactic uniqueness of metavariable solutions but dependently typed languages use a weaker notion of equality 
+	%- full thesis coming soonTM
 \end{frame}

### 2023-03-20 paper-thesis/chapters/10-coc.tex
Update presentation
Work a bit on thesis

@@
 \begin{lstlisting}
-typeOfExpr(s, Let(n, v, b)) = typeOfExpr(s', b) :-
-    typeOfExpr(s, v) == vt, sPutSubst(s, n, (s, v)) == s'.
+typeOfExpr (s, Let(x, e, b)) = bt :-
+	typeOfExpr (s, e) == et, 
+	typeOfExpr (sPutSubst (s, x, (s, e)), b) == bt.
 \end{lstlisting}
+
+Each premise in the inference rule is a premise in the Statix code, and the conclusion of the inference rule is the declaration of the Statix rule.
 

### 2023-03-20 paper-thesis/chapters/11-variable-capture.tex
Work a bit on thesis

@@
 \end{lstlisting}
-The implementation so far would tell you it is \verb|T : Type -> T : T -> T| \todo{why is this wrong}.
+The implementation so far would tell you it is \verb|T : Type -> T : T -> T|.
 Given the scoping rules of the language, that is equivalent to \verb|T : Type -> x : T -> x|.
-However, the correct answer would be \verb|T : Type -> x : T -> T|.
+Note that \verb|x| is not a type, so this does not even type check.
+The correct answer would be \verb|T : Type -> x : T -> T|.
 There is no way of expressing this type without renaming a variable.

### 2023-03-20 paper-thesis/chapters/21-adding-inference.tex
Add todo
Work on CoC chapter
Add chapter transitions
Work a bit on thesis

@@
 Inference is an important feature of dependent programming languages, that allows redundant parts of programs to be left out.
-For example, it allows you to infer the arguments of a function, if they can be inferred by the arguments that follow, like here where the type of the argument can be inferred, since we pass it \verb|true| which is a boolean:
+For example, it allows you to infer the argument type of a function, if it can be inferred by the usage of the function, like in this example where the type of the argument of the polymorphic identity function can be inferred, since we pass it \verb|true| which is a boolean:
 \begin{lstlisting}
@@
 
-We call it \emph{term inference} rather than \emph{type inference}, because we can infer values other than types. For example, it can infer that the \verb|_| in this example must be \verb|true|:
+The syntax we chose is that if we want a value to be inferred, we replace it with an underscore.
+Other languages such as Agda allow for \emph{implicit arguments}, which are arguments that can be left out, which will then be inferred~\cite{agda}~\cite{coq}.
+This is just syntactic sugar over the \emph{placeholders} we implement.
+
+Furthermore, we call the inference algorithm \emph{term inference} rather than \emph{type inference}, because it can infer values other than types.
+For example, it can infer that the placeholder in this example must be \verb|true|:
 \begin{lstlisting}
@@
 
-\section{Different algorithms for inference}
+\section{Different Algorithms for Inference}
 \label{strength-inference}
@@
 
-\section{Analysis of the approximation}
+\section{Analysis of the Approximation}
 \label{analysing-inference}
@@
 \end{lstlisting}
+
+\todo{Mention Equational Unification as a solution}
+
+We have now shown how to add inference to the implementation. Next, we will add Inductive Datatypes (chapter~\ref{ch:datatypes}) and Universes (chapter~\ref{ch:universes}) to the implementation.

### 2023-03-20 paper-thesis/chapters/21-adding-inference.tex
Improve more text

@@
 One algorithm for unification is \emph{first-order unification} (FOU), where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the the value free variable to be equal to the value of the non-free variable.
-There are some situations in which this approach fails, but in most real-world scenarios it works perfectly.
+There are some situations in which this approach fails to infer a term, but in most real-world scenarios it works very well.
 For example, it can infer both programs in the introduction of this chapter, but it fails to infer the following program:
@@
 
-A more powerful algorithm that could solve this is \emph{higher-order pattern unification} as defined by Miller~\cite{Miller89}, which is implemented in Agda.
+A more powerful algorithm that could solve this is \emph{higher-order pattern unification} as defined by Miller~\cite{Miller89}, which is implemented in Agda~\cite{agda}.
 Implementing this in Statix is theoretically possible\footnote{Since Statix is turing-complete}, but would be quite difficult to do in practice.
@@
 We would like to avoid implementing an algorithm at all, instead using Statix' built-in first-order unification to do the type inference for us.
-Implementing an inference algorithm in Statix is theoretically possible but this would be a lot of ugly code (since Statix is not a general-purpose programming language), and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.
+Implementing an inference algorithm in Statix is theoretically possible but this would not be a clean implementation (since Statix is a domain specific language that is not meant to implement these algorithms), and the goal is to use Statix in a way that is clean and declarative, not to do optimal inference.
 
 However, we cannot immediately use Statix' built-in first-order unification (which acts in the meta language, Statix) to implement first-order unification in the object language.
-Ideally when implementing beta equality we would match on $\beq{e_1}{e_2}$ where $e_1$ is a free variable, but Statix does not allow for querying whether variables are free.
+Ideally when implementing beta equality when we encounter $\beq{e_1}{e_2}$ we would check if $e_1$ is a free variable, but Statix does not allow for querying whether variables are free.
 
@@
 
-\section{Implementing AFOU} \todo{Do we want all these examples in Statix or math notation?}
+\section{Implementing AFOU}
 \label{implementing-inference}
 
-First, we introduce a new constructor \verb|Infer : Expr -> Expr|, which denotes the variables which could be free.
-The constructor is introduced when we encounter a \verb|_| variable, a marker that something needs to be inferred.
+First, we introduce a new constructor \verb|Infer : Expr -> Expr|, which denotes the variables which could be free. The constructor is introduced when we encounter a placeholder.
 \begin{lstlisting}
@@
 
-Note that the type of \verb|typeOfExpr| has changed, it now returns two expressions, the first being the same expression that was passed in except with \verb|_| variables replaced with \verb|Infer| constructors, and the second being the type.
+Note that the type of \verb|typeOfExpr| has changed, it now returns two expressions, the first being the same expression that was passed in except with placeholders replaced with \verb|Infer| constructors, and the second being the type.
 \begin{lstlisting}
@@
 
-The only difference between AFOU and FOU is case \ref{inf-case2} (for functions) and \ref{inf-case4}.
+The only difference between AFOU and FOU is case \ref{inf-case2} (for functions) and case \ref{inf-case4}.
 
-In case \ref{inf-case2}, this will only infer the value if it has not been inferred yet.
-This means that if the same value is inferred from two different places, this will infer two different values and fail.
+In case \ref{inf-case2}, we force the name of the function constructors to be equal.
+Ideally, we would generate a fresh name for the variable if and only if no name has been generated for it yet, but there is no way to tell whether this is the case so we use this approximation.
 
@@
 
-\todo{Mention Equational Unification as a solution}
+We could solve this problem in two ways:
+
+\begin{enumerate}
+	\item We could always beta-reduce terms completely to their normal form, then terms that are beta-equal are also identical up to alpha equality.
+This would solve the problem in case \ref{inf-case4}, but the problem of case \ref{inf-case2} would remain.
+Furthermore, doing all this beta reduction is terrible for the performance of the type checker: Beta-reducing terms may take an arbitrary amount of time, though it is guaranteed to terminate since the Calculus of Constructions is strongly normalizing~\cite{Coquand_Huet_1988}.
+	\item Ideally, we want a declarative definition of the problem.
+To do this, we need to be able to declare that two terms that are not identical are equal, which is possible under \emph{equational unification}~\cite[Section 2.1]{Siekmann86}.
+If this system was added to Statix, this would make implementing more powerful inference algorithms cleanly possible.
+\end{enumerate}
 

### 2023-03-20 paper-thesis/chapters/22-adding-datatypes.tex
Improve more text

@@
 
-Another useful feature is support for inductive data types. An example of a simple inductive datatype is the \verb|Nat| type:
+Another useful feature is support for inductive data types.
+Inductive data types declare types that are defined by a list of constructors that produce the type.
+An example of a simple inductive datatype is the \verb|Nat| type:
 \begin{lstlisting}
@@
 \end{lstlisting}
+The data type has two constructors: \verb|zero| represents the natural number zero, \verb|suc| takes a natural number and represents the successor of that number.
+For example, \verb|suc (suc zero)| represents 2.
+
 We will design data types to have the same features as in Agda:
@@
 \item Data types may be recursive, that is the data type may take itself as one of the constructor arguments. This can be seen in the definition of \verb|Nat| above.
-\item Parameters, which are datatypes that are polymorphic, and are required to be the same for all constructors, such as
+\item Data types may have \emph{parameters}, which are datatypes that are polymorphic over a certain value, that is required to be the same for all constructors, such as:
 \begin{lstlisting}
@@
 \end{lstlisting}
-\item Indices, which are datatypes that are polymorphic, that may vary from constructor to constructor
+\item Data types may have \emph{indices}, which are datatypes that are polymorphic over a certain value, that may vary from constructor to constructor, such as:
 \begin{lstlisting}
@@
 \end{lstlisting}
-\item We can also combine parameters and indices, for example to create the generic \verb|Eq| type. Note that parameters and indices may also depend on earlier parameters and indices.
+\item We can also combine parameters and indices, for example to create the polymorphic \verb|Eq| type, which represents a proof that two values are equal.
+Note that the type of parameters and indices may depend on the value earlier parameters and indices.
+For example, \verb|e1| and \verb|e2| have type \verb|T|.
 \begin{lstlisting}
@@
 
-A difference between Agda and our implementation is that Agda has case matching as a native construct, whereas we chose to use eliminators~\cite{eliminators}.
+A difference between Agda and our implementation is that Agda has case matching as a native construct, whereas we chose to use eliminators as described in Inductive families~\cite{eliminators}.
 An eliminator is a function that can be used to case match on a data type.
@@
 parameters 
--> P: (indices -> N params indices -> Type) 
+-> P: (indices -> v: N params indices -> Type) 
 -> constructors
@@
 \begin{itemize}
-	\item \verb|parameters| are the parameters of the data type we want to eliminate, since these are constant among all constructors those can be at the start.
+	\item \verb|parameters| are the parameters of the data type we want to eliminate, since these are constant among all constructors those should be specified at the start.
 	\item \verb|P| is the type that this eliminator will return. It may depend on the value \verb|v| that is eliminated.
@@
 \label{ch:datatypes:typechecking}
-
+In this section we will explain how to type check data type declarations using Statix.
 The definition of \verb|type_check| for an inductive data type \verb|N| is, conceptually:
@@
 Call the end of this chain \verb|s3|.
+The scope \verb|s3| is not used in the rest of this process, but creating this chain is required to check the types of the indices.
 
@@
 \end{enumerate}
-For example, for the following inductive datatype:
+For example, figure \ref{scope-graph-eq} shows the scopes that were created during the type checking of the following inductive datatype:
 \begin{lstlisting}
@@
 	refl : e : T -> Eq e e;
-...
 \end{lstlisting}
-Figure \ref{scope-graph-eq} shows the scope graph generated by this data type.
 \begin{figure}[!ht]
@@
 
-The type of an eliminator was discussed in section \ref{ch:datatypes:intro}. When we type check an \verb|elim N| expression, the following happens:
+This section discusses how to type check eliminators.
+The type of an eliminator was already discussed in section \ref{ch:datatypes:intro}.
+When we type check an \verb|elim N| expression, the following happens:
 \begin{enumerate}
@@
 This gives us access to the parameters, indices and constructors of the data type that is being eliminated.
-	\item Using some tedious logic, create the type discussed in section \ref{ch:datatypes:intro}. This ends up being around 100 lines of Statix code.
+	\item Using some tedious but conceptually simple Statix rules, create the type discussed in section \ref{ch:datatypes:intro}. This ends up being around 100 lines of Statix code.
 \end{enumerate}
@@
 
+This is all that needs to happen to define inductive data types. Next, we will add positivity checking.
+
 \section{Positivity Checking}
@@
 
+To show how this is can go wrong, consider the following example:
+\begin{lstlisting}
+data False : Set where
+
+data Bad : Set where
+	bad : (Bad → False) → Bad;
+
+self-app : Bad → False
+self-app (bad f) = f (bad f)
+
+absurd : False
+absurd = self-app (bad self-app)
+\end{lstlisting}
+
+Under the curry-howard correspondence, \verb|Bad| states that \verb|Bad| is true if and only if \verb|Bad| implies false.
+Since such a statement is inconsistent, we can use it to prove \verb|False|.
+To prevent such statements, we don't allow inductive datatypes to refer to themselves negatively, checking this is called \emph{positivity checking}.
+This solution is not complete (it rejects valid data types), but it is the solution most often used in practice, such as in Agda~\cite{agda}, Lean~\cite{lean} and Coq~\cite{coq}.
+
+The exact condition that is checked is that if a constructor has a function as its argument, then the argument type (\verb|A| in the example below) can not refer to the declared data type.
+
 \begin{lstlisting}
@@
 	--     A       B       C
-	-- A is in a negative position, B and C are OK
 \end{lstlisting}
 
-The implementation is simply an extra check during the type checking of data type declarations. It checks that the datatype only refers to itself in positive positions.
+The implementation is simply an extra check during the type checking of data type declarations, which is straight-forward to implement in Statix.
 
-We have now shown how to add inductive datatypes to the implementation. Next, we will Universes (chapter~\ref{ch:universes}) to the implementation.
+We have now shown how to add inductive datatypes to the implementation. Next, we will add  Universes (chapter~\ref{ch:universes}) to the implementation.

### 2023-03-20 paper-thesis/chapters/23-adding-universes.tex
Update 23 and 24

--- 2023-02-08 paper-thesis/chapters/23-adding-universes.tex
@@
 Our language so far has the property that the type of \verb|Type| is \verb|Type|, this is called \emph{type in type}.
-This allows for \emph{Girard's paradox}, which means that the logic created by this language is inconsistent: we can prove false.
+This allows for \emph{Girard's paradox}, which means that the logic created by this language is inconsistent: we can prove false~\cite{girard}.
 
@@
 
-The solution to this problem is introducing universes into the language. We now have the following hierarchy of types:
+The solution to this problem is to introduce universes into the language. We now have the following hierarchy of types:
 \begin{lstlisting}
@@
 
-\section{Implementing Universes in the CoC}
+\section{Implementing Universes in the Calculus of Constructions}
 
-The constructor of Type has been changed to \verb|Type : int -> Expr|.
-We need to update all occurrences of \verb|Type| in figure \ref{fig:type-check-rules}.
+The \verb|Type| constructor has to be changed to \verb|Type : int -> Expr|, taking the universe of the type as an argument.
+We then need to update all occurrences of \verb|Type| in figure \ref{fig:type-check-rules}.
 Figure \ref{fig:type-check-rules-universes} contains all the rules that require a change.
@@
 \begin{enumerate}
-	\item The rule for type-checking a \verb|Type| expression, which now is explicit about the universes of each \verb|Type|.
-	\item The rule for type-checking \verb|FnType|, which takes the maximum of the two arguments.
-	\item The rule for type-checking \verb|FnConstruct|, which ignores the universe of its argument type, and only verifies that it is a type.
+	\item The rule for type-checking a \verb|Type| expression. The universe of \verb|Type u| is \verb|u + 1|. 
+	\item The rule for type-checking \verb|FnType|. The universe of \verb|FnType| is the maximum of the universe of the argument and return type.
+	\item A trivial change to the rule for type-checking \verb|FnConstruct|, which ignores the universe of its argument type.
 \end{enumerate}
@@
 		} {
-			\toe{s}{\co{FnType}(x, a, b)}{\co{Type}(max(u_a, u_b))}
+			\toe{s}{\co{FnType}(x, a, b)}{\co{Type}(\co{max}(u_a, u_b))}
 		}
@@
 \end{enumerate}
+
+We have now finished the type checker of the language. Next, we will discuss how to implement the back end of the language.

### 2023-03-20 paper-thesis/chapters/24-backend.tex
Update 23 and 24

@@
 \section{Interpreter}
-It is valuable to define an interpreter for our language, because it is a way to run a program without relying on the definition of the language you are compiling to.
-For a dependently typed language specifically it is easy to add on, as we already needed to define the dynamic semantics of our language for type checking anyways.
+It is valuable to define an interpreter for our language, because it is a way to run a program without relying on another language to compile to.
+For a dependently typed language specifically, it is easy to develop an interpreter, as we already needed to define the dynamic semantics of our language for type checking anyways.
 
@@
 
+We defined the dynamic semantics in Statix, which comes with some advantages and disadvantages over defining it in Stratego.
+
 An advantage of implementing the dynamic semantics in Statix instead of Stratego is that we can use scope graphs as a part of our dynamic semantics rules.
+As discussed in chapter~\ref{chap:baselang}, scope graphs are a natural way to store the environment of the language during evaluation.
+When dynamic semantics is defined in Stratego, a different representation of the environment would be needed.
 
-A disadvantage is that Stratego is specifically optimized for these transformations~\cite{fcpmc}. Therefore, the resulting interpreter will be slower when implemented in Statix. \todo{comparison}
+A disadvantage is that Stratego is a domain specific language for transformations, and it has some features that would make the definitions of beta reductions shorter.
+For example, to define \verb|betaReduce| in terms of \verb|betaReduceHead| we need to match the constructor and apply \verb|betaReduceHead| recursively.
+Stratego has the \verb|topdown| traversal strategy that does the same thing without needing to be explicitly defined on all constructors.
+
+Another disadvantage is that Stratego is specifically optimized for these transformations, for example as discussed in "Optimising First-Class Pattern Match Compilation"~\cite{fcpmc}.
+Therefore, the resulting interpreter will be slower when implemented in Statix.
+This is however just a constant factor overhead, it does not affect the time complexity of the algorithm.
 
 \section{Compiler}
-The alternative to writing an interpreter is writing a compiler to another language.
+An alternative to writing an interpreter is writing a compiler to another language.
 In order to explore this avenue, we wrote a compiler to Clojure, a dynamically typed dialect of Lisp that runs on the JVM~\cite{clojure}.
@@
 
+The compiler is created while keeping in mind the Agda to Scheme compiler, available at \url{https://github.com/jespercockx/agda2scheme}.
+This is also a compiler from a dependently typed language (Agda) to a LISP (Scheme).
+We make some of the same choices as this compiler, as these choices have been proven to work.
+
+\subsection{Calculus of Constructions}
 Both languages support first-class functions, so we can map most constructs of our language directly onto constructs of Clojure.
-Therefore, most of our compiler has rules that look like the example below, where the constructors are compiled recursively using the \verb|compile-expr| strategy, ignoring the specified types.
+Therefore, most of our compiler rules look like the example below, where the constructors are compiled recursively using the \verb|compile-expr| strategy, ignoring the specified types.
 The constructors starting with \verb|C| are constructors from Clojure.
@@
 Since our language has no way to match on types, types can actually be completely eliminated during type checking.
-The easiest way to do this is to compile all types to an arbitrary value (we chose the integer 0), since the value is irrelevant anyways.
+The easiest way to do this is to compile all types to an arbitrary value (we chose the string "TYPE"), since the value is irrelevant anyways.
 A more efficient solution could be to implement type erasure~\cite[Section 23.6]{tapl}, where type-level functions are completely removed.
 We decided against this as the goal of this section is to show that we can compile the language with ease, not to write an efficient compiler.
+This optimization was also not implemented in the Agda to Scheme compiler.
 
--analysis is hard, agda doesn't do it
-- if compiled language doesn't matter
-- erasure annotations (idris/agda)
+\subsection{Inductive Data Types}
 
 Finally, inductive data types are a bit harder to compile, as Clojure does not have an equivalent to sum types on which we can match. \todo{finish this after implementing}
-- agda to scheme compiler
 - atoms voor constructors

### 2023-03-20 paper-thesis/chapters/12-editor-services.tex
More changes to the text

@@
 
+When a language is implemented in Spoofax, we give a declarative definition of the language.
+Spoofax then generates the parser, type checker and compiler for us.
+An additional benefit of these declarative definitions is that Spoofax can use them to generate editor services, such as \verb|go to definition|, \verb|renaming| and \verb|semantic code completion|.
+
 We explored how semantic code completion presented by Pelsmaeker et al.~\cite{codecompletion} applies to dependently typed languages.
 Code completion is an editor service in IDEs that proposes code fragments for the user to insert at the caret position in their code.
+We chose to explore code completion specifically because deciding if a fragment is relevant to propose requires reasoning about the types of the fragments, which may be more difficult in the case of dependent types.
+We will show that Spoofax can use the declarative definition correctly and that it has no problem with semantic code completion for dependently typed languages.
 
 \section{Setup required}
-To set up editor services, we followed the steps in the Spoofax documentation \cite{semcomplet}. To be precise, we followed the following steps:
+To set up editor services, we followed the steps of the "How to Enable Semantic Code Completion" guide of the Spoofax documentation \cite{semcomplet}. To be precise, we followed the following steps:
 \begin{enumerate}
@@
 \end{lstlisting}
-This would return the following suggestions: (Note that \verb|f| and \verb|Type| are not suggested, since they cannot have type \verb|Bool|)
+Requesting suggestions from spoofax for this placeholder gives the following suggestions:
  \begin{itemize}
@@
 
+Note that \verb|f| and \verb|Type| are not suggested, since they cannot have type \verb|Bool|.
+Function application (the \verb|[[Expr]] [[Expr]]| suggestion) is suggested since functions can return booleans, similar to let bindings.
+We have tried quite a few cases and we conclude that semantic code completion works well for dependently typed languages.
+
 We have now discussed semantic code completion. Next, we will compare the implementation with one in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi}).

### 2023-03-22 pres-evcs/presentation.tex
Work a bit on thesis
Update presentation

@@
 	\end{exampleblock}
-% - n is a RUNTIME value, it does not have to be known at compile time
 % - n occurs insie a type `Vec`!
@@
 \begin{itemize}
+	\item Will it be easier than doing it in Haskell?
 	\item Defining system F was a challenge\footnote{Hendrik van Antwerpen et al. Scopes as types.}, will it even be possible?
-	\item Will it be easier than doing it in Haskell?
 \end{itemize}
@@
 	\begin{exampleblock}{Example 1}
-		\texttt{($\backslash$v: Type. v) T}
+		\texttt{($\backslash$v: Type. v) Type}
 	\end{exampleblock}
@@
 	\end{exampleblock}
+
+	% most simple dependently typed language: lambda calculus
+	% what other languages use as a core
 \end{frame}
@@
 	\begin{block}{How do we use scopes?}
-		A scope is used as a replacement for an environment and a context..\\
-		One relation \texttt{name $\rightarrow$ NameEntry}, NameEntry is either:
+		A scope is used as a replacement for an environment and a context.\\
+		One edge \verb|p|. One relation \texttt{name $\rightarrow$ NameEntry}, NameEntry is either:
 		\begin{itemize}
@@
 		\item Statix can still use improvements
+		% - Higher order 
 	\end{itemize}

### 2023-03-27 paper-thesis/chapters/24-backend.tex
More changes to the text
More writing work from today

@@
 
-Finally, inductive data types are a bit harder to compile, as Clojure does not have an equivalent to sum types on which we can match. \todo{finish this after implementing}
+Finally, inductive data types are a bit harder to compile, as Clojure does not have an equivalent to sum types on which we can match.
+Instead, we compile inductive datatypes to a pair (constructor, list of constructor arguments).
+
+\todo{finish this after implementing}
 - atoms voor constructors
+
+The implementation of the language is now complete, both the type checker and the back-end.
+We will now discuss editor services (chapter \ref{chap:editor-services}) and compare the implementation with implementations in Haskell (chapter \ref{ch:comp-haskell}) and LambdaPi (chapter \ref{ch:comp-lambdapi}).

### 2023-03-27 paper-thesis/chapters/30-comparing-with-haskell.tex
More writing work from today

@@
 In this chapter, we implement the language defined in chapter~\ref{chap:baselang} in Haskell, and then compare the implementation with the implementation in Statix.
-We want the design of the implementation in Haskell to be similar to conventional implementations of dependently typed languages, so we can compare the Statix implementation with them.
+We want the design of the implementation in Haskell to be representative of conventional implementations of dependently typed languages, so we can use it to compare the Statix implementation with them.
 
 \section{Defining the AST}
-To implement the calculus of constructions in Haskell, we first define the \verb|Expr| datatype.
+To implement the calculus of constructions in Haskell, we first define the inductive datatype \verb|Expr|.
 We chose to define the language using De Bruijn indices, since this is the convention when implementing dependently typed languages, and the goal of this implementation in Haskell is to compare the Statix implementation with conventional ones.
@@
 \end{lstlisting}
+There is a constructor for each constructor in our language.
+The difference with the implementation in Spoofax is that the \verb|Let|, \verb|FnType|, and \verb|FnConstruct| no longer store an identifier, instead they introduce a new binding into their body, which are accessible by their De Bruijn indices.
 
@@
 \begin{tikzpicture}[scopegraph, node distance = 2.5em and 2.5em]
-	\node[scope] (s1) {$s1$};
+	\node[scope] (s0) {$s0$};
+
+	\node[scope, right of = s0, xshift=5em] (s1) {$s1$};
 	\node[type, above of = s1] (s1l) {$\texttt{T : Type()}$};
@@
 
+	\draw (s1) edge[] (s0);
 	\draw (s2) edge[] (s1);
@@
 
+\section{Defining Type Checking}
+
+Next we need to define type checking. We define a function \verb|tc|:
+\begin{lstlisting}
+tc :: SExpr -> Either String Expr
+\end{lstlisting}
+
+The function returns a \verb|String| in case of an error, or a \verb|Expr| which is the type of the expression.
+This is different than in Statix, where errors are automatically created during type checking by Statix.
+We use the power of the \verb|Either String| monad to our advantage, so we can define our function using \emph{do blocks}.
+For example, compare the implementation of \verb|Let| in figure \ref{fig:type-check-rules} (repeated below for convenience) with the implementation in Haskell:
+
+\begin{lstlisting}
+tc (s, Let e b) = do
+	_ <- tc (s, e)
+	bt <- tc (sPutSubst s (s, e), b)
+	pure (shift (-1) 0 bt)
+\end{lstlisting}
+
+\begin{mathpar}
+		\inferrule{
+	\toe{s}{e}{t_e}
+	\\ \toe{\co{sPutSubst}(s, x, (s, e))}{b}{t_b}
+}{
+	\toe{s}{\co{Let}(x, e, b)}{t_b}
+}
+\end{mathpar}
+
+We are using do blocks to pass errors through the type checking function.
+The only significant difference with the inference rule is that we need to shift the resulting type by one, since the binding of the let-bound variable does not exist at the type level.
+This shifting is required for variables too, other than this the inference rules are converted directly to cases of the Haskell function.
+
 \section{Defining Beta Reduction}
 
-Now we want to define beta head reduction. Remember that in Statix, beta head reduction is a relation with the signature
+During type checking evaluation may be needed, so we need to define beta reduction. Remember that in Statix, beta head reduction is a relation with the signature:
 \begin{lstlisting}
@@
 
+Now we can define beta head reduction as cases of \verb|brh_|, returning \verb|Nothing| if the expression cannot be reduced, in which case \verb|brh| will take the original expression.
+The cases correspond directly to the cases of figure \ref{fig:beta-head-reduce-rules}.
+
 \section{Comparison}

### 2023-03-27 paper-thesis/chapters/31-comparing-with-lambdapi.tex
More writing work from today

@@
 
-There exist several \emph{logical frameworks}, tools designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}.
+In chapter \ref{ch:inference} we discussed how to implement inference algorithms, and we concluded in section \ref{analysing-inference} that adding equational unification to Statix would make inference algorithms easier to implement.
+In this chapter we explore this further, by implementing our dependently typed language in a logical framework.
+
+A logical framework is a language that provides a means to define a type system.
+There exist several \emph{logical frameworks}, designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}.
 Since these tools are designed specifically for the task, implementing the type system takes less effort in them compared to Spoofax, but for other tasks such as defining a parser or editor services they are not as well equipped.
@@
 
-Note that these types can only represent type-correct terms, \emph{intrinsically typed} terms.
+Note that these constructors can only represent type-correct terms, \emph{intrinsically typed} terms.
 This is useful because it means the meta language does the type checking for us.
@@
 These could easily be automatically generated by a type checker but are tedious to specify manually.
-
-It seems like it should be possible to infer these, but lambdapi fails to infer even the simplest ones. \todo{not sure why}.
+It seems like it should be possible to infer these, but lambdapi fails to do so.
+Further research is required to find out why lambdapi does not infer these.
 
@@
 
-We define reduction rules to reduce the symbols we defined in the previous section into lambdapi.
+We now define the \emph{reduction rules}, which define how to reduce the language.
+A rule $a \hookrightarrow b$ states that \verb|a| reduces to \verb|b|.
+We reduce the constructs in our language to constructors in lambdapi, this is called \emph{embedding} the language.
 Lambdapi can then evaluate the rules using its own semantics, not requiring any additional rules similar to the ones in figure \ref{fig:beta-head-reduce-rules}.

### 2023-03-27 paper-thesis/chapters/31-comparing-with-lambdapi.tex
Improve more text

@@
 Lambdapi can then evaluate the rules using its own semantics, not requiring any additional rules similar to the ones in figure \ref{fig:beta-head-reduce-rules}.
+It is also possible to define the rules in \ref{fig:beta-head-reduce-rules} using reduction rules, but this way of defining reduction rules is much simpler.
 The reduction rules are given in figure \ref{fig:lp-rules}.
@@
 One program which would not type-check with this approach is the following.
-It fails to compile since it cannot know that \verb|b| is a boolean, which is required by the definition of f.
-\footnote{Assuming we introduce booleans into the language, there are examples that don't require booleans but they are a bit more difficult to understand}.
+It fails to compile since it cannot know that \verb|b| is a boolean, which is required by the definition of f \footnote{Assuming we introduce booleans into the language, there are examples that don't require booleans but they are a bit more difficult to understand}.
 \begin{lstlisting}
@@
 \end{lstlisting}
+
+We could solve this problem by not defining our program as an embedding, instead defining it using the reduction rules in figure \ref{fig:beta-head-reduce-rules}.
+This results in a definition very close to the one in Statix, except that instead of a beta reduction relation we use the reductions built-in to the language, which solves the problem.
+
+We've now discussed equational unification using lambdapi, which is a feature that we suggest could improve Statix. Other possible improvements are discussed in chapter \ref{ch:ergonomics}.

### 2023-03-27 paper-thesis/chapters/40-discussing-statix.tex
Improve more text

@@
 \begin{enumerate}
+	\item Allow specifying reduction rules, and then implement equality under reduction in Statix. This feature was discussed extensively in chapter \ref{ch:comp-lambdapi}.
+
 	\item Nested constraints: Defining a constraint that can use the metavariables of a parent constraint. This works similar to nested function definitions. For example, for type checking datatypes:
@@
 	The definition of \texttt{isStrictlyPositive} can then use all the metavariables of \\ \texttt{DataTypeDecl} without having to explicitly pass them.
-
-	\item In order to implement more powerful inference algorithms, it would be useful to have a mechanism to check if metavariables have been instantiated or not.
-Since the order in which Statix evaluates constraints is so dynamic, we're not entirely sure how this mechanism would be implemented.
-More research is needed on this.
 

### 2023-03-28 paper-thesis/chapters/01-introduction.tex
More writing of the future work

@@
 \section*{Contributions}
-This goal of this thesis is to investigate how well Statix is fit for the task of defining a dependently-typed language.
+The primary contribution of this thesis is to investigate how well Statix is fit for the task of defining a dependently-typed language.
 The goal is not only to show that Statix can implement it, but also that the implementation is more concise than in a general purpose language.
@@
 However, this language is not very practical to write actual complex programs in, so to improve this situation we will extend the language with typical features of dependently typed languages.
-We will do this in the following steps:
+We will provide the following additional contributions:
 \begin{itemize}

### 2023-03-28 paper-thesis/chapters/98-relatedwork.tex
More writing of the future work

@@
 
-In chapter \ref{ch:comp-haskell} and \ref{ch:comp-lambdapi} we already compared our implementation with two different implementations. In this chapter we discuss remaining related work.
-TODO TODO TODO
+We now discuss some work that is related to the topics in this thesis.
+This thesis implements a dependently typed language in Statix.
+We compare this implementation with implementations of other languages in Statix, and with implementations of other dependently typed languages.
 
@@
 
-The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope.
+The implementation in this paper requires performing substitutions in types immediately, as types don't have a corresponding scope to store substitutions.
 Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types}~\cite{nameres} present an implementation of System F in Statix that does lazy substitutions, by using scopes as types.
-It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can be arbitrary terms.
+Each scope represents a type, and scopes have edges to other scopes in case of quantifications and substitutions.
+Further research is needed to see if this approach could also apply to dependently typed languages, where types can contain arbitrary terms.
 
@@
 Construction Language Implementations~\cite{Rouvoet21} describes how to create a declarative specification of a language in Statix, and then to use a dependently typed language to integrate the specification of well-typing in the representation of the program that is being interpreted or transformed.
-The work from this thesis and ours could be combined, by using our dependently typed languages to do this verification.
+The work from this thesis and ours could be combined, by using our dependently typed language implemented in Statix, to do this verification, as another Statix DSL.
+This could make the presented approach integrate better with the ecosystem of Spoofax.
 
@@
 
-Below are some dependently typed languages, we discuss what make these languages unique and if these languages could be implemented in Statix:
+Below are some more implementations dependently typed languages, we discuss what make these languages unique and if these languages could be implemented in Statix:
 
@@
 In chapter~\ref{ch:comp-haskell} we compared the implementation in Statix with an implementation in Haskell.
-The pi-forall language~\cite{pi_forall} is a language with a similar complexity to the language presented in this paper.
+The pi-forall language~\cite{pi_forall} is a language with a similar complexity to the language presented in this thesis.
 In principle, the implementations are very similar.
-For example, the inference rules of pi-forall are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper.
-The primary difference is that they use a bidirectional type system~\cite{bidirectional}, whereas this paper uses Statix' unification.
+For example, the type checking rules of pi-forall are similar to the type checking rules presented in figure \ref{fig:type-check-rules} from this paper.
+The primary difference is that they use a bidirectional type system~\cite{bidirectional} to do some inference, whereas this paper uses Statix' unification.
 
-\paragraph{Other}
+\paragraph{Agda}
+
+Agda is one of the most used dependently typed languages~\cite{agda}.
+It is implemented in Haskell, with full editor support for Emacs, VS Code, Atom and Vim.
+Interacting with the language is often done by giving commands to a language server, for example one can press \verb|Ctrl+C Ctrl+A| to do automatic proof search.
+These commands give feedback in a separate view in the editor.
+
+Other dependently typed languages such as Lean and Coq have similar views to interact with the language.
+In general, when working with dependently typed languages there is a lot of these kinds of interactions between the IDE and the developer.
+Spoofax currently does not have support for defining shortcuts or adding these views to Eclipse, but there are no fundamental issues that would prevent this from being implemented.

### 2023-03-28 paper-thesis/chapters/99-conclusion.tex
More writing of the future work

@@
 We implemented this algorithm in a concise way.
+In chapter~\ref{ch:comp-lambdapi} we explore how we could define the language in a language supporting equational unification, adding this to Statix would allows for a more powerful inference algorithm to be cleanly implemented.
 
@@
 While the language as implemented currently is fully usable, there are still some open questions.
+
+\subsection*{Adding more features to the language}
+
+In this thesis we created a language that already has a lot of features: Inductive datatypes, universes, and a basic inference algorithm.
+In chapter~\ref{chap:bools} we also specified the general steps that are needed to extend the language with new constructs.
+However, in comparison with languages such as Agda a lot of features are still missing, among which are:
+
+\begin{itemize}
+	\item Support for universe polymorphism~\cite[Universe Levels]{agda}, which allow functions and inductive data types to be polymorphic over a certain universe level.
+For example, this allows to define a \verb|List| data type that can be both a list of values and a list of types without defining it twice.
+We see no reason why this couldn't be cleanly implemented.
+	\item Another feature of Agda is implementing dependent pattern matching, as described in "Eliminating Pattern Matching without K"~\cite{patternmatching}.
+This is requires a desugaring stage, in which matches are desugared to eliminators, which should be implemented in Stratego before type checking.
+This should also be possible to implement, but it may be quite a bit of work.
+	\item Inference in our language is currently implemented by explicitly specifying a placeholder \verb|_| where inference is required.
+In Agda we can specify an implicit argument, which is automatically inferred unless explicitly provided.
+We see no reason why this couldn't be cleanly implemented.
+	\item Instance Arguments are a special kind of implicit arguments, that are solved by the instance resolution algorithm of Agda~\cite[Instance Arguments]{agda}.
+They can be used similarly to type class constraints, in order to restrict the data types over which a term is polymorphic, for example requiring the terms to have an implementation of equality.
+The instance resolution algorithm should be possible to implementing by traversing the scope graph (since it is linear), and checking whether each let binding is a possible solution similar to the Agda implementation.
+	\item Agda has holes, which allows the user of the language to build proofs interactively.
+One of these interactions is proof search, when triggered Agda will try to find a term that fits the expected type of a hole.
+The interactive part of this was discussed in chapter~\ref{ch:relatedwork}, this is not yet supported by Spoofax.
+After support for this is added, a proof search algorithm could be implemented in Stratego, such as the structurally recursive unification algorithm described by McBride~\cite{proofsearch}.
+	\item Agda allows function arguments to be marked as irrelevant, meaning that the value will be erased at runtime.
+An even stronger version of this feature is implementing a quantitive type system, such as in Idris~2~\cite{idris2}, which combines linear and dependent types.
+Implementing this in Spoofax should be possible, but it may be quite a bit of work as it is a complex type system.
+	\item Agda has mixfix operators, which allows to define custom operators such as \verb|a and b|.
+This requires parsing to happen in multiple stages, where the first stage finds the mixfix operator definitions and the second stage parses expressions using them.
+This results in quite a complex parser, implementing these multiple stages in Spoofax using SDF3 is not possible without fundamental changes to the way that Spoofax works.
+One solution would be using an \emph{adaptive parsing algorithm}, which allows modification of the grammar during parsing.
+More research is needed to explore the possibilities of solving this problem.
+	\item Modules
+	\item Recursion
+\end{itemize}

### 2023-03-29 paper-thesis/chapters/01-introduction.tex
Finish future work section

@@
 
+These contributions are important for two reasons.
+Firstly from the perspective of a Spoofax researcher, developing a language with a complex type system in Statix tests the boundaries of what Statix can do, and how Statix could be further improved to support these usecases better.
+On the other hand, from the perspective of a researcher in dependent type systems, if we can show that Statix provides a good way to quickly implement a dependently typed language, then Spoofax could help with rapid prototyping of dependently typed languages.
+
 \section*{Discussions}
 
-Furthermore, the thesis contains some discussions:
+Furthermore, the thesis contains some discussions about the contributions:
 \begin{itemize}

### 2023-03-29 paper-thesis/chapters/99-conclusion.tex
Finish future work section

@@
 For example, this allows to define a \verb|List| data type that can be both a list of values and a list of types without defining it twice.
-We see no reason why this couldn't be cleanly implemented.
+We see no reason why this couldn't be cleanly implemented in Statix.
 	\item Another feature of Agda is implementing dependent pattern matching, as described in "Eliminating Pattern Matching without K"~\cite{patternmatching}.
@@
 In Agda we can specify an implicit argument, which is automatically inferred unless explicitly provided.
-We see no reason why this couldn't be cleanly implemented.
+We see no reason why this couldn't be cleanly implemented in Statix.
 	\item Instance Arguments are a special kind of implicit arguments, that are solved by the instance resolution algorithm of Agda~\cite[Instance Arguments]{agda}.
@@
 More research is needed to explore the possibilities of solving this problem.
-	\item Modules
-	\item Recursion
+	\item Another thing that could be explored is creating a module system for a dependently typed language in Spoofax.
+At the time of writing, Agda has a module system which is quite powerful (including parameterized modules), but the implementation is quite naive by substituting the modules into the code when imported~\cite[Modules]{agda}.
+Spoofax's scope graphs allow to create a module system with ease, simply by adding extra \emph{import} edges between scopes.
+Future research is needed on if Agda's module system could be implemented using scope graphs.
+	\item Currently the language in the paper does not allow for recursive functions.
+Agda allows for functions that are structurally recursive, and uses termination checking to enforce this property.
+We see no reason why this couldn't be cleanly implemented in Statix, similar to the implementation of positivity checking in chapter~\ref{ch:datatypes}.
 \end{itemize}
+
+\subsection*{Implementing other complex type systems in Statix}
+
+In this paper we have shown that a dependent type system can be implemented in Statix, one of the goals being to show the power of Statix.
+To further show this goal, more different kinds of type systems could be implemented in Statix:
+
+\begin{itemize}
+	\item There has not yet been an implementation of a sub-structural type system in Spoofax.
+A sub-structural type system is a type system that places restrictions on how often variables can be used.
+Now that languages with sub-structural type systems, such as Rust, are becoming popular, we should see if the goal of Statix to cover a broad range of type systems, still holds true for these.
+	\item Effect systems are a formal system that describe the side effects of a program.
+These allow programmers to specify what side effects a function has, which can then be verified by a type checker.
+Researching whether an effect system could be cleanly specified in Statix may lead to more insights on the limits of Statix.
+\end{itemize}

### 2023-04-03 paper-thesis/chapters/000-abstract.tex
First part of feedback from Jesper

@@
-\noindent Static type systems can greatly enhance the quality of programs, but implementing a type checker is challenging and error-prone.
+\noindent Static type systems can greatly enhance the quality of programs, but implementing a type checker for them is challenging and error-prone.
 The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system.
@@
 
-In this thesis, we present an specification of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
+In this thesis, we present a specification of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments.
 While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.

### 2023-04-03 paper-thesis/chapters/01-introduction.tex
First part of feedback from Jesper

@@
 While we keep building more and more complex programming languages, their type checkers are still often written in general purpose languages.
-This takes a lot of effort, and it is easier to make mistakes.
+This takes a lot of effort, and it is easy to make mistakes.
 Instead of writing the type checker in a general purpose language, in this thesis we will systematically derive a type checker from a high-level, declarative specification.
+A declarative specification defines what the semantics of the language are, rather than specifying the imperative steps to type check the language.
 This allows for a cleaner implementation that is easy to extend and maintain.
@@
 \begin{itemize}
-	\item We show the language is easily extendable, by extending it with booleans among other features (Chapter~\ref{chap:bools})
-	\item We show how to add inference of implicit arguments to the implementation (Chapter~\ref{ch:inference})
-	\item We show how to add support for inductive datatypes to the implementation (Chapter~\ref{ch:datatypes})
-	\item We show how to add support for universes to the implementation (Chapter~\ref{ch:universes})
+	\item We show the language is easily extendable, by extending it with booleans among other features (chapter~\ref{chap:bools}).
+	\item We show how to add inference of implicit arguments to the implementation (chapter~\ref{ch:inference}).
+	\item We show how to add support for inductive datatypes to the implementation (chapter~\ref{ch:datatypes}).
+	\item We show how to add support for universes to the implementation (chapter~\ref{ch:universes}).
 \end{itemize}
@@
 \begin{itemize}
-	\item We discuss how well semantic code completion works for our language (chapter~\ref{chap:editor-services})
-	\item We compare our implementation with an implementation of the same language in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi})
-	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (chapter~\ref{ch:ergonomics})
-	\item We discuss related work (chapter~\ref{ch:relatedwork}) and future work (chapter~\ref{ch:conclusion})
+	\item We discuss how well semantic code completion, which is an editor service provided by Spoofax that completes holes in expressions, works for our language (chapter~\ref{chap:editor-services}).
+	\item We compare our implementation with an implementation of the same language in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi}).
+	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (chapter~\ref{ch:ergonomics}).
+	\item We discuss related work (chapter~\ref{ch:relatedwork}) and future work (chapter~\ref{ch:conclusion}).
 \end{itemize}
@@
 
-The source code for this thesis can be found online \\ \url{https://github.com/JonathanBrouwer/master-thesis}
+The source code for this thesis can be found online \\ \url{https://github.com/JonathanBrouwer/master-thesis}.

### 2023-04-03 paper-thesis/chapters/03-background-dependent-types.tex
First part of feedback from Jesper

@@
 
-This chapter will explain the concepts behind dependent types. Readers already familiar with Dependent Types are recommended to only skim through this chapter.
+This chapter will explain the concepts behind dependent types. Readers already familiar with dependent types are recommended to only skim through this chapter.
 
@@
 This means it is possible to have a type such as \verb|Vec n|, which is a vector of exactly length \verb|n|.
-Note that the value of \verb|n| does not have to be known at compile time, it is more powerful than that.
+Note that the value of \verb|n| does not have to be known at compile time, it is therefore more powerful than functions that can be polymorphic over a value.
 
@@
 This append function appends two \verb|Vecs| of equal length, returning a \verb|Vec| of length \verb|n + n|.
-This is an example of a type-level computation.
+This is an example of a type-level expression.
 
-Dependent types are useful because dependent type systems allow for the Curry-Howard Correspondence~\cite{chc}.
-Using this correspondence, types correspond to propositions and a term of a type correspond to a proof of a proposition.
-Thus if our type system is sound\footnote{The type system we will define in chapter~\ref{chap:baselang} is not sound, we will fix this in chapter~\ref{ch:universes}.}, that is, we cannot create a term of an empty type, then we can use our type systems to prove things.
+In dependent type systems we can use the Curry-Howard Correspondence to prove properties of our code~\cite{chc}.
+Under this correspondence, types correspond to propositions and a term of a type correspond to a proof of a proposition.
+Thus if our type system is type-sound\footnote{The type system we will define in chapter~\ref{chap:baselang} is not type-sound, we will fix this by introducing universes in chapter~\ref{ch:universes}.}, that is, we cannot create a term of an empty type, then we can use our type systems to prove things.
+The type checker will then verify that our proofs are correct.
 
@@
 The \textit{Calculus of Constructions} is the language that has all three of these features.
-Now that we have explained the concepts behind Spoofax and Dependent Types, we will further explain the Calculus of Constructions and implement it in Spoofax in chapter~\ref{chap:baselang}.
+Now that we have explained the concepts behind Spoofax and dependent types, we will further explain the Calculus of Constructions and implement it in Spoofax in chapter~\ref{chap:baselang}.
 Then in the chapters after that, we will extend the language with more features.

### 2023-04-03 paper-thesis/chapters/10-coc.tex
First part of feedback from Jesper

@@
 and substituting \verb|x| in \verb|b| with \verb|e|.
-Beta reduction applies this rule anywhere in the term, whereas beta head-reduction only applies this rule at the head (outermost expression) of the term, and produces a term in beta-head normal form.
+Beta reduction applies this rule everywhere in the term, whereas beta head-reduction only applies this rule at the outermost expression of the term, and produces a term in beta-head normal form, which means the outermost constructor is the same as the outermost constructor of the normal form.
 
@@
 The machine can head evaluate lambda expressions with a call-by-name semantics.
-This is a strategy under which the leftmost, outermost term is always reduced first~\cite{tapl}.
+This is a strategy under which the leftmost, outermost beta-redex is always reduced first~\cite{tapl}.
 It works by keeping a stack of all arguments that have not been applied yet.
@@
 
-In conventional dependently typed languages, evaluation is often done using De Bruijn indices.
+In conventional dependently typed languages, variables are often represented using de Bruijn indices.
 De Bruijn representation~\cite[Section 6.1]{tapl} uses the distance from a binder to identify a variable.
 In this representation, alpha equivalence is the same as syntactic equivalence, which can simplify the manipulation of terms.
-However, we chose to use names rather than De Bruijn indices, because scope graphs work based on names.
-Using De Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations (which are Spoofax annotations that declare one name as being a use of another name that is a definition).
+However, we chose to use names rather than de Bruijn indices, because scope graphs work based on names.
+Using de Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations (which are Spoofax annotations that declare one name as being a use of another name that is a definition).
 
@@
 Additionally, we define \verb|betaReduce| which fully beta reduces a term.
-It works by first calling \verb|betaReduceHead| and then matching on the head, calling \verb|betaReduce| on the sub-expressions of the head recursively.
+It works by first calling \verb|betaReduceHead| and then matching on the outermost constructor, calling \verb|betaReduce| on the sub-expressions of the outermost constructor recursively.
 
@@
 \begin{enumerate}
-	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
-	\item $\beq{\scope{s_1}{e_1}}{\scope{s_2}{e_2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
-	\item $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ is the same as \verb|betaReduceHead((s1, e1), ps) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
-	\item $\bred{s_1}{e_1}{e_2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
-	\item $\scope{\co{sEmpty}}{e}$ is the same as $e$ (empty scopes can be left out)
+	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|.
+	\item $\beq{\scope{s_1}{e_1}}{\scope{s_2}{e_2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|.
+	\item $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ is the same as \verb|betaReduceHead((s1, e1), ps) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms}).
+	\item $\bred{s_1}{e_1}{e_2}$ is the same as \verb|betaReduce((s1, e1)) == e2|.
+	\item $\scope{\co{sEmpty}}{e}$ is the same as $e$ (empty scopes can be left out).
 \end{enumerate}

### 2023-04-03 paper-thesis/chapters/30-comparing-with-haskell.tex
First part of feedback from Jesper

@@
 To implement the calculus of constructions in Haskell, we first define the inductive datatype \verb|Expr|.
-We chose to define the language using De Bruijn indices, since this is the convention when implementing dependently typed languages, and the goal of this implementation in Haskell is to compare the Statix implementation with conventional ones.
+We chose to define the language using de Bruijn indices, since this is the convention when implementing dependently typed languages, and the goal of this implementation in Haskell is to compare the Statix implementation with conventional ones.
 \begin{lstlisting}
@@
 There is a constructor for each constructor in our language.
-The difference with the implementation in Spoofax is that the \verb|Let|, \verb|FnType|, and \verb|FnConstruct| no longer store an identifier, instead they introduce a new binding into their body, which are accessible by their De Bruijn indices.
+The difference with the implementation in Spoofax is that the \verb|Let|, \verb|FnType|, and \verb|FnConstruct| no longer store an identifier, instead they introduce a new binding into their body, which are accessible by their de Bruijn indices.
 
@@
 We still need to store both function arguments and substitutions, which we called \verb|NType| and \verb|NSubst| in chapter~\ref{chap:baselang}.
-We will use the same names, and store the environment as a list, with the head of the list representing De Bruijn index 0.
+We will use the same names, and store the environment as a list, with the head of the list representing de Bruijn index 0.
 Finally, we define the type of a scoped expression \verb|SExpr|, as a tuple of an environment and an expression.

### 2023-04-03 paper-thesis/chapters/10-coc.tex
Second part of feedback from Jesper

@@
 The relation associates a \verb|NameEntry| with each name in the scope graph.
-The \verb|NameEntry| can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.
-The Statix definition of these concepts is given below:
+The \verb|NameEntry| can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.\footnote{As a performance optimization, we could store the type of the substitution on the NSubst as well, this avoids the need to possibly type check the value again in the future.} The Statix definition of these concepts is given below:
 \begin{lstlisting}
@@
 It works by keeping a stack of all arguments that have not been applied yet.
-This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation, which is an alternative we will discuss in section \ref{sec:coc-subst}.
+This turned out to be the more natural way of expressing this compared to recursive evaluation relation, which is an alternative we will discuss in section \ref{sec:coc-recursive}.
 
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$}]{Beta head-reduction rules}
@@
 This substitution is called \verb|AlphaEqVars : ID * ID -> Expr|.
-The combination of the ID on the left and right hand side of the equality guarantees that the substitution is unique.
-In figure \ref{fig:beta-eq-rules} we show how \verb|AlphaEqVars| is used to determine equality of functions.
+The combination of the ID on the left and right hand is used to identify the variables.\footnote{Technically we only need the left or the right hand side here, but we use both since we can use them to get better error messages.} Currently this approach has problems with variable capture here: multiple variables with the same name that are asserted to be alpha equal can introduce the same \verb|AlphaEqVars| into the scope.
+This is solved in chapter \ref{chap:namecolls}.
+In figure \ref{fig:alpha-eq-rules} we show how \verb|AlphaEqVars| is used to determine alpha-equality $\aeq{e1}{e2}$ of terms.
+We can then define beta-equality $\beq{e1}{e2}$ by first beta-reducing terms and then using the alpha-equality relation.
 
-\begin{figure}[h]
-	\figuresection[\fbox{$\beq{\scope{s_1}{e_1}}{\scope{s_2}{e_2}}$}]{Beta equality rules}
+\begin{figure}[ht]
+	\figuresection[\fbox{$\aeq{\scope{s_1}{e_1}}{\scope{s_2}{e_2}}$}]{Alpha equality rules}
 
@@
 		} {
-			\beq
+			\aeq
 			{ \co{AlphaEqVars}(x_1, x_2) }
@@
 		\inferrule{
-			\beq{\scope{s_1}{a_1}}{\scope{s_2}{a_2}}
+			\aeq{\scope{s_1}{a_1}}{\scope{s_2}{a_2}}
 			\\\\ s_1' = \co{sPutSubst}(s_1, x_1, (\co{sEmpty}(), \co{AlphaEqVars}(x_1, x_2)))
 			\\\\ s_2' = \co{sPutSubst}(s_2, x_2, (\co{sEmpty}(), \co{AlphaEqVars}(x_1, x_2)))
-			\\\\ \beq{
+			\\\\ \aeq{
 				\scope{s_1'}{b_1}
@@
 		} {
-			\beq
+			\aeq
 			{ \scope{s_1}{\co{FnType}(x_1, a_1, b_1)} }
@@
 		} {
-			\beq
+			\aeq
 			{ \scope{s_1}{\co{FnConstruct}(x_1, a_1, b_1)} }
@@
 
-	\caption{Rules for beta equality in the Calculus of Constructions}
-	\label{fig:beta-eq-rules}
+	\caption{Rules for alpha equality in the Calculus of Constructions}
+	\label{fig:alpha-eq-rules}
 \end{figure}
@@
 The goal of this beta reduce is to make the term into a term that does not need an environment (by substituting all let bindings).
-A full beta reduce is not necessary, but this is merely a performance optimization.
+A full beta reduce is not necessary, merely substituting all the values in the environment would be enough, but this is merely a performance optimization.
 
-The inference rules in figures \ref{fig:beta-head-reduce-rules}, \ref{fig:beta-eq-rules}, and \ref{fig:type-check-rules} can be directly translated to Statix rules.
+The inference rules in figures \ref{fig:beta-head-reduce-rules}, \ref{fig:alpha-eq-rules}, and \ref{fig:type-check-rules} can be directly translated to Statix rules.
 For example, the rule for \verb|Let| bindings in figure \ref{fig:type-check-rules} is expressed like this in Statix:
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 	\figuresection[\fbox{$\toe{s}{e}{t}$}]{Type checking rules}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 	\figuresection[\fbox{$b_0 = \co{FnDestruct}(b_1, \co{Var}("T")))$}]{Type checking let}
@@
 
-\section{Discussion of a Substitution-Based Approach}
-\label{sec:coc-subst}
+\section{Discussion of a Recursive Approach}
+\label{sec:coc-recursive}
 
-An alternative for a Krivine machine, which keeps a stack of arguments it has encountered, is a substitution-based relation.
-This beta-reduces a \verb|FnDestruct| by doing a nested beta-reduction of the function, and substituting into that, as is shown in figure \ref{fig:subst-approach}.
+An alternative for a Krivine machine, which keeps a stack of arguments it has encountered, is a recursive relation.
+This beta-reduces a \verb|FnDestruct| by doing a nested beta-reduction of the function first, and substituting the argument into the body afterwards, as is shown in figure \ref{fig:subst-approach}.
+These rules were successfully implemented in Statix.
 
 Although these rules look cleaner, they are more complicated to implement, requiring a separate relation to check if \verb|f| is a \verb|FnConstruct| or something else.
-For the pure calculus of constructions this still works quite well, but when adding inductive datatypes (chapter~\ref{ch:datatypes}) the rules required become a lot more complex than those for a Krivine machine.
+For the calculus of constructions this still works quite well, but when adding inductive datatypes (chapter~\ref{ch:datatypes}) the rules required become a lot more complex than those for a Krivine machine.
+The primary problem is that eliminators can take an arbitrary of arguments depending on the datatype that all need to be available to choose the right eliminating function.
+Therefore we chose to use the Krivine machine instead.
 
-An additional benefit of Krivine machines is that the run-time performance of them tends to be a bit better, though this is only a constant factor and not a time-complexity improvement.
-
-\begin{figure}[h]
+\begin{figure}[ht]
 	\figuresection[\fbox{$\bhr{\scope{s_1}{e_1}}{}{\scope{s_2}{e_2}}$}]{A substitution-based approach for beta reduction}

### 2023-04-03 paper-thesis/chapters/20-adding-booleans.tex
Work a bit on thesis
First part of feedback from Jesper
Third part of feedback from Jesper

@@
 One for if \verb|c| evaluates to true, one for if \verb|c| evaluates to false, and finally a third case for other cases (such as when \verb|c| is a variable that does not have a substitution).
-These rules are stated below: (Remember that \verb|rebuild| is the rule introduced in chapter~\ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s)
+These rules are stated below: (Remember that \verb|rebuild| is the rule introduced in chapter~\ref{chap:baselang}, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s).
 
@@
 At the current stage, this is useful for testing the language.
-For example, this is a test with a postulate:
+For an example of a test with postulates, we assert that this program evaluates to Bool:
 \begin{lstlisting}
@@
 
-\section{Type Assert}
+\section{Type Assertions}
 
@@
 \begin{enumerate}
-	\item Define the parsing rules for the new feature
+	\item Define the parsing rules for the new feature.
 	\item Create a new file, \verb|tp_[feature].stx| and import this file in the main \verb|type_check.stx| file.
@@
 Also define cases for \verb|expectBetaEq| and \verb|betaReduce| if necessary.
-This is only necessary iff the new constructors are not always eliminated by \verb|betaReduceHead|.
+This is only necessary if the new constructors are not always eliminated by \verb|betaReduceHead|.
 	\item In the new file, define a case for the \verb|typeOfExpr| rule for the constructors that were added.
@@
 \begin{itemize}
-	\item Inference (chapter~\ref{ch:inference})
-	\item Inductive Datatypes (chapter~\ref{ch:datatypes})
-	\item Universes (chapter~\ref{ch:universes})
+	\item Inference (chapter~\ref{ch:inference}).
+	\item Inductive Datatypes (chapter~\ref{ch:datatypes}).
+	\item Universes (chapter~\ref{ch:universes}).
 \end{itemize}

### 2023-04-03 paper-thesis/chapters/21-adding-inference.tex
More writing work from today
More writing of the future work
Third part of feedback from Jesper

@@
 Inference is an important feature of dependent programming languages, that allows redundant parts of programs to be left out.
+This reduces the annotation burden in comparison with explicitly typed languages.
 For example, it allows you to infer the argument type of a function, if it can be inferred by the usage of the function, like in this example where the type of the argument of the polymorphic identity function can be inferred, since we pass it \verb|true| which is a boolean:
@@
 
-There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inferences than others.
-One algorithm for unification is \emph{first-order unification} (FOU), where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free variable, we set the the value free variable to be equal to the value of the non-free variable.
+There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inference problems than others.
+One algorithm for unification is \emph{first-order unification} (FOU), where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free metavariable, we set the the value free metavariable to be equal to the value of the non-free metavariable.
 There are some situations in which this approach fails to infer a term, but in most real-world scenarios it works very well.
@@
 \x : Type.
-\g: (_: (f x) -> Bool).
+\g: (f x) -> Bool.
 g true
@@
 	\label{inf-case2}
-	Ideally, we would generate a new name iff \verb|e2| is a free variable, otherwise using the already generated name.
+	Ideally, we would generate a new name iff \verb|e2| is a unresolved metavariable, otherwise using the already generated name.
 We can't query whether \verb|e2| is free, so as a best-attempt we always assume that \verb|e2| is free and generate a new name, this will fail in some cases.
@@
 If this system was added to Statix, this would make implementing more powerful inference algorithms cleanly possible.
+This idea is explored further in chapter \ref{ch:comp-lambdapi}.
 \end{enumerate}

### 2023-04-03 paper-thesis/chapters/22-adding-datatypes.tex
Update 23 and 24
Third part of feedback from Jesper

@@
 
-A difference between Agda and our implementation is that Agda has case matching as a native construct, whereas we chose to use eliminators as described in Inductive families~\cite{eliminators}.
+A difference between Agda and our implementation is that Agda has pattern matching as a native construct, whereas we chose to use eliminators as described in Inductive families~\cite{eliminators}.
 An eliminator is a function that can be used to case match on a data type.
@@
 -> P: (indices -> v: N params indices -> Type) 
--> constructors
+-> methods
 -> indices
@@
 	\item \verb|P| is the type that this eliminator will return. It may depend on the value \verb|v| that is eliminated.
-	\item \verb|constructors| is a function that eliminates each of the constructors of the datatype.
-To eliminate a constructor of the form \verb|C : args -> N params indices| it generates a function \verb|args -> P indices (C pars args)|.
+	\item \verb|methods| is for each of the constructors of the datatype, a function that eliminates it.
+To eliminate a constructor of the form \verb|C : args -> N params indices| it requires a function \verb|args -> P indices (C pars args)|.
 For recursive datatypes, a previous case is generated, such as the \verb|P x| in the eliminator of \verb|Nat| above.
@@
 	\item Create a new scope \verb|s4| with \verb|s1| as the parent.
-Then type-check each constructor with \verb|s2| as scope (so that the constructors can depend on parameters, but cannot depend on indices), and declare the type of that constructor in \verb|s4|.
-The parameters are automatically added to the type of the constructor.
+Then type-check each constructor with \verb|s2| as scope (so that the constructors can depend on parameters, but cannot depend on indices), and declare the constructor in \verb|s4|.
+We must also ensure that each constructor returns the datatype it belongs to, with matching parameters, though the indices may of course vary.
 
@@
 When defining a recursive data type, we would like our datatype to be \emph{strictly positive}, meaning that it can only be recursive on positive positions.
-If this rule would not exist, we would be able to create non-terminating functions using the datatype.
-Since under the curry-howard correspondence this allows to prove false, this is undesirable.
+If this check would not exist, we would be able to create non-terminating functions using the datatype.
+Since under the Curry Howard correspondence this allows to prove false, this is undesirable.
 

### 2023-04-03 paper-thesis/chapters/31-comparing-with-lambdapi.tex
Third part of feedback from Jesper

@@
 	\end{lstlisting}
-	\caption{Definition of less powerful Let in lampdapi}
+	\caption{Definition of less powerful Let in LambdaPi}
 	\label{fig:lp-let}
@@
 
+\section{Comparison}
+
+In this chapter we've defined the calculus of constructions by embedding it into LambdaPi.
+This is a declarative definition, similarly to Statix and unlike Haskell.
+However, since we are embedding the language into LambdaPi, the dynamic semantics of the language are now defined in terms of the dynamic semantics of LambdaPi.
+This means that the features our language can have are limited by the features of LambdaPi, this means it was impossible to define \verb|Let| bindings the way we wanted to.
+
+The static semantics of the language are defined by the way we create our terms.
+This is quite flexible, though less flexible than Statix since for example we have no control over the scoping rules (such as shadowing) of the language, which would be controlled by scope graphs in Statix.
+
+The conclusion is that embedding a language in LambdaPi is primarily useful if we want to verify the implementation of our language is correct, which is one of the primary use cases of LambdaPi.
+Using it as a compiler for the language is not the intention behind LambdaPi, which is clear from the limitations we've mentioned above.
+We believe this comparison was still insightful, as it shows how compact an implementation in Statix could be if Statix would have equational unification.
+
 We've now discussed equational unification using lambdapi, which is a feature that we suggest could improve Statix. Other possible improvements are discussed in chapter \ref{ch:ergonomics}.

### 2023-04-03 paper-thesis/chapters/98-relatedwork.tex
More feedback

@@
 Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types}~\cite{nameres} present an implementation of System F in Statix that does lazy substitutions, by using scopes as types.
-Each scope represents a type, and scopes have edges to other scopes in case of quantifications and substitutions.
+Each type is represented by a scope, and these scopes have edges to other scopes in case of quantifications and substitutions.
+Our Statix implementation does lazy substitutions in the expressions of our language, but not in the types.
+This is a minor compromise since types tend to stay quite small, but it is still a compromise.
 Further research is needed to see if this approach could also apply to dependently typed languages, where types can contain arbitrary terms.
@@
 
-Construction Language Implementations~\cite{Rouvoet21} describes how to create a declarative specification of a language in Statix, and then to use a dependently typed language to integrate the specification of well-typing in the representation of the program that is being interpreted or transformed.
+Arjen Rouvoet~\cite{Rouvoet21} describes how to create a declarative specification of a language in Statix, and then to use a dependently typed language to integrate the specification of well-typing in the representation of the program that is being interpreted or transformed.
 The work from this thesis and ours could be combined, by using our dependently typed language implemented in Statix, to do this verification, as another Statix DSL.
@@
 The primary difference is that they use a bidirectional type system~\cite{bidirectional} to do some inference, whereas this paper uses Statix' unification.
+Another difference is that pi-forall does not use de Bruijn indices but instead uses the Unbound library in Haskell.
 

### 2023-04-03 paper-thesis/chapters/99-conclusion.tex
More feedback
Bit of work on the conclusion

@@
 
-This thesis presents an implementation of a dependently typed language in Statix. Our aim was to anther the following research questions:
+This thesis presents an implementation of a dependently typed language in Statix. Our aim was to answer the following research questions:
 
@@
 For example, this allows to define a \verb|List| data type that can be both a list of values and a list of types without defining it twice.
-We see no reason why this couldn't be cleanly implemented in Statix.
-	\item Another feature of Agda is implementing dependent pattern matching, as described in "Eliminating Pattern Matching without K"~\cite{patternmatching}.
-This is requires a desugaring stage, in which matches are desugared to eliminators, which should be implemented in Stratego before type checking.
-This should also be possible to implement, but it may be quite a bit of work.
+Agda has quite a rather involved definition of equality between universe levels, which may be difficult to implement in Statix, but at least a simple version of universe polymorphism should be possible to implement in Statix.
+	\item Another feature of Agda is implementing dependent pattern matching, as described by Coquand~\cite{patternmatching}.
+Dependent pattern matching is complex because it requires interleaved type checking and desugaring, which is problematic since desugaring is usually done in Stratego, while type checking is done in Statix.
+The solution may be to do the desugaring in Statix instead, but at some point this is not ideal, since Statix is not meant for this.
+More research is needed on a good solution for this problem.
 	\item Inference in our language is currently implemented by explicitly specifying a placeholder \verb|_| where inference is required.
 In Agda we can specify an implicit argument, which is automatically inferred unless explicitly provided.
-We see no reason why this couldn't be cleanly implemented in Statix.
+It can be tricky to determine when an implicit argument has to be inserted, since this may require some type information~\cite{implicit_arguments}.
+More research is needed on whether the solution from this paper can be implemented in Statix.
 	\item Instance Arguments are a special kind of implicit arguments, that are solved by the instance resolution algorithm of Agda~\cite[Instance Arguments]{agda}.
@@
 After support for this is added, a proof search algorithm could be implemented in Stratego, such as the structurally recursive unification algorithm described by McBride~\cite{proofsearch}.
-	\item Agda allows function arguments to be marked as irrelevant, meaning that the value will be erased at runtime.
+Running the algorithm may require running type checking code, so here we encounter the interleaving problem again.
+More research is needed on whether proof search can be implemented in Spoofax.
+	\item \todo{irrelevance}
+	\item Agda allows function arguments to be marked as erased, meaning that the value will be erased at runtime.
 An even stronger version of this feature is implementing a quantitive type system, such as in Idris~2~\cite{idris2}, which combines linear and dependent types.
@@
 Researching whether an effect system could be cleanly specified in Statix may lead to more insights on the limits of Statix.
+	\item \todo{quantitive type system}
 \end{itemize}

### 2023-04-03 paper-thesis/chapters/99-conclusion.tex
Finish Jesper's feedback

@@
 More research is needed on whether proof search can be implemented in Spoofax.
-	\item \todo{irrelevance}
+	\item Agda allows for compile-time irrelevance annotations, which are a marker stating that the value of a certain variable is irrelevant, only that the type is non-empty matters~\cite[Irrelevance]{agda}.
+This can simplify some proofs, since all values of this type are considered equal.
+We see no reason why this couldn't be implemented in Statix.
 	\item Agda allows function arguments to be marked as erased, meaning that the value will be erased at runtime.
+This is different than irrelevance annotations, as the value is still available at compile time.
 An even stronger version of this feature is implementing a quantitive type system, such as in Idris~2~\cite{idris2}, which combines linear and dependent types.
 Implementing this in Spoofax should be possible, but it may be quite a bit of work as it is a complex type system.
-	\item Agda has mixfix operators, which allows to define custom operators such as \verb|a and b|.
-This requires parsing to happen in multiple stages, where the first stage finds the mixfix operator definitions and the second stage parses expressions using them.
+	\item Agda has mixfix operators, which allows to define custom operators such as \verb|a and b|~\cite[Mixfix Operators]{agda}.
+This requires parsing to happen in multiple stages, where the first stage finds the mixfix operator definitions and the second stage parses expressions using them~\cite{mixfix}.
 This results in quite a complex parser, implementing these multiple stages in Spoofax using SDF3 is not possible without fundamental changes to the way that Spoofax works.
@@
 	\item Currently the language in the paper does not allow for recursive functions.
-Agda allows for functions that are structurally recursive, and uses termination checking to enforce this property.
+Agda allows for functions that are structurally recursive (as well as mutually structurally recursive), and uses termination checking to enforce this property.
 We see no reason why this couldn't be cleanly implemented in Statix, similar to the implementation of positivity checking in chapter~\ref{ch:datatypes}.
@@
 Researching whether an effect system could be cleanly specified in Statix may lead to more insights on the limits of Statix.
-	\item \todo{quantitive type system}
+	\item As mentioned above, implementing a quantitive type system, such as in Idris~2~\cite{idris2}, would be an interesting extension to this thesis. 
 \end{itemize}

### 2023-04-10 paper-thesis/chapters/01-introduction.tex
Aron's feedback

@@
 These contributions are important for two reasons.
-Firstly from the perspective of a Spoofax researcher, developing a language with a complex type system in Statix tests the boundaries of what Statix can do, and how Statix could be further improved to support these usecases better.
+Firstly from the perspective of a Spoofax researcher, developing a language with a complex type system in Statix tests the boundaries of what Statix can do, and how Statix could be further improved to support these use-cases better.
 On the other hand, from the perspective of a researcher in dependent type systems, if we can show that Statix provides a good way to quickly implement a dependently typed language, then Spoofax could help with rapid prototyping of dependently typed languages.

### 2023-04-10 paper-thesis/chapters/02-background-spoofax.tex
First part of feedback from Jesper
Aron's feedback

@@
 general-purpose programming languages.
-Each aspect of the language is defined in one the meta-languages, each with their own purpose.
+Each aspect of the language is defined in one the meta-languages, each with its own purpose.
 From these specifications Spoofax generates a parser, type checker and other useful tools.
@@
 The constraints are generated by predicates.
-A predicate defines a relation on terms and scopes, it is of the form $\mathtt{p(t_1, ..., t_n) :-\,\,c_1, ..., c_n}$.
+A predicate defines a relation on terms and scopes.
+It is of the form $\mathtt{p(t_1, ..., t_n) :-\,\,c_1, ..., c_n}$.
 Each constraint $c_1...c_n$ is another predicate or an equality constraint.
@@
 Statix also has scope graphs.
-Scope graphs are used to represent a programs' name and type information~\cite{scopes_as_types}~\cite{nameres}.
+Scope graphs are used to represent a programs' name and type information~\cite{scopes_as_types, nameres}.
 A scope graph is a graph that consists of the following components:

### 2023-04-10 paper-thesis/chapters/03-background-dependent-types.tex
Finish Jesper's feedback
Aron's feedback

@@
 
-The biggest advantage of dependent types is that they increase the expressiveness of a type system a lot, for example:
+The biggest advantage of dependent types is that they increase the expressiveness of a type system, for example:
 \begin{lstlisting}
-append : (A: Set) -> (n : Nat) -> Vec A n -> Vec A n -> Vec A (n + n)
+append : (A : Type) -> (n : Nat) -> Vec A n -> Vec A n -> Vec A (n + n)
 \end{lstlisting}

### 2023-04-10 paper-thesis/chapters/11-variable-capture.tex
Aron's feedback

@@
 \noindent
-The same thing happens, the body of the \verb|FnConstruct| is typechecked with a new scope.
+The same thing happens, the body of the \verb|FnConstruct| is type-checked with a new scope.
 Note that \verb|Var("T")| in the type of the second \verb|T| is ambiguous, does it refer to the first or second node?
@@
 This doesn't work for a variety of reasons.
-First of all, it doesn't actually solve the problem.
+This doesn't actually solve the problem.
 Names can be duplicated during beta reduction of terms, so we still don't have the guarantee that each variable has a unique name.
-Furthermore, this changes the names in the AST, so it breaks editor services in the same was as De Bruijn Indices did.
+Furthermore, unless we keep track of error messages
 
@@
    ID.Syn = RID
+\end{lstlisting}
+\begin{lstlisting}
 signature constructors

### 2023-04-10 paper-thesis/chapters/12-editor-services.tex
Aron's feedback

@@
 Spoofax then generates the parser, type checker and compiler for us.
-An additional benefit of these declarative definitions is that Spoofax can use them to generate editor services, such as \verb|go to definition|, \verb|renaming| and \verb|semantic code completion|.
+An additional benefit of these declarative definitions is that Spoofax can use them to generate editor services, such as \emph{go to definition}, \emph{renaming} and \emph{semantic code completion}.
 

### 2023-04-10 paper-thesis/chapters/20-adding-booleans.tex
Aron's feedback

@@
 	{ \overline{p} }
-	{ \co{rebuild}((s, \co{BoolIf}(c, b1, b2)), \overline{p}) }
+	{ \co{rebuild}((s, \co{BoolIf}(c', b1, b2)), \overline{p}) }
 }
@@
 If we decide that we don't like a feature after all, we can remove it simply by unimporting the file.
-This problem is similar to the expression problem\cite{expression_problem}, where we want to extend datatypes and their associated behaviour, which Statix solves very well.
+This problem is similar to the expression problem\cite{expression_problem, extensability}, where we want to extend datatypes and their associated behaviour, which Statix solves partially.
+Statix does not fulfill the requirement that files can be individually recompiled.
 

### 2023-04-10 paper-thesis/chapters/21-adding-inference.tex
Aron's feedback

@@
 The syntax we chose is that if we want a value to be inferred, we replace it with an underscore.
-Other languages such as Agda allow for \emph{implicit arguments}, which are arguments that can be left out, which will then be inferred~\cite{agda}~\cite{coq}.
+Other languages such as Agda allow for \emph{implicit arguments}, which are arguments that can be left out, which will then be inferred~\cite{agda, coq}.
 This is just syntactic sugar over the \emph{placeholders} we implement.
@@
 
-There are a lot of different algorithms for inference\cite{typeinference}, some algorithms can solve more inference problems than others.
+There are a lot of different algorithms for inference~\cite{typeinference}, some algorithms can solve more inference problems than others.
 One algorithm for unification is \emph{first-order unification} (FOU), where if at any point during type checking we assert that $\beq{e_1}{e_2}$ and either $e_1$ or $e_2$ is a free metavariable, we set the the value free metavariable to be equal to the value of the non-free metavariable.
@@
 Ideally when implementing beta equality when we encounter $\beq{e_1}{e_2}$ we would check if $e_1$ is a free variable, but Statix does not allow for querying whether variables are free.
+The reason that Statix does not allow this is that it is non-trivial to guarantee confluence if this feature is added to the language.
 
@@
 	\label{inf-case2}
-	Ideally, we would generate a new name iff \verb|e2| is a unresolved metavariable, otherwise using the already generated name.
+	Ideally, we would generate a new name if \verb|e2| is a unresolved metavariable, otherwise using the already generated name.
 We can't query whether \verb|e2| is free, so as a best-attempt we always assume that \verb|e2| is free and generate a new name, this will fail in some cases.
@@
 However, these programs are surprisingly uncommon considering how rough the approximations are.
-They mostly fail when we attempt to infer function values, and two functions are inferred that are beta-equal, and they are beta-equal but not identical.
+Inference may fail when we attempt to infer function values, and two functions are inferred that are beta-equal, and they are beta-equal but not identical.
 An example of a program where these approximation fails is:

### 2023-04-10 paper-thesis/chapters/22-adding-datatypes.tex
Aron's feedback

@@
 \begin{lstlisting}
-parameters 
+params 
 -> P: (indices -> v: N params indices -> Type) 
@@
 \begin{itemize}
-	\item \verb|parameters| are the parameters of the data type we want to eliminate, since these are constant among all constructors those should be specified at the start.
+	\item \verb|params| are the parameters of the data type we want to eliminate, since these are constant among all constructors those should be specified at the start.
 	\item \verb|P| is the type that this eliminator will return. It may depend on the value \verb|v| that is eliminated.
 	\item \verb|methods| is for each of the constructors of the datatype, a function that eliminates it.
-To eliminate a constructor of the form \verb|C : args -> N params indices| it requires a function \verb|args -> P indices (C pars args)|.
+To eliminate a constructor of the form \verb|C : args -> N params indices| it requires a function \verb|args -> P indices (C params args)|.
 For recursive datatypes, a previous case is generated, such as the \verb|P x| in the eliminator of \verb|Nat| above.
@@
 Call the end of this chain \verb|s3|.
-The scope \verb|s3| is not used in the rest of this process, but creating this chain is required to check the types of the indices.
+The scope \verb|s3| is not used in the rest of this process (since constructors cannot refer to indices), but creating this chain is required to check the types of the indices.
 

### 2023-04-10 paper-thesis/chapters/24-backend.tex
Third part of feedback from Jesper
Aron's feedback

@@
 
-Luckily, Stratego has a strategy built in that runs a Statix predicate on a term: \verb|stx-evaluate|.
+Luckily, Statix exposes an API to Stratego that can run a Statix predicate on a term: \verb|stx-evaluate|.
 Using this strategy, the actual implementation of our interpreter in Stratego is just one line: Match an expression \verb|e| and run the functional predicate \verb|interpret| from the \verb|main.stx| file on it.
@@
 Therefore, the resulting interpreter will be slower when implemented in Statix.
-This is however just a constant factor overhead, it does not affect the time complexity of the algorithm.
 
@@
 Types require a bit more attention.
-Since our language has no way to match on types, types can actually be completely eliminated during type checking.
+Since our language has no way to match on types, types can actually be completely eliminated during compilation.
 The easiest way to do this is to compile all types to an arbitrary value (we chose the string "TYPE"), since the value is irrelevant anyways.

### 2023-04-10 paper-thesis/chapters/40-discussing-statix.tex
Third part of feedback from Jesper
Aron's feedback

@@
 \begin{enumerate}
-	\item Allow specifying reduction rules, and then implement equality under reduction in Statix. This feature was discussed extensively in chapter \ref{ch:comp-lambdapi}.
+	\item Allow specifying reduction rules, and then implement equality under reduction in Statix.
+This feature was discussed extensively in chapter \ref{ch:comp-lambdapi}.
+This would require extensive changes to the core language of Statix, so this is challenging to implement.
 
@@
 	The definition of \texttt{isStrictlyPositive} can then use all the metavariables of \\ \texttt{DataTypeDecl} without having to explicitly pass them.
+This feature could be desugared by moving the relation to the global scope and explicitly passing the metavariables of the parent constraint when it is used.
 
 	\item Generic constraints.
-For example, we created two separate relations for reversing lists of different types, making these one generic function would've been better.
-This was the only case where this feature would've been useful for this thesis, but it may be more useful for other work.
+For example, we created two separate relations for reversing lists of different types, making these one generic function would have been better.
+This was the only case where this feature would have been useful for this thesis, but it may be more useful for other work.
+This could be implemented by monomorphization or by changing the core language.
 \end{enumerate}
@@
 However, support for unit tests is still lacking.
-The to create unit tests is using Statix tests, allowing assertions on specific relations.
+The way to create unit tests for the type checker is using Statix tests, allowing assertions on specific relations.
 But these tests are not integrated with SPT tests (they don't even reside in the test folder) and there is no way to run all Statix tests, to ensure no regressions happened.
@@
 Having support for a debugger would help tremendously with this.
+
+	The way that this debugger could work is that you could start by entering a certain constraint \verb|typeCheck(...)|, the debugger would then show the current value of each metavariable in scope, and each constraint that is yet to be expanded.
+Since there is no fixed evaluation order in Statix, we could even allow the programmer to choose which constraint is evaluated next, which then generates more constraints.
 \end{enumerate}

### 2023-04-10 paper-thesis/chapters/98-relatedwork.tex
Aron's feedback

@@
 The implementation in this paper requires performing substitutions in types immediately, as types don't have a corresponding scope to store substitutions.
-Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types}~\cite{nameres} present an implementation of System F in Statix that does lazy substitutions, by using scopes as types.
+Van Antwerpen et al.~\cite[sect 2.5]{scopes_as_types, nameres} present an implementation of System F in Statix that does lazy substitutions, by using scopes as types.
 Each type is represented by a scope, and these scopes have edges to other scopes in case of quantifications and substitutions.

### 2023-04-10 paper-thesis/chapters/99-conclusion.tex
Aron's feedback

@@
 Agda has quite a rather involved definition of equality between universe levels, which may be difficult to implement in Statix, but at least a simple version of universe polymorphism should be possible to implement in Statix.
-	\item Another feature of Agda is implementing dependent pattern matching, as described by Coquand~\cite{patternmatching}.
+	\item Another feature of Agda is dependent pattern matching, as described by Coquand~\cite{patternmatching}.
 Dependent pattern matching is complex because it requires interleaved type checking and desugaring, which is problematic since desugaring is usually done in Stratego, while type checking is done in Statix.
@@
 After support for this is added, a proof search algorithm could be implemented in Stratego, such as the structurally recursive unification algorithm described by McBride~\cite{proofsearch}.
+The holes system that was introduced for semantic code completion~(chapter~\ref{chap:editor-services}) could be re-used for this purpose.
 Running the algorithm may require running type checking code, so here we encounter the interleaving problem again.

### 2023-04-12 paper-thesis/chapters/24-backend.tex
Inductive data types in the backend

@@
 
-Finally, inductive data types are a bit harder to compile, as Clojure does not have an equivalent to sum types on which we can match.
+Finally, inductive data types are a bit more challenging, as Clojure does not have an equivalent to sum types on which we can match.
 Instead, we compile inductive datatypes to a pair (constructor, list of constructor arguments).
+An eliminator can then be compiled to a function from this pair, that matches on which constructor the datatype has, and calls the corresponding argument of the eliminator function.
+This is the same approach that the Agda to Scheme compiler takes.
 
-\todo{finish this after implementing}
-- atoms voor constructors
+We did not end up implementing this, data types are not supported by the current implementation of the compiler.
+However, we see no reason why this couldn't be implemented identically to the Agda to Scheme compiler.
 

### 2023-04-12 paper-thesis/chapters/01-introduction.tex
Minor improvements to the thesis

@@
 The primary contribution of this thesis is to investigate how well Statix is fit for the task of defining a dependently-typed language.
-The goal is not only to show that Statix can implement it, but also that the implementation is more concise than in a general purpose language.
+The goal is not only to show that Statix can implement it, but also that the implementation is more concise than in a general purpose language (\textbf{Research Question 1}).
 We will start by implementing the Calculus of Constructions~\cite{Coquand_Huet_1988}, a lambda calculus with dependent types, as this is one of the most simple dependently typed languages (chapter~\ref{chap:baselang}).
@@
 \begin{itemize}
-	\item We show the language is easily extendable, by extending it with booleans among other features (chapter~\ref{chap:bools}).
-	\item We show how to add inference of implicit arguments to the implementation (chapter~\ref{ch:inference}).
-	\item We show how to add support for inductive datatypes to the implementation (chapter~\ref{ch:datatypes}).
-	\item We show how to add support for universes to the implementation (chapter~\ref{ch:universes}).
+	\item \textbf{RQ2:} We show the language is easily extendable, by extending it with booleans among other features (chapter~\ref{chap:bools}).
+	\item \textbf{RQ3:} We show how to add inference of implicit arguments to the implementation (chapter~\ref{ch:inference}).
+	\item \textbf{RQ4:} We show how to add support for inductive datatypes to the implementation (chapter~\ref{ch:datatypes}).
+	\item \textbf{RQ5:} We show how to add support for universes to the implementation (chapter~\ref{ch:universes}).
 \end{itemize}

### 2023-04-12 paper-thesis/chapters/99-conclusion.tex
Minor improvements to the thesis

@@
 \paragraph{RQ4: Can we add support for inductive data types to the implementation?}
-chapter~\ref{ch:datatypes} adds support for inductive data types with parameters and indices.
+Chapter~\ref{ch:datatypes} adds support for inductive data types with parameters and indices.
 We give steps for type-checking data type declarations.

### 2023-04-19 pres-thesis/presentation.tex
Start on thesis presentation
Work on thesis defense presentation

--- 2023-03-22 pres-evcs/presentation.tex
@@
 \usepackage{qrcode}
+\usepackage{listings}
 
@@
 
+\begin{frame}[fragile]{Background: What is Spoofax?}
+	A language designer's workbench with everything you need to design a programming language.
+	\begin{itemize}
+		\item Declarative, using meta-languages
+	\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Background: What is Spoofax?}
+SDF3, Statix and Stratego (and more)
+
+\begin{figure}
+	\includegraphics[width=0.32\linewidth]{img/semantics}
+	\includegraphics[width=0.32\linewidth]{img/syntax}
+	\includegraphics[width=0.32\linewidth]{img/transformation}
+\end{figure}
+\end{frame}
+
+\begin{frame}[fragile]{Background: What is Spoofax?}
+
+\begin{exampleblock}{Example}
+\begin{lstlisting}
+int x = 5;
+int y = x + 3;
+return x + y;
+\end{lstlisting}
+\end{exampleblock}
+
+\begin{figure}
+	\includegraphics[width=0.7\linewidth]{img/screenshot005}
+\end{figure}
+
+% The logic I just explained would be written down in Statix
+% Like: When you encounter a let, create a new scope...
+
+\end{frame}
+
 \begin{frame}[fragile]{Background: What are Dependent Types?}
@@
 % - Challenge: Type checking may require evaluating arbitrary terms to decide equality
-	\item Proof assistants
-	\item For example: Agda, Coq, Lean, ...
-% Can be used to write proofs about code
+\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Background: What are Dependent Types?}
+	Type checking requires evaluation
+	\begin{exampleblock}{Example 1}
+		\texttt{let T = if false then Int else Bool end;\\let b: T = true;}
+	\end{exampleblock}
+\end{frame}
+
+\begin{frame}[fragile]{Background: What are Dependent Types?}
+	Why are dependent types useful?
+	\begin{itemize}
+		\item Proof Assistants: Agda, Lean, Coq, etc
 	\end{itemize}
 
+	% We're looking only at signatures
+	\begin{exampleblock}{Sorted lists}
+\texttt{sort : List t -> List t\\sort\_sorted : (v : List t) -> IsSorted (sort v)}
+	\end{exampleblock}
 \end{frame}
@@
 
-	\begin{exampleblock}{Example 1}
-		\texttt{($\backslash$v: Type. v) Type}
-	\end{exampleblock}
+	\begin{block}{Syntax Definition}
+		\begin{figure}
+			\includegraphics[width=0.7\linewidth]{img/screenshot006}
+		\end{figure}
+	\end{block}
 
-	\begin{exampleblock}{Example 2}
+	\begin{exampleblock}{Example}
 		\texttt{let f = $\backslash$T: Type. $\backslash$x: T. x; \\
@@
 \begin{frame}[fragile]{Type Checking}
-	\begin{block}{Type checking relation}
-		\texttt{typeOfExpr : scope * Expr -> Expr}
-	\end{block}
-	% Returns an Expr, not a Type
-	% Scope -> Two different types of nodes
+	\begin{figure}
+		\includegraphics[width=0.9\linewidth]{img/screenshot007}
+	\end{figure}
+	% We define the language using inference rules
+	% How can we express those rules in Statix?
+\end{frame}
+
+\begin{frame}[fragile]{Type Checking: From inference rules to Statix code}
+	\begin{figure}
+		\centering
+		\includegraphics[width=0.8\linewidth]{img/screenshot004}
+	\end{figure}
+
+	\begin{figure}
+		\centering
+		\begin{exampleblock}{Equivalent Statix code} 
+			\texttt{typeOfExpr (s, Let(x, e, b)) = bt :-\\
+				\,\, typeOfExpr (s, e) == et, \\
+				\,\, typeOfExpr (sPutSubst (s, x, (s, e)), b) == bt}
+		\end{exampleblock}
+	\end{figure}
+
+\end{frame}
+
+\begin{frame}[fragile]{Type Checking}
 	\begin{block}{How do we use scopes?}
@@
 	\end{block}
-\end{frame}
-
-\begin{frame}[fragile]{Type Checking: From inference rules to Statix code}
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\linewidth]{img/screenshot004}
-\end{figure}
-
-\begin{figure}
-	\centering
-	\begin{exampleblock}{Equivalent Statix code} 
-		\texttt{typeOfExpr (s, Let(x, e, b)) = bt :-\\
-			\,\, typeOfExpr (s, e) == et, \\
-			\,\, typeOfExpr (sPutSubst (s, x, (s, e)), b) == bt}
-	\end{exampleblock}
-\end{figure}
-
 \end{frame}

### 2023-04-20 pres-thesis/presentation.tex
More slides on core

@@
 \begin{frame}[fragile]{Background: What is Spoofax?}
-SDF3, Statix and Stratego (and more)
+\begin{figure}
+	\includegraphics[width=0.8\linewidth]{img/syntax}
+\end{figure}
+\end{frame}
 
+\begin{frame}[fragile]{Background: What is Spoofax?}
 	\begin{figure}
-	\includegraphics[width=0.32\linewidth]{img/semantics}
-	\includegraphics[width=0.32\linewidth]{img/syntax}
-	\includegraphics[width=0.32\linewidth]{img/transformation}
+		\includegraphics[width=0.8\linewidth]{img/semantics}
+	\end{figure}
+\end{frame}
+
+\begin{frame}[fragile]{Background: What is Spoofax?}
+	\begin{figure}
+		\includegraphics[width=0.8\linewidth]{img/transformation}
 	\end{figure}
@@
 	\begin{exampleblock}{Example 1}
-		\texttt{let T = if false then Int else Bool end;\\let b: T = true;}
+		\texttt{let T: Type = if false then Int else Bool end;\\let b: T = true;}
 	\end{exampleblock}
@@
 	\end{figure}
-
 \end{frame}
 
-\begin{frame}[fragile]{Type Checking}
+\begin{frame}[fragile]{Type Checking: Environments \& Context}
+	\begin{exampleblock}{Example}
+		\texttt{let T: Type = Bool;\\
+			$\backslash$b: T. ???
+		}
+	\end{exampleblock}
+
+	\begin{block}{What information is needed?}
+		\begin{enumerate}
+			\item T = Bool (Stored in Environment)
+			\item x : T (Stored in Context)
+		\end{enumerate}
+	\end{block}
+\end{frame}
+
+\begin{frame}[fragile]{Type Checking: Environments \& Context}
 	\begin{block}{How do we use scopes?}
 		A scope is used as a replacement for an environment and a context.\\
-		One edge \verb|p|. One relation \texttt{name $\rightarrow$ NameEntry}, NameEntry is either:
+		\begin{itemize}
+			\item One type of edge: \verb|p|.
+			\item One type of relation: \verb|NameEntry|, which can be:
 				\begin{itemize}
@@
 					\item NSubst: Stores a substitution (Corresponds to an environment)
+				\end{itemize}
 		\end{itemize}
@@
 \begin{frame}[fragile]{Type Checking: Requires Evaluation}
-	\begin{exampleblock}{Example 1}
+	\begin{exampleblock}{Example from earlier}
 		\texttt{let T = if false then Int else Bool end;\\let b: T = true;}
@@
 	\end{block}
+\end{frame}
+
+\begin{frame}[fragile]{Type Checking: Requires Evaluation}
+	\begin{figure}
+		\includegraphics[width=0.9\linewidth]{screenshot001}
+	\end{figure}
+
 \end{frame}

### 2023-04-20 pres-thesis/presentation.tex
More work

@@
 	\end{figure}
+\end{frame}
 
+\begin{frame}[fragile]{Type Checking: Variable Capturing}
+	todo
 \end{frame}
@@
 
+\begin{frame}{Comparison with Haskell}
+	\begin{block}{Type checking}
+		\texttt{data EnvEntry = NType Expr | NSubst SExpr
+			\\type Env = [EnvEntry]
+			\\type SExpr = (Env , Expr)
+			\\tc :: SExpr -> Either String Expr}
+	\end{block}
+
+	\begin{block}{Differences}
+		\begin{itemize}
+			\item Distribution of definitions over files
+			\item De Bruijn Indices vs Names
+			\item Inference built-in to Statix
+		\end{itemize}
+	\end{block}
+\end{frame}
+
+\begin{frame}{Term Inference}
+	\begin{example}\texttt{let id = ($\backslash$T : Type. $\backslash$x: T. x);
+	\\id \_ true}
+	\end{example}
+	\begin{block}{How it works}
+		\begin{itemize}
+			\item When we encounter \texttt{expectBetaEq(e1, e2)} and \texttt{e1} or \texttt{e2} is free, we can infer it!
+			\item But in Statix we can't query whether variables are free.
+			\item Solution: Wrap each free constructor in \texttt{Infer}.
+		\end{itemize}
+	\end{block}
+\end{frame}
+
+\begin{frame}{Term Inference: Equational Unification}
+	\begin{itemize}
+		\item We can declare two terms to be equal if they satisfy a certain relation (such as beta equality)
+		\item We do this using \emph{reduction rules}, such as:
+		\begin{exampleblock}{Example of a rewrite rule}
+			($\backslash$ x : T. b) a => b[x := a]
+		\end{exampleblock}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Comparison with LambdaPi}
+	\begin{itemize}
+		\item A \emph{logical framework} in which you can embed other programming languages
+		\item Supports rewrite rules
+		\item Cannot define \emph{Let} because of limitations of LambdaP
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Comparison with LambdaPi: Example}
+
+\end{frame}
+
+\begin{frame}{Inductive Data Types}
+	todo
+\end{frame}
+
+\begin{frame}{Interpreter \& Compiler}
+	todo
+\end{frame}
+
 \begin{frame}[fragile]{Conclusions}

### 2023-04-22 pres-thesis/presentation.tex
Finish presentation draft

@@
 \institute[]{Delft University of Technology, The Netherlands}
-\author{Jonathan Brouwer \and Jesper Cockx \and Aron Zwaan}
+\author{Jonathan Brouwer}
 
@@
 \begin{frame}[fragile]{Type Checking: Variable Capturing}
-	todo
+	\begin{exampleblock}{What is the type of this expression?}
+		\texttt{$\backslash$T : Type. $\backslash$T : T. T}
+	\end{exampleblock}
+\end{frame}
+
+\begin{frame}[fragile]{Type Checking: Variable Capturing}
+	\begin{exampleblock}{What is the type of this expression?}
+		\texttt{$\backslash$T : Type. $\backslash$T : T. T}
+	\end{exampleblock}
+
+	\begin{block}{The type}
+		\texttt{T : Type -> T : T -> T}
+	\end{block}
+
+	\begin{block}{Equivalent type under renaming: Not correct!}
+		\texttt{T : Type -> x : T -> x}
+	\end{block}
+\end{frame}
+
+\begin{frame}[fragile]{Type Checking: Variable Capturing}
+	\begin{block}{Solutions}
+		\begin{enumerate}
+			\item De Bruijn indices
+			\item Uniquifying names
+			\item Capture-avoiding substitution
+			\item \textbf{Using scopes to distinguish names}
+		\end{enumerate}
+	\end{block}
 \end{frame}
@@
 
+\begin{frame}{On Extensibility}
+	\begin{block}{How to extend the language}
+		\begin{enumerate}
+			\item Define the parsing rules for the new feature
+			\item Create a new .stx file for the type-checking rules (and import it)
+			\item Define the relations (typeOfExpr, betaReduceHead, expectBetaEq and betaReduce) in this file using the defining inference rules
+		\end{enumerate}
+	\end{block}
+\end{frame}
+
 \begin{frame}{Term Inference}
@@
 		\item Supports rewrite rules
-		\item Cannot define \emph{Let} because of limitations of LambdaP
+		\item Cannot define \emph{Let} because of limitations of LambdaPi
 	\end{itemize}
@@
 \begin{frame}{Comparison with LambdaPi: Example}
+	\begin{exampleblock}{Defining FnConstruct}
+		\texttt{symbol FnConstruct :
+		\\\hspace{1em}Π (A : TmSort ),
+		\\\hspace{1em}Π (B : TmType A → TmSort ),
+		\\\hspace{1em}Π (f : Π (x : TmType A), TmType (B x)),
+		\\\hspace{1em}TmType ( FnType A B);
+		}
 
-\end{frame}
+	\end{exampleblock}
 
-\begin{frame}{Inductive Data Types}
-	todo
-\end{frame}
-
-\begin{frame}{Interpreter \& Compiler}
-	todo
+	\begin{exampleblock}{Reduction rule for FnConstruct}
+		\texttt{rule FnConstruct \_ \_ \$f → \$f;}
+	\end{exampleblock}
 \end{frame}

### 2023-04-23 pres-thesis/presentation.tex
Update presentation.tex
Changes to presentation based on first test runs

@@
 \begin{frame}[fragile]{Background: What is Spoofax?}
-	\begin{figure}
-		\includegraphics[width=0.8\linewidth]{img/transformation}
-	\end{figure}
-\end{frame}
-
-\begin{frame}[fragile]{Background: What is Spoofax?}
 
@@
 
+\begin{frame}[fragile]{Background: What is Spoofax?}
+	\begin{figure}
+		\includegraphics[width=0.8\linewidth]{img/transformation}
+	\end{figure}
+	% - used for interpreters (3+5=8) or compiler (a+b compiles to add instruction in x86)
+	% - = DYNAMIC SEMANTICS
+\end{frame}
+
 \begin{frame}[fragile]{Background: What are Dependent Types?}
@@
 	\begin{exampleblock}{Example}
-		\texttt{concat : (A: Set) -> (n m : Nat) -> Vec A n -> Vec A m
+		\texttt{concat : (A: Type) -> (n m : Nat) -> Vec A n -> Vec A m
 			\\ \hspace*{48pt} -> Vec A (n + m)}
@@
 
-	\begin{figure}
-		\centering
+	\begin{block}{Signature of \emph{typeOfExpr}}
+		\texttt{typeOfExpr : scope * Expr -> Expr}
+	\end{block}
+
 	\begin{exampleblock}{Equivalent Statix code} 
@@
 	\end{exampleblock}
-	\end{figure}
 \end{frame}
@@
 			\item T = Bool (Stored in Environment)
-			\item x : T (Stored in Context)
+			\item b : T (Stored in Context)
 		\end{enumerate}
@@
 	\begin{block}{How do we use scopes?}
-		A scope is used as a replacement for an environment and a context.\\
-		\begin{itemize}
-			\item One type of edge: \verb|p|.
-			\item One type of relation: \verb|NameEntry|, which can be:
-				\begin{itemize}
-					\item NType: Stores a type (Corresponds to a context)
-					\item NSubst: Stores a substitution (Corresponds to an environment)
-				\end{itemize}
-		\end{itemize}
+		Scope graphs are used as a replacement for an environment and a context.
 	\end{block}
+
+	\begin{figure}
+		\includegraphics[width=0.7\linewidth]{screenshot004}
+	\end{figure}
+
 \end{frame}
@@
 	\end{figure}
+
+	% Talk about the fact that we're now defining evaluation in Statix
 \end{frame}
@@
 		\end{enumerate}
+	\end{block}
+
+	\begin{block}{The type}
+		\texttt{T : Type -> T : T -> T}
+	\end{block}
+
+	\begin{block}{The type (fixed)}
+		\texttt{T : Type -> T : (T, s0) -> (T, s0)}
 	\end{block}
@@
 		\begin{enumerate}
-			\item Define the parsing rules for the new feature
-			\item Create a new .stx file for the type-checking rules (and import it)
-			\item Define the relations (typeOfExpr, betaReduceHead, expectBetaEq and betaReduce) in this file using the defining inference rules
+			\item Define parsing rules
+			\item Create a new .stx file
+			\item Define the relations: \texttt{typeOfExpr, betaReduceHead, expectBetaEq and betaReduce}
 		\end{enumerate}
@@
 		\begin{itemize}
-			\item When we encounter \texttt{expectBetaEq(e1, e2)} and \texttt{e1} or \texttt{e2} is free, we can infer it!
+			\item When \texttt{expectBetaEq(e1, e2)} and \texttt{e1} or \texttt{e2} is free, infer!
 			\item But in Statix we can't query whether variables are free.
@@
 		\end{exampleblock}
+		\item Comparison with LambdaPi
 	\end{itemize}
-\end{frame}
-
-\begin{frame}{Comparison with LambdaPi}
-	\begin{itemize}
-		\item A \emph{logical framework} in which you can embed other programming languages
-		\item Supports rewrite rules
-		\item Cannot define \emph{Let} because of limitations of LambdaPi
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Comparison with LambdaPi: Example}
-	\begin{exampleblock}{Defining FnConstruct}
-		\texttt{symbol FnConstruct :
-		\\\hspace{1em}Π (A : TmSort ),
-		\\\hspace{1em}Π (B : TmType A → TmSort ),
-		\\\hspace{1em}Π (f : Π (x : TmType A), TmType (B x)),
-		\\\hspace{1em}TmType ( FnType A B);
-		}
-
-	\end{exampleblock}
-
-	\begin{exampleblock}{Reduction rule for FnConstruct}
-		\texttt{rule FnConstruct \_ \_ \$f → \$f;}
-	\end{exampleblock}
 \end{frame}

### 2023-04-24 pres-thesis/presentation.tex
Minor changes based on more practice runs
typo
Changes based on meeting

@@
 \begin{lstlisting}
-int x = 5;
-int y = x + 3;
+let x = 5;
+let y = x + 3;
 return x + y;
@@
 \begin{frame}[fragile]{Background: What are Dependent Types?}
-	Type checking requires evaluation
-	\begin{exampleblock}{Example 1}
-		\texttt{let T: Type = if false then Int else Bool end;\\let b: T = true;}
-	\end{exampleblock}
-\end{frame}
-
-\begin{frame}[fragile]{Background: What are Dependent Types?}
 	Why are dependent types useful?
@@
 
+\begin{frame}[fragile]{Background: What are Dependent Types?}
+	Type checking requires evaluation
+	\begin{exampleblock}{Example 1}
+		\texttt{let T: Type = if false then Int else Bool end;\\let b: T = true;}
+	\end{exampleblock}
+\end{frame}
+
 \begin{frame}[fragile]{Research Question}
@@
 	\item Will it be easier than doing it in Haskell?
-	\item Defining system F was a challenge\footnote{Hendrik van Antwerpen et al. Scopes as types.}, will it even be possible?
+	\item Prior work: Defining System F\footnote{Hendrik van Antwerpen et al. Scopes as types.}
 \end{itemize}
@@
 		\texttt{let f = $\backslash$T: Type. $\backslash$x: T. x; \\
-f (T: Type -> Type) ($\backslash$y: Type. y)
+f (Type -> Type) ($\backslash$y: Type. y)
 		}
@@
 	% what other languages use as a core
-\end{frame}
-
-\begin{frame}[fragile]{Type Checking}
-	\begin{figure}
-		\includegraphics[width=0.9\linewidth]{img/screenshot007}
-	\end{figure}
-	% We define the language using inference rules
-	% How can we express those rules in Statix?
 \end{frame}
@@
 	\begin{figure}
-		\includegraphics[width=0.7\linewidth]{screenshot004}
+		\includegraphics[width=0.7\linewidth]{screenshot005}
 	\end{figure}
@@
 	\end{exampleblock}
-	\begin{block}{Evaluation relation} 
+	\begin{block}{Evaluation relations} 
 		\texttt{betaHeadReduce : scope * Expr -> scope * Expr} \\
 		\texttt{betaReduce : scope * Expr -> Expr} \\
-		\texttt{exectBetaEq : (scope * Expr) * (scope * Expr)}
+		\texttt{expectBetaEq : (scope * Expr) * (scope * Expr)}
 	\end{block}
-\end{frame}
-
-\begin{frame}[fragile]{Type Checking: Requires Evaluation}
-	\begin{figure}
-		\includegraphics[width=0.9\linewidth]{screenshot001}
-	\end{figure}
 
@@
 
-\begin{frame}{Comparison with Haskell}
-	\begin{block}{Type checking}
-		\texttt{data EnvEntry = NType Expr | NSubst SExpr
-			\\type Env = [EnvEntry]
-			\\type SExpr = (Env , Expr)
-			\\tc :: SExpr -> Either String Expr}
-	\end{block}
-
-	\begin{block}{Differences}
-		\begin{itemize}
-			\item Distribution of definitions over files
-			\item De Bruijn Indices vs Names
-			\item Inference built-in to Statix
-		\end{itemize}
-	\end{block}
-\end{frame}
-
 \begin{frame}{On Extensibility}
@@
 		\end{enumerate}
+	\end{block}
+\end{frame}
+
+\begin{frame}{Comparison with Haskell}
+	\begin{block}{Type checking}
+		\texttt{data EnvEntry = NType Expr | NSubst (Env, Expr)
+			\\type Env = [EnvEntry]
+			\\tc :: Env -> Expr -> Either String Expr}
+	\end{block}
+
+	\begin{block}{Differences}
+		\begin{itemize}
+			\item Distribution of definitions over files
+			\item De Bruijn Indices vs Names
+			\item Definition in Statix is approximately 20\% longer
+			\item Inference built-in to Statix
+		\end{itemize}
 	\end{block}
@@
 \begin{frame}[fragile]{Conclusions}
-	Spoofax is a great tool for developing dependently typed languages!
 	\begin{itemize}
-		\item We can use scopes to represent environments and contexts
-		\item Statix can still use improvements
+		\item \textbf{Spoofax perspective:} Support for complex inference could be improved, for example using equational unification.
+		\item \textbf{Dependent types perspective:} This already functions as a great tool for prototyping dependently typed languages!
 		% - Higher order 
 	\end{itemize}
-
-	%- using scopes as env+context  
-	%- statix enforces syntactic uniqueness of metavariable solutions but dependently typed languages use a weaker notion of equality 
-	%- full thesis coming soonTM
 \end{frame}
 
-% block
-% exampleblock
-% alertblock
-
 \end{document}

### 2023-04-24 pres-thesis/presentation.tex
Minor changes to the presentation
Change conclusion a bit

@@
 }
+
+\begin{frame}[fragile]{Table of Contents}
+	\begin{itemize}
+		\item Background
+		\item Why is this important?
+		\item \textbf{Primary Contribution:} Calculus of Constructions
+		\item Extensibility
+		\item Comparison with implementation in Haskell
+		\item Inference
+	\end{itemize}
+\end{frame}
 
@@
 \begin{enumerate}
-	\item Implemented Inference
+	\item \textbf{Implemented Inference}
 	\item Implemented Inductive Data Types
@@
 \begin{enumerate}
-	\item Comparison with implementation in Haskell
+	\item \textbf{Comparison with implementation in Haskell}
 	\item Comparison with implementation in LambdaPi
@@
 	\begin{itemize}
-		\item \textbf{Spoofax perspective:} Support for complex inference could be improved, for example using equational unification.
-		\item \textbf{Dependent types perspective:} This already functions as a great tool for prototyping dependently typed languages!
-		% - Higher order 
+		\item Statix is already as a great tool for prototyping dependently typed languages!
+		\item Support for complex inference could be improved, for example using equational unification.
 	\end{itemize}
